---
title: "Effect of DDX55 knockdown (ENCODE) on functional miR-29 targets defined using miR-29 inhibition and over-expression experiments"
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
```

# Load libraries

```{r}
library(fgsea)
library(ggplot2)
library(glue)
```

# Load the miRNA-binding matrix

```{r}
binding <- read.csv(here("data", "miRNA_seed_matches_3UTRs2000bp.csv"), row.names = 1, header = TRUE)
```

# Define miR-29 predicted targets

```{r}
pred_tars <- row.names(binding)[binding[, "miR.29abcd"] > 0]
```

# Load the differential gene expression data for the DDX55 KD experiment from ENCODE (HepG2 cells). Retain only non-NA entries

```{r}
DDX55_KD <- read.table(here("data", "ENCFF410VPN.tsv"))
row.names(DDX55_KD) <- gsub("\\..*", "", row.names(DDX55_KD))
DDX55_KD$padj <- p.adjust(DDX55_KD$pvalue, method = "fdr")
DDX55_KD <- DDX55_KD[ !is.na(DDX55_KD$padj), ] 
```

# Load the differential expression analysis results for the miR-29 mimic experiment in **786-O cells** (from the DiffRAC paper https://doi.org/10.1038/s42003-022-03796-w). Retain only non-NA entries

```{r}
mimic_786O <- readRDS(here("data", "siDDX55_res_condition_Mimic.miR_vs_Mimic.neg.rds"))
mimic_786O <- mimic_786O[ !is.na(mimic_786O$padj), ]
```

# Load the differential expression analysis results for the miR-29 mimic experiment in **A-498 cells** (from the DiffRAC paper https://doi.org/10.1038/s42003-022-03796-w). Retain only non-NA entries

```{r}
mimic_A498 <- readRDS(here("data", "DiffRAC_miR29_validation_mimic_res_condition_miR29_mimic_vs_miR29_mimic_ctrl.rds"))
mimic_A498 <- mimic_A498[ !is.na(mimic_A498$padj), ]
```

# Load the differential expression analysis results for the miR-29 **inhibition** experiment in ACHN cells (from the DiffRAC paper https://doi.org/10.1038/s42003-022-03796-w). Retain only non-NA entries

```{r}
inhib_ACHN <- readRDS(here("data", "DiffRAC_miR29_validation_inhibition_res_condition_miR29_inhibition_vs_miR29_inhibition_ctrl.rds"))
inhib_ACHN <- inhib_ACHN[ !is.na(inhib_ACHN$padj), ]
```

# Define functional miR-29 targets: Genes that are predicted targets of miR-29 and satisfy at least one of these categories:
- Significantly downregulated in at least one of the two miR-29 mimic experiment
- Significantly upregulated in the miR-29 inhibition experiment

```{r}
down_786O <- row.names(mimic_786O)[mimic_786O$padj < 0.05 & mimic_786O$log2FoldChange < 0]
down_A498 <- row.names(mimic_A498)[mimic_A498$padj < 0.05 & mimic_A498$log2FoldChange < 0]
up_ACHN <- row.names(inhib_ACHN)[mimic_A498$padj < 0.05 & mimic_A498$log2FoldChange > 0]
tars <- pred_tars[pred_tars %in% down_786O | pred_tars %in% down_A498 | pred_tars %in% up_ACHN]
```

# GSEA to define the enrichment of these miR-29 targets among up/down-regulated genes in the DDX55 KD from ENCODE (HepG2).

```{r, fig.width=3, fig.height=2.5}
genelist <- setNames(list(tars), "miR-29 functional targets")
stat <- setNames(DDX55_KD$stat, row.names(DDX55_KD))

fgseaRes <- fgsea(pathways = genelist, stats = stat, minSize = 10)

# GSEA Plot
p <- plotEnrichment(genelist[[1]], stat, gseaParam = 1, ticksSize = 0.05) +
  labs(title = "DDX55 KD - miR-29",
       x = paste("Rank-ordered genes in\nDDX55 KD (", length(genelist[[1]]), ")", sep=""),
       y = "Enrichment score (ES)")

# Get y-axis limits to position annotation slightly above center
y_lims <- ggplot_build(p)$layout$panel_scales_y[[1]]$range$range
y_pos <- (y_lims[1] + y_lims[2]) / 3 # Slightly above center on y-axis

p <- p +
  annotate("text",
           x = length(stat) * 0.6,
           y = y_pos,
           label = glue("ES = {signif(fgseaRes$ES[1], 3)}\np = {signif(fgseaRes$pval[1], 3)}"),
           hjust = 0, vjust = 0.5,
           size = 3, fontface = "plain")

# Change enrichment curve color
p$layers[[1]]$aes_params$colour <- "blue"

# Remove gridlines
p <- p +
  theme_classic() +
  theme(panel.grid = element_blank())

# Save and print
ggsave(here("DDX55_miR29_overexpression_inhibition", "output", "figures",
            "GSEA_functional_miR-29_targets_rank_genes_ENCODE_DDX55_KD.pdf"),
       p, width = 3, height = 2.5)

print(p)
```

```{r}
sessionInfo()
```
