---
title: "shRNA positions associated with changes in mRNA stability in ENCODE RBP KD samples"
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries

```{r}
library(Biostrings)
library(stringr)
library(seqRFLP)
library(data.table)
library(gplots)
library(here)
```

First, we look at associations between counts of shRNA 6-mers/7-mers in 3'UTRs and mRNA stability. 

Load shRNA sequences file

```{r}
RNAi_seqs <- read.table(here("ENCODE_RBP_KD", "data", "RNAi_sequences.txt"), header = T)

# Format the Experiment names appropriately
RNAi_seqs$Experiment <- gsub("/experiments/", "", RNAi_seqs$Experiment)
RNAi_seqs$Experiment <- gsub("/", "", RNAi_seqs$Experiment)

# Format the shrNA sequence appropriately
RNAi_seqs$RNAi_sequence <- gsub("-", "", RNAi_seqs$RNAi_sequence)
```

Load the 3'UTR sequences

```{r}
UTR_seq <- as.character(readDNAStringSet(here("ENCODE_RBP_KD", "data", "3UTR_seq.fa")))
names(UTR_seq) <- gsub(".*gene=", "", names(UTR_seq))
```

Trim to keep the first 2000bps of each 3'UTR

```{r}
UTR_seq <- unlist(lapply(UTR_seq, function(x) substring(x, 1, 2000)))
```

Load bias and batch corrected stability estimates

```{r}
stab_HepG2 <- read.table(here("ENCODE_RBP_KD", "data", "residuals_ddf_dfi_batches_HepG2_ROBUST.txt"), header=TRUE, row.names=1)
stab_K562 <- read.table(here("ENCODE_RBP_KD", "data", "residuals_ddf_dfi_batches_K562_ROBUST.txt"), header=TRUE, row.names=1)
```

Filter stability and UTR sequences tables to keep the same genes in the same order

```{r}
common_genes_HepG2 <- intersect(row.names(stab_HepG2), names(UTR_seq))
common_genes_K562 <- intersect(row.names(stab_K562), names(UTR_seq))

UTR_HepG2_fltr <- UTR_seq[common_genes_HepG2]
stab_HepG2_fltr <- stab_HepG2[common_genes_HepG2, ]

UTR_K562_fltr <- UTR_seq[common_genes_K562]
stab_K562_fltr <- stab_K562[common_genes_K562, ]
```

For each cell line, for each sample (and so for each shRNA), use a 6nt/7nt-length sliding window to get all 6nt/7nt sequences in the shRNA and count their occurence in the 3'UTRs. Then we assess the associations between counts of shRNA 6-mers/7-mers in 3'UTRs and mRNA stability (bias- and batch-corrected). We do this separately for 6-mers and 7-mers, so see the comments in the code for description of the next steps.

THE CSV FILES THAT RESULT FROM THIS SECTION ARE ALREADY INCLUDED, so you can move on to the next section without having to run this one (it takes a very long time to run).

```{r}
for(len in c(6,7)){
  for (cell in c("HepG2", "K562"))
  {
    # UTR sequences for the cell line
    UTRs <- get(paste("UTR_", cell, "_fltr", sep=""))
    
    # Stability data for the cell line
    stab <- get(paste("stab_", cell, "_fltr", sep=""))
    
    # For each sample
    for (i in 1:ncol(stab))
    {
      # Define the experiment
      experiment <- gsub("_.*","", colnames(stab)[i])
      
      # If the experiment is not in the shRNA sequences table, write stability as NA and remove after this part
      if (!(experiment %in% RNAi_seqs$Experiment))
      {
        next
        print("Experiment not found in the shRNA sequences table")
      }
      
      # Define the shRNA sequence
      shRNA <- RNAi_seqs[RNAi_seqs$Experiment == experiment, ]$RNAi_sequence
      
      # Use a sliding window to get all 6-mers/7-mers in the shRNAs
      k <- len
      n <- nchar(shRNA)
      mers <- substring(shRNA, 1:(n-k+1), k:n)
      
      # Reverse complement to get target sequences
      mers <- unlist(lapply(mers, revComp))
      
      # Make a table to write the 6-mer/7-mer counts, with the genes as rows and the 6-mers/7-mers as columns
      mers_counts <- as.data.frame(matrix(NA, ncol = length(mers), nrow=nrow(stab)))
      row.names(mers_counts) <- row.names(stab)
      colnames(mers_counts) <- mers
      
      # Fill the table with counts of each 6-mer/7-mer in the 3'UTR sequences for each gene
      for (g in 1:nrow(mers_counts))
      {
        UTR <- UTRs[row.names(mers_counts)[g]]
        for (mer in colnames(mers_counts))
        {
          mers_counts[g, mer] <- str_count(UTR, mer)
        }
      }
      
      # Delete columns with the same 6-mer/7-mer sequence, but record their position and corresponding sequence first
      del_cols <- which(duplicated(colnames(mers_counts)))
      del_cols_mers <- colnames(mers_counts)[duplicated(colnames(mers_counts))]
      mers_counts_2 <- mers_counts[, !(duplicated(colnames(mers_counts)))]
      
      # Save the count table for later use
      write.csv(mers_counts_2, file=here("ENCODE_RBP_KD", "output", "supplementary_tables", paste(experiment, "_shRNA_", len, "nts_counts.csv", sep="")))
      
      # Throw error if there are columns with all 0s
      if (sum(colSums(mers_counts_2) == 0) > 0)
      {
        print("ERROR: Some kmers don't occur in some UTR(s)")
        break
      }
      
      # Perform glm on sample stability data using the kmers as predictor
      fit <- glm(stab[, i] ~ data.matrix(mers_counts_2), na.action=na.exclude)
      
      # Fill in the table of regression coefficients
      coeffs <- as.data.frame(coef(fit)[-1])
      row.names(coeffs) <- gsub(".*)", "", row.names(coeffs))
      pvals <- as.data.frame(coef(summary(fit))[,4][-1])
      row.names(pvals) <- gsub(".*)", "", row.names(pvals))
      
      # Put back the kmers that were deleted, in both the coefficients and p-values tables
      coeffs_complete <- cbind(row.names(coeffs), coeffs[,1])
      coeffs_complete_names <- row.names(coeffs)
      
      pvals_complete <- cbind(row.names(pvals), pvals[,1])
      pvals_complete_names <- row.names(pvals)
      
      if (length(del_cols) > 0)
      {
        for (x in 1:length(del_cols))
        {
          coeffs_complete_names <- c(coeffs_complete_names[1:(del_cols[x]-1)], del_cols_mers[x], coeffs_complete_names[(del_cols[x]):length(coeffs_complete_names)])
          coeffs_complete <- rbind(coeffs_complete[1:(del_cols[x]-1), ], coeffs_complete[coeffs_complete[,1] == del_cols_mers[x], ], coeffs_complete[(del_cols[x]):nrow(coeffs_complete), ])
        }
        
        for (x in 1:length(del_cols))
        {
          pvals_complete_names <- c(pvals_complete_names[1:(del_cols[x]-1)], del_cols_mers[x], pvals_complete_names[(del_cols[x]):length(pvals_complete_names)])
          pvals_complete <- rbind(pvals_complete[1:(del_cols[x]-1), ], pvals_complete[pvals_complete[,1] == del_cols_mers[x], ], pvals_complete[(del_cols[x]):nrow(pvals_complete), ])
        }      
      }
      
      # Save the tables of coefficients and p-values
      write.csv(coeffs_complete, file=here("ENCODE_RBP_KD", "output", "supplementary_tables", paste(colnames(stab)[i], "_regressionCorrectionStabShRNA", len, "nts_coeffs.csv", sep="")))
      write.csv(pvals_complete, file=here("ENCODE_RBP_KD", "output", "supplementary_tables", paste(colnames(stab)[i], "_regressionCorrectionStabShRNA", len, "nts_pvals.csv", sep="")))
    }
  }
}
  
```

Now we want to identify and visualize the shRNA positions most predictive of mRNA stability. We repeat the same procedure for 6-mers and 7-mers separately, so see comments in the code for next steps.

```{r}

for(len in c(6,7)){
 
   # Load the regression coefficients from the previous section for all samples
  coeff_files <- list.files(here("ENCODE_RBP_KD", "data"), pattern = paste("regressionCorrectionStabShRNA", len, "nts_coeffs", sep=""), full.names = T)
  coeff_data <- lapply(coeff_files, fread, header=T)
  coeff_data <- lapply(coeff_data, function(x) as.vector(as.data.frame(x)[,3]))
  max_len <- as.numeric(max(names(table(unlist(lapply(coeff_data, length))))))
  
  # Max shRNA length is 54, some are shorter, so fill the vectors of coefficients with NAs so that all have length 54
  
  for (i in 1:length(coeff_data))
  {
    init_len <- length(coeff_data[[i]])
    nb_missing <- max_len - init_len
    coeff_data[[i]] <- c(coeff_data[[i]], rep(NA, nb_missing))
  }
    
  # Make a big table with all of the coefficients
  coeffs <- do.call("rbind", coeff_data)
  sample_names <- gsub("\\./", "", coeff_files)
  sample_names <- gsub("_regressionCorrectionStabShRNA.*", "", sample_names)
  row.names(coeffs) <- sample_names
  colnames(coeffs) <- c(1:max_len)
    
  # Replace NAs with 0s
  coeffs[is.na(coeffs)] <- 0
  
  # Load the p-values from the previous section for all samples
  pval_files <- list.files(here("ENCODE_RBP_KD", "data"), pattern = paste("regressionCorrectionStabShRNA", len, "nts_pvals", sep=""), full.names = T)
  pval_data <- lapply(pval_files, fread, header=T)
  pval_data <- lapply(pval_data, function(x) as.vector(as.data.frame(x)[,3]))
  max_len <- as.numeric(max(names(table(unlist(lapply(pval_data, length))))))
  
  # Max shRNA length is 54, some are shorter, so fill the p-values vectors with NAs so that all have length 54, like we did for coefficients
  
  for (i in 1:length(pval_data))
  {
    init_len <- length(pval_data[[i]])
    nb_missing <- max_len - init_len
    pval_data[[i]] <- c(pval_data[[i]], rep(NA, nb_missing))
  }
    
  # Make a big matrix with all of the p-values  
  pvals <- do.call("rbind", pval_data)
  sample_names <- gsub("\\./", "", pval_files)
  sample_names <- gsub("_regressionCorrectionStabShRNA.*", "", sample_names)
  row.names(pvals) <- sample_names
  colnames(pvals) <- c(1:max_len)
  
  # Transform to signed log10(p-values)
  signs <- sign(coeffs)
  slog10pvals <- abs(log10(pvals))*signs
    
  # Replace NAs with 0s
  slog10pvals[is.na(slog10pvals)] <- 0
    
  # Make a heatmap
  palette <- colorRampPalette(c("blue", "white", "red"))
  tryCatch({
    pdf(here("ENCODE_RBP_KD", "output", "figures", paste("ENCODE_RBPKD_samples_shRNA_position", len, "nts_associationStability_signedlog10pvals.pdf", sep="")), height = 2, width = 7)
    heatmap.2(slog10pvals, trace="none", col=palette, cexRow=0.001, cexCol=0.5,  breaks=seq(-10,10,length.out=100), mar=c(2,2), key=F, Colv = F, 
              hclust=function(x) hclust(x,method="complete"), distfun=function(x) as.dist((1-cor(t(x)))/2), density.info = "none", lwid = c(1,10))
    dev.off()
  }, error=function(e){})
}
  
```

![Heatmap of associations between shRNA 6-mer counts and mRNA stability](ENCODE_RBPKD_samples_shRNA_position6nts_associationStability_signedlog10pvals.pdf){width=65%}

![Heatmap of associations between shRNA 7-mer counts and mRNA stability](ENCODE_RBPKD_samples_shRNA_position7nts_associationStability_signedlog10pvals.pdf){width=65%} 

We see that for 6-mers (top), positions 34-38 on shRNA are associated with decreased mRNA stability. It is very similar for 7-mers (bottom), except that the region that seems to mediate the off-target effects is from position 34 to 37 instead of 34 to 38.