---
title: "shRNA positions associated with changes in mRNA stability in ENCODE RBP KD samples"
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
set.seed(1)
```

# Load libraries

```{r}
library(Biostrings)
library(data.table)
library(gplots)
```

# First, we look at associations between counts of shRNA 6-mers/7-mers in 3'UTRs and mRNA stability. 

## Load shRNA sequences file

```{r}
RNAi_seqs <- read.table(here("ENCODE_RBP_KD", "data", "RNAi_sequences.txt"), header = T)

# Format the Experiment names appropriately
RNAi_seqs$Experiment <- gsub("/experiments/", "", RNAi_seqs$Experiment)
RNAi_seqs$Experiment <- gsub("/", "", RNAi_seqs$Experiment)

# Format the shRNA sequence appropriately
RNAi_seqs$RNAi_sequence <- gsub("-", "", RNAi_seqs$RNAi_sequence)
```

## Load the 3'UTR sequences

```{r}
UTR_seq <- as.character(readDNAStringSet(here("ENCODE_RBP_KD", "data", "3UTR_seq.fa")))
names(UTR_seq) <- gsub(".*gene=", "", names(UTR_seq))
```

## Trim to keep the first 2000bps of each 3'UTR

```{r}
UTR_seq <- unlist(lapply(UTR_seq, function(x) substring(x, 1, 2000)))
```

## Load bias and batch corrected stability estimates

```{r}
stab_HepG2 <- read.csv(here("ENCODE_RBP_KD", "output", "supplementary_tables", paste0("ENCODE_RBP_KD_bias_batch_corrected_stability_", "HepG2", ".csv")), row.names=1)
stab_K562 <- read.csv(here("ENCODE_RBP_KD", "output", "supplementary_tables", paste0("ENCODE_RBP_KD_bias_batch_corrected_stability_", "K562", ".csv")), row.names=1)
```

## Filter stability and UTR sequences tables to keep the same genes in the same order

```{r}
common_genes_HepG2 <- intersect(row.names(stab_HepG2), names(UTR_seq))
common_genes_K562 <- intersect(row.names(stab_K562), names(UTR_seq))

UTR_HepG2_fltr <- UTR_seq[common_genes_HepG2]
stab_HepG2_fltr <- stab_HepG2[common_genes_HepG2, ]

UTR_K562_fltr <- UTR_seq[common_genes_K562]
stab_K562_fltr <- stab_K562[common_genes_K562, ]
```

## Associations between counts of shRNA 6-mers/7-mers in 3'UTRs and mRNA stability

For each cell line, for each sample (and so for each shRNA), use a 6nt/7nt-length sliding window to get all 6nt/7nt sequences in the shRNA and count their occurrence in the 3'UTRs. Then we assess the associations between counts of shRNA 6-mers/7-mers in 3'UTRs and mRNA stability (bias- and batch-corrected). We do this separately for 6-mers and 7-mers, so see the comments in the code for description of the next steps.

```{r}
# Helper function to reverse complement DNA sequences
revComp <- function(seq) {
  as.character(reverseComplement(DNAString(seq)))
}

# Helper function to count kmers efficiently across all genes at once
count_kmers <- function(subject, kmers) {

  counts_mat <- sapply(kmers, function(kmer) {
    vcountPattern(
      pattern = kmer,
      subject = subject,
      fixed = TRUE
    )
  })

  if (is.vector(counts_mat)) {
    counts_mat <- matrix(counts_mat, ncol = 1)
  }

  rownames(counts_mat) <- names(subject)
  colnames(counts_mat) <- kmers

  counts_mat
}

# Storage for regression results 
regression_results <- list()

for(len in c(6, 7)) {
  regression_results[[paste0("len_", len)]] <- list()
  
  for (cell in c("HepG2", "K562")) {
    # UTR sequences for the cell line
    UTRs <- switch(
      cell,
      HepG2 = UTR_HepG2_fltr,
      K562  = UTR_K562_fltr
    )
    
    # Convert once for efficient pattern matching
    UTRs_dna <- DNAStringSet(UTRs)
    names(UTRs_dna) <- names(UTRs)
  
    # Stability data for the cell line
    stab <- switch(
      cell,
      HepG2 = stab_HepG2_fltr,
      K562  = stab_K562_fltr
    )
  
    # For each sample
    for(i in seq_len(ncol(stab))) {
      # Define the experiment
      experiment <- gsub("_.*", "", colnames(stab)[i])
      
      # Skip if experiment not in shRNA sequences table
      if (!(experiment %in% RNAi_seqs$Experiment)) {
        next
      }
      
      # Define the shRNA sequence
      shRNA <- RNAi_seqs[RNAi_seqs$Experiment == experiment, ]$RNAi_sequence
      stopifnot(length(shRNA) == 1)
      
      # Use a sliding window to get all k-mers in the shRNA
      k <- len
      n <- nchar(shRNA)
      mers <- substring(shRNA, 1:(n - k + 1), k:n)
      
      # Reverse complement to get target sequences
      mers <- unlist(lapply(mers, revComp))
      
      # Track which kmers are unique (duplicates will be restored later)
      dup_mask <- duplicated(mers)
      unique_mers <- mers[!dup_mask]
      
      # Count occurrences of each unique k-mer in every gene's 3'UTR
      mers_counts <- count_kmers(UTRs_dna, unique_mers)
      
      # Stop processing if a k-mer is absent from all 3'UTRs
      if (sum(colSums(mers_counts) == 0) > 0) {
        stop("ERROR: kmers doesn't occur in any UTR(s)")
      }
      
      # Fit a linear regression model on sample stability data using the kmers counts as predictor
      fit <- glm(stab[, i] ~ data.matrix(mers_counts), na.action = na.exclude)
      
      # Extract regression coefficients and p-values
      fit_summary <- coef(summary(fit))
      coeffs <- as.data.frame(fit_summary[-1, "Estimate", drop = FALSE])
      pvals <- as.data.frame(fit_summary[-1, "Pr(>|t|)", drop = FALSE])
      
      # Clean up row names
      rownames(coeffs) <- gsub(".*)", "", rownames(coeffs))
      rownames(pvals) <- gsub(".*)", "", rownames(pvals))
      
      # Reinsert duplicated k-mers so regression coefficients align with their original positions along the shRNA
      restore_idx <- match(mers, unique_mers)
      
      coeffs <- coeffs[restore_idx, , drop = FALSE]
      pvals  <- pvals[restore_idx, , drop = FALSE]
      
      # Store coefficient and p-value vectors in memory (indexed by position along shRNA)
      sample_name <- colnames(stab)[i]
      regression_results[[paste0("len_", len)]][[sample_name]] <- list(
        coeffs = coeffs[, 1],
        pvals = pvals[, 1]
      )
    }
  }
}
```

# Visualize the shRNA positions most strongly associated with mRNA stability

We visualize which shRNA positions (1-54) are most predictive of mRNA stability changes.

```{r}
for(len in c(6, 7)) {
  # Extract regression results for the kmer length
  results_key <- paste0("len_", len)
  sample_results <- regression_results[[results_key]]
  
  # Get all sample names and determine max shRNA length
  sample_names <- names(sample_results)
  max_len <- max(sapply(sample_results, function(x) length(x$coeffs)))
  
  # Initialize matrices for all coefficients and p-values
  coeffs <- matrix(0, nrow = length(sample_names), ncol = max_len)
  pvals <- matrix(0, nrow = length(sample_names), ncol = max_len)
  rownames(coeffs) <- sample_names
  rownames(pvals) <- sample_names
  colnames(coeffs) <- as.character(1:max_len)
  colnames(pvals) <- as.character(1:max_len)
  
  # Pad shorter shRNAs with zeros so all samples share a common matrix dimension
  for (i in seq_along(sample_names)) {
    sample_name <- sample_names[i]
    coeff_vec <- sample_results[[sample_name]]$coeffs
    pval_vec <- sample_results[[sample_name]]$pvals
    
    coeffs[i, 1:length(coeff_vec)] <- coeff_vec
    pvals[i, 1:length(pval_vec)] <- pval_vec
  }
  
  # Compute signed log10(p-values) with direction from coefficients
  signs <- sign(coeffs)
  slog10pvals <- abs(log10(pvals)) * signs
  slog10pvals[is.na(slog10pvals)] <- 0
  
  # Generate heatmap
  palette <- colorRampPalette(c("blue", "white", "red"))
  tryCatch({
    pdf(here("ENCODE_RBP_KD", "output", "figures", 
             paste("ENCODE_RBP_KD_shRNA_", len, "mers_positions_vsStab_signedlog10pvals.pdf", sep="")), 
        height = 2, width = 7)
    heatmap.2(slog10pvals, trace = "none", col = palette, cexRow = 0.001, cexCol = 0.5,
              breaks = seq(-10, 10, length.out = 100), mar = c(2, 2), key = FALSE, Colv = FALSE,
              hclust = function(x) hclust(x, method = "complete"), 
              distfun = function(x) as.dist((1 - cor(t(x))) / 2), 
              density.info = "none", lwid = c(1, 10))
    dev.off()
  }, error = function(e) {})
}
  
```

We see that for 6-mers (top), positions 34-38 on shRNAs are associated with decreased mRNA stability. It is very similar for 7-mers (bottom), except that the region that seems to mediate the off-target effects is from position 34 to 37 instead of 34 to 38.

```{r}
sessionInfo()
```
