---
title: "shRNA positions associated with changes in mRNA stability in ENCODE RBP KD samples"
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
set.seed(1)
```

# Load libraries

```{r}
library(Biostrings)
library(data.table)
library(gplots)
```

# Load shRNA sequences file

```{r}
RNAi_seqs <- read.table(here("ENCODE_RBP_KD", "data", "RNAi_sequences.txt"), header = T)

# Format the Experiment names appropriately
RNAi_seqs$Experiment <- gsub("/experiments/", "", RNAi_seqs$Experiment)
RNAi_seqs$Experiment <- gsub("/", "", RNAi_seqs$Experiment)

# Format the shRNA sequence appropriately
RNAi_seqs$RNAi_sequence <- gsub("-", "", RNAi_seqs$RNAi_sequence)
```

# Load the 3'UTR sequences

```{r}
UTR_seq <- as.character(readDNAStringSet(here("ENCODE_RBP_KD", "data", "3UTR_seq.fa")))
names(UTR_seq) <- gsub(".*gene=", "", names(UTR_seq))
```

# Trim to keep the first 2000bps of each 3'UTR

```{r}
UTR_seq <- unlist(lapply(UTR_seq, function(x) substring(x, 1, 2000)))
```

# Load bias and batch corrected stability estimates

```{r}
stab_HepG2 <- read.csv(here("ENCODE_RBP_KD", "output", "supplementary_tables", paste0("ENCODE_RBP_KD_bias_batch_corrected_stability_", "HepG2", ".csv")), row.names=1)
stab_K562 <- read.csv(here("ENCODE_RBP_KD", "output", "supplementary_tables", paste0("ENCODE_RBP_KD_bias_batch_corrected_stability_", "K562", ".csv")), row.names=1)
```

# Filter stability and UTR sequences tables to keep the same genes in the same order

```{r}
common_genes_HepG2 <- intersect(row.names(stab_HepG2), names(UTR_seq))
common_genes_K562 <- intersect(row.names(stab_K562), names(UTR_seq))

UTR_HepG2_fltr <- UTR_seq[common_genes_HepG2]
stab_HepG2_fltr <- stab_HepG2[common_genes_HepG2, ]

UTR_K562_fltr <- UTR_seq[common_genes_K562]
stab_K562_fltr <- stab_K562[common_genes_K562, ]
```

# Summary of steps to identify shRNA positions associated with off-target effects and correct stability estimates for these off-target effects

1. Test associations between counts of shRNA 6-mers/7-mers in 3'UTRs and mRNA stability:
    - For each cell line, for each sample (and so for each shRNA), use a 6nt/7nt-length sliding window to get all 6nt/7nt sequences in the shRNA and count their occurrence in the 3'UTRs. Then we assess the associations between counts of shRNA 6-mers/7-mers in 3'UTRs and mRNA stability (bias- and batch-corrected). We do this separately for 6-mers and 7-mers. Retain regression coefficients and P-values
2. Identify positions of shRNAs that are predictive of stability changes:
  - Multiple testing corrections for P-values obtained at step 1 
  - Filter for significant hits
  - For each shRNA position, determine if there are more significant hits than expected by chance (Fisher’s test)
  - Repeat for 6-mers and 7-mers separately
3. Correct stability estimates for shRNA off-target effects 
  - Using a regression with 6-mer and 7-mer counts from the positions associated with off-target effects as predictors, retain residuals as corrected stability data

# Test associations between counts of shRNA 6-mers/7-mers in 3'UTRs and mRNA stability

```{r}
# Helper function to reverse complement DNA sequences
revComp <- function(seq) {
  as.character(reverseComplement(DNAString(seq)))
}

# Helper function to count kmers efficiently across all genes at once
count_kmers <- function(subject, kmers) {

  counts_mat <- sapply(kmers, function(kmer) {
    vcountPattern(
      pattern = kmer,
      subject = subject,
      fixed = TRUE
    )
  })

  if (is.vector(counts_mat)) {
    counts_mat <- matrix(counts_mat, ncol = 1)
  }

  rownames(counts_mat) <- names(subject)
  colnames(counts_mat) <- kmers

  counts_mat
}

# Storage for regression results 
regression_results <- list()

for(len in c(6, 7)) {
  regression_results[[paste0("len_", len)]] <- list()
  
  for (cell in c("HepG2", "K562")) {
    # UTR sequences for the cell line
    UTRs <- switch(
      cell,
      HepG2 = UTR_HepG2_fltr,
      K562  = UTR_K562_fltr
    )
    
    # Convert once for efficient pattern matching
    UTRs_dna <- DNAStringSet(UTRs)
    names(UTRs_dna) <- names(UTRs)
  
    # Stability data for the cell line
    stab <- switch(
      cell,
      HepG2 = stab_HepG2_fltr,
      K562  = stab_K562_fltr
    )
  
    # For each sample
    for(i in seq_len(ncol(stab))) {
      # Define the experiment
      experiment <- gsub("_.*", "", colnames(stab)[i])
      
      # Skip if experiment not in shRNA sequences table
      if (!(experiment %in% RNAi_seqs$Experiment)) {
        next
      }
      
      # Define the shRNA sequence
      shRNA <- RNAi_seqs[RNAi_seqs$Experiment == experiment, ]$RNAi_sequence
      stopifnot(length(shRNA) == 1)
      
      # Use a sliding window to get all k-mers in the shRNA
      k <- len
      n <- nchar(shRNA)
      mers <- substring(shRNA, 1:(n - k + 1), k:n)
      
      # Reverse complement to get target sequences
      mers <- unlist(lapply(mers, revComp))
      
      # Track which kmers are unique (duplicates will be restored later)
      dup_mask <- duplicated(mers)
      unique_mers <- mers[!dup_mask]
      
      # Count occurrences of each unique k-mer in every gene's 3'UTR
      mers_counts <- count_kmers(UTRs_dna, unique_mers)
      
      # Stop processing if a k-mer is absent from all 3'UTRs
      if (sum(colSums(mers_counts) == 0) > 0) {
        stop("ERROR: kmers doesn't occur in any UTR(s)")
      }
      
      # Fit a linear regression model on sample stability data using the kmers counts as predictor
      fit <- glm(stab[, i] ~ data.matrix(mers_counts), na.action = na.exclude)
      
      # Extract regression coefficients and p-values
      fit_summary <- coef(summary(fit))
      coeffs <- as.data.frame(fit_summary[-1, "Estimate", drop = FALSE])
      pvals <- as.data.frame(fit_summary[-1, "Pr(>|t|)", drop = FALSE])
      
      # Clean up row names
      rownames(coeffs) <- gsub(".*)", "", rownames(coeffs))
      rownames(pvals) <- gsub(".*)", "", rownames(pvals))
      
      # Reinsert duplicated k-mers so regression coefficients align with their original positions along the shRNA
      restore_idx <- match(mers, unique_mers)
      
      coeffs <- coeffs[restore_idx, , drop = FALSE]
      pvals  <- pvals[restore_idx, , drop = FALSE]
      
      # Store coefficient and p-value vectors in memory (indexed by position along shRNA)
      sample_name <- colnames(stab)[i]
      regression_results[[paste0("len_", len)]][[sample_name]] <- list(
        coeffs = coeffs[, 1],
        pvals = pvals[, 1]
      )
    }
  }
}

saveRDS(regression_results, here("ENCODE_RBP_KD", "output", "rds", "regression_results.rds"))

```

# Visualize the shRNA positions most strongly associated with mRNA stability

We visualize which shRNA positions (1-54) are most predictive of mRNA stability changes.

```{r}
for(len in c(6, 7)) {
  # Extract regression results for the kmer length
  results_key <- paste0("len_", len)
  sample_results <- regression_results[[results_key]]
  
  # Get all sample names and determine max shRNA length
  sample_names <- names(sample_results)
  max_len <- max(sapply(sample_results, function(x) length(x$coeffs)))
  
  # Initialize matrices for all coefficients and p-values
  coeffs <- matrix(0, nrow = length(sample_names), ncol = max_len)
  pvals <- matrix(0, nrow = length(sample_names), ncol = max_len)
  rownames(coeffs) <- sample_names
  rownames(pvals) <- sample_names
  colnames(coeffs) <- as.character(1:max_len)
  colnames(pvals) <- as.character(1:max_len)
  
  # Pad shorter shRNAs with zeros so all samples share a common matrix dimension
  for (i in seq_along(sample_names)) {
    sample_name <- sample_names[i]
    coeff_vec <- sample_results[[sample_name]]$coeffs
    pval_vec <- sample_results[[sample_name]]$pvals
    
    coeffs[i, 1:length(coeff_vec)] <- coeff_vec
    pvals[i, 1:length(pval_vec)] <- pval_vec
  }
  
  # Compute signed log10(p-values) with direction from coefficients
  signs <- sign(coeffs)
  slog10pvals <- abs(log10(pvals)) * signs
  slog10pvals[is.na(slog10pvals)] <- 0
  
  # Generate heatmap
  palette <- colorRampPalette(c("blue", "white", "red"))
  tryCatch({
    pdf(here("ENCODE_RBP_KD", "output", "figures", 
             paste("ENCODE_RBP_KD_shRNA_", len, "mers_positions_vsStab_signedlog10pvals.pdf", sep="")), 
        height = 2, width = 7)
    heatmap.2(slog10pvals, trace = "none", col = palette, cexRow = 0.001, cexCol = 0.5,
              breaks = seq(-10, 10, length.out = 100), mar = c(2, 2), key = FALSE, Colv = FALSE,
              hclust = function(x) hclust(x, method = "complete"), 
              distfun = function(x) as.dist((1 - cor(t(x))) / 2), 
              density.info = "none", lwid = c(1, 10))
    dev.off()
  }, error = function(e) {})
}
  
```

We see that for 6-mers (top), positions 34-38 on shRNAs are associated with decreased mRNA stability. It is very similar for 7-mers (bottom), except that the region that seems to mediate the off-target effects is from position 34 to 37 instead of 34 to 38.

# Identify positions of shRNAs that are predictive of stability changes (Fisher's exact test)

Using a Fisher's test to determine for each position, if there are more significant associations (FDR < 0.05) that we would expect by chance. Repeat separately for the coefficients of 6-mers and 7-mers. Then in the regression to correct stability estimates, only use the 6-mers and 7-mers of these positions.

```{r, eval = F}

<==
wd = "/Users/gabrielleperron/Documents/R_workdir_ENCODE"
setwd(wd)

############## 6 mers sig positions ################

# Load p-values from above regression
pvals_files <- list.files(path = wd, pattern = "_regressionCorrectionStabShRNA6nts_pvals.csv", full.names = T)
pvals_data <- lapply(pvals_files, fread)
head(pvals_data[[1]])
dim(pvals_data[[1]])
pvals_data <- lapply(pvals_data, function(x) as.vector(as.data.frame(x)[,3]))
head(pvals_data[[1]])


# Fill in the vectors so that they have equal lengths (add NAs)
max_len <- as.numeric(max(names(table(unlist(lapply(pvals_data, length))))))
for (i in 1:length(pvals_data))
{
  init_len <- length(pvals_data[[i]])
  nb_missing <- max_len - init_len
  pvals_data[[i]] <- c(pvals_data[[i]], rep(NA, nb_missing))
}
table(unlist(lapply(pvals_data, length)))

# Bind pvalue vectors from each sample in a matrix
pvals <- do.call(rbind, pvals_data)
head(pvals[,1:5])
dim(pvals)

# FDR correction
pvals_vector <- c(pvals)
pvals_vector_FDR <- p.adjust(pvals_vector, method="fdr")
FDR <- matrix(pvals_vector_FDR, nrow = nrow(pvals), ncol = ncol(pvals))
head(FDR[,1:5])

# Replace sig adjusted pvals with 1, rest with 0
FDR_bin <- FDR
FDR_bin[FDR_bin < 0.5] <- 1
FDR_bin[FDR >= 0.5] <- 0
FDR_bin[is.na(FDR)] <- 0
head(FDR_bin[,1:5])

# For each position, check if there are more 1s than expect by chance using a Fisher's test
fisher_pval <- rep(NA, max_len)
fisher_or <- rep(NA, max_len)
for(x in 1:max_len)
{
  nb_sig_posX <- sum(FDR_bin[, x] == 1, na.rm = T)
  nb_sig_otherPos <- sum(FDR_bin[, -(x)] == 1, na.rm = T)
  nb_NS_posX <- sum(FDR_bin[, x] == 0, na.rm = T)
  nb_NS_otherPos <- sum(FDR_bin[, -(x)] == 0, na.rm = T)
  cont_tbl <- cbind(c(nb_sig_posX, nb_sig_otherPos), c(nb_NS_posX, nb_NS_otherPos))
  fisher_pval[x] <- fisher.test(cont_tbl)$p.value
  fisher_or[x] <- fisher.test(cont_tbl)$estimate
}

# Check positions that have significantly MORE hits than expected by random
sig_pos_6mers <- which(fisher_pval < 0.05 & fisher_or > 1)


############## 7 mers sig positions ################

# Load p-values from above regression
pvals_files <- list.files(path = wd, pattern = "_regressionCorrectionStabShRNA7nts_pvals.csv", full.names = T)
pvals_data <- lapply(pvals_files, fread)
head(pvals_data[[1]])
dim(pvals_data[[1]])
pvals_data <- lapply(pvals_data, function(x) as.vector(as.data.frame(x)[,3]))
head(pvals_data[[1]])

# Fill in the vectors so that they have equal lengths (add NAs)
max_len <- as.numeric(max(names(table(unlist(lapply(pvals_data, length))))))
for (i in 1:length(pvals_data))
{
  init_len <- length(pvals_data[[i]])
  nb_missing <- max_len - init_len
  pvals_data[[i]] <- c(pvals_data[[i]], rep(NA, nb_missing))
}
table(unlist(lapply(pvals_data, length)))

# Bind pvalue vectors from each sample in a matrix
pvals <- do.call(rbind, pvals_data)
head(pvals[,1:5])
dim(pvals)

# FDR correction
pvals_vector <- c(pvals)
pvals_vector_FDR <- p.adjust(pvals_vector, method="fdr")
FDR <- matrix(pvals_vector_FDR, nrow = nrow(pvals), ncol = ncol(pvals))
head(FDR[,1:5])

# Replace sig adjusted pvals with 1, rest with 0
FDR_bin <- FDR
FDR_bin[FDR_bin < 0.5] <- 1
FDR_bin[FDR >= 0.5] <- 0
FDR_bin[is.na(FDR)] <- 0
head(FDR_bin[,1:5])

# For each position, check if there are more 1s than expect by chance using a Fisher's test
fisher_pval <- rep(NA, max_len)
fisher_or <- rep(NA, max_len)
for(x in 1:max_len)
{
  nb_sig_posX <- sum(FDR_bin[, x] == 1, na.rm = T)
  nb_sig_otherPos <- sum(FDR_bin[, -(x)] == 1, na.rm = T)
  nb_NS_posX <- sum(FDR_bin[, x] == 0, na.rm = T)
  nb_NS_otherPos <- sum(FDR_bin[, -(x)] == 0, na.rm = T)
  cont_tbl <- cbind(c(nb_sig_posX, nb_sig_otherPos), c(nb_NS_posX, nb_NS_otherPos))
  fisher_pval[x] <- fisher.test(cont_tbl)$p.value
  fisher_or[x] <- fisher.test(cont_tbl)$estimate
}

# Check positions that have significantly MORE hits than expected by random
sig_pos_7mers <- which(fisher_pval < 0.05 & fisher_or > 1)
```

# Correct stability estimates for shRNA off-target effects

Using a regression with 6-mer and 7-mer counts from the positions associated with off-target effects as predictors, retain residuals as corrected stability data

```{r}

############## Stability correction ################

library(Biostrings)
library(stringr)
library(seqRFLP)

# Load RNAi sequences file
RNAi_seqs <- read.table("RNAi_sequences.txt", header = T)

## Format the Experiment accession
RNAi_seqs$Experiment <- gsub("/experiments/", "", RNAi_seqs$Experiment)
RNAi_seqs$Experiment <- gsub("/", "", RNAi_seqs$Experiment)

## Format the RNAi sequence
RNAi_seqs$RNAi_sequence <- gsub("-", "", RNAi_seqs$RNAi_sequence)

# Load 3'UTR sequences
UTR_seq <- as.character(readDNAStringSet("/Users/gabrielleperron/Documents/feature_tables/3UTR_seq.fa"))
names(UTR_seq) <- gsub(".*gene=", "", names(UTR_seq))

# Trim first 2000bps of each 3'UTRs
UTR_seq <- unlist(lapply(UTR_seq, function(x) substring(x, 1, 2000)))

# Load bias-corrected stability estimates
  
stab_HepG2 <- read.table("residuals_ddf_dfi_batches_HepG2_ROBUST.txt", header=TRUE, row.names=1)
stab_K562 <- read.table("residuals_ddf_dfi_batches_K562_ROBUST.txt", header=TRUE, row.names=1)

# Filter stab and UTR seq tables to keep same genes in same order
common_genes_HepG2 <- intersect(row.names(stab_HepG2), names(UTR_seq))
common_genes_K562 <- intersect(row.names(stab_K562), names(UTR_seq))

UTR_HepG2_fltr <- UTR_seq[common_genes_HepG2]
stab_HepG2_fltr <- stab_HepG2[common_genes_HepG2, ]

UTR_K562_fltr <- UTR_seq[common_genes_K562]
stab_K562_fltr <- stab_K562[common_genes_K562, ]

# Create a corrected stability table
corr_stab_HepG2 <- stab_HepG2_fltr
corr_stab_HepG2[] <- NA

corr_stab_K562 <- stab_K562_fltr
corr_stab_K562[] <- NA

# For each sample, get bias-corrected stab and Use a 6nt-length sliding window to get all 6nt sequences in the shRNA and count their occurence in the 3'UTRs

for (cell in c("HepG2", "K562"))
{
  # Define UTR seq set
  UTRs <- get(paste("UTR_", cell, "_fltr", sep=""))
  
  # Define stab set
  stab <- get(paste("stab_", cell, "_fltr", sep=""))

  # Define corrected stab table
  corr_stab <- get(paste("corr_stab_", cell, sep=""))
  
  # For each sample
  for (i in 1:ncol(stab))
  {
    # Define the experiment
    experiment <- gsub("_.*","", colnames(stab)[i])
    
    # If the experiment is not in the RNAi seq table, put stab as NA and remove after this part
    if (!(experiment %in% RNAi_seqs$Experiment))
    {
      corr_stab[,i] <- NA
      assign(paste("corr_stab_", cell, sep=""), corr_stab)
      next
      print("Experiment not found in the RNAi seq table")
    }
    
    # Load the count tables for 6-mers and 7-mers
    counts_6mers <- read.csv(paste(experiment, "_shRNA_6nts_counts.csv", sep=""), row.names=1)
    counts_7mers <- read.csv(paste(experiment, "_shRNA_7nts_counts.csv", sep=""), row.names=1)
    
    # Filter for the sig positions, if they exist
    counts_6mers <- counts_6mers[, which((1:ncol(counts_6mers)) %in% sig_pos_6mers)]
    counts_7mers <- counts_7mers[, which((1:ncol(counts_7mers)) %in% sig_pos_7mers)]
    
    # Merge 6-mers and 7-mers tables
    counts_mers <- cbind(counts_6mers, counts_7mers)

    # Make sure that mer counts and stab tables have same genes in same order
    if(all(row.names(counts_mers) != row.names(stab)))
    {
      print("ERROR: Different genes/order in stability and 6-mer counts tables")
      break
    }
    
    # Perform glm on sample stab data using this as predictor
    fit <- glm(stab[, i] ~ data.matrix(counts_mers), na.action=na.exclude)
    
    # Put the residuals in their table, same for coeffs and pvals
    corr_stab[,i] <- residuals(fit)
    assign(paste("corr_stab_", cell, sep=""), corr_stab)
    
    coeffs <- as.data.frame(coef(fit)[-1])
    row.names(coeffs) <- gsub(".*)", "", row.names(coeffs))
    pvals <- as.data.frame(coef(summary(fit))[,4][-1])
    row.names(pvals) <- gsub(".*)", "", row.names(pvals))
  }
}

# Remove samples with all NAs as corrected stability (RNAi sequence not available)
corr_stab_HepG2 <- corr_stab_HepG2[ , colSums(is.na(corr_stab_HepG2)) < nrow(corr_stab_HepG2)]
corr_stab_K562 <- corr_stab_K562[ , colSums(is.na(corr_stab_K562)) < nrow(corr_stab_K562)]
  
# Save corrected stab tables
write.csv(corr_stab_HepG2, file = "RBPKD_stab_biasCorr_corrShRNA6nts7nts_sigPosFishers_HepG2.csv")
write.csv(corr_stab_K562, file = "RBPKD_stab_biasCorr_corrShRNA6nts7nts_sigPosFishers_K562.csv")

```

```{r}
sessionInfo()
```
