---
title: "Inferring miRNA activity in ENCODE RBP KD samples (without correction for shRNA
  off-target effects)"
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
set.seed(1)
```

# Load libraries 

```{r}
library(tidyverse)
library(scales)
library(ggnewscale)
```

# Load bias and batch corrected stability data, 3'UTR nucleotide composition and length annotations, and miRNA binding network

```{r}
stab_HepG2 <- read.csv(here("ENCODE_RBP_KD", "output", "supplementary_tables", "ENCODE_RBP_KD_bias_batch_corrected_stability_HepG2.csv"))
stab_K562 <- read.csv(here("ENCODE_RBP_KD", "output", "supplementary_tables", "ENCODE_RBP_KD_bias_batch_corrected_stability_K562.csv"))

UTR_nt <- read.table(here("ENCODE_RBP_KD", "data", "3UTR_nt_composition.txt"), header = TRUE, row.names = NULL)
UTR_length <- read.table(here("ENCODE_RBP_KD", "data", "3UTR_length.txt"), header = TRUE, row.names = NULL)
affinity_miRNAs <- read.table(here("ENCODE_RBP_KD", "data", "3UTR_miRNA_affinity.txt"), header = TRUE, row.names = NULL)
```

# Prepare GLM data

Merge 3'UTR annotations with stability data for each cell line

```{r}
UTR <- merge(UTR_nt, UTR_length, by = 1)
UTR_affinity <- merge(UTR, affinity_miRNAs, by = 1)

glmMatrix_HepG2 <- merge(stab_HepG2, UTR_affinity, by = 1)
glmMatrix_K562 <- merge(stab_K562, UTR_affinity, by = 1)
```

# Fit GLM models 

Run GLM for each RBP KD sample, extracting coefficients and p-values

```{r}
fit_glm_sample <- function(glm_data, sample_name, cell_line) {
  # glm_data: complete data frame with response and all predictors
  # sample_name: name of the RBP KD sample 
  # cell_line: "HepG2" or "K562"
  
  response_col <- which(colnames(glm_data) == sample_name)
  predictors_confounders_cols <- which(colnames(glm_data) != "X" & !grepl("ENCSR", colnames(glm_data)))
  
  # Formula: response ~ confounders and miRNA affinity predictors
  # Intercept included automatically
  formula_str <- paste0(
    colnames(glm_data)[response_col], 
    " ~ ",
    paste(colnames(glm_data)[predictors_confounders_cols], collapse = " + ")
  )
  
  fit <- glm(as.formula(formula_str), data = glm_data, na.action = na.exclude)
  coef_summary <- coef(summary(fit))
  
  # Extract coefficients and p-values for predictors only (miRNA cols, skip intercept and confounders)
  predictor_rows <- which(grepl("^(miR\\.|let\\.)", rownames(coef_summary)))
  
  tibble(
    miRNA = rownames(coef_summary)[predictor_rows],
    coefficient = coef_summary[predictor_rows, 1],
    pvalue = coef_summary[predictor_rows, 4],
    sample_name = sample_name,
    cell_line = cell_line
  )
}

# Apply to all RBP KD samples in both cell lines
rbp_sample_names_HepG2 <- colnames(stab_HepG2)[-1]
rbp_sample_names_K562 <- colnames(stab_K562)[-1]

results_HepG2 <- map_df(rbp_sample_names_HepG2, 
                         ~fit_glm_sample(glmMatrix_HepG2, .x, "HepG2"))
results_K562 <- map_df(rbp_sample_names_K562, 
                        ~fit_glm_sample(glmMatrix_K562, .x, "K562"))

# Combine results from both cell lines
glm_results <- bind_rows(results_HepG2, results_K562) |>
  mutate(
    # Switch signs so coefficients represent miRNA activity (not target stability)
    coefficient = -coefficient,
    # Convert from ln to log2 scale
    coefficient_log2 = log2(exp(coefficient))
  )

head(glm_results)
```

# Multiple testing correction (separately for each cell line)

```{r}
glm_results <- glm_results |>
  group_by(cell_line) |>
  mutate(FDR = p.adjust(pvalue, method = "fdr")) |>
  ungroup()
```

# Reformat miRNA names

```{r}
glm_results$miRNA <- glm_results$miRNA |>
  str_extract("^((?:[^A-Z]*[A-Z]){1}[^A-Z]*)") |>
  str_remove("_.*") |>
  str_remove("\\.$") |>
  str_replace_all("\\.", "-")

glm_results$miRNA[str_detect(glm_results$miRNA, "let-7")] <- "let-7"

# Define mapping from original names to canonical names
mirna_name_mapping <- tibble(
  original = c(
    "miR-33ab", "miR-26ab", "miR-29abcd", "miR-27abc", "miR-133abc",
    "miR-199ab-5p", "miR-376abd", "miR-23abc", "miR-19ab", "miR-182",
    "miR-758", "miR-22", "miR-150", "miR-124", "miR-96", "miR-370",
    "miR-17", "miR-218", "miR-21", "miR-25", "miR-376c", "miR-10abc",
    "miR-143", "miR-34ac", "miR-338", "miR-377", "miR-495", "miR-544",
    "miR-18ab", "miR-299", "miR-328a", "let-7"
  ),
  canonical = c(
    "miR-33-5p", "miR-26-5p", "miR-29-3p", "miR-27-3p", "miR-133-3p",
    "miR-199-5p", "miR-376-3p", "miR-23-3p", "miR-19-3p", "miR-182-5p",
    "miR-758-3p", "miR-22-3p", "miR-150-5p", "miR-124-3p", "miR-96-5p",
    "miR-370-3p", "miR-17-5p", "miR-218-5p", "miR-21-5p", "miR-25-3p",
    "miR-376c-3p", "miR-10-5p", "miR-143-3p", "miR-34-5p", "miR-338-3p",
    "miR-377-3p", "miR-495-3p", "miR-544a-3p", "miR-18-5p", "miR-299-3p",
    "miR-328-3p", "let-7"
  )
)

glm_results <- glm_results |>
  left_join(mirna_name_mapping, by = c("miRNA" = "original")) |>
  mutate(miRNA = coalesce(canonical, miRNA)) |>
  select(-canonical)
```

# Reformat RBP and sample names

```{r}
extract_rbp_info <- function(sample_name) {
  # Extract RBP name and replicate info from ENCODE naming convention
  # Format: ENCSR000WWW_XX_YY_ZZ where XX is the RBP, YY is the replicate (11 or 21) and ZZ is cell line code
  
  parts <- unlist(strsplit(sample_name, "_"))
  rbp_id <- parts[2]
  rep_og <- parts[3]
  
  # Parse replicate number
  rep <- case_when(
    rep_og == "11" ~ "rep1",
    rep_og == "21" ~ "rep2",
    TRUE ~ NA_character_
  )

  sample_name2 <- paste0(rbp_id, "_", rep)
  
  sample_name2
}

glm_results <- glm_results |>
  mutate(sample_name2 = map_chr(sample_name, extract_rbp_info))

glm_results$RBP <- str_split_i(glm_results$sample_name, "_", 2)
```

# Save results

```{r}
# Save all results in long-format tibble 
write.csv(glm_results, file = here("ENCODE_RBP_KD", "output", "supplementary_tables", "ENCODE_RBP_KD_miRNA_activity_noCorrOffTargetEffects.csv"), row.names = FALSE)
```

# Identify RBP-miRNA pairs with FDR < 0.05 in at least one sample of at least one cell line

```{r}
significant_interactions <- glm_results |>
  filter(FDR < 0.05) |>
  distinct(RBP, miRNA, cell_line) 

significant_interactions

nrow(unique(significant_interactions[, colnames(significant_interactions) != "cell_line"]))
```

# For these candidate interactions, verify that all activity changes estimates have the same sign across all samples where the RBP is KD, within the corresponding cell line, and retrieve the types of interactions

```{r}
sign_check <- significant_interactions |>
  inner_join(
    glm_results |>
      select(RBP, miRNA, cell_line, coefficient),
    by = c("RBP", "miRNA", "cell_line")
  ) |>
  group_by(RBP, miRNA, cell_line) |>
  summarise(
    n = n(),
    n_signs = n_distinct(sign(coefficient)),
    all_same_sign = n_signs == 1,
    Type = case_when(
      # If miRNA activity is up when RBP is KD, it means that the RBP normally inhibits the miRNA
      all_same_sign & first(sign(coefficient)) > 0 ~ "-",
      # If miRNA activity is down when RBP is KD, it means that the RBP normally increases the activity of the miRNA
      all_same_sign & first(sign(coefficient)) < 0 ~ "+",
      TRUE ~ "mixed"
    ),
    .groups = "drop"
  )

sign_check

stopifnot(all(sign_check$all_same_sign == T))

candidate_interactions <- sign_check[sign_check$all_same_sign == T, c("RBP", "miRNA", "cell_line", "Type")]

saveRDS(candidate_interactions, here("ENCODE_RBP_KD", "output", "rds", "ENCODE_RBP_KD_initial_interactions.rds"))
```

# Prepare data for heatmap

```{r}
# Helper function to create sample × miRNA matrices
make_matrix <- function(data, value_col) {
  data |>
    select(sample_name, miRNA, {{ value_col }}) |>
    pivot_wider(
      names_from = miRNA,
      values_from = {{ value_col }},
      values_fn = mean,
      values_fill = NA
    ) |>
    column_to_rownames("sample_name") |>
    as.matrix()
}

# Matrix of all coefficients
heatmap_all <- make_matrix(glm_results, coefficient)

# Matrix for display: non-significant interactions set to zero
significant_pairs <- candidate_interactions |>
  distinct(RBP, miRNA)

heatmap_display <- glm_results |>
  left_join(
    significant_pairs |> mutate(significant = TRUE),
    by = c("RBP", "miRNA")
  ) |>
  mutate(
    coefficient = if_else(is.na(significant), 0, coefficient)
  ) |>
  select(sample_name, miRNA, coefficient) |>
  pivot_wider(
    names_from = miRNA,
    values_from = coefficient,
    values_fn = mean,
    values_fill = 0
  ) |>
  column_to_rownames("sample_name") |>
  as.matrix()

# Remove rows and columns with no significant interactions
keep_rows <- rowSums(heatmap_display, na.rm = TRUE) != 0
keep_cols <- colSums(heatmap_display, na.rm = TRUE) != 0

heatmap_display <- heatmap_display[keep_rows, keep_cols]
heatmap_all <- heatmap_all[rownames(heatmap_display), colnames(heatmap_display)]

# Matrix of signed log10 p-values for clustering
signed_log_pval_matrix <- glm_results |>
  mutate(
    signed_log10_pval = abs(log10(pvalue)) * sign(coefficient)
  ) |>
  make_matrix(signed_log10_pval)

signed_log_pval_matrix <- signed_log_pval_matrix[
  rownames(heatmap_display),
  colnames(heatmap_display)
]

# Cluster samples using signed log10 p-values
dist_matrix <- as.dist((1 - cor(t(signed_log_pval_matrix))) / 2)
hc <- hclust(dist_matrix, method = "complete")

# Reorder rows by clustering
dat_clustered <- heatmap_all[rev(hc$order), ]

# Sort columns by row of maximum absolute coefficient
col_max_row <- apply(abs(dat_clustered), 2, which.max)
dat_sorted <- dat_clustered[, order(col_max_row)]

# Create mapping of ENCODE names to display labels
row_label_map <- glm_results |>
  select(sample_name, sample_name2) |>
  distinct() |>
  deframe()

ordered_labels <- row_label_map[rownames(dat_sorted)]

# Prepare plotting data
plot_data <- glm_results |>
  filter(
    !is.na(miRNA),
    sample_name %in% rownames(dat_sorted),
    miRNA %in% colnames(dat_sorted)
  ) |>
  mutate(
    coefficient_log2 = pmin(abs(coefficient_log2), 0.35),
    signed_log10_pval = pmax(
      pmin(abs(log10(pvalue)) * sign(coefficient), 5),
      -5
    ),
    sample_name = factor(sample_name, levels = rownames(dat_sorted)),
    miRNA = factor(miRNA, levels = colnames(dat_sorted))
  )
```

# Create heatmap

```{r, fig.dim = c(15, 35), fig.align = 'center'}
row_annotation <- plot_data |>
  distinct(sample_name, cell_line)

annot_x <- length(levels(plot_data$miRNA)) + 1

p <- ggplot() +

  # Heatmap
  geom_point(
    data = plot_data,
    aes(
      x = miRNA,
      y = fct_rev(sample_name),
      size = coefficient_log2,
      fill = signed_log10_pval
    ),
    shape = 21,
    color = "transparent"
  ) +
  
  scale_size(
    "miRNA activity",
    range = c(0.5, 8),
    breaks = seq(0, 0.35, 0.05),
    limits = c(0, 0.35),
    guide = guide_legend(
      override.aes = list(
        shape = 21,
        fill = "grey70",
        colour = "black",
        stroke = 0.3
      )
    )
  ) + 

  scale_fill_gradient2(
    "Signed log10(P-value)",
    limits = c(-5, 5),
    low = "blue",
    mid = "white",
    high = "red"
  ) +

  ggnewscale::new_scale_fill() +

  # Row annotation
  geom_tile(
    data = row_annotation,
    aes(
      x = annot_x,
      y = fct_rev(sample_name),
      fill = cell_line
    ),
    width = 0.8,      
    height = 0.95
  ) + 

  scale_fill_manual(
    "Cell line",
    values = c(
      HepG2 = "#9ecae1",  # light blue
      K562  = "#fdae6b"   # orange
    )
  ) +

  scale_x_discrete(
    limits = c(levels(plot_data$miRNA), "")
  ) +

  scale_y_discrete(labels = rev(ordered_labels)) +

  labs(
    x = "miRNAs",
    y = "RBP KD samples"
  ) +

  theme(
    legend.key.size = unit(1.5, "lines"),
    text = element_text(size = 20),
    axis.text = element_text(size = 14),
    axis.text.x = element_text(angle = 90, size = 16, vjust = 0.5, hjust = 1),
    legend.position = "left",
    panel.background = element_rect(fill = "white", colour = "white")
  )
p

pdf(here("ENCODE_RBP_KD", "output", "figures", 
         "ENCODE_RBP_KD_interactions_noCorrOffTargetEffects_heatmap_allRepsAndCellsOfsigHits.pdf"),
    width = 15, height = 35)
p
dev.off()
```

# Create dendrogram figure

```{r}
pdf(here("ENCODE_RBP_KD", "output", "figures",
         "ENCODE_RBP_KD_interactions_noCorrOffTargetEffects_heatmap_allRepsAndCellsOfsigHits_dendrogram.pdf"),
    width = 10, height = 3)

plot(hc, hang = -1, cex.lab = 0.5, cex = 0.2, xlab = "")

dev.off()
```

```{r}
sessionInfo()
```
