---
title: "Inferring miRNA activity in ENCODE RBP KD samples (without correction for shRNA
  off-target effects)"
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
set.seed(1)
```

# Load libraries 

```{r}
library(gplots)
library(plyr)
library(scales)
library(reshape2)
library(ggplot2)
```

# Load bias and batch corrected mRNA stability data 

```{r}
stab_HepG2 <- read.csv(here("ENCODE_RBP_KD", "output", "supplementary_tables", "ENCODE_RBP_KD_bias_batch_corrected_stability_HepG2.csv"))
stab_K562 <- read.csv(here("ENCODE_RBP_KD", "output", "supplementary_tables", "ENCODE_RBP_KD_bias_batch_corrected_stability_K562.csv"))
```

# Load 3'UTR length and nucleotide composition annotations

```{r}
UTR_nt <- read.table(here("ENCODE_RBP_KD", "data", "3UTR_nt_composition.txt"), header=TRUE, row.names=NULL)
UTR_length <- read.table(here("ENCODE_RBP_KD", "data", "3UTR_length.txt"), header=TRUE, row.names=NULL)
```

# Load miRNA-mRNA predicted binding network

```{r}
affinity_miRNAs <- read.table(here("ENCODE_RBP_KD", "data", "3UTR_miRNA_affinity.txt"), header=TRUE, row.names=NULL)
```

# Merge the above tables based on gene IDs, separately for each cell line

```{r}
UTR <- merge(UTR_nt, UTR_length, by=1)
UTR_affinity <- merge(UTR, affinity_miRNAs, by=1)
glmMatrix_HepG2 <- merge(stab_HepG2, UTR_affinity, by=1)
glmMatrix_K562 <- merge(stab_K562, UTR_affinity, by=1)

```

# Create tables where the results from glm will be stored

```{r}
pval_HepG2 <- matrix(NA, nrow=ncol(stab_HepG2)-1, ncol=ncol(UTR_affinity)-1)
rownames(pval_HepG2) = colnames(stab_HepG2)[-1]
colnames(pval_HepG2) = colnames(UTR_affinity)[-1]

pval_K562 <- matrix(NA, nrow=ncol(stab_K562)-1, ncol=ncol(UTR_affinity)-1)
rownames(pval_K562) = colnames(stab_K562)[-1]
colnames(pval_K562) = colnames(UTR_affinity)[-1]

coeff_HepG2 <- pval_HepG2
coeff_K562 <- pval_K562
```

# In each RBP KD sample, infer miRNA activity

Run glm where the response variable is the mRNA stability, the predictor variable is the miRNA-mRNA predicted binding network, and the confounding variables are 3'UTR_length, 3'UTR nucleotide and dinucleotide composition

```{r}
for (i in seq_len(ncol(stab_HepG2))[-1]){
  fit_HepG2 <- glm(as.matrix(glmMatrix_HepG2[,i]) ~ as.matrix(glmMatrix_HepG2[,413:ncol(glmMatrix_HepG2)]), na.action=na.exclude)
  pval_HepG2[i-1,] <- as.vector(coef(summary(fit_HepG2))[,4])[-1]
  coeff_HepG2[i-1,] <- as.vector(coef(summary(fit_HepG2))[,1])[-1]
}

for (i in seq_len(ncol(stab_K562))[-1]){
  fit_K562 <- glm(as.matrix(glmMatrix_K562[,i]) ~ as.matrix(glmMatrix_K562[,426:ncol(glmMatrix_K562)]), na.action=na.exclude)
  pval_K562[i-1,] <- as.vector(coef(summary(fit_K562))[,4])[-1]
  coeff_K562[i-1,] <- as.vector(coef(summary(fit_K562))[,1])[-1]
}
```

# Discard 3'UTR variables that are no longer needed

Retain only the miRNA binding variables 

```{r}
pval_HepG2 <- pval_HepG2[,22:ncol(pval_HepG2)]
pval_K562 <- pval_K562[,22:ncol(pval_K562)]

coeff_HepG2 <- coeff_HepG2[,22:ncol(coeff_HepG2)]
coeff_K562 <- coeff_K562[,22:ncol(coeff_K562)]
```

# Multiple testing correction

```{r}
pval_HepG2_vector <- c(pval_HepG2)
pval_HepG2_vector_FDR <- p.adjust(pval_HepG2_vector, method="fdr")
FDR_HepG2 <- matrix(pval_HepG2_vector_FDR, nrow = nrow(pval_HepG2),ncol = ncol(pval_HepG2))

pval_K562_vector <- c(pval_K562)
pval_K562_vector_FDR <- p.adjust(pval_K562_vector, method="fdr")
FDR_K562 <- matrix(pval_K562_vector_FDR, nrow = nrow(pval_K562),ncol = ncol(pval_K562))

colnames(FDR_HepG2) <- colnames(pval_HepG2)
row.names(FDR_HepG2) <- row.names(pval_HepG2)

colnames(FDR_K562) <- colnames(pval_K562)
row.names(FDR_K562) <- row.names(pval_K562)
```

# Switch the signs of regression coefficients so that they represent miRNA activity and not target stability (miRNAs usually destabilize their targets)

```{r}
coeff_HepG2 <- -(coeff_HepG2)
coeff_K562 <- -(coeff_K562)
```

# Save results

```{r}
write.csv(FDR_HepG2, file="FDR_glm_HepG2.csv")
write.csv(FDR_K562, file="FDR_glm_K562.csv")

write.csv(coeff_HepG2, file="coeff_glm_HepG2.csv")
write.csv(coeff_K562, file="coeff_glm_K562.csv")

write.csv(pval_HepG2, file="pvalues_glm_HepG2.csv")
write.csv(pval_K562, file="pvalues_glm_K562.csv")

```

# Merge result matrices to have all results from both cell lines 

```{r}
coeff_HepG2andK562 <- rbind.fill(as.data.frame(coeff_HepG2), as.data.frame(coeff_K562))
names_coeff <- c(row.names(coeff_HepG2), row.names(coeff_K562))
row.names(coeff_HepG2andK562) <- names_coeff

pval_HepG2andK562 <- rbind.fill(as.data.frame(pval_HepG2), as.data.frame(pval_K562))
names_pval <- c(row.names(pval_HepG2), row.names(pval_K562))
row.names(pval_HepG2andK562) <- names_pval

FDR_HepG2andK562 <- rbind.fill(as.data.frame(FDR_HepG2), as.data.frame(FDR_K562))
names_FDR <- c(row.names(FDR_HepG2), row.names(FDR_K562))
row.names(FDR_HepG2andK562) <- names_FDR

```

# Reformating miRNA names

```{r}
mirs <- colnames(coeff_HepG2andK562)
for (m in mirs)
{
  d <- m
  d2 <- unlist(strsplit(d, "")) 
  d3 <- grep("[A-Z]", d2)[2]
  d4 <- paste0(c(d2[1:(d3-1)]), collapse="")
  d5 <- unlist(strsplit(d4, "_"))[1]
  d6 <- gsub("\\.$", "", d5)
  d7 <- gsub("\\.", "-", d6)
  mirs[mirs == m] <- d7
}
mirs[grep("let-7", mirs)] <- "let-7"
colnames(coeff_HepG2andK562) <- mirs
colnames(pval_HepG2andK562) <- mirs
colnames(FDR_HepG2andK562) <- mirs

```

# List of RBPs that have RBP KD samples present in the data

```{r}
extract_string = function(x, n, i){
    do.call(c, lapply(x, function(X)
        paste(unlist(strsplit(X, "_"))[(n+1):(i)], collapse = "_")))
} # To get string betwen nth and ith instance of the delimiter "_", from https://stackoverflow.com/questions/44934703/r-extract-string-between-nth-and-ith-instance-of-delimiter#44934922

RBPs <- row.names(coeff_HepG2andK562)
RBPs <- extract_string(RBPs, 1, 4)
RBPs[grep("_11_", RBPs)] <- paste(RBPs[grep("_11_", RBPs)], "_rep1", sep="")
RBPs[grep("_21_", RBPs)] <- paste(RBPs[grep("_21_", RBPs)], "_rep2", sep="")
RBPs <- gsub("_21_", "_", RBPs)
RBPs <- gsub("_11_", "_", RBPs)
RBPs <- gsub("_.*", "", RBPs)

```

# List of interactions that have a significant hit in at least one replicate of at least one cell line

```{r}
list_sig <- list()
x = 1
for (r in 1:nrow(FDR_HepG2andK562))
{
  for (cl in 1:ncol(FDR_HepG2andK562))
  {
    RBP <- RBPs[r]
    miR <- mirs[cl]
    if (FDR_HepG2andK562[r, cl] < 0.05)
    {
      list_sig[[x]] <- c(RBP, miR)
      x = x + 1
    }
  }
}
list_sig <- unique(list_sig)

```

# Log-transform the p-values and create a matrix of signed log10(p-values)

Where values for any replicates of interactions that don't have at least one significant hit are replaced with 0s

```{r}
log_pval_HepG2andK562 <- log10(pval_HepG2andK562)

signs <- coeff_HepG2andK562
signs[signs < 0] <- -1
signs[signs > 0] <- 1

signed_log_pval_HepG2andK562 <- abs(log_pval_HepG2andK562) * signs

signed_log_pval_FDR <- data.matrix(signed_log_pval_HepG2andK562)
for (r in 1:nrow(signed_log_pval_HepG2andK562))
{
  for (cl in 1:ncol(signed_log_pval_HepG2andK562))
  {
    RBP <- RBPs[r]
    miR <- mirs[cl]
    int <- c(RBP, miR)
    if (!(list(int) %in% list_sig))
    {
      signed_log_pval_FDR[r, cl] <- 0
    }
  }
}

```

# Remove rows and columns that only contain 0s

```{r}
signed_log_pval_FDR <- signed_log_pval_FDR[rowSums(signed_log_pval_FDR) != 0, ]
signed_log_pval_FDR <- signed_log_pval_FDR[, colSums(signed_log_pval_FDR) != 0]
```

# The previous steps were done for filtering, now we get back all of the values

```{r}
signed_log_pval_FDR <- signed_log_pval_HepG2andK562[match(row.names(signed_log_pval_FDR), row.names(signed_log_pval_HepG2andK562)), match(colnames(signed_log_pval_FDR), colnames(signed_log_pval_HepG2andK562))]
signed_log_pval_FDR <- data.matrix(signed_log_pval_FDR)
```

# Sort the data to get a diagonal of significant hits when making the heatmap

```{r}
htmp <- heatmap.2(data.matrix(signed_log_pval_FDR), hclust=function(x) hclust(x,method="complete"), distfun=function(x) as.dist((1-cor(t(x)))/2), Colv = F, dendrogram = "row")
dat_clustered <- signed_log_pval_FDR[rev(htmp$rowInd), htmp$colInd]

# Define column maximums
col_max <- apply(abs(dat_clustered),2,max)

# Define which row has each maximum (absolute value)
which_row_col_max <- function (x)
{
  row_col_max <- as.vector(which(abs(dat_clustered) == x, arr.ind = TRUE))[1] # Row where this col max is found
  return(row_col_max)
}
col_max_row <- unlist(lapply(col_max, which_row_col_max) )

# Sort column maximums on corresponding row number, get the column order from that 
cols_sorted <- seq(1:length(col_max_row))[order(col_max_row)]

# Sort data columns based on this list
dat_clustered_sorted <- dat_clustered[, cols_sorted]

```

# Get the coefficients sorted in the same way and with the same values removed, and get the absolute value.

```{r}
coeffs_fltr_srt <- coeff_HepG2andK562[match(row.names(dat_clustered_sorted), row.names(coeff_HepG2andK562)), match(colnames(dat_clustered_sorted), colnames(coeff_HepG2andK562))]
coeffs_FDR <- data.matrix(coeffs_fltr_srt)
coeffs_FDR <- abs(coeffs_FDR)
```

# Also get back the signed log10(p-values) and sort in the same way

```{r}
signed_log_pval_HepG2andK562 <- signed_log_pval_HepG2andK562[match(row.names(dat_clustered_sorted), row.names(coeff_HepG2andK562)), match(colnames(dat_clustered_sorted), colnames(coeff_HepG2andK562))]
signed_log_pval_HepG2andK562 <- data.matrix(signed_log_pval_HepG2andK562)
```

# Melt dataframes and combine in preparation for making the heatmap

```{r}
signed_log_pval_HepG2andK562 <- as.data.frame(cbind(row.names(signed_log_pval_HepG2andK562), signed_log_pval_HepG2andK562))
row.names(signed_log_pval_HepG2andK562) <- NULL
colnames(signed_log_pval_HepG2andK562)[1] <- "RBP KD samples"
pval_melt <- melt(signed_log_pval_HepG2andK562, id.vars = "RBP KD samples")
names(pval_melt)[2:3] <- c("miRNAs", "signedLog10pval")

coeffs_FDR <- as.data.frame(cbind(row.names(coeffs_FDR), coeffs_FDR))
row.names(coeffs_FDR) <- NULL
colnames(coeffs_FDR)[1] <- "RBP KD samples"
coeffs_melt <- melt(coeffs_FDR, id.vars = "RBP KD samples")
names(coeffs_melt)[2:3] <- c("miRNAs", "coeffs")

dat_htmp <- cbind(pval_melt, coeffs_melt$coeffs)
colnames(dat_htmp)[4] <- "coeffs"
dat_htmp$signedLog10pval <- as.numeric(as.vector(dat_htmp$signedLog10pval))
dat_htmp$coeffs <- as.numeric(as.vector(dat_htmp$coeffs))

dat_htmp$`RBP KD samples` <- factor(dat_htmp$`RBP KD samples`, levels = as.character(signed_log_pval_HepG2andK562$`RBP KD samples`))
dat_htmp$miRNAs <- factor(dat_htmp$miRNAs, levels = as.character(colnames(signed_log_pval_HepG2andK562)))

```

# Reformat sample names 

```{r}
rowNames <- extract_string(as.vector(signed_log_pval_HepG2andK562$`RBP KD samples`), 1, 4)
rowNames[grep("_11_", rowNames)] <- paste(rowNames[grep("_11_", rowNames)], "_rep1", sep="")
rowNames[grep("_21_", rowNames)] <- paste(rowNames[grep("_21_", rowNames)], "_rep2", sep="")
rowNames <- gsub("_21_", "_", rowNames)
rowNames <- gsub("_11_", "_", rowNames)
rowNames <- gsub("_[^_]+_", "_", rowNames)
```

# The coefficients are represent ln(FC) of miRNA activity right now. Convert to log2 scale.

```{r}
dat_htmp$coeffs <- log2(exp(dat_htmp$coeffs))
```

# Replace with correct names of the miRNA familes

```{r}
lbls <- dat_htmp$miRNAs
lbls <- gsub("miR-33ab$", "miR-33-5p", lbls)
lbls <- gsub("miR-26ab$", "miR-26-5p", lbls)
lbls <- gsub("miR-29abcd$", "miR-29-3p", lbls)
lbls <- gsub("miR-27abc$", "miR-27-3p", lbls)
lbls <- gsub("miR-133abc$", "miR-133-3p", lbls)
lbls <- gsub("miR-199ab-5p$", "miR-199-5p", lbls)
lbls <- gsub("miR-376abd$", "miR-376-3p", lbls)
lbls <- gsub("miR-23abc$", "miR-23-3p", lbls)
lbls <- gsub("miR-19ab$", "miR-19-3p", lbls)
lbls <- gsub("miR-182$", "miR-182-5p", lbls)
lbls <- gsub("miR-758$", "miR-758-3p", lbls)
lbls <- gsub("miR-22$", "miR-22-3p", lbls)
lbls <- gsub("miR-150$", "miR-150-5p", lbls)
lbls <- gsub("miR-124$", "miR-124-3p", lbls)
lbls <- gsub("miR-96$", "miR-96-5p", lbls)
lbls <- gsub("miR-370$", "miR-370-3p", lbls)
lbls <- gsub("miR-17$", "miR-17-5p", lbls)
lbls <- gsub("miR-218$", "miR-218-5p", lbls)
lbls <- gsub("miR-21$", "miR-21-5p", lbls)
lbls <- gsub("miR-25$", "miR-25-3p", lbls)
lbls <- gsub("miR-376c$", "miR-376c-3p", lbls)
lbls <- gsub("miR-10abc$", "miR-10-5p", lbls)
lbls <- gsub("miR-143$", "miR-143-3p", lbls)
lbls <- gsub("miR-34ac$", "miR-34-5p", lbls)
lbls <- gsub("miR-338$", "miR-338-3p", lbls)
lbls <- gsub("miR-377$", "miR-377-3p", lbls)
lbls <- gsub("miR-495$", "miR-495-3p", lbls)
lbls <- gsub("miR-544$", "miR-544a-3p", lbls)
lbls <- gsub("miR-18ab$", "miR-18-5p", lbls)
lbls <- gsub("miR-299$", "miR-299-3p", lbls)
lbls <- gsub("miR-328a$", "miR-328-3p", lbls)
dat_htmp$miRNAs <- lbls
```

# Set the p-value and regression coefficient limits for the heatmap color scales

```{r}
pval_lim <- 5
dat_htmp$signedLog10pval[dat_htmp$signedLog10pval > pval_lim] <- pval_lim
dat_htmp$signedLog10pval[dat_htmp$signedLog10pval < -pval_lim] <- -pval_lim
coeff_lim <- 0.35
dat_htmp$coeffs[dat_htmp$coeffs > coeff_lim] <- coeff_lim
dat_htmp$miRNAs <- factor(dat_htmp$miRNAs, levels=unique(dat_htmp$miRNAs))
```

# Make the heatmap

```{r}
pdf(here("ENCODE_RBP_KD", "output", "figures", "miRs-RBPs_heatmap_signedPvaluesANDcoeff_HepG2andK562_allRepsAndCellsOfsigHits_sorted.pdf"), width = 15, height = 35)
ggplot(dat_htmp) +
geom_point(aes(
  x = miRNAs, 
  size = coeffs, 
  fill = signedLog10pval,
  shape = 21, 
  color = "transparent",
  y = factor(`RBP KD samples`, levels = rev(as.character(signed_log_pval_HepG2andK562$`RBP KD samples`))))) +
scale_shape_identity() +
scale_colour_manual(values=c("transparent", "transparent"))+
scale_y_discrete(label = rev(rowNames)) +
scale_fill_gradient2(limits=c(-pval_lim,pval_lim), low="blue", mid="white", high="red") +  # oob=squish
scale_size("miRNA activity",range = c(0.5, 8), breaks = seq(0, coeff_lim, 0.05), limits=c(0, coeff_lim))+
labs(x="miRNAs",y="RBP KD samples") +
theme(legend.key.size = unit(1.5, 'lines'), text = element_text(size=20), axis.text = element_text(size=14), axis.text.x = element_text(angle=90, size=16, vjust=0.5, hjust=1),legend.position="left", panel.background = element_rect(fill = 'white', colour = 'white'))
dev.off()
```

# And make a dendrogram for the heatmap that can be manually added

```{r}
pdf(here("ENCODE_RBP_KD", "output", "figures", "miRs-RBPs_heatmap_signedPvaluesANDcoeff_HepG2andK562_allRepsAndCellsOfsigHits_sorted_dendrogram.pdf"), width = 10, height = 3)
hc <- as.hclust( htmp$rowDendrogram )
plot(hc, hang = -1, cex.lab=0.5, cex=0.2, xlab="")
dev.off()
```

```{r}
sessionInfo()
```


