---
title: 'Inferring miRNA activity in ENCODE RBP KD samples (with correction for shRNA off-target effects'
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
set.seed(1)
```

# Load libraries

```{r}
library(tidyverse)
library(scales)
```

# Summary of steps for identifying candidate RBP-miRNA interactions

  - Perform new inference of miRNA activity in RBP KD samples using stability data corrected for shRNA off-target effects
  - Consider results of all original significant hits (candidate interactions, FDR < 0.05, from before correction of stability data for shRNA off-target effects)
  - Retrieved corresponding new P-values
  - FDR correction on this subset of P-values
  - Filter for new significant hits (FDR < 0.05)
  - Identify candidate RBP-miRNA interactions: New significant hits where the other replicates are in agreement in terms of direction of interaction (sign of regression coefficients)

# Load the differential mRNA stability after correction for bias, batches, and shRNA off-target effects, 3'UTR nucleotide composition and length annotations, and miRNA seed matches

```{r}
stab_HepG2 <- read.csv(here("ENCODE_RBP_KD", "output", "supplementary_tables", "ENCODE_RBP_KD_stab_corr_bias_batch_shRNAoffTargetEffects_HepG2.csv"))
stab_K562 <- read.csv(here("ENCODE_RBP_KD", "output", "supplementary_tables", "ENCODE_RBP_KD_stab_corr_bias_batch_shRNAoffTargetEffects_K562.csv"))

UTR_nt <- read.table(here("ENCODE_RBP_KD", "data", "3UTR_nt_composition_new.txt"), header = TRUE, row.names = NULL)
UTR_length <- read.table(here("ENCODE_RBP_KD", "data", "3UTR_length_new.txt"), header = TRUE, row.names = NULL)
miRNAs_seed_matches <- read.csv(here("ENCODE_RBP_KD", "data", "miRNA_seed_matches_3UTRs2000bp.csv"), header = TRUE)
```

# Prepare GLM data

Merge 3'UTR annotations with stability data for each cell line

```{r}
UTR <- merge(UTR_nt, UTR_length, by = 1)
UTR_miRNAs <- merge(UTR, miRNAs_seed_matches, by = 1)

glmMatrix_HepG2 <- merge(stab_HepG2, UTR_miRNAs, by = 1)
glmMatrix_K562 <- merge(stab_K562, UTR_miRNAs, by = 1)
```

# Fit GLM models 

Run GLM for each RBP KD sample, extracting coefficients and p-values

```{r}
fit_glm_sample <- function(glm_data, sample_name, cell_line) {
  # glm_data: complete data frame with response and all predictors
  # sample_name: name of the RBP KD sample 
  # cell_line: "HepG2" or "K562"
  
  response_col <- which(colnames(glm_data) == sample_name)
  predictors_confounders_cols <- which(colnames(glm_data) != "X" & !grepl("ENCSR", colnames(glm_data)))
  
  # Formula: response ~ confounders and miRNA affinity predictors
  # Intercept included automatically
  formula_str <- paste0(
    colnames(glm_data)[response_col], 
    " ~ ",
    paste(colnames(glm_data)[predictors_confounders_cols], collapse = " + ")
  )
  
  fit <- glm(as.formula(formula_str), data = glm_data, na.action = na.exclude)
  coef_summary <- coef(summary(fit))
  
  # Extract coefficients and p-values for predictors only (miRNA cols, skip intercept and confounders)
  predictor_rows <- which(grepl("^(miR\\.|let\\.)", rownames(coef_summary)))
  
  tibble(
    miRNA = rownames(coef_summary)[predictor_rows],
    coefficient = coef_summary[predictor_rows, 1],
    pvalue = coef_summary[predictor_rows, 4],
    sample_name = sample_name,
    cell_line = cell_line
  )
}

# Apply to all RBP KD samples in both cell lines
rbp_sample_names_HepG2 <- colnames(stab_HepG2)[-1]
rbp_sample_names_K562 <- colnames(stab_K562)[-1]

results_HepG2 <- map_df(rbp_sample_names_HepG2, 
                         ~fit_glm_sample(glmMatrix_HepG2, .x, "HepG2"))
results_K562 <- map_df(rbp_sample_names_K562, 
                        ~fit_glm_sample(glmMatrix_K562, .x, "K562"))

# Combine results from both cell lines
glm_results <- bind_rows(results_HepG2, results_K562) |>
  mutate(
    # Switch signs so coefficients represent miRNA activity (not target stability)
    coefficient = -coefficient,
    # Convert from ln to log2 scale
    coefficient_log2 = log2(exp(coefficient))
  )

head(glm_results)
```

# Multiple testing correction (separately for each cell line)

```{r}
glm_results <- glm_results |>
  group_by(cell_line) |>
  mutate(FDR = p.adjust(pvalue, method = "fdr")) |>
  ungroup()
```

# Reformat miRNA names

```{r}
glm_results$miRNA <- gsub("\\.", "-", glm_results$miRNA)

# Define mapping from original names to canonical names
mirna_name_mapping <- tibble(
  original = c(
    "miR-33ab", "miR-26ab", "miR-29abcd", "miR-27abc", "miR-133abc",
    "miR-199ab-5p", "miR-376abd", "miR-23abc", "miR-19ab", "miR-182",
    "miR-758", "miR-22", "miR-150", "miR-124", "miR-96", "miR-370",
    "miR-17", "miR-218", "miR-21", "miR-25", "miR-376c", "miR-10abc",
    "miR-143", "miR-34ac", "miR-338", "miR-377", "miR-495", "miR-544",
    "miR-18ab", "miR-299", "miR-328a", "let-7"
  ),
  canonical = c(
    "miR-33-5p", "miR-26-5p", "miR-29-3p", "miR-27-3p", "miR-133-3p",
    "miR-199-5p", "miR-376-3p", "miR-23-3p", "miR-19-3p", "miR-182-5p",
    "miR-758-3p", "miR-22-3p", "miR-150-5p", "miR-124-3p", "miR-96-5p",
    "miR-370-3p", "miR-17-5p", "miR-218-5p", "miR-21-5p", "miR-25-3p",
    "miR-376c-3p", "miR-10-5p", "miR-143-3p", "miR-34-5p", "miR-338-3p",
    "miR-377-3p", "miR-495-3p", "miR-544a-3p", "miR-18-5p", "miR-299-3p",
    "miR-328-3p", "let-7"
  )
)

glm_results <- glm_results |>
  left_join(mirna_name_mapping, by = c("miRNA" = "original")) |>
  mutate(miRNA = coalesce(canonical, miRNA)) |>
  select(-canonical)
```

# Reformat RBP and sample names

```{r}
extract_rbp_info <- function(sample_name) {
  # Extract RBP name and replicate info from ENCODE naming convention
  # Format: ENCSR000WWW_XX_YY_ZZ where XX is the RBP, YY is the replicate (11 or 21) and ZZ is cell line code
  
  parts <- unlist(strsplit(sample_name, "_"))
  rbp_id <- parts[2]
  rep_og <- parts[3]
  
  # Parse replicate number
  rep <- case_when(
    rep_og == "11" ~ "rep1",
    rep_og == "21" ~ "rep2",
    TRUE ~ NA_character_
  )

  sample_name2 <- paste0(rbp_id, "_", rep)
  
  sample_name2
}

glm_results <- glm_results |>
  mutate(sample_name2 = map_chr(sample_name, extract_rbp_info))

glm_results$RBP <- str_split_i(glm_results$sample_name, "_", 2)
```

# Save results

```{r}
# Save all results in long-format tibble 
write.csv(glm_results, file = here("ENCODE_RBP_KD", "output", "supplementary_tables", "ENCODE_RBP_KD_miRNA_activity.csv"), row.names = FALSE)
```

# Load candidate RBP-miRNA interactions from miRNA activity inference before correction of mRNA stability estimates for shRNA off-target effects

```{r}
orig_interactions <- readRDS(here("ENCODE_RBP_KD", "output", "rds", "ENCODE_RBP_KD_initial_interactions.rds"))
```

