---
title: "mRNA stability in RBP KD data"
author: "Gabrielle Perron"
date: "January 13th, 2017"
output:
  pdf_document:
    highlight: haddock
  html_document:
    highlight: haddock
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Regression for 6-mers only

```{r, eval = F}

library(Biostrings)
library(stringr)
library(seqRFLP)

wd = "/Users/gabrielleperron/Documents/R_workdir_ENCODE"
setwd(wd)

# Load RNAi sequences file
RNAi_seqs <- read.table("RNAi_sequences.txt", header = T)

## Format the Experiment accession
RNAi_seqs$Experiment <- gsub("/experiments/", "", RNAi_seqs$Experiment)
RNAi_seqs$Experiment <- gsub("/", "", RNAi_seqs$Experiment)

## Format the RNAi sequence
RNAi_seqs$RNAi_sequence <- gsub("-", "", RNAi_seqs$RNAi_sequence)

# Load 3'UTR sequences
UTR_seq <- as.character(readDNAStringSet("/Users/gabrielleperron/Documents/feature_tables/3UTR_seq.fa"))
names(UTR_seq) <- gsub(".*gene=", "", names(UTR_seq))

# Trim first 2000bps of each 3'UTRs
UTR_seq <- unlist(lapply(UTR_seq, function(x) substring(x, 1, 2000)))

# Load bias-corrected stability estimates
  
stab_HepG2 <- read.table("residuals_ddf_dfi_batches_HepG2_ROBUST.txt", header=TRUE, row.names=1)
stab_K562 <- read.table("residuals_ddf_dfi_batches_K562_ROBUST.txt", header=TRUE, row.names=1)

# Filter stab and UTR seq tables to keep same genes in same order
common_genes_HepG2 <- intersect(row.names(stab_HepG2), names(UTR_seq))
common_genes_K562 <- intersect(row.names(stab_K562), names(UTR_seq))

UTR_HepG2_fltr <- UTR_seq[common_genes_HepG2]
stab_HepG2_fltr <- stab_HepG2[common_genes_HepG2, ]

UTR_K562_fltr <- UTR_seq[common_genes_K562]
stab_K562_fltr <- stab_K562[common_genes_K562, ]

# Create a corrected stability table
corr_stab_HepG2 <- stab_HepG2_fltr
corr_stab_HepG2[] <- NA

corr_stab_K562 <- stab_K562_fltr
corr_stab_K562[] <- NA

# For each sample, get bias-corrected stab and Use a 6nt-length sliding window to get all 6nt sequences in the shRNA and count their occurence in the 3'UTRs

for (cell in c("HepG2", "K562"))
{
  # Define UTR seq set
  UTRs <- get(paste("UTR_", cell, "_fltr", sep=""))
  
  # Define stab set
  stab <- get(paste("stab_", cell, "_fltr", sep=""))

  # Define corrected stab table
  corr_stab <- get(paste("corr_stab_", cell, sep=""))
  
  # For each sample
  for (i in 1:ncol(stab))
  {
    print(i)
    # Define the experiment
    experiment <- gsub("_.*","", colnames(stab)[i])
    
    # If the experiment is not in the RNAi seq table, put stab as NA and remove after this part
    if (!(experiment %in% RNAi_seqs$Experiment))
    {
      corr_stab[,i] <- NA
      assign(paste("corr_stab_", cell, sep=""), corr_stab)
      next
      print("Experiment not found in the RNAi seq table")
    }
    
    # Define the shRNA sequence
    shRNA <- RNAi_seqs[RNAi_seqs$Experiment == experiment, ]$RNAi_sequence
    
    # Use a 6nt-length sliding window to get all 6nt sequences in the shRNAs
    k <- 6
    n <- nchar(shRNA)
    mers <- substring(shRNA, 1:(n-k+1), k:n)
    
    # Reverse complement to get target sequences
    mers <- unlist(lapply(mers, revComp))
    
    # Make a table with the genes as rows and the 6nts as columns
    mers_counts <- as.data.frame(matrix(NA, ncol = length(mers), nrow=nrow(stab)))
    row.names(mers_counts) <- row.names(stab)
    colnames(mers_counts) <- mers
    
    # Fill the table with counts of each 6nt in the 3'UTR sequences for each gene
    for (g in 1:nrow(mers_counts))
    {
      UTR <- UTRs[row.names(mers_counts)[g]]
      for (mer in colnames(mers_counts))
      {
        mers_counts[g, mer] <- str_count(UTR, mer)
      }
    }
    
    # Delete columns with same 6nt sequence, but record their position and 6mer first
    del_cols <- which(duplicated(colnames(mers_counts)))
    del_cols_mers <- colnames(mers_counts)[duplicated(colnames(mers_counts))]
    mers_counts_2 <- mers_counts[, !(duplicated(colnames(mers_counts)))]
    
    # Save the table
    write.csv(mers_counts_2, file=paste(experiment, "_shRNA_6nts_counts.csv", sep=""))
    
    # Throw error if there are columns with all 0s
    # del_cols <- sort(c(del_cols, which(colSums(mers_counts_2) == 0)))
    # mers_counts_2 <- mers_counts_2[, colSums(mers_counts_2) > 0]
    if (sum(colSums(mers_counts_2) == 0) > 0)
    {
      print("ERROR: Some 6mers don't occur in some UTR(s)")
      break
    }
    
    # Perform glm on sample stab data using this as predictor
    fit <- glm(stab[, i] ~ data.matrix(mers_counts_2), na.action=na.exclude)
    
    # Put the residuals in their table, same for coeffs and pvals
    corr_stab[,i] <- residuals(fit)
    assign(paste("corr_stab_", cell, sep=""), corr_stab)
    
    coeffs <- as.data.frame(coef(fit)[-1])
    row.names(coeffs) <- gsub(".*)", "", row.names(coeffs))
    pvals <- as.data.frame(coef(summary(fit))[,4][-1])
    row.names(pvals) <- gsub(".*)", "", row.names(pvals))
    
    # Put back 6mers that were deleted, for coeffs and pvals
    coeffs_complete <- cbind(row.names(coeffs), coeffs[,1])
    coeffs_complete_names <- row.names(coeffs)
    
    pvals_complete <- cbind(row.names(pvals), pvals[,1])
    pvals_complete_names <- row.names(pvals)
    
    if (length(del_cols) > 0)
    {
      for (x in 1:length(del_cols))
      {
        coeffs_complete_names <- c(coeffs_complete_names[1:(del_cols[x]-1)], del_cols_mers[x], coeffs_complete_names[(del_cols[x]):length(coeffs_complete_names)])
        coeffs_complete <- rbind(coeffs_complete[1:(del_cols[x]-1), ], coeffs_complete[coeffs_complete[,1] == del_cols_mers[x], ], coeffs_complete[(del_cols[x]):nrow(coeffs_complete), ])
      }
      
      for (x in 1:length(del_cols))
      {
        pvals_complete_names <- c(pvals_complete_names[1:(del_cols[x]-1)], del_cols_mers[x], pvals_complete_names[(del_cols[x]):length(pvals_complete_names)])
        pvals_complete <- rbind(pvals_complete[1:(del_cols[x]-1), ], pvals_complete[pvals_complete[,1] == del_cols_mers[x], ], pvals_complete[(del_cols[x]):nrow(pvals_complete), ])
      }      
    }
    
    # Save the coeffs and pvals
    write.csv(coeffs_complete, file=paste(colnames(stab)[i], "_regressionCorrectionStabShRNA6nts_coeffs.csv", sep=""))
    write.csv(pvals_complete, file=paste(colnames(stab)[i], "_regressionCorrectionStabShRNA6nts_pvals.csv", sep=""))
  }
}

# Remove samples with all NAs as corrected stability (RNAi sequence not available)
corr_stab_HepG2 <- corr_stab_HepG2[ , colSums(is.na(corr_stab_HepG2)) < nrow(corr_stab_HepG2)]
corr_stab_K562 <- corr_stab_K562[ , colSums(is.na(corr_stab_K562)) < nrow(corr_stab_K562)]
  
# Save corrected stab tables
# write.csv(corr_stab_HepG2, file = "RBPKD_stab_biasCorr_corrShRNA6nts_HepG2.csv")
# write.csv(corr_stab_K562, file = "RBPKD_stab_biasCorr_corrShRNA6nts_K562.csv")

```

### Regression for 7-mers only

```{r, eval = F}

library(Biostrings)
library(stringr)
library(seqRFLP)

wd = "/Users/gabrielleperron/Documents/R_workdir_ENCODE"
setwd(wd)

# Load RNAi sequences file
RNAi_seqs <- read.table("RNAi_sequences.txt", header = T)

## Format the Experiment accession
RNAi_seqs$Experiment <- gsub("/experiments/", "", RNAi_seqs$Experiment)
RNAi_seqs$Experiment <- gsub("/", "", RNAi_seqs$Experiment)

## Format the RNAi sequence
RNAi_seqs$RNAi_sequence <- gsub("-", "", RNAi_seqs$RNAi_sequence)

# Load 3'UTR sequences
UTR_seq <- as.character(readDNAStringSet("/Users/gabrielleperron/Documents/feature_tables/3UTR_seq.fa"))
names(UTR_seq) <- gsub(".*gene=", "", names(UTR_seq))

# Trim first 2000bps of each 3'UTRs
UTR_seq <- unlist(lapply(UTR_seq, function(x) substring(x, 1, 2000)))

# Load bias-corrected stability estimates
  
stab_HepG2 <- read.table("residuals_ddf_dfi_batches_HepG2_ROBUST.txt", header=TRUE, row.names=1)
stab_K562 <- read.table("residuals_ddf_dfi_batches_K562_ROBUST.txt", header=TRUE, row.names=1)

# Filter stab and UTR seq tables to keep same genes in same order
common_genes_HepG2 <- intersect(row.names(stab_HepG2), names(UTR_seq))
common_genes_K562 <- intersect(row.names(stab_K562), names(UTR_seq))

UTR_HepG2_fltr <- UTR_seq[common_genes_HepG2]
stab_HepG2_fltr <- stab_HepG2[common_genes_HepG2, ]

UTR_K562_fltr <- UTR_seq[common_genes_K562]
stab_K562_fltr <- stab_K562[common_genes_K562, ]

# Create a corrected stability table
corr_stab_HepG2 <- stab_HepG2_fltr
corr_stab_HepG2[] <- NA

corr_stab_K562 <- stab_K562_fltr
corr_stab_K562[] <- NA

# For each sample, get bias-corrected stab and Use a 7nt-length sliding window to get all 7nt sequences in the shRNA and count their occurence in the 3'UTRs

for (cell in c("HepG2", "K562"))
{
  # Define UTR seq set
  UTRs <- get(paste("UTR_", cell, "_fltr", sep=""))
  
  # Define stab set
  stab <- get(paste("stab_", cell, "_fltr", sep=""))

  # Define corrected stab table
  corr_stab <- get(paste("corr_stab_", cell, sep=""))
  
  # For each sample
  for (i in 1:ncol(stab))
  {
    print(i)
    # Define the experiment
    experiment <- gsub("_.*","", colnames(stab)[i])
    
    # If the experiment is not in the RNAi seq table, put stab as NA and remove after this part
    if (!(experiment %in% RNAi_seqs$Experiment))
    {
      corr_stab[,i] <- NA
      assign(paste("corr_stab_", cell, sep=""), corr_stab)
      next
      print("Experiment not found in the RNAi seq table")
    }
    
    # Define the shRNA sequence
    shRNA <- RNAi_seqs[RNAi_seqs$Experiment == experiment, ]$RNAi_sequence
    
    # Use a 7nt-length sliding window to get all 7nt sequences in the shRNAs
    k <- 7
    n <- nchar(shRNA)
    mers <- substring(shRNA, 1:(n-k+1), k:n)
    
    # Reverse complement to get target sequences
    mers <- unlist(lapply(mers, revComp))
    
    # Make a table with the genes as rows and the 7nts as columns
    mers_counts <- as.data.frame(matrix(NA, ncol = length(mers), nrow=nrow(stab)))
    row.names(mers_counts) <- row.names(stab)
    colnames(mers_counts) <- mers
    
    # Fill the table with counts of each 7nt in the 3'UTR sequences for each gene
    for (g in 1:nrow(mers_counts))
    {
      UTR <- UTRs[row.names(mers_counts)[g]]
      for (mer in colnames(mers_counts))
      {
        mers_counts[g, mer] <- str_count(UTR, mer)
      }
    }
    
    # Delete columns with same 7nt sequence, but record their position and 6mer first
    del_cols <- which(duplicated(colnames(mers_counts)))
    del_cols_mers <- colnames(mers_counts)[duplicated(colnames(mers_counts))]
    mers_counts_2 <- mers_counts[, !(duplicated(colnames(mers_counts)))]
    
    # Save the table
    write.csv(mers_counts_2, file=paste(experiment, "_shRNA_7nts_counts.csv", sep=""))
    
    # Throw error if there are columns with all 0s
    # del_cols <- sort(c(del_cols, which(colSums(mers_counts_2) == 0)))
    # mers_counts_2 <- mers_counts_2[, colSums(mers_counts_2) > 0]
    if (sum(colSums(mers_counts_2) == 0) > 0)
    {
      print("ERROR: Some 6mers don't occur in some UTR(s)")
      break
    }
    
    # Perform glm on sample stab data using this as predictor
    fit <- glm(stab[, i] ~ data.matrix(mers_counts_2), na.action=na.exclude)
    
    # Put the residuals in their table, same for coeffs and pvals
    corr_stab[,i] <- residuals(fit)
    assign(paste("corr_stab_", cell, sep=""), corr_stab)
    
    coeffs <- as.data.frame(coef(fit)[-1])
    row.names(coeffs) <- gsub(".*)", "", row.names(coeffs))
    pvals <- as.data.frame(coef(summary(fit))[,4][-1])
    row.names(pvals) <- gsub(".*)", "", row.names(pvals))
    
    # Put back 6mers that were deleted, for coeffs and pvals
    coeffs_complete <- cbind(row.names(coeffs), coeffs[,1])
    coeffs_complete_names <- row.names(coeffs)
    
    pvals_complete <- cbind(row.names(pvals), pvals[,1])
    pvals_complete_names <- row.names(pvals)
    
    if (length(del_cols) > 0)
    {
      for (x in 1:length(del_cols))
      {
        coeffs_complete_names <- c(coeffs_complete_names[1:(del_cols[x]-1)], del_cols_mers[x], coeffs_complete_names[(del_cols[x]):length(coeffs_complete_names)])
        coeffs_complete <- rbind(coeffs_complete[1:(del_cols[x]-1), ], coeffs_complete[coeffs_complete[,1] == del_cols_mers[x], ], coeffs_complete[(del_cols[x]):nrow(coeffs_complete), ])
      }
      
      for (x in 1:length(del_cols))
      {
        pvals_complete_names <- c(pvals_complete_names[1:(del_cols[x]-1)], del_cols_mers[x], pvals_complete_names[(del_cols[x]):length(pvals_complete_names)])
        pvals_complete <- rbind(pvals_complete[1:(del_cols[x]-1), ], pvals_complete[pvals_complete[,1] == del_cols_mers[x], ], pvals_complete[(del_cols[x]):nrow(pvals_complete), ])
      }      
    }
    
    # Save the coeffs and pvals
    write.csv(coeffs_complete, file=paste(colnames(stab)[i], "_regressionCorrectionStabShRNA7nts_coeffs.csv", sep=""))
    write.csv(pvals_complete, file=paste(colnames(stab)[i], "_regressionCorrectionStabShRNA7nts_pvals.csv", sep=""))
  }
}

# Remove samples with all NAs as corrected stability (RNAi sequence not available)
corr_stab_HepG2 <- corr_stab_HepG2[ , colSums(is.na(corr_stab_HepG2)) < nrow(corr_stab_HepG2)]
corr_stab_K562 <- corr_stab_K562[ , colSums(is.na(corr_stab_K562)) < nrow(corr_stab_K562)]
  
# Save corrected stab tables
# write.csv(corr_stab_HepG2, file = "RBPKD_stab_biasCorr_corrShRNA7nts_HepG2.csv")
# write.csv(corr_stab_K562, file = "RBPKD_stab_biasCorr_corrShRNA7nts_K562.csv")

```

### Checking the positions of shRNAs most predictive of mRNA stability

Heatmap with all samples as rows and shRNA position as column, fill with the regression coefficients from the 6nt correction above

#### For 6mers

```{r, eval = F}

library(data.table)
library(gplots)

wd = "/Users/gabrielleperron/Documents/R_workdir_ENCODE"
setwd(wd)

# Load all coeff vectors for all samples

coeff_files <- list.files(wd, pattern = "regressionCorrectionStabShRNA6nts_coeffs", full.names = T)
coeff_data <- lapply(coeff_files, fread, header=T)
coeff_data <- lapply(coeff_data, function(x) as.vector(as.data.frame(x)[,3]))
table(unlist(lapply(coeff_data, length)))
max_len <- as.numeric(max(names(table(unlist(lapply(coeff_data, length))))))

# Max shRNA length is 54, some are shorter, so fill the coeff vectors with NAs so that all have length 54

for (i in 1:length(coeff_data))
{
  init_len <- length(coeff_data[[i]])
  nb_missing <- max_len - init_len
  coeff_data[[i]] <- c(coeff_data[[i]], rep(NA, nb_missing))
}
table(unlist(lapply(coeff_data, length)))
  
# Make a result matrix with ncol = max ncol from the coeff files 
res <- do.call("rbind", coeff_data)
sample_names <- gsub(wd, "", coeff_files)
sample_names <- gsub("/", "", sample_names)
sample_names <- gsub("_regressionCorrectionStabShRNA6nts_coeffs.csv", "", sample_names)
row.names(res) <- sample_names
colnames(res) <- c(1:max_len)
  
# Replace NAs with 0s
res[is.na(res)] <- 0
  
# Make a heatmap
palette <- colorRampPalette(c("blue", "white", "red"))
pdf("ENCODE_RBPKD_samples_shRNA_position6ntsPredStab.pdf", height = 2, width = 7)
heatmap.2(res, trace="none", col=palette, cexRow=0.001, cexCol=0.5,  breaks=seq(-0.1,0.1,length.out=200), mar=c(2,2), keysize=0.5, Colv = F, 
          hclust=function(x) hclust(x,method="complete"), distfun=function(x) as.dist((1-cor(t(x)))/2), density.info = "none", key.par = list(cex=3), lwid = c(1,10))
dev.off()


################# Testing same for signed log10 pvals ###################

coeffs <- res

# Load all coeff vectors for all samples

pval_files <- list.files(wd, pattern = "regressionCorrectionStabShRNA6nts_pvals", full.names = T)
pval_data <- lapply(pval_files, fread, header=T)
pval_data <- lapply(pval_data, function(x) as.vector(as.data.frame(x)[,3]))
table(unlist(lapply(pval_data, length)))
max_len <- as.numeric(max(names(table(unlist(lapply(pval_data, length))))))

# Max shRNA length is 54, some are shorter, so fill the pval vectors with NAs so that all have length 54

for (i in 1:length(pval_data))
{
  init_len <- length(pval_data[[i]])
  nb_missing <- max_len - init_len
  pval_data[[i]] <- c(pval_data[[i]], rep(NA, nb_missing))
}
table(unlist(lapply(pval_data, length)))
  
# Make a result matrix with ncol = max ncol from the pval files 
res <- do.call("rbind", pval_data)
sample_names <- gsub(wd, "", pval_files)
sample_names <- gsub("/", "", sample_names)
sample_names <- gsub("_regressionCorrectionStabShRNA6nts_pvals.csv", "", sample_names)
row.names(res) <- sample_names
colnames(res) <- c(1:max_len)

# Transformed to signed log10
signs <- sign(coeffs)
head(signs[,1:5])
slog10pvals <- abs(log10(res))*signs
head(slog10pvals[,1:5])
  
# Replace NAs with 0s
slog10pvals[is.na(slog10pvals)] <- 0
  
# Make a heatmap
palette <- colorRampPalette(c("blue", "white", "red"))
pdf("ENCODE_RBPKD_samples_shRNA_position6ntsPredStab_signedlog10pvals.pdf", height = 2, width = 7)
heatmap.2(slog10pvals, trace="none", col=palette, cexRow=0.001, cexCol=0.5,  breaks=seq(-10,10,length.out=100), mar=c(2,2), keysize=0.5, Colv = F, 
          hclust=function(x) hclust(x,method="complete"), distfun=function(x) as.dist((1-cor(t(x)))/2), density.info = "none", key.par = list(cex=3), lwid = c(1,10))
dev.off()


################# Testing same for signed log10 FDR-corrected pvals ###################

# FDR correction
pvals_vector <- c(res)
pvals_vector_FDR <- p.adjust(pvals_vector, method="fdr")
FDR <- matrix(pvals_vector_FDR, nrow = nrow(res), ncol = ncol(res))

# Transformed to signed log10
slog10FDR <- abs(log10(FDR))*signs
  
# Replace NAs with 0s
slog10FDR[is.na(slog10FDR)] <- 0

# Make a heatmap
palette <- colorRampPalette(c("blue", "white", "red"))
pdf("ENCODE_RBPKD_samples_shRNA_position6ntsPredStab_signedlog10FDR.pdf", height = 2, width = 7)
heatmap.2(slog10FDR, trace="none", col=palette, cexRow=0.001, cexCol=0.5,  breaks=seq(-10,10,length.out=100), mar=c(2,2), keysize=0.5, Colv = F, 
          hclust=function(x) hclust(x,method="complete"), distfun=function(x) as.dist((1-cor(t(x)))/2), density.info = "none", key.par = list(cex=3), lwid = c(1,10))
dev.off()

```

See that nt postitions 34-38 on shRNA is associated with decreased mRNA stability

#### For 7mers

```{r, eval = F}

library(data.table)
library(gplots)

wd = "/Users/gperron/Documents/R_workdir_ENCODE"
setwd(wd)

# Load all coeff vectors for all samples

coeff_files <- list.files(wd, pattern = "regressionCorrectionStabShRNA7nts_coeffs", full.names = T)
coeff_data <- lapply(coeff_files, fread, header=T)
coeff_data <- lapply(coeff_data, function(x) as.vector(as.data.frame(x)[,3]))
table(unlist(lapply(coeff_data, length)))
max_len <- as.numeric(max(names(table(unlist(lapply(coeff_data, length))))))

# Max shRNA length is 54, some are shorter, so fill the coeff vectors with NAs so that all have length 54

for (i in 1:length(coeff_data))
{
  init_len <- length(coeff_data[[i]])
  nb_missing <- max_len - init_len
  coeff_data[[i]] <- c(coeff_data[[i]], rep(NA, nb_missing))
}
table(unlist(lapply(coeff_data, length)))
  
# Make a result matrix with ncol = max ncol from the coeff files 
res <- do.call("rbind", coeff_data)
sample_names <- gsub(wd, "", coeff_files)
sample_names <- gsub("/", "", sample_names)
sample_names <- gsub("_regressionCorrectionStabShRNA7nts_coeffs.csv", "", sample_names)
row.names(res) <- sample_names
colnames(res) <- c(1:max_len)
  
# Replace NAs with 0s
res[is.na(res)] <- 0
  
# Make a heatmap
palette <- colorRampPalette(c("blue", "white", "red"))
pdf("ENCODE_RBPKD_samples_shRNA_position7ntsPredStab.pdf", height = 2, width = 7)
heatmap.2(res, trace="none", col=palette, cexRow=0.001, cexCol=0.5,  breaks=seq(-0.1,0.1,length.out=200), mar=c(2,2), keysize=0.5, Colv = F, 
          hclust=function(x) hclust(x,method="complete"), distfun=function(x) as.dist((1-cor(t(x)))/2), density.info = "none", key.par = list(cex=3), lwid = c(1,10))
dev.off()


################# Testing same for signed log10 pvals ###################

coeffs <- res

# Load all coeff vectors for all samples

pval_files <- list.files(wd, pattern = "regressionCorrectionStabShRNA7nts_pvals", full.names = T)
pval_data <- lapply(pval_files, fread, header=T)
pval_data <- lapply(pval_data, function(x) as.vector(as.data.frame(x)[,3]))
table(unlist(lapply(pval_data, length)))
max_len <- as.numeric(max(names(table(unlist(lapply(pval_data, length))))))

# Max shRNA length is 54, some are shorter, so fill the pval vectors with NAs so that all have length 54

for (i in 1:length(pval_data))
{
  init_len <- length(pval_data[[i]])
  nb_missing <- max_len - init_len
  pval_data[[i]] <- c(pval_data[[i]], rep(NA, nb_missing))
}
table(unlist(lapply(pval_data, length)))
  
# Make a result matrix with ncol = max ncol from the pval files 
res <- do.call("rbind", pval_data)
sample_names <- gsub(wd, "", pval_files)
sample_names <- gsub("/", "", sample_names)
sample_names <- gsub("_regressionCorrectionStabShRNA7nts_pvals.csv", "", sample_names)
row.names(res) <- sample_names
colnames(res) <- c(1:max_len)

# Transformed to signed log10
signs <- sign(coeffs)
head(signs[,1:5])
slog10pvals <- abs(log10(res))*signs
head(slog10pvals[,1:5])
  
# Replace NAs with 0s
slog10pvals[is.na(slog10pvals)] <- 0
  
# Make a heatmap
palette <- colorRampPalette(c("blue", "white", "red"))
pdf("ENCODE_RBPKD_samples_shRNA_position7ntsPredStab_signedlog10pvals.pdf", height = 2, width = 7)
heatmap.2(slog10pvals, trace="none", col=palette, cexRow=0.001, cexCol=0.5,  breaks=seq(-10,10,length.out=100), mar=c(2,2), keysize=0.5, Colv = F, 
          hclust=function(x) hclust(x,method="complete"), distfun=function(x) as.dist((1-cor(t(x)))/2), density.info = "none", key.par = list(cex=3), lwid = c(1,10))
dev.off()


################# Testing same for signed log10 FDR-corrected pvals ###################

# FDR correction
pvals_vector <- c(res)
pvals_vector_FDR <- p.adjust(pvals_vector, method="fdr")
FDR <- matrix(pvals_vector_FDR, nrow = nrow(res), ncol = ncol(res))

# Transformed to signed log10
slog10FDR <- abs(log10(FDR))*signs
  
# Replace NAs with 0s
slog10FDR[is.na(slog10FDR)] <- 0

# Make a heatmap
palette <- colorRampPalette(c("blue", "white", "red"))
pdf("ENCODE_RBPKD_samples_shRNA_position7ntsPredStab_signedlog10FDR.pdf", height = 2, width = 7)
heatmap.2(slog10FDR, trace="none", col=palette, cexRow=0.001, cexCol=0.5,  breaks=seq(-10,10,length.out=100), mar=c(2,2), keysize=0.5, Colv = F, 
          hclust=function(x) hclust(x,method="complete"), distfun=function(x) as.dist((1-cor(t(x)))/2), density.info = "none", key.par = list(cex=3), lwid = c(1,10))
dev.off()

```

So the same heatmap made instead with stability estimates corrected for the shRNA 7mers looks like this (attached). It is very similar, except that the region that seems to mediate the off-target effets is from position 34 to 37 instead of 34 to 38.

### Correction of stability estimates, using 6mers and 7mers from positions associated with off-target effects

Using a Fisher's test to determine for each position, if there are more significant associations (FDR < 0.05) that we would expect by chance. Repeat separatelly for the coefficients of 6mers and 7mers. Then in the regression to correct stability estimates, only use the 6mers and 7mers of these positions.


```{r, eval = F}

wd = "/Users/gabrielleperron/Documents/R_workdir_ENCODE"
setwd(wd)

############## 6 mers sig positions ################

# Load p-values from above regression
pvals_files <- list.files(path = wd, pattern = "_regressionCorrectionStabShRNA6nts_pvals.csv", full.names = T)
pvals_data <- lapply(pvals_files, fread)
head(pvals_data[[1]])
dim(pvals_data[[1]])
pvals_data <- lapply(pvals_data, function(x) as.vector(as.data.frame(x)[,3]))
head(pvals_data[[1]])


# Fill in the vectors so that they have equal lengths (add NAs)
max_len <- as.numeric(max(names(table(unlist(lapply(pvals_data, length))))))
for (i in 1:length(pvals_data))
{
  init_len <- length(pvals_data[[i]])
  nb_missing <- max_len - init_len
  pvals_data[[i]] <- c(pvals_data[[i]], rep(NA, nb_missing))
}
table(unlist(lapply(pvals_data, length)))

# Bind pvalue vectors from each sample in a matrix
pvals <- do.call(rbind, pvals_data)
head(pvals[,1:5])
dim(pvals)

# FDR correction
pvals_vector <- c(pvals)
pvals_vector_FDR <- p.adjust(pvals_vector, method="fdr")
FDR <- matrix(pvals_vector_FDR, nrow = nrow(pvals), ncol = ncol(pvals))
head(FDR[,1:5])

# Replace sig adjusted pvals with 1, rest with 0
FDR_bin <- FDR
FDR_bin[FDR_bin < 0.5] <- 1
FDR_bin[FDR >= 0.5] <- 0
FDR_bin[is.na(FDR)] <- 0
head(FDR_bin[,1:5])

# For each position, check if there are more 1s than expect by chance using a Fisher's test
fisher_pval <- rep(NA, max_len)
fisher_or <- rep(NA, max_len)
for(x in 1:max_len)
{
  nb_sig_posX <- sum(FDR_bin[, x] == 1, na.rm = T)
  nb_sig_otherPos <- sum(FDR_bin[, -(x)] == 1, na.rm = T)
  nb_NS_posX <- sum(FDR_bin[, x] == 0, na.rm = T)
  nb_NS_otherPos <- sum(FDR_bin[, -(x)] == 0, na.rm = T)
  cont_tbl <- cbind(c(nb_sig_posX, nb_sig_otherPos), c(nb_NS_posX, nb_NS_otherPos))
  fisher_pval[x] <- fisher.test(cont_tbl)$p.value
  fisher_or[x] <- fisher.test(cont_tbl)$estimate
}

# Check positions that have significantly MORE hits than expected by random
sig_pos_6mers <- which(fisher_pval < 0.05 & fisher_or > 1)


############## 7 mers sig positions ################

# Load p-values from above regression
pvals_files <- list.files(path = wd, pattern = "_regressionCorrectionStabShRNA7nts_pvals.csv", full.names = T)
pvals_data <- lapply(pvals_files, fread)
head(pvals_data[[1]])
dim(pvals_data[[1]])
pvals_data <- lapply(pvals_data, function(x) as.vector(as.data.frame(x)[,3]))
head(pvals_data[[1]])

# Fill in the vectors so that they have equal lengths (add NAs)
max_len <- as.numeric(max(names(table(unlist(lapply(pvals_data, length))))))
for (i in 1:length(pvals_data))
{
  init_len <- length(pvals_data[[i]])
  nb_missing <- max_len - init_len
  pvals_data[[i]] <- c(pvals_data[[i]], rep(NA, nb_missing))
}
table(unlist(lapply(pvals_data, length)))

# Bind pvalue vectors from each sample in a matrix
pvals <- do.call(rbind, pvals_data)
head(pvals[,1:5])
dim(pvals)

# FDR correction
pvals_vector <- c(pvals)
pvals_vector_FDR <- p.adjust(pvals_vector, method="fdr")
FDR <- matrix(pvals_vector_FDR, nrow = nrow(pvals), ncol = ncol(pvals))
head(FDR[,1:5])

# Replace sig adjusted pvals with 1, rest with 0
FDR_bin <- FDR
FDR_bin[FDR_bin < 0.5] <- 1
FDR_bin[FDR >= 0.5] <- 0
FDR_bin[is.na(FDR)] <- 0
head(FDR_bin[,1:5])

# For each position, check if there are more 1s than expect by chance using a Fisher's test
fisher_pval <- rep(NA, max_len)
fisher_or <- rep(NA, max_len)
for(x in 1:max_len)
{
  nb_sig_posX <- sum(FDR_bin[, x] == 1, na.rm = T)
  nb_sig_otherPos <- sum(FDR_bin[, -(x)] == 1, na.rm = T)
  nb_NS_posX <- sum(FDR_bin[, x] == 0, na.rm = T)
  nb_NS_otherPos <- sum(FDR_bin[, -(x)] == 0, na.rm = T)
  cont_tbl <- cbind(c(nb_sig_posX, nb_sig_otherPos), c(nb_NS_posX, nb_NS_otherPos))
  fisher_pval[x] <- fisher.test(cont_tbl)$p.value
  fisher_or[x] <- fisher.test(cont_tbl)$estimate
}

# Check positions that have significantly MORE hits than expected by random
sig_pos_7mers <- which(fisher_pval < 0.05 & fisher_or > 1)


############## Stability correction ################

library(Biostrings)
library(stringr)
library(seqRFLP)

# Load RNAi sequences file
RNAi_seqs <- read.table("RNAi_sequences.txt", header = T)

## Format the Experiment accession
RNAi_seqs$Experiment <- gsub("/experiments/", "", RNAi_seqs$Experiment)
RNAi_seqs$Experiment <- gsub("/", "", RNAi_seqs$Experiment)

## Format the RNAi sequence
RNAi_seqs$RNAi_sequence <- gsub("-", "", RNAi_seqs$RNAi_sequence)

# Load 3'UTR sequences
UTR_seq <- as.character(readDNAStringSet("/Users/gabrielleperron/Documents/feature_tables/3UTR_seq.fa"))
names(UTR_seq) <- gsub(".*gene=", "", names(UTR_seq))

# Trim first 2000bps of each 3'UTRs
UTR_seq <- unlist(lapply(UTR_seq, function(x) substring(x, 1, 2000)))

# Load bias-corrected stability estimates
  
stab_HepG2 <- read.table("residuals_ddf_dfi_batches_HepG2_ROBUST.txt", header=TRUE, row.names=1)
stab_K562 <- read.table("residuals_ddf_dfi_batches_K562_ROBUST.txt", header=TRUE, row.names=1)

# Filter stab and UTR seq tables to keep same genes in same order
common_genes_HepG2 <- intersect(row.names(stab_HepG2), names(UTR_seq))
common_genes_K562 <- intersect(row.names(stab_K562), names(UTR_seq))

UTR_HepG2_fltr <- UTR_seq[common_genes_HepG2]
stab_HepG2_fltr <- stab_HepG2[common_genes_HepG2, ]

UTR_K562_fltr <- UTR_seq[common_genes_K562]
stab_K562_fltr <- stab_K562[common_genes_K562, ]

# Create a corrected stability table
corr_stab_HepG2 <- stab_HepG2_fltr
corr_stab_HepG2[] <- NA

corr_stab_K562 <- stab_K562_fltr
corr_stab_K562[] <- NA

# For each sample, get bias-corrected stab and Use a 6nt-length sliding window to get all 6nt sequences in the shRNA and count their occurence in the 3'UTRs

for (cell in c("HepG2", "K562"))
{
  # Define UTR seq set
  UTRs <- get(paste("UTR_", cell, "_fltr", sep=""))
  
  # Define stab set
  stab <- get(paste("stab_", cell, "_fltr", sep=""))

  # Define corrected stab table
  corr_stab <- get(paste("corr_stab_", cell, sep=""))
  
  # For each sample
  for (i in 1:ncol(stab))
  {
    # Define the experiment
    experiment <- gsub("_.*","", colnames(stab)[i])
    
    # If the experiment is not in the RNAi seq table, put stab as NA and remove after this part
    if (!(experiment %in% RNAi_seqs$Experiment))
    {
      corr_stab[,i] <- NA
      assign(paste("corr_stab_", cell, sep=""), corr_stab)
      next
      print("Experiment not found in the RNAi seq table")
    }
    
    # Load the count tables for 6mers and 7mers
    counts_6mers <- read.csv(paste(experiment, "_shRNA_6nts_counts.csv", sep=""), row.names=1)
    counts_7mers <- read.csv(paste(experiment, "_shRNA_7nts_counts.csv", sep=""), row.names=1)
    
    # Filter for the sig positions, if they exist
    counts_6mers <- counts_6mers[, which((1:ncol(counts_6mers)) %in% sig_pos_6mers)]
    counts_7mers <- counts_7mers[, which((1:ncol(counts_7mers)) %in% sig_pos_7mers)]
    
    # Merge 6mers and 7mers tables
    counts_mers <- cbind(counts_6mers, counts_7mers)

    # Make sure that mer counts and stab tables have same genes in same order
    if(all(row.names(counts_mers) != row.names(stab)))
    {
      print("ERROR: Different genes/order in stability and 6mer counts tables")
      break
    }
    
    # Perform glm on sample stab data using this as predictor
    fit <- glm(stab[, i] ~ data.matrix(counts_mers), na.action=na.exclude)
    
    # Put the residuals in their table, same for coeffs and pvals
    corr_stab[,i] <- residuals(fit)
    assign(paste("corr_stab_", cell, sep=""), corr_stab)
    
    coeffs <- as.data.frame(coef(fit)[-1])
    row.names(coeffs) <- gsub(".*)", "", row.names(coeffs))
    pvals <- as.data.frame(coef(summary(fit))[,4][-1])
    row.names(pvals) <- gsub(".*)", "", row.names(pvals))
  }
}

# Remove samples with all NAs as corrected stability (RNAi sequence not available)
corr_stab_HepG2 <- corr_stab_HepG2[ , colSums(is.na(corr_stab_HepG2)) < nrow(corr_stab_HepG2)]
corr_stab_K562 <- corr_stab_K562[ , colSums(is.na(corr_stab_K562)) < nrow(corr_stab_K562)]
  
# Save corrected stab tables
write.csv(corr_stab_HepG2, file = "RBPKD_stab_biasCorr_corrShRNA6nts7nts_sigPosFishers_HepG2.csv")
write.csv(corr_stab_K562, file = "RBPKD_stab_biasCorr_corrShRNA6nts7nts_sigPosFishers_K562.csv")

```

