---
title: "Comparing transcriptome changes induced CRISPR-mediated RBP knockouts with those observed upon shRNA-mediated inhibition "
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
set.seed(1)
```

# Load libraries

```{r}
library(DESeq2)
library(ggplot2)
library(EnhancedVolcano)
library(fgsea)
library(tidyverse)
library(MASS)
library(glue)
library(matrixStats)
```

# CRISPR RBP KO sample metadata 

```{r}
# Extract condition from filename
extract_condition <- function(files) {
  condition <- gsub("HEK293T-*", "", files)
  condition <- gsub("\\.GRCh38.*", "", condition)
  condition <- gsub("_.*", "", condition)
  condition[grepl("empty", condition)] <- "empty"
  condition
}

sampleFiles <- grep("GRCh38.gene", list.files(here("CRISPR_RBP_KO", "data")), value = TRUE)
sampleNames <- gsub("\\..*", "", sampleFiles)
sampleCondition <- extract_condition(sampleFiles)

sampleTable <- data.frame(
  sampleName = sampleNames,
  fileName = sampleFiles,
  condition = sampleCondition
)
sampleTable
```

# Build the DESeqDataSet 

```{r}
ddsHTSeq <- DESeqDataSetFromHTSeqCount(
  sampleTable = sampleTable,
  directory = here("CRISPR_RBP_KO", "data"),
  design = ~ condition
)
ddsHTSeq
```

# DGEA results

```{r}
ddsHTSeq$group <- factor(ddsHTSeq$condition)
design(ddsHTSeq) <- ~ group
ddsHTSeq <- DESeq(ddsHTSeq)

# Extract and name DESeq results for a given RBP
extract_deseq_results <- function(dds, rbp_name) {
  res <- as.data.frame(results(dds, contrast = c("group", rbp_name, "empty")))
  res
}

res_lst <- list(
  DDX55_KOvsEmpty = extract_deseq_results(ddsHTSeq, "DDX55"),
  PTBP1_KOvsEmpty = extract_deseq_results(ddsHTSeq, "PTBP1"),
  SUPV3L1_KOvsEmpty = extract_deseq_results(ddsHTSeq, "SUPV3L1")
)

# Save result tables
write.csv(res_lst$DDX55_KOvsEmpty, file = here("CRISPR_RBP_KO", "output", "supplementary_tables", "CRISPR_RBP_KO_DDX55_KOvsEmpty_DGEA.csv"))
write.csv(res_lst$PTBP1_KOvsEmpty, file = here("CRISPR_RBP_KO", "output", "supplementary_tables", "CRISPR_RBP_KO_PTBP1_KOvsEmpty_DGEA.csv"))
write.csv(res_lst$SUPV3L1_KOvsEmpty, file = here("CRISPR_RBP_KO", "output", "supplementary_tables", "CRISPR_RBP_KO_SUPV3L1_KOvsEmpty_DGEA.csv"))
```

# Prepare normalized counts

```{r}
# Variance-stabilized counts for visualization
vsd <- vst(ddsHTSeq, blind = TRUE)
norm_counts <- assay(vsd)
```

# PCA analysis for RBP knockdowns

```{r, fig.dim = c(4,2.5), fig.align = 'center'}
# Filter counts for a specific RBP
filter_counts <- function(counts, exclude_patterns) {

  cols_to_remove <- unique(unlist(lapply(
    c("WT", exclude_patterns),
    function(pat) grep(pat, colnames(counts))
  )))

  counts_filt <- counts[, -cols_to_remove, drop = FALSE]

  # Remove genes with zero variance
  counts_filt[apply(counts_filt, 1, var) > 0, , drop = FALSE]
}

# Create PCA plot with tidy evaluation
plot_pca <- function(pca_obj, conditions, color_mapping, title) {
  # Calculate variance explained by each PC
  var_explained <- pca_obj$sdev^2 / sum(pca_obj$sdev^2) * 100
  
  pca_data <- as.data.frame(pca_obj$x) |>
    mutate(
      condition = conditions,
      color = color_mapping[conditions]
    )
  
  ggplot(pca_data, aes(x = PC1, y = PC2, color = condition)) +
    geom_point(size = 3) +
    scale_color_manual(values = setNames(
      color_mapping[unique(conditions)],
      unique(conditions)
    )) +
    labs(
      title = title,
      x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
      y = paste0("PC2 (", round(var_explained[2], 1), "%)")
    ) +
    theme_minimal() +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "#e5e5e5", color = NA),
      panel.grid.major = element_line(color = "white", linewidth = 0.4),
      panel.grid.minor = element_line(color = "white", linewidth = 0.2),
      plot.title = element_text(hjust = 0.5, size = 13),
      legend.position = "right"
    )
}

# Wrapper for complete PCA analysis
# exclude_patterns is a regex
run_pca <- function(norm_counts, exclude_patterns, rbp_name, color, output_file) {
  counts_filt <- filter_counts(norm_counts, exclude_patterns)
  conds <- gsub("_.*", "", colnames(counts_filt))
  conds[grepl("empty", conds)] <- "HEK293T-empty"
  
  color_map <- setNames(
    c(color, "black"),
    c(paste0("HEK293T-", rbp_name), "HEK293T-empty")
  )
  
  pca_obj <- prcomp(
    t(counts_filt),
    center = TRUE,
    scale. = FALSE
  )
  
  p <- plot_pca(pca_obj, conds, color_map, rbp_name)
  
  ggsave(output_file, p, width = 4, height = 2.5)
  p
}

# Generate PCA plots
run_pca(norm_counts, "PTBP1|SUPV3L1", "DDX55", "red", 
                 here("CRISPR_RBP_KO", "output", "figures", "CRISPR_RBP_KO_PCA_DDX55.pdf"))

run_pca(norm_counts, "DDX55|SUPV3L1", "PTBP1", "orange",
                 here("CRISPR_RBP_KO", "output", "figures", "CRISPR_RBP_KO_PCA_PTBP1.pdf"))

run_pca(norm_counts, "DDX55|PTBP1", "SUPV3L1", "magenta",
                 here("CRISPR_RBP_KO", "output", "figures", "CRISPR_RBP_KO_PCA_SUPV3L1.pdf"))
```

# Volcano plots of differential expression

```{r, fig.dim = c(6,7), fig.align = 'center'}
# Create and save volcano plot
plot_volcano <- function(res_df, title, output_file) {
  p <- EnhancedVolcano(
    res_df,
    lab = "",
    x = "log2FoldChange",
    y = "pvalue",
    title = title,
    ylim = c(0, 16)
  )
  ggsave(output_file, p, width = 6, height = 7)
  p
}

plot_volcano(
  res_lst$PTBP1_KOvsEmpty,
  "PTBP1 KO vs empty vector control",
  here("CRISPR_RBP_KO", "output", "figures", "CRISPR_RBP_KO_volcano_PTBP1.pdf")
)

plot_volcano(
  res_lst$SUPV3L1_KOvsEmpty,
  "SUPV3L1 KO vs empty vector control",
  here("CRISPR_RBP_KO", "output", "figures", "CRISPR_RBP_KO_volcano_SUPV3L1.pdf")
)
```

# Comparison of miRNA target stability/expression in RBP KOs compared to RBP KDs from ENCODE

Density plots comparing DGEA (W/O WT) in the KOs to differential stability in ENCODE KDs, and showing the enrichment of miR binding sites

```{r}
# Load data and define comparison parameters
stab_HepG2 <- read.csv(here("ENCODE_RBP_KD", "output", "supplementary_tables", "ENCODE_RBP_KD_stab_corr_bias_batch_shRNAoffTargetEffects_HepG2.csv"), row.names=1)
stab_K562 <- read.csv(here("ENCODE_RBP_KD", "output", "supplementary_tables", "ENCODE_RBP_KD_stab_corr_bias_batch_shRNAoffTargetEffects_K562.csv"), row.names=1)
miR_int <- read.csv(here("ENCODE_RBP_KD", "data", "miRNA_seed_matches_3UTRs2000bp.csv"), header = TRUE, row.names=1)

# RBP-miRNA comparisons: RBP, miRNA, condition, KD_ID, cell_type, direction
comp_tbl <- tribble(
  ~RBP, ~miRNA, ~condition, ~KD_ID, ~cell_type, ~direction,
  "PTBP1", "miR.29abcd", "PTBP1_KOvsEmpty", "ENCSR064DXG", "HepG2", "-",
  "SUPV3L1", "miR.19ab", "SUPV3L1_KOvsEmpty", "ENCSR778SIU", "K562", "+",
  "SUPV3L1", "miR.17", "SUPV3L1_KOvsEmpty", "ENCSR778SIU", "K562", "+"
)

stab_list <- list(HepG2 = stab_HepG2, K562 = stab_K562)
cond_name <- "KO vs control"
```

```{r, fig.width=3.3, fig.height=3}
# Density comparison plots
plot_density_comparison <- function(positive_data, negative_data, rbp_name, lims = c(-2, 2, -1.5, 1.5)) {
  f1 <- MASS::kde2d(as.numeric(positive_data$KO), as.numeric(positive_data$KD), 
                    lims = lims, n = 500, h = 0.5)
  f2 <- MASS::kde2d(as.numeric(negative_data$KO), as.numeric(negative_data$KD), 
                    lims = lims, n = 500, h = 0.5)
  
  # Set margins: bottom, left, top, right
  old_par <- par(mar = c(5, 6.5, 2, 1))
  on.exit(par(old_par))
  
  image(f2$x, f2$y, f1$z - f2$z, 
        zlim = c(-0.35, 0.35),
        col = colorRampPalette(c("blue", "white", "red"))(100),
        xlab = glue("Differential expression\n{cond_name} (log2FC)"),
        ylab = glue("Differential stability\n{rbp_name} KD (log2FC)"))
  abline(v = 0, lty = 2, col = "black", lwd = 0.5)
  abline(h = 0, lty = 2, col = "black", lwd = 0.5)
  
  # Return invisible to allow printing
  invisible(NULL)
}

# Generate density plots
for (i in seq_len(nrow(comp_tbl))) {
  params <- comp_tbl[i, ]
  stab_kd <- stab_list[[params$cell_type]]
  
  # Get common genes across all comparisons
  common_genes <- Reduce(intersect, c(lapply(res_lst, rownames), list(rownames(stab_kd))))
  
  # Compute median KD stability (log2FC) for genes shared with KO analysis
  stopifnot(length(grep(params$KD_ID, colnames(stab_kd))) > 0)
  stab_subset <- stab_kd[common_genes, grep(params$KD_ID, colnames(stab_kd)), drop = FALSE]
  kd_stab <- log2(exp(data.matrix(stab_subset))) |>
    matrixStats::rowMedians()
  
  ko_expr <- res_lst[[params$condition]][common_genes, "log2FoldChange"]
  
  # Identify miRNA targets
  targets <- rownames(miR_int)[miR_int[[params$miRNA]] > 0]
  
  # Create comparison table
  comp_data <- tibble(
    gene = common_genes,
    KD = kd_stab,
    KO = ko_expr,
    is_target = as.numeric(gene %in% targets)
  ) |>
    drop_na()
  
  # Save density difference plot to PDF
  pdf(here("CRISPR_RBP_KO", "output", "figures", 
           glue("CRISPR_RBP_KO_density_{params$RBP}_{params$miRNA}.pdf")),
      width = 3.3, height = 3)
  plot_density_comparison(
    comp_data |> filter(is_target == 1),
    comp_data |> filter(is_target == 0),
    params$RBP
  )
  dev.off()
  
  # Print plot to console
  plot_density_comparison(
    comp_data |> filter(is_target == 1),
    comp_data |> filter(is_target == 0),
    params$RBP
  )
}
```

# GSEA for RBP-responsive miRNA targets in the DGEA data of RBP KOs

RBP-responsive miRNA targets = miRNA targets behaving as expected in the KDs (not KOs).

```{r, fig.width=3, fig.height=2.5}
# Test enrichment of RBP-responsive miRNA targets among KO differential expression and plot
run_gsea_analysis <- function(params, stab_list, res_lst, miR_int) {
  stab_kd <- stab_list[[params$cell_type]]
  
  # Get common genes and prepare data
  ko_results <- res_lst[[params$condition]][!is.na(res_lst[[params$condition]]$padj), ]
  stopifnot(
    all(c("log2FoldChange", "stat", "padj") %in% colnames(ko_results))
  )
  common_genes <- intersect(rownames(ko_results), rownames(stab_kd))
  
  # Compute stability medians
  stab_subset <- stab_kd[common_genes, grep(params$KD_ID, colnames(stab_kd)), drop = FALSE]
  kd_stab <- log2(exp(data.matrix(stab_subset))) |>
    matrixStats::rowMedians()
  
  # Extract stats vector
  stat <- setNames(ko_results[common_genes, "stat"], 
                   rownames(ko_results[common_genes, ]))
  
  # Identify miRNA targets with expected behavior in KD
  targets <- rownames(miR_int)[miR_int[[params$miRNA]] > 0]
  threshold <- 0
  direction <- if(params$direction=="-") `<` else `>`
  
  goi <- tibble(gene = common_genes, kd_stab = kd_stab) |>
    mutate(is_target = gene %in% targets,
           behaves_as_expected = is_target & direction(kd_stab, threshold)) |>
    filter(behaves_as_expected) |>
    pull(gene)
  
  # Run GSEA
  genelist <- setNames(list(goi), "miRtarsExpChangeKD")
  gsea_res <- fgsea(pathways = genelist, stats = stat, minSize = 10)
  
  # Create plot with annotations
  p <- fgsea::plotEnrichment(goi, stat, gseaParam = 1, ticksSize = 0.05) +
    labs(title = glue("{params$RBP} KO - {params$miRNA}"),
         x = paste("Rank-ordered genes (", length(genelist[[1]]), ")", sep=""), 
         y = "Enrichment score (ES)")
  
  # Get y-axis limits to position annotation at center
  y_lims <- ggplot_build(p)$layout$panel_scales_y[[1]]$range$range
  y_pos <- (y_lims[1] + y_lims[2]) / 1.8  # Center on y-axis
  
  p <- p +
    annotate("text", 
             x = length(stat) * 0.6,
             y = y_pos,
             label = glue("ES = {signif(gsea_res$ES[1], 3)}\np = {signif(gsea_res$pval[1], 3)}"),
             hjust = 0, vjust = 0.5,
             size = 3, fontface = "plain")
  
  # Change enrichment curve color
  p$layers[[1]]$aes_params$colour <- "blue"
  
  # Remove gridlines
  p <- p +
    theme_classic() +
    theme(panel.grid = element_blank())
  
  # Save and print
  ggsave(here("CRISPR_RBP_KO", "output", "figures", 
              glue("CRISPR_RBP_KO_GSEA_{params$RBP}_{params$miRNA}.pdf")),
         p, width = 3, height = 2.5)
  
  print(p)
}

# Generate GSEA plots for all comparisons
for (i in seq_len(nrow(comp_tbl))) {
  run_gsea_analysis(comp_tbl[i, ], stab_list, res_lst, miR_int)
}
```

```{r}
sessionInfo()
```