---
title: "Correlations between miRNA activity and RBP expression across tissues and cell lines from ENCODE"
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
set.seed(1)
```

# Load libraries

```{r}
library(DESeq2)
library(ggplot2)
library(EnhancedVolcano)
library(fgsea)
library(tidyverse)
```

# CRISPR RBP KO sample metadata 

```{r}
# Extract condition from filename
extract_condition <- function(files) {
  condition <- gsub("HEK293T-*", "", files)
  condition <- gsub("\\.GRCh38.*", "", condition)
  condition <- gsub("_.*", "", condition)
  condition[grepl("empty", condition)] <- "empty"
  condition
}

sampleFiles <- grep("GRCh38.gene", list.files(here("CRISPR_RBP_KO", "data")), value = TRUE)
sampleNames <- gsub("\\..*", "", sampleFiles)
sampleCondition <- extract_condition(sampleFiles)

sampleTable <- data.frame(
  sampleName = sampleNames,
  fileName = sampleFiles,
  condition = sampleCondition
)
sampleTable
```

# Build the DESeqDataSet 

```{r}
ddsHTSeq <- DESeqDataSetFromHTSeqCount(
  sampleTable = sampleTable,
  directory = here("CRISPR_RBP_KO", "data"),
  design = ~ condition
)
ddsHTSeq
```

# DGEA results

```{r}
ddsHTSeq$group <- factor(ddsHTSeq$condition)
design(ddsHTSeq) <- ~ group
ddsHTSeq <- DESeq(ddsHTSeq)

# Extract and name DESeq results for a given RBP
extract_deseq_results <- function(dds, rbp_name, gene_id) {
  res <- as.data.frame(results(dds, contrast = c("group", rbp_name, "empty")))
  res
}

res_lst <- list(
  DDX55_KOvsEmpty = extract_deseq_results(ddsHTSeq, "DDX55", "ENSG00000111364"),
  PTBP1_KOvsEmpty = extract_deseq_results(ddsHTSeq, "PTBP1", "ENSG00000011304"),
  SUPV3L1_KOvsEmpty = extract_deseq_results(ddsHTSeq, "SUPV3L1", "ENSG00000156502")
)
```

# Prepare normalized counts

```{r}
# Center counts by row mean
center_counts <- function(counts) {
  t(apply(counts, 1, function(x) x - mean(x)))
}

norm_counts <- getVarianceStabilizedData(ddsHTSeq)
```

# PCA analysis for RBP knockdowns

```{r, fig.dim = c(4,2.5), fig.align = 'center'}
# Filter and prepare counts for a specific RBP
filter_and_normalize_counts <- function(counts, exclude_patterns) {
  cols_to_remove <- unique(unlist(lapply(
    c("WT", exclude_patterns),
    function(pat) grep(pat, colnames(counts))
  )))
  
  counts_filt <- counts[, -cols_to_remove]
  norm <- center_counts(counts_filt)
  # Remove rows with all zeros
  norm[rowSums(norm) > 0, ]
}

# Create PCA plot with tidy evaluation
plot_pca <- function(pca_obj, conditions, color_mapping, title) {
  # Calculate variance explained by each PC
  var_explained <- pca_obj$sdev^2 / sum(pca_obj$sdev^2) * 100
  
  pca_data <- as.data.frame(pca_obj$x) |>
    mutate(
      condition = conditions,
      color = color_mapping[conditions]
    )
  
  ggplot(pca_data, aes(x = PC1, y = PC2, color = condition)) +
    geom_point(size = 3) +
    scale_color_manual(values = setNames(
      color_mapping[unique(conditions)],
      unique(conditions)
    )) +
    labs(
      title = title,
      x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
      y = paste0("PC2 (", round(var_explained[2], 1), "%)")
    ) +
    theme_minimal() +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "#e5e5e5", color = NA),
      panel.grid.major = element_line(color = "white", linewidth = 0.4),
      panel.grid.minor = element_line(color = "white", linewidth = 0.2),
      plot.title = element_text(hjust = 0.5, size = 13),
      legend.position = "right"
    )
}

# Wrapper for complete PCA analysis
run_pca <- function(norm_counts, exclude_patterns, rbp_name, color, output_file) {
  norm_rowAvg <- filter_and_normalize_counts(norm_counts, exclude_patterns)
  conds <- gsub("_.*", "", colnames(norm_rowAvg))
  conds[grepl("empty", conds)] <- "HEK293T-empty"
  
  color_map <- setNames(
    c(color, "black"),
    c(paste0("HEK293T-", rbp_name), "HEK293T-empty")
  )
  
  pca_obj <- prcomp(t(norm_rowAvg), scale. = FALSE, center = FALSE)
  
  p <- plot_pca(pca_obj, conds, color_map, rbp_name)
  
  ggsave(output_file, p, width = 4, height = 2.5)
  p
}

# Generate PCA plots
run_pca(norm_counts, "PTBP1|SUPV3L1", "DDX55", "red", 
                 here("CRISPR_RBP_KO", "output", "figures", "CRISPR_RBP_KO_PCA_DDX55.pdf"))

run_pca(norm_counts, "DDX55|SUPV3L1", "PTBP1", "orange",
                 here("CRISPR_RBP_KO", "output", "figures", "CRISPR_RBP_KO_PCA_PTBP1.pdf"))

run_pca(norm_counts, "DDX55|PTBP1", "SUPV3L1", "magenta",
                 here("CRISPR_RBP_KO", "output", "figures", "CRISPR_RBP_KO_PCA_SUPV3L1.pdf"))
```

# Volcano plots of differential expression

```{r, fig.dim = c(6,7), fig.align = 'center'}
# Create and save volcano plot
plot_volcano <- function(res_df, title, output_file) {
  p <- EnhancedVolcano(
    res_df,
    lab = "",
    x = "log2FoldChange",
    y = "pvalue",
    title = title,
    ylim = c(0, 16)
  )
  ggsave(output_file, p, width = 6, height = 7)
  p
}

plot_volcano(
  res_lst$PTBP1_KOvsEmpty,
  "PTBP1 KO vs empty vector control",
  here("CRISPR_RBP_KO", "output", "figures", "CRISPR_RBP_KO_volcano_PTBP1.pdf")
)

plot_volcano(
  res_lst$SUPV3L1_KOvsEmpty,
  "SUPV3L1 KO vs empty vector control",
  here("CRISPR_RBP_KO", "output", "figures", "CRISPR_RBP_KO_volcano_SUPV3L1.pdf")
)

<== finish reviewing all code above, check why two types of normalized counts used, etc. 
```

# Comparison of miRNA target stability/expression in RBP KOs compared to RBP KDs from ENCODE

Density plots comparing DGEA (W/O WT) in the KOs to differential stability in ENCODE KDs, and showing the enrichment of miR binding sites

Load the differential mRNA stability for ENCODE RBP KDs

```{r}
stab_HepG2 <- read.csv(paste("RBPKD_stab_biasCorr_corrShRNA6nts7nts_sigPosFishers_HepG2.csv", sep=""), row.names=1)
stab_K562 <- read.csv(paste("RBPKD_stab_biasCorr_corrShRNA6nts7nts_sigPosFishers_K562.csv", sep=""), row.names=1)
```

Make a list of RBPs, miRNAs, conditions, KD IDs and cell types for the candidates to test

```{r}
comp_lst=list(c("PTBP1", "miR.29abcd", "PTBP1_KOvsEmpty", "ENCSR064DXG", "HepG2", "-"),
              c("SUPV3L1", "miR.19ab", "SUPV3L1_KOvsEmpty", "ENCSR778SIU", "K562",  "+"),
              c("SUPV3L1", "miR.17", "SUPV3L1_KOvsEmpty", "ENCSR778SIU", "K562",  "+"))

cond_name <- "KO vs control" 
```

Load the miRNA predicted binding matrix

```{r}
miR_int <- read.csv("miRNA_seed_matches_3UTRs2000bp.csv", header=TRUE, row.names=1)
```

Make density plots showing the change in expression/stability of miRNA targets in the two different RBP inhibition experiments. See specific steps as indicated in the comments within the code.

```{r}
for (i in 1:length(comp_lst)){
  # Define the RBP, miRNA, condition and KD
  RBP = comp_lst[[i]][1]
  miRNA = comp_lst[[i]][2]
  cond = comp_lst[[i]][3]
  KD = comp_lst[[i]][4]
  cell = comp_lst[[i]][5]
  
  if(cell == "HepG2"){
    stab_KD = stab_HepG2
  } else {
    stab_KD = stab_K562
  }
   
  # Filtering to keep the same genes
  common_genes <- Reduce(intersect, c(lapply(res_lst, row.names), list(row.names(stab_KD))))
  
  # Convert to log2FC
  stab_KD <- log2(exp(stab_KD))
  
  # KD of RBP of interest, medians
  KD_RBP <- stab_KD[common_genes, grep(KD, colnames(stab_KD))]
  KD_RBP <- rowMedians(data.matrix(KD_RBP))
  
  # KO of RBP of interest
  KO_RBP <- res_lst[[cond]][common_genes, ]$log2FoldChange
    
  # Define miRNA targets
  targets <- row.names(miR_int)[miR_int[, miRNA] > 0]
  
  # Make table with median stab changes in both conditions and column indicating if genes are miRNA targets
  x <- as.data.frame(cbind(common_genes, KD_RBP, KO_RBP))
  colnames(x)[2:3] <- c("KD", "KO") 
  x$miR_target <- NA
  x$miR_target <- as.numeric(x$common_genes %in% targets)
  
  # Filter for rows w/o NAs
  x <- x[rowSums(is.na(x)) == 0, ]
  
  # separate positive cases from negative cases
  positive <- x[x$miR_target==1,]
  negative <- x[x$miR_target==0,]
  
  # 2D densities of each of positive and negative cases
  # plot the difference of the densities
  lims=c(-2,2,-1.5,1.5)
  f1 <- kde2d(as.numeric(as.vector(positive$KO)), as.numeric(as.vector(positive$KD)), lims=lims,n=500,h=0.5)
  f2 <- kde2d(as.numeric(as.vector(negative$KO)), as.numeric(as.vector(negative$KD)), lims=lims,n=500,h=0.5)
  image(f2$x,f2$y,f1$z-f2$z,zlim=c(-0.35,0.35),col=colorRampPalette(c("blue", "white","red"))(100), xlab = paste("Differential expression\n", cond_name, " (log2FC)", sep=""), ylab = paste("Differential stability \n", RBP, " KD (log2FC)", sep="")) +
  abline(v=0, lty=2, col="black", lwd=0.5) +
  abline(h=0, lty=2, col="black", lwd=0.5)
}
```

# GSEA for RBP-responsive miRNA targets in the DGEA data of RBP KOs

RBP-responsive miRNA targets = miRNA targets behaving as expected in the KDs.

See the comments within the code for a description of the specific steps.

```{r}
for (i in 1:length(comp_lst)){
  # Define the RBP, miR, condition and KD
  RBP = comp_lst[[i]][1]
  miRNA = comp_lst[[i]][2]
  cond = comp_lst[[i]][3]
  KD = comp_lst[[i]][4]
  cell = comp_lst[[i]][5]
  type_int = comp_lst[[i]][6]

  if(cell == "HepG2"){
    stab_KD = stab_HepG2
  } else {
    stab_KD = stab_K562
  }
  
  # Remove NAs
  KO_RBP <- res_lst[[cond]][!(is.na(res_lst[[cond]]$padj)), ] 
  
  # Filtering to keep the same genes
  common_genes <- intersect(row.names(KO_RBP), row.names(stab_KD))
  
  # KD of RBP of interest, medians
  KD_RBP <- stab_KD[common_genes, grep(KD, colnames(stab_KD))]
  KD_RBP <- rowMedians(data.matrix(KD_RBP))

  # KO of RBP of interest
  KO_RBP <- KO_RBP[common_genes, ]
  log2FCs <- KO_RBP$log2FoldChange
  names(log2FCs) <- row.names(KO_RBP)
  stat <- KO_RBP$stat
  names(stat) <- row.names(KO_RBP)
    
  # Define miRNA targets
  targets <- row.names(miR_int)[miR_int[, miRNA] > 0]
  
  # Make table with median stab changes in both conditions and column indicating if genes are miRNA targets
  x <- as.data.frame(cbind(common_genes, KD_RBP, log2FCs))
  colnames(x)[2:3] <- c("KD", "KO") 
  x$miR_target <- NA
  x$miR_target <- as.numeric(x$common_genes %in% targets)
  
  # Filter for rows w/o NAs
  x <- x[rowSums(is.na(x)) == 0, ]
  
  # Add whter the gene is a target and if it behaves as expected in KD
  x$GOI <- 0
  if(type_int == "-"){
    for (j in 1:nrow(x)){
      if (x$miR_target[j] == 1 & x$KD[j] < 0){
        x$GOI[j] <- 1
      }
    }
  } else {
    for (j in 1:nrow(x)){
      if (x$miR_target[j] == 1 & x$KD[j] > 0){
        x$GOI[j] <- 1
      }
    }
  }
    
  genelist <- list(x$common_genes[x$GOI == 1])
  names(genelist) <- "miRtarsExpChangeKD"
  fgseaRes <- fgsea(pathways = genelist,
    stats = stat,
    minSize=10)
  print(plotEnrichment(genelist[[1]], stat, gseaParam = 1, ticksSize = 0.05))
}
```

The plots correspond, in order, to PTBP1-miR-29, SUPV3L1-miR-19, and SUPV3L1-miR-17

```{r}
sessionInfo()
```