---
title: "Inferrence of mRNA stability and miRNA activity across tissues and cell lines from ENCODE"
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
set.seed(1)
```

# Load libraries

```{r}
library(MASS)
library(data.table)
library(tidyverse)
```

# Load the ddf (estimates of changes in stability), ddf_stdev (standard deviations) and dfi (estimates of change in pre-mRNA abundance) data obtained using BETAS 

```{r, echo = F}
# Helper function to load and convert BETAS data 
load_betas_data <- function(filename) {
  fread(here("ENCODE_tissues_cell_lines", "data", filename)) |>
    column_to_rownames(var = "V1") |>
    data.matrix()
}

ddf <- load_betas_data("ENCODE_totalRNA-seq.BETAS_ddf.csv")
ddf_stdev <- load_betas_data("ENCODE_totalRNA-seq.BETAS_ddf_stdev.csv")
dfi <- load_betas_data("ENCODE_totalRNA-seq.BETAS_dfi.csv")
```

# Define std cutoff to retain closest to 10,000 genes as possible, and apply that filter

```{r}
dim(ddf)

# Helper function to find cutoff that retains ~10,000 genes
find_optimal_cutoff <- function(data, stdev, row_threshold = 0.8, col_threshold = 0.2, target_genes = 10000, increment = 0.05) {
  cutoff <- 0.1
  data_filtered <- NULL
  data_with_rows_kept <- NULL
  
  while (TRUE) {
    cat("Cutoff being tested:", cutoff, "\n")
    
    # Mask low-confidence values
    data_masked <- data
    data_masked[stdev > cutoff] <- NA
    
    # Keep rows (genes) with at least 80% non-NA values 
    row_data_completeness <- rowSums(!is.na(data_masked)) / ncol(data_masked)
    passing_rows <- which(row_data_completeness >= row_threshold)
    data_with_rows_kept <- data_masked[passing_rows, ]
    
    # Remove samples (columns) with > 20% NAs
    col_data_completeness <- colSums(is.na(data_with_rows_kept)) / nrow(data_with_rows_kept)
    passing_cols <- col_data_completeness <= col_threshold
    data_filtered <- data_with_rows_kept[, passing_cols]
    
    n_genes <- nrow(data_filtered)
    
    if (n_genes >= target_genes) break
    cutoff <- cutoff + increment
  }
  
  # Return the first cutoff that retains at least the target number of genes
  final_cutoff <- cutoff
  removed_samples <- colnames(data_with_rows_kept)[!colnames(data_with_rows_kept) %in% colnames(data_filtered)]
  
  list(data = data_filtered, cutoff = final_cutoff, n_genes = n_genes, 
       removed_samples = removed_samples)
}

cutoff_result <- find_optimal_cutoff(ddf, ddf_stdev)
ddf_fltr <- cutoff_result$data
cutoff <- cutoff_result$cutoff

cat("Final cutoff:", cutoff, "\n")
cat("Number of genes remaining:", cutoff_result$n_genes, "\n")

# Filter dfi to keep same genes and samples as ddf_fltr
dfi_fltr <- dfi[rownames(ddf_fltr), colnames(ddf_fltr)]

cat("ddf_fltr dimensions:", nrow(ddf_fltr), "x", ncol(ddf_fltr), "\n")
cat("dfi_fltr dimensions:", nrow(dfi_fltr), "x", ncol(dfi_fltr), "\n")
```

# Load the sample metadata

```{r}
met <- read.table(here("ENCODE_tissues_cell_lines", "data", "ENCODE_total_RNA-seq_experiments_metadata.txt"), header = F)
```

# Bias correction of differential stability estimates

Fit a robust linear regression for each gene, modelling stability changes (ddf) as a function of transcriptional changes (dfi), and retain the residuals as bias-corrected stability estimates

```{r}
stab_cor <- do.call(rbind, lapply(seq_len(nrow(ddf_fltr)), function(i) {
  fit <- rlm(ddf_fltr[i,] ~ dfi_fltr[i,], na.action=na.exclude, maxit=200)
  residuals(fit)
}))

rownames(stab_cor) <- rownames(ddf_fltr)
colnames(stab_cor) <- colnames(ddf_fltr)
```

# Save the bias-corrected stability data

```{r}
write.csv(stab_cor, here("ENCODE_tissues_cell_lines", "output", "supplementary_tables", "ENCODE_tissues_cell_lines_stability_bias_corr.csv"))
```

# Load the 3'UTR nucleotide composition and length annotations, and miRNA seed matches

```{r}
UTR_nt <- read.table(here("ENCODE_RBP_KD", "data", "3UTR_nt_composition_new.txt"), header = TRUE, row.names = NULL)
UTR_length <- read.table(here("ENCODE_RBP_KD", "data", "3UTR_length_new.txt"), header = TRUE, row.names = NULL)
# miR_binding <- read.table(here("ENCODE_tissues_cell_lines", "data", "3UTR_miRNA_affinity_new.txt"), , header = TRUE, row.names = NULL)
miR_binding <- read.csv(here("ENCODE_RBP_KD", "data", "miRNA_seed_matches_3UTRs2000bp.csv"), header = TRUE)
```

# Prepare GLM data

Merge 3'UTR annotations with stability data 

```{r}
stab_cor <- as.data.frame(stab_cor) |>
  rownames_to_column(var = "Gene_id")
UTR <- merge(UTR_nt, UTR_length, by=1)
UTR_miRNA <- merge(UTR, miR_binding, by=1)
glmMatrix <- merge(stab_cor, UTR_miRNA, by=1)
```

# Remove columns that are only NAs

```{r}
glmMatrix <- glmMatrix |>
    select(where(~ !all(is.na(.))))
```

# Fit GLM models 

Run GLM for each sample, extracting coefficients and P-values

```{r}
fit_glm_sample <- function(glm_data, sample_name) {
  # glm_data: complete data frame with response and all predictors
  # sample_name: name of the sample 
  
  predictor_names <- names(glm_data)[
    names(glm_data) != "Gene_id" &
      !grepl("ENCSR", names(glm_data))
  ]
  # Formula: response ~ confounders and miRNA affinity predictors
  # Intercept included automatically
  formula_str <- paste0(
    sample_name,
    " ~ ",
    paste(predictor_names, collapse = " + ")
  )
  
  fit <- glm(as.formula(formula_str), data = glm_data, na.action = na.exclude)
  coef_summary <- coef(summary(fit))
  
  # Extract coefficients and p-values for predictors only (miRNA cols, skip intercept and confounders)
  predictor_rows <- which(grepl("^(miR\\.|let\\.)", rownames(coef_summary)))
  
  tibble(
    miRNA = rownames(coef_summary)[predictor_rows],
    coefficient = coef_summary[predictor_rows, 1],
    pvalue = coef_summary[predictor_rows, 4],
    sample_name = sample_name
  )
}

# Apply to all samples in the stability data
glm_results <-
    map_dfr(
        colnames(stab_cor)[-1],
        ~ fit_glm_sample(glmMatrix, .x)
    )
```

# Convert inferred target stability effects into miRNA activity estimates by reversing the coefficient sign, express coefficients on a log2 scale

```{r}
glm_results <- glm_results |>
  mutate(
    coefficient = -coefficient,
    coefficient_log2 = log2(exp(coefficient))
  )

head(glm_results)
```

# Multiple testing correction 

```{r}
glm_results <- glm_results |>
  mutate(FDR = p.adjust(pvalue, method = "fdr"))
```

# Reformat miRNA names

```{r}
glm_results$miRNA <- gsub("\\.[A-Z].*", "", glm_results$miRNA)
glm_results$miRNA <- gsub("_.*", "", glm_results$miRNA)
glm_results$miRNA <- gsub("\\.", "-", glm_results$miRNA)

# Define mapping from original names to canonical names
mirna_name_mapping <- tibble(
  original = c(
    "miR-33ab", "miR-26ab", "miR-29abcd", "miR-27abc", "miR-133abc",
    "miR-199ab-5p", "miR-376abd", "miR-23abc", "miR-19ab", "miR-182",
    "miR-758", "miR-22", "miR-150", "miR-124", "miR-96", "miR-370",
    "miR-17", "miR-218", "miR-21", "miR-25", "miR-376c", "miR-10abc",
    "miR-143", "miR-34ac", "miR-338", "miR-377", "miR-495", "miR-544",
    "miR-18ab", "miR-299", "miR-328a", "let-7"
  ),
  canonical = c(
    "miR-33-5p", "miR-26-5p", "miR-29-3p", "miR-27-3p", "miR-133-3p",
    "miR-199-5p", "miR-376-3p", "miR-23-3p", "miR-19-3p", "miR-182-5p",
    "miR-758-3p", "miR-22-3p", "miR-150-5p", "miR-124-3p", "miR-96-5p",
    "miR-370-3p", "miR-17-5p", "miR-218-5p", "miR-21-5p", "miR-25-3p",
    "miR-376c-3p", "miR-10-5p", "miR-143-3p", "miR-34-5p", "miR-338-3p",
    "miR-377-3p", "miR-495-3p", "miR-544a-3p", "miR-18-5p", "miR-299-3p",
    "miR-328-3p", "let-7"
  )
)

glm_results <- glm_results |>
  left_join(mirna_name_mapping, by = c("miRNA" = "original")) |>
  mutate(miRNA = coalesce(canonical, miRNA)) |>
  select(-canonical)
```

# Save results

```{r}
# Save all results in long-format tibble 
write.csv(glm_results, file = here("ENCODE_tissues_cell_lines", "output", "supplementary_tables", "ENCODE_tissues_cell_lines_miRNA_activity.csv"), row.names = FALSE)
```

```{r}
sessionInfo()
```
