---
title: "Correlations between miRNA activity and RBP expression across tissues and cell lines from ENCODE"
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
set.seed(1)
```

# Load libraries

```{r}
library(MASS)
library(dplyr)
library(data.table)
library(tibble)
```

# Load the ddf (estimates of changes in stability), ddf_stdev (standard deviations) and dfi (estimates of change in pre-mRNA abundance) data obtained using BETAS 

```{r, echo = F}
# Helper function to load and convert BETAS data 
load_betas_data <- function(filename) {
  fread(here("ENCODE_tissues_cell_lines", "data", filename)) |>
    column_to_rownames(var = "V1") |>
    data.matrix()
}

ddf <- load_betas_data("ENCODE_totalRNA-seq.BETAS_ddf.csv")
ddf_stdev <- load_betas_data("ENCODE_totalRNA-seq.BETAS_ddf_stdev.csv")
dfi <- load_betas_data("ENCODE_totalRNA-seq.BETAS_dfi.csv")
```

# Define std cutoff to retain closest to 10,000 genes as possible, and apply that filter

```{r}
dim(ddf)

# Helper function to find cutoff that retains ~10,000 genes
find_optimal_cutoff <- function(data, stdev, row_threshold = 0.8, col_threshold = 0.2, target_genes = 10000, increment = 0.05) {
  cutoff <- 0.1
  data_filtered <- NULL
  data_with_rows_kept <- NULL
  
  while (TRUE) {
    cat("Cutoff being tested:", cutoff, "\n")
    
    # Mask low-confidence values
    data_masked <- data
    data_masked[stdev > cutoff] <- NA
    
    # Keep rows (genes) with at least 80% non-NA values 
    row_data_completeness <- rowSums(!is.na(data_masked)) / ncol(data_masked)
    passing_rows <- which(row_data_completeness >= row_threshold)
    data_with_rows_kept <- data_masked[passing_rows, ]
    
    # Remove samples (columns) with > 20% NAs
    col_data_completeness <- colSums(is.na(data_with_rows_kept)) / nrow(data_with_rows_kept)
    passing_cols <- col_data_completeness <= col_threshold
    data_filtered <- data_with_rows_kept[, passing_cols]
    
    n_genes <- nrow(data_filtered)
    
    if (n_genes >= target_genes) break
    cutoff <- cutoff + increment
  }
  
  # Return the cutoff that just exceeded target, minus increment
  final_cutoff <- cutoff - increment
  removed_samples <- colnames(data_with_rows_kept)[!colnames(data_with_rows_kept) %in% colnames(data_filtered)]
  
  list(data = data_filtered, cutoff = final_cutoff, n_genes = n_genes, 
       removed_samples = removed_samples)
}

cutoff_result <- find_optimal_cutoff(ddf, ddf_stdev)
ddf_fltr <- cutoff_result$data
cutoff <- cutoff_result$cutoff

cat("Final cutoff:", cutoff, "\n")
cat("Number of genes remaining:", cutoff_result$n_genes, "\n")

# Filter dfi to keep same genes and samples as ddf_fltr
dfi_fltr <- dfi[rownames(ddf_fltr), colnames(ddf_fltr)]

cat("ddf_fltr dimensions:", nrow(ddf_fltr), "x", ncol(ddf_fltr), "\n")
cat("dfi_fltr dimensions:", nrow(dfi_fltr), "x", ncol(dfi_fltr), "\n")
```

# Load the sample metadata

```{r}
met <- read.table(here("ENCODE_tissues_cell_lines", "data", "ENCODE_total_RNA-seq_experiments_metadata.txt"), header = F)
```

# Bias correction of differential stability estimates



```{r}

<== 
# Create a matrix that will be filled with bias-corrected stability 

stab_cor <- matrix(NA, nrow=nrow(ddf_final), ncol=ncol(ddf_final))
rownames(stab_cor) = rownames(ddf_final)
colnames(stab_cor) = colnames(ddf_final)
```

Perform bias correction using a robust linear regression for each gene (rows), using the ddf as the response variable and the dfi as the confounding variable

```{r}
for(i in seq_len(nrow(ddf_final))) {
  fit <- rlm(ddf_final[i,] ~ dfi_final[i,], na.action=na.exclude, maxit=200)
  stab_cor[i,] <- residuals(fit)
}
```

Save the bias corrected stability data

```{r}
write.table(stab_cor, file = here("ENCODE_tissues_cell_lines", "output", "supplementary_tables", "ENCODE_total_RNAseq_stability_biasCorrected.txt"), sep ="\t", row.names=TRUE, col.names=TRUE)
```

# Infer changes in miRNA activiy

Load the 3'UTR annotations containing 3'UTR lengths and nucleotide composition

```{r}
UTR_nt <- read.table(here("ENCODE_tissues_cell_lines", "data", "3UTR_nt_composition_new.txt"), header=TRUE, row.names=NULL)
UTR_length <- read.table(here("ENCODE_tissues_cell_lines", "data", "3UTR_length_new.txt"), header=TRUE, row.names=NULL)
```

Load the predicted miRNA-mRNA interactions

```{r}
miR_binding <- read.table(here("ENCODE_tissues_cell_lines", "data", "3UTR_miRNA_affinity_new.txt"), header=TRUE, row.names=NULL)
```

Merge tables based on gene IDs (separate merged tables for each cell line)

```{r}
stab_cor <- cbind(row.names(stab_cor), stab_cor)
UTR <- merge(UTR_nt, UTR_length, by=1)
UTR_miRNA <- merge(UTR, miR_binding, by=1)
glmMatrix <- merge(stab_cor, UTR_miRNA, by=1)
```

Remove columns that are only NAs

```{r}
glmMatrix <- glmMatrix[ , !apply(is.na(glmMatrix),2,all)]
```

Create matrices where the results (regression coefficients and p-values) from miRNA activity inference will be outputed

```{r}
nb_samples_filtered <- ncol(glmMatrix) - ncol(UTR_miRNA)
pval <- matrix(NA, nrow=nb_samples_filtered, ncol=ncol(UTR_miRNA)-1)
rownames(pval) = colnames(glmMatrix)[2:(nb_samples_filtered + 1)]
colnames(pval) = colnames(UTR_miRNA)[-1]

coeff <- pval
```

For each sample, infer differential miRNA activity using differential mRNA stability as the response and miRNA binding predictions as the predictor, with 3'UTR length and nucleotide composition as confounders

```{r}
for (i in 2:(nb_samples_filtered +1)){
  fit <- glm(as.numeric(as.vector(glmMatrix[,i])) ~ as.matrix(glmMatrix[,(nb_samples_filtered+2):ncol(glmMatrix)]), na.action=na.exclude)
  
  coeff_full <- as.data.frame(coeff[i-1,]) # Define an empty coefficient table for that sample
  coeff_tmp <- as.data.frame(coef(summary(fit))[-1,1]) # Get the outputed coefficients for that sample
  row.names(coeff_tmp) <- gsub("as.matrix\\(glmMatrix\\[, \\(nb_samples_filtered \\+ 2):ncol\\(glmMatrix\\)\\]\\)", "", row.names(coeff_tmp))
  coeff_full[,1] <- coeff_tmp[match(row.names(coeff_full), row.names(coeff_tmp)), 1] # Fill in the sample coefficients table with the available coefficients, missing ones remain as NAs
  coeff[i-1,] <- as.vector(coeff_full[,1]) # Fill the big coefficients table with the coefficient of the sample

  pval_full <- as.data.frame(pval[i-1,]) # Define an empty p-values table for that sample
  pval_tmp <- as.data.frame(coef(summary(fit))[-1,4]) # Get the outputed p-value for that sample
  row.names(pval_tmp) <- gsub("as.matrix\\(glmMatrix\\[, \\(nb_samples_filtered \\+ 2):ncol\\(glmMatrix\\)\\]\\)", "", row.names(pval_tmp))
  pval_full[,1] <- pval_tmp[match(row.names(pval_full), row.names(pval_tmp)), 1] # Fill in the sample p-values table with the available p-value, missing ones remain NAs
  pval[i-1,] <- as.vector(pval_full[,1]) # Fill the big p-values table with the p-value of the sample
} 
```

Remove the columns for 3'UTR length and composition from the coefficients and p-values tables

```{r}
pval <- pval[,22:ncol(pval)]           
coeff <- coeff[,22:ncol(coeff)]
```

FDR correction

```{r}
pval_vector <- c(pval)
pval_vector_FDR <- p.adjust(pval_vector, method="fdr")
FDR <- matrix(pval_vector_FDR, nrow = nrow(pval),ncol = ncol(pval))

colnames(FDR) <- colnames(pval)
row.names(FDR) <- row.names(pval)
```

Switch signs of coefficients so that they represent miRNA activity and not target mRNA stability

```{r}
coeff <- -(coeff)
```

Reformat miRNA names and save the results

```{r}
s1 <- gsub("\\.[A-Z].*", "", colnames(coeff))
s2 <- gsub("_.*", "", s1)

colnames(coeff) <- s2
colnames(pval) <- s2
colnames(FDR) <- s2

write.csv(coeff, file=here("ENCODE_tissues_cell_lines", "output", "supplementary_tables", "ENCODE_totalRNA_miRNA_activity.csv"))
write.csv(pval, file=here("ENCODE_tissues_cell_lines", "output", "supplementary_tables", "ENCODE_totalRNA_miRNA_activity_pval.csv"))
write.csv(FDR, file=here("ENCODE_tissues_cell_lines", "output", "supplementary_tables", "ENCODE_totalRNA_miRNA_activity_FDR.csv"))
```

# Correlation RBP expression vs miRNA activity 

Load the list of candidate interactions

```{r}
miRNA_RBPs <- read.csv(here("ENCODE_tissues_cell_lines", "data", "RBPKD_miRNAs_newSigInts_HepG2andOrK562_corrShRNA6nts7ntsSigPosFishers_oneSigRepAndSameDirOtherReps.csv"))[,-1]
miRNA_RBPs$miRNA <- as.vector(miRNA_RBPs$miRNA)
miRNA_RBPs$RBP <- as.vector(miRNA_RBPs$RBP)
```

Reformat miRNA names

```{r}
miRNA_RBPs$miRNA <- gsub("\\.", "-", miRNA_RBPs$miRNA)
colnames(coeff) <- gsub("\\.", "-", colnames(coeff))
```

Get the ensembl IDs of the RBPs

```{r}
ensembl = useMart("ensembl",dataset="hsapiens_gene_ensembl")
gene_names <- as.vector(miRNA_RBPs$RBP)
annotation <- getBM(attributes = c('ensembl_gene_id', 'hgnc_symbol'),
                        filters = 'hgnc_symbol',
                        values = gene_names,
                        mart = ensembl)
annotation <- annotation[!(is.na(annotation$ensembl_gene_id)),]
```

Load gene expression profiles (FPKM)

```{r}
FPKM <- as.data.frame(fread(here("ENCODE_tissues_cell_lines", "data", "ENCODE_total_RNA-seq_gene_FPKM.csv")))
row.names(FPKM) <- FPKM$gene_id
FPKM <- FPKM[, c(-1, -2)]
```

Filter to keep same samples as miRNA activity table

```{r}
common_samples <- intersect(colnames(FPKM), row.names(coeff))
FPKM <- FPKM[, common_samples] 
coeff <- coeff[common_samples, ]
```

Get the RBP gene names and add in a new column of the interactions table

```{r}
miRNA_RBPs$RBP_ID <- NA
for(i in 1:nrow(miRNA_RBPs)){
  RBPname <- miRNA_RBPs[i, ]$RBP
  RBPids <- annotation[annotation$hgnc_symbol == RBPname, ]$ensembl_gene_id
  RBPid <- RBPids[RBPids %in% row.names(FPKM)]
  miRNA_RBPs$RBP_ID[i] <- RBPid
}
```

Make an example scatter plot comparing the expression of SUPV3L1 to the activity of miR-17

```{r}
spec_miRNA_act <- coeff[, "miR-17"]
spec_RBP_expr <- unlist(FPKM["ENSG00000156502", ])
common_samples <- intersect(names(spec_RBP_expr), names(spec_miRNA_act))
spec_RBP_expr_matched <- spec_RBP_expr[common_samples]
spec_miRNA_activity_matched <- spec_miRNA_act[common_samples]

pdf(here("ENCODE_tissues_cell_lines", "output", "figures", "ENCODE_totalRNAseq_SUPV3L1_expr_VS_miR-17_act_scatterplot.pdf"), height=4, width=4)
plot(spec_RBP_expr_matched, spec_miRNA_activity_matched, pch = 16, cex=0.5, xlab="SUPV3L1 expression (log2FPKM)", ylab="miR-17-5p activity")
abline(lm(spec_miRNA_activity_matched ~ spec_RBP_expr_matched), col = 4, lwd = 2)
text(0.8, 0.12, paste("r = ", round(cor.test(spec_RBP_expr_matched, spec_miRNA_activity_matched)$estimate, 3), sep=""))
dev.off()
```

![Scatter plot of SUPV3L1 expression vs miR-17 activity](ENCODE_totalRNAseq_SUPV3L1_expr_VS_miR-17_act_scatterplot.pdf){width=65%}

Now we want to look at every candidate interactions. Create a table to ouput the results from the correlations

```{r}
result_table <- data.frame(matrix(ncol = 3, nrow = nrow(miRNA_RBPs)))
colnames(result_table) <- c("pair", "correlation_coefficient", "pvalue")
result_table$pair <- paste(miRNA_RBPs$RBP, miRNA_RBPs$miRNA, sep="_")
result_table <- as.data.frame(cbind(result_table, miRNA_RBPs))
```

Take the correlation between miRNA activity and RBP expression across the samples, for each candidate interaction. See the steps as indicated in the comments within the code.

```{r}
for (p in 1:nrow(miRNA_RBPs))
{
  # Define miRNA and RBP of interest
  miRNA <- as.vector(miRNA_RBPs[p, ]$miRNA)
  RBP_name <- as.vector(miRNA_RBPs[p, ]$RBP)
  RBP_ID <- as.vector(miRNA_RBPs[p, ]$RBP_ID)
  
  # Select the activity for the miRNA of interest
  spec_miRNA_act <- coeff[, miRNA]
  
  # Skip if RBP is not in the expression data
  if (!(RBP_ID %in% row.names(FPKM)))
  {
    print(RBP_ID)
    print("RBP not in gene expression matrix")
    next
  }
  spec_RBP_expr <- unlist(FPKM[RBP_ID, ])

  # Filter RBP expression and miRNA activity to have same samples in same order
  common_samples <- intersect(names(spec_RBP_expr), names(spec_miRNA_act))
  spec_RBP_expr_matched <- spec_RBP_expr[common_samples]
  spec_miRNA_activity_matched <- spec_miRNA_act[common_samples]

  corr <- cor.test(spec_RBP_expr_matched, spec_miRNA_activity_matched)
  
  result_table[p,]$correlation_coefficient <- corr$estimate
  result_table[p,]$pvalue <- corr$p.value
}
```

Remove the rows with NAs and order by coefficient

```{r}
result_table <- result_table[!(is.na(result_table$correlation_coefficient)), ]
result_table <- result_table[order(result_table$correlation_coefficient), ]
```

FDR correction

```{r}
result_table$FDR <- p.adjust(result_table$pvalue, method = "fdr")
```

Reformat the miRNA names

```{r}
result_table$pair <- gsub("miR-33ab$", "- miR-33-5p", result_table$pair)
result_table$pair <- gsub("miR-26ab$", "- miR-26-5p", result_table$pair)
result_table$pair <- gsub("miR-29abcd$", "- miR-29-3p", result_table$pair)
result_table$pair <- gsub("miR-27abc$", "- miR-27-3p", result_table$pair)
result_table$pair <- gsub("miR-133abc$", "- miR-133-3p", result_table$pair)
result_table$pair <- gsub("miR-199ab-5p$", "- miR-199-5p", result_table$pair)
result_table$pair <- gsub("miR-376abd$", "- miR-376-3p", result_table$pair)
result_table$pair <- gsub("miR-23abc$", "- miR-23-3p", result_table$pair)
result_table$pair <- gsub("miR-19ab$", "- miR-19-3p", result_table$pair)
result_table$pair <- gsub("miR-182$", "- miR-182-5p", result_table$pair)
result_table$pair <- gsub("miR-758$", "- miR-758-3p", result_table$pair)
result_table$pair <- gsub("miR-22$", "- miR-22-3p", result_table$pair)
result_table$pair <- gsub("miR-150$", "- miR-150-5p", result_table$pair)
result_table$pair <- gsub("miR-124$", "- miR-124-3p", result_table$pair)
result_table$pair <- gsub("miR-96$", "- miR-96-5p", result_table$pair)
result_table$pair <- gsub("miR-370$", "- miR-370-3p", result_table$pair)
result_table$pair <- gsub("miR-17$", "- miR-17-5p", result_table$pair)
result_table$pair <- gsub("miR-218$", "- miR-218-5p", result_table$pair)
result_table$pair <- gsub("miR-21$", "- miR-21-5p", result_table$pair)
result_table$pair <- gsub("miR-25$", "- miR-25-3p", result_table$pair)
result_table$pair <- gsub("_-", " -", result_table$pair)
result_table$pair <- gsub("_", " - ", result_table$pair)
```

Include the type of interaction (+/-) in the labels

```{r}
result_table$sign <- gsub("-1", "-", result_table$sign) 
result_table$sign <- gsub("1", "+", result_table$sign)
```

Assign colors to each label reflecting the type of interaction

```{r}
cols[grep("-", cols)] <- "darkblue"
cols[grep("\\+", cols)] <- "red"
```

Draw the barplot

```{r}
pdf(here("ENCODE_tissues_cell_lines", "output", "figures", "ENCODE_totalRNAseq_corrMirActVsRBPexpr.pdf"), height=4, width=5)
par(mar=c(7,5,1,1))
x <- barplot(result_table$correlation_coefficient, 
        ylab = "Correlation miRNA activity VS \n RBP expression (Pearson)",
        ylim = c(-0.4, 0.6),
        cex.names = 0.6,
        cex.axis = 0.6,
        cex.lab=0.6,
        col=ifelse(result_table$FDR < 0.05, "cornflowerblue","gray"))
# text(x=x, y=-0.38, labels = paste(result_table$pair,result_table$sign), srt=90, xpd=TRUE, cex=0.5, adj=c(1,0), col = cols)
text(x=x+0.25, y=-0.38, labels = result_table$pair, srt=90, xpd=TRUE, cex=0.5, adj=c(1,0), col = cols)
legend(x=1, y = 0.5, legend = c("padj < 0.05", "NS"), fill = c("cornflowerblue","gray"), cex=0.6)
dev.off()
```

![Bar plot of RBP expression vs miRNA activity](ENCODE_totalRNAseq_corrMirActVsRBPexpr.pdf){width=65%}
