---
title: "Correlations between miRNA activity and RBP expression across tissues and cell lines from ENCODE"
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
set.seed(1)
```

# Load libraries

```{r}
library(biomaRt)
library(data.table)
library(tidyverse)
```

# Load miRNA activity for ENCODE tissues and cell lines

```{r}
miR_act <- read.csv(file = here("ENCODE_tissues_cell_lines", "output", "supplementary_tables", "ENCODE_tissues_cell_lines_miRNA_activity.csv"))
```

# Load the list of candidate interactions from ENCODE RBP KDs

```{r}
miRNA_RBPs <- readRDS(here("ENCODE_RBP_KD", "output", "rds", "ENCODE_RBP_KD_interactions.rds"))
```

# Get the ensembl IDs of the RBPs

```{r}
# ensembl = useMart("ensembl", dataset="hsapiens_gene_ensembl")
# gene_names <- as.vector(miRNA_RBPs$RBP)
# annotation <- getBM(attributes = c('ensembl_gene_id', 'hgnc_symbol'),
#                         filters = 'hgnc_symbol',
#                         values = gene_names,
#                         mart = ensembl)
# annotation <- annotation[!(is.na(annotation$ensembl_gene_id)),]
# 
# # Check which RBPs get more than 1 ID
# unique(annotation$hgnc_symbol[duplicated(annotation$hgnc_symbol)])
# 
# # Manually retain the main ensembl ID for these and delete the others
# annotation <- annotation[!(annotation$ensembl_gene_id %in% c("ENSG00000276072", "ENSG00000236149", "ENSG00000232169", "ENSG00000236342", "ENSG00000225989", "ENSG00000231129", "ENSG00000206490", "ENSG00000206268", "ENSG00000231044", "ENSG00000206357", "ENSG00000229363", "ENSG00000233801")), ]
# 
# stopifnot(!anyDuplicated(annotation$ensembl_gene_id))
# 
# saveRDS(annotation, here("ENCODE_tissues_cell_lines", "output", "rds", "ENCODE_tissues_cell_lines_RBP_IDs.rds"),)
annotation <- readRDS(here("ENCODE_tissues_cell_lines", "output", "rds", "ENCODE_tissues_cell_lines_RBP_IDs.rds"))
```

# Add Ensembl gene IDs for each RBP to the interactions table

```{r}
miRNA_RBPs$RBP_ID <- annotation$ensembl_gene_id[match(miRNA_RBPs$RBP, annotation$hgnc_symbol)]
```

# Load gene expression profiles (FPKM)

```{r}
FPKM <- fread(here("ENCODE_tissues_cell_lines", "data", "ENCODE_total_RNA-seq_gene_FPKM.csv"))
FPKM <- FPKM[,-1]
FPKM <- column_to_rownames(FPKM, var = "gene_id")
```

# Restrict both datasets to samples present in both expression and miRNA activity data

```{r}
common_samples <- intersect(colnames(FPKM), miR_act$sample_name)
FPKM <- FPKM[, common_samples] 
miR_act <- miR_act |>
  filter(sample_name %in% common_samples)
```

# Make an example scatter plot comparing the expression of SUPV3L1 to the activity of miR-17

```{r, fig.dim = c(4,4), fig.align = 'center'}
moi <- "miR-17-5p"
roi <- "SUPV3L1"
roi_id <- annotation$ensembl_gene_id[annotation$hgnc_symbol == roi]
moi_act <- miR_act[miR_act$miRNA == moi, ]
roi_expr <- unlist(FPKM[roi_id, ])
common_samples <- intersect(names(roi_expr), moi_act$sample_name)
roi_expr <- roi_expr[common_samples]
moi_act <- moi_act[match(common_samples, moi_act$sample_name), ]
stopifnot(all(moi_act$sample_name == names(roi_expr)))

pdf(here("ENCODE_tissues_cell_lines", "output", "figures", "ENCODE_tissues_cell_lines_SUPV3L1_vs_miR17_scatterplot.pdf"), height=4, width=4)
plot(roi_expr, moi_act$coefficient_log2, pch = 16, cex=0.5, xlab=paste0(roi, " expression log2(FPKM)"), ylab=paste0(moi, " activity"))
abline(lm(moi_act$coefficient_log2 ~ roi_expr), col = 4, lwd = 2)
text(0.8, 0.12, paste("r = ", round(cor.test(roi_expr, moi_act$coefficient_log2)$estimate, 3), sep=""))
dev.off()

plot(roi_expr, moi_act$coefficient_log2, pch = 16, cex=0.5, xlab=paste0(roi, " expression log2(FPKM)"), ylab=paste0(moi, " activity"))
abline(lm(moi_act$coefficient_log2 ~ roi_expr), col = 4, lwd = 2)
text(0.8, 0.12, paste("r = ", round(cor.test(roi_expr, moi_act$coefficient_log2)$estimate, 3), sep=""))
```

# Compute correlations between RBP expression and miRNA activity for all candidate interaction pairs

```{r}
# Helper function to compute correlation for a single pair
compute_correlation <- function(rbp_id, mirna, mirna_activity_df) {
  # Skip if RBP is not in the expression data
  if (!(rbp_id %in% rownames(FPKM))) {
    return(NULL)
  }
  
  # Get RBP expression and miRNA activity for matching samples
  rbp_expr <- as.numeric(FPKM[rbp_id, ])
  names(rbp_expr) <- colnames(FPKM)
  
  mirna_data <- mirna_activity_df |> 
    filter(miRNA == mirna) 
  
  if (nrow(mirna_data) == 0) {
    return(NULL)
  }
  
  mirna_samples <- mirna_data$sample_name
  mirna_vals <- mirna_data$coefficient_log2
  
  # Find common samples and match order
  common_samples <- intersect(names(rbp_expr), mirna_samples)
  if (length(common_samples) == 0) {
    return(NULL)
  }
  
  rbp_matched <- rbp_expr[common_samples]
  mirna_matched <- mirna_vals[match(common_samples, mirna_samples)]
  
  # Compute Pearson correlation between RBP expression and miRNA activity
  corr <- cor.test(rbp_matched, mirna_matched)
  tibble(
    correlation_coefficient = corr$estimate,
    pvalue = corr$p.value
  )
}

# Compute expression–activity correlations for all candidate RBP–miRNA pairs
corr_table <- miRNA_RBPs |>
  mutate(
    pair = paste(RBP, miRNA, sep = "_")
  ) |>
  mutate(
    corr_data = pmap(
      list(RBP_ID, miRNA),
      compute_correlation,
      mirna_activity_df = miR_act
    )
  ) |>
  unnest(corr_data, keep_empty = TRUE) |>
  filter(!is.na(correlation_coefficient)) |>
  arrange(correlation_coefficient) |>
  mutate(
    FDR = p.adjust(pvalue, method = "fdr"),
    sign = Type
  ) |>
  as.data.frame()

# Number of candidate interactions that show the expected correlation
nrow(corr_table)

high_conf_ints <- corr_table[(corr_table$sign == "+" & sign(corr_table$correlation_coefficient) == 1) |
                              (corr_table$sign == "-" & sign(corr_table$correlation_coefficient) == -1), ]
nrow(high_conf_ints)

saveRDS(high_conf_ints, here("ENCODE_tissues_cell_lines", "output", "rds", "ENCODE_tissues_cell_lines_high_conf_ints.rds"))


# Number of candidate interactions with significant corration and |r| > 0.2 that show the expected correlation
top_corr <- corr_table[abs(corr_table$correlation_coefficient) > 0.2 & corr_table$FDR < 0.05, ]
nrow(top_corr)

nrow(top_corr[top_corr$sign == "+" & sign(top_corr$correlation_coefficient) == 1, ]) + 
  nrow(top_corr[top_corr$sign == "-" & sign(top_corr$correlation_coefficient) == -1, ])
```

# Barplot

```{r, fig.dim = c(6.6,3.8), fig.align = 'center'}
# Color labels according to the interaction sign inferred from the RBP knockdown analysis
cols <- ifelse(corr_table$sign == "-", "darkblue", "red")

pdf(here("ENCODE_tissues_cell_lines", "output", "figures", "ENCODE_tissues_cell_lines_corr_miRs_RBPs.pdf"), height=3.8, width=6.6)
par(mar=c(7,5,1,1))
x <- barplot(corr_table$correlation_coefficient, 
        ylab = "miRNA activity vs RBP expression \n(Pearson r)",
        ylim = c(-0.4, 0.6),
        cex.names = 0.8,
        cex.axis = 0.8,
        cex.lab=0.8,
        col=ifelse(corr_table$FDR < 0.05, "cornflowerblue","gray"))
text(x=x+0.25, y=-0.38, labels = corr_table$pair, srt=90, xpd=TRUE, cex=0.8, adj=c(1,0), col = cols)
legend(x=1, y = 0.5, legend = c("Adjusted P < 0.05", "NS"), fill = c("cornflowerblue","gray"), cex=0.8)
dev.off()

x <- barplot(corr_table$correlation_coefficient, 
        ylab = "miRNA activity vs RBP expression \n(Pearson r)",
        ylim = c(-0.4, 0.6),
        cex.names = 0.8,
        cex.axis = 0.8,
        cex.lab=0.8,
        col=ifelse(corr_table$FDR < 0.05, "cornflowerblue","gray"))
text(x=x+0.25, y=-0.38, labels = corr_table$pair, srt=90, xpd=TRUE, cex=0.8, adj=c(1,0), col = cols)
legend(x=1, y = 0.5, legend = c("Adjusted P < 0.05", "NS"), fill = c("cornflowerblue","gray"), cex=0.8)
```

```{r}
sessionInfo()
```