---
title: "Estimating mRNA stability changes in TCGA tumors"
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
set.seed(1)
```

# Load libraries

```{r}
library(MASS)
library(glue)
```

# Load the ddf (estimates of changes in stability), ddf_stdev (standard deviations) and dfi (estimates of change in pre-mRNA abundance) data obtained using BETAS 

We are analyzing all 17 cancer types that had at least 5 normal samples.

```{r}
# Extract cancer IDs from BETAS_ddf files 
cancer_IDs <- list.files(here("data"), pattern = "^TCGA-.*\\.BETAS_ddf\\.csv$") |>
  sub("TCGA-", "", x = _) |>
  sub(".BETAS_ddf.csv", "", x = _)

cat("Cancer IDs found:", paste(cancer_IDs, collapse = ", "), "\n")

# Helper function to load and convert CSV to numeric matrix
load_numeric_matrix <- function(filename) {
  read.csv(filename, row.names = 1) |>
    as.matrix() |>
    `mode<-`("numeric")
}
```

# Define std cutoff to retain closest to 10,000 genes as possible, and apply that filter

```{r}
# Helper function to find cutoff that retains ~10,000 genes
find_optimal_cutoff <- function(data, stdev, row_threshold = 0.8, col_threshold = 0.2, target_genes = 10000, increment = 0.05) {
  cutoff <- 0.1
  
  while (TRUE) {
    print(cutoff)
    # Mask low-confidence values
    data_masked <- data
    data_masked[stdev > cutoff] <- NA
    
    # Keep rows (genes) with at least 80% non-NA values in normal and tumor samples
    normal_cols <- grep("\\.11$", colnames(data_masked))
    tumor_cols <- grep("\\.01$", colnames(data_masked))
    
    row_completeness_normal <- rowSums(!is.na(data_masked[, normal_cols])) / length(normal_cols)
    row_completeness_tumor <- rowSums(!is.na(data_masked[, tumor_cols])) / length(tumor_cols)
    
    passing_rows <- intersect(
      which(row_completeness_normal >= row_threshold),
      which(row_completeness_tumor >= row_threshold)
    )
    
    data_with_rows_kept <- data_masked[passing_rows, ]
    
    # Remove samples (columns) with > 20% NAs
    col_completeness <- colSums(!is.na(data_with_rows_kept)) / nrow(data_with_rows_kept)
    passing_cols <- col_completeness > (1 - col_threshold)
    data_filtered <- data_with_rows_kept[, passing_cols]
    
    n_genes <- nrow(data_filtered)
    
    if (n_genes >= target_genes) break
    cutoff <- cutoff + increment
  }
  
  removed_samples <- colnames(data_with_rows_kept)[!colnames(data_with_rows_kept) %in% colnames(data_filtered)]
  
  list(data = data_filtered, cutoff = cutoff, n_genes = n_genes, removed_samples = removed_samples)
}
```

# Perform bias correction and generate partial residuals for all cancer types

```{r}
# Helper function to infer sample type from TCGA barcode suffix (.01 = tumor, .11 = normal)
extract_sample_type <- function(sample_names) {
  ifelse(grepl("\\.01$", sample_names), 1, 0)
}

# Helper function to perform robust regression and extract outputs
fit_rlm_row <- function(gene_idx, ddf_data, dfi_data, stype_vec) {
  fit <- rlm(ddf_data[gene_idx, ] ~ dfi_data[gene_idx, ] + stype_vec, 
             na.action = na.exclude, maxit = 100)
  coefs <- coef(fit)
  resids <- residuals(fit)
  partial <- resids + (coefs[3] * stype_vec)  # partial residual includes sample type effect
  
  list(residual = resids, coeff_dfi = coefs[2], coeff_stype = coefs[3], 
       partial_residual = partial)
}

# Process each cancer type
process_cancer_type <- function(cancer_ID) {
  cat("\n=== Processing", cancer_ID, "===\n")
  
  # Load data matrices
  ddf <- load_numeric_matrix(here("data", glue("TCGA-{cancer_ID}.BETAS_ddf.csv")))
  ddf_stdev <- load_numeric_matrix(here("data", glue("TCGA-{cancer_ID}.BETAS_ddf_stdev.csv")))
  dfi <- load_numeric_matrix(here("data", glue("TCGA-{cancer_ID}.BETAS_dfi.csv")))
  
  cat("Original dimensions - ddf:", nrow(ddf), "x", ncol(ddf), "\n")
  
  # Mask low-confidence ddf estimates and remove genes/samples with excessive missingness
  cutoff_result <- find_optimal_cutoff(ddf, ddf_stdev)
  ddf_fltr <- cutoff_result$data
  dfi_fltr <- dfi[rownames(ddf_fltr), colnames(ddf_fltr)]
  stype <- extract_sample_type(colnames(ddf_fltr))
  
  cat("Final cutoff:", cutoff_result$cutoff, "\n")
  cat("Filtered dimensions - ddf:", nrow(ddf_fltr), "x", ncol(ddf_fltr), "\n")
  
  # Fit robust regression for each gene
  cat("Fitting robust regression models...\n")
  results <- lapply(seq_len(nrow(ddf_fltr)), 
                   fit_rlm_row, ddf_fltr, dfi_fltr, stype)
  
  # Combine per-gene partial residuals into a gene × sample matrix
  partial_res <- do.call(rbind, lapply(results, `[[`, "partial_residual"))
  rownames(partial_res) <- rownames(ddf_fltr)
  colnames(partial_res) <- colnames(ddf_fltr)
  
  # Save results
  output_dir <- here("TCGA", "output", "supplementary_tables")
  output_file <- file.path(output_dir, 
    glue("TCGA-{cancer_ID}_stability_bias_corr.csv"))
  write.csv(partial_res, file = output_file)
  
  partial_res
}

# Process all cancer types
partial_residuals_list <- setNames(
  lapply(cancer_IDs, process_cancer_type),
  cancer_IDs
)
```

```{r}
sessionInfo()
```
