---
title: "Differential miRNA activity in TCGA tumors"
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
set.seed(1)
```

# Load libraries

```{r}
library(tibble)
library(dplyr)
library(purrr)
library(glue)
```

# Cancer types to analyze

```{r}
# Extract cancer IDs from BETAS_ddf files 
cancer_IDs <- list.files(here("TCGA", "data"), pattern = "BETAS_ddf\\.csv$") |>
  sub("TCGA-", "", x = _) |>
  sub(".BETAS_ddf.csv", "", x = _)

cat("Cancer IDs found:", paste(cancer_IDs, collapse = ", "), "\n")
```

# Load the 3'UTR nucleotide composition and length annotations, and miRNA seed matches

```{r}
UTR_nt <- read.table(here("ENCODE_RBP_KD", "data", "3UTR_nt_composition_new.txt"), header = TRUE, row.names = NULL)
UTR_length <- read.table(here("ENCODE_RBP_KD", "data", "3UTR_length_new.txt"), header = TRUE, row.names = NULL)
miR_binding <- read.csv(here("ENCODE_RBP_KD", "data", "miRNA_seed_matches_3UTRs2000bp.csv"), header = TRUE)
```

# Define miRNA name mapping

```{r}
mirna_name_mapping <- tibble(
  original = c(
    "miR-33ab", "miR-26ab", "miR-29abcd", "miR-27abc", "miR-133abc",
    "miR-199ab-5p", "miR-376abd", "miR-23abc", "miR-19ab", "miR-182",
    "miR-758", "miR-22", "miR-150", "miR-124", "miR-96", "miR-370",
    "miR-17", "miR-218", "miR-21", "miR-25", "miR-376c", "miR-10abc",
    "miR-143", "miR-34ac", "miR-338", "miR-377", "miR-495", "miR-544",
    "miR-18ab", "miR-299", "miR-328a", "let-7"
  ),
  canonical = c(
    "miR-33-5p", "miR-26-5p", "miR-29-3p", "miR-27-3p", "miR-133-3p",
    "miR-199-5p", "miR-376-3p", "miR-23-3p", "miR-19-3p", "miR-182-5p",
    "miR-758-3p", "miR-22-3p", "miR-150-5p", "miR-124-3p", "miR-96-5p",
    "miR-370-3p", "miR-17-5p", "miR-218-5p", "miR-21-5p", "miR-25-3p",
    "miR-376c-3p", "miR-10-5p", "miR-143-3p", "miR-34-5p", "miR-338-3p",
    "miR-377-3p", "miR-495-3p", "miR-544a-3p", "miR-18-5p", "miR-299-3p",
    "miR-328-3p", "let-7"
  )
)
```

# Define GLM fitting function

```{r}
fit_glm_sample <- function(glm_data, sample_name, cancer_ID) {
  # glm_data: complete data frame with response and all predictors
  # sample_name: name of the sample 
  # cancer_ID: cancer type identifier
  
  predictor_names <- names(glm_data)[
    names(glm_data) != "Gene_id" &
      !grepl("^TCGA\\.", names(glm_data))
  ]
  # Formula: response ~ confounders and miRNA affinity predictors
  # Intercept included automatically
  formula_str <- paste0(
    sample_name,
    " ~ ",
    paste(predictor_names, collapse = " + ")
  )
  
  fit <- glm(as.formula(formula_str), data = glm_data, na.action = na.exclude)
  coef_summary <- coef(summary(fit))
  
  # Extract coefficients and p-values for predictors only (miRNA cols, skip intercept and confounders)
  predictor_rows <- which(grepl("^(miR\\.|let\\.)", rownames(coef_summary)))
  
  tibble(
    miRNA = rownames(coef_summary)[predictor_rows],
    coefficient = coef_summary[predictor_rows, 1],
    pvalue = coef_summary[predictor_rows, 4],
    sample_name = sample_name,
    cancer_type = cancer_ID
  )
}
```

# Process all cancer types, fit glm and correction for multiple testing

```{r}
stab_dir <- here("TCGA", "output", "supplementary_tables")

# Prepare 3'UTR annotations (constant across cancer types)
UTR <- merge(UTR_nt, UTR_length, by = 1)
UTR_miRNA <- merge(UTR, miR_binding, by = 1)

# Function to process a single cancer type
process_cancer_type <- function(cancer_ID) {
  cat("Processing cancer type:", cancer_ID, "\n")
  
  # Load stability data for this cancer type
  stab_file <- file.path(stab_dir, glue("TCGA-{cancer_ID}_stability_bias_corr.csv"))
  stability <- read.csv(stab_file)
  
  # Extract sample names and count tumor/normal samples
  sample_names <- colnames(stability)[-1]
  n_tumor <- sum(grepl("\\.01$", sample_names))
  n_normal <- sum(grepl("\\.11$", sample_names))
  
  # Merge stability with 3'UTR annotations
  glmMatrix <- merge(stability, UTR_miRNA, by = 1)
  colnames(glmMatrix)[1] <- "Gene_id"
  
  # Remove columns that are only NAs
  glmMatrix <- glmMatrix |>
    select(where(~ !all(is.na(.))))
  
  # Fit GLM for all samples in this cancer type and apply FDR correction
  glm_data <- map_dfr(
    colnames(stability)[-1],
    ~ fit_glm_sample(glmMatrix, .x, cancer_ID)
  ) |>
    mutate(FDR = p.adjust(pvalue, method = "fdr"))
  
  # Return both GLM results and sample counts
  list(
    glm_results = glm_data,
    sample_counts = tibble(cancer_ID = cancer_ID, n_tumor = n_tumor, n_normal = n_normal)
  )
}

# Apply to all cancer IDs and combine results
all_results <- map(cancer_IDs, process_cancer_type)

# Extract GLM results and sample counts from the list
glm_results <- map_dfr(all_results, "glm_results")
sample_counts <- map_dfr(all_results, "sample_counts")
```

# Sample counts table by cancer type

```{r}
sample_counts

sum(sample_counts$n_tumor)
sum(sample_counts$n_normal)

length(sample_counts$cancer_ID)
```

# Convert inferred target stability effects into miRNA activity estimates by reversing the coefficient sign, express coefficients on a log2 scale

```{r}
glm_results <- glm_results |>
  mutate(
    coefficient = -coefficient,
    coefficient_log2 = log2(exp(coefficient))
  )

head(glm_results)
```

# Reformat miRNA names

```{r}
glm_results$miRNA <- gsub("\\.[A-Z].*", "", glm_results$miRNA)
glm_results$miRNA <- gsub("_.*", "", glm_results$miRNA)
glm_results$miRNA <- gsub("\\.", "-", glm_results$miRNA)

glm_results <- glm_results |>
  left_join(mirna_name_mapping, by = c("miRNA" = "original")) |>
  mutate(miRNA = coalesce(canonical, miRNA)) |>
  select(-canonical)
```

# Compute signed log10(p-values)

```{r}
glm_results$signed_logP <- -log10(glm_results$pvalue)
```

# Save results

```{r}
# Save all results in long-format tibble 
write.csv(glm_results, file = here("TCGA", "output", "supplementary_tables", "TCGA_miRNA_activity.csv"), row.names = FALSE)
```

# Determine the miRNAs that have significant changes in activity in at least 10 tumor samples for each cancer type

```{r}
sig_miRNAs <- glm_results %>%
  filter(grepl("\\.01$", sample_name)) %>%   # Keep only primary tumor samples
  filter(FDR < 0.05) %>%
  group_by(cancer_type, miRNA) %>%
  summarise(
    n_samples = n_distinct(sample_name),
    median_coefficient = median(coefficient, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_samples >= 10) %>%
  arrange(cancer_type, desc(n_samples))

sig_miRNAs

# Save
write.csv(sig_miRNAs, here("TCGA", "output", "supplementary_tables", "TCGA_sig_mirs_atleast10Tsamples_allCancers.csv"), row.names = F)
```


# Visualizing significantly dysregulated miRNAs

<==

Initialize tables that will contain the percentages of samples with a significant dysregulation for each miRNA, in each cancer. Separate tables for upregulated and downregulated miRNAs

```{r}
res_up <- data.matrix(matrix(NA, ncol = length(unique(act$cancers_vector)), nrow = (ncol(act) - 1)))
colnames(res_up) <- unique(act$cancers_vector)
row.names(res_up) <- colnames(act)[-ncol(act)]

res_down <- res_up
```

Fill in the table using the miRNA activity data

```{r}
for (miR in row.names(res_down)){
  for (cancer in colnames(res_down)){
    fdr_fltr <- FDR[FDR$cancers_vector == cancer, miR]
    act_fltr <- act[act$cancers_vector == cancer, miR]
    percent_up <- sum(fdr_fltr < 0.05 & act_fltr > 0, na.rm = T)/length(fdr_fltr)*100
    percent_down <- sum(fdr_fltr < 0.05 & act_fltr < 0, na.rm = T)/length(fdr_fltr)*100
    res_up[miR, cancer] <- percent_up
    res_down[miR, cancer] <- percent_down
  }
}
```

Remove miRNAs with only 0s from both tables

```{r}
rm <- which(rowSums(res_up) == 0 & rowSums(res_down) == 0)
res_up <- res_up[-rm, ]
res_down <- res_down[-rm, ]
```

Replace percentages with sqrt(%), to get the circle area (instead of diameter) proportional to the percentage, for making the heatmap

```{r}
res_up <- sqrt(res_up)
res_down <- sqrt(res_down)
```

Modify the name format of the miRNAs (short form of names, only the first miRNA in the family is retained (e.g. miR-17-5p, not miR-17-50/20....))

```{r}
lbls <- row.names(res_up)
lbls <- gsub("miR-33ab$", "miR-33-5p", lbls)
lbls <- gsub("miR-26ab$", "miR-26-5p", lbls)
lbls <- gsub("miR-29abcd$", "miR-29-3p", lbls)
lbls <- gsub("miR-27abc$", "miR-27-3p", lbls)
lbls <- gsub("miR-133abc$", "miR-133-3p", lbls)
lbls <- gsub("miR-199ab-5p$", "miR-199-5p", lbls)
lbls <- gsub("miR-376abd$", "miR-376-3p", lbls)
lbls <- gsub("miR-23abc$", "miR-23-3p", lbls)
lbls <- gsub("miR-19ab$", "miR-19-3p", lbls)
lbls <- gsub("miR-182$", "miR-182-5p", lbls)
lbls <- gsub("miR-758$", "miR-758-3p", lbls)
lbls <- gsub("miR-22$", "miR-22-3p", lbls)
lbls <- gsub("miR-150$", "miR-150-5p", lbls)
lbls <- gsub("miR-124$", "miR-124-3p", lbls)
lbls <- gsub("miR-96$", "miR-96-5p", lbls)
lbls <- gsub("miR-370$", "miR-370-3p", lbls)
lbls <- gsub("miR-17$", "miR-17-5p", lbls)
lbls <- gsub("miR-218$", "miR-218-5p", lbls)
lbls <- gsub("miR-21$", "miR-21-5p", lbls)
lbls <- gsub("miR-25$", "miR-25-3p", lbls)
lbls <- gsub("miR-376c$", "miR-376c-3p", lbls)
lbls <- gsub("miR-10abc$", "miR-10-5p", lbls)
lbls <- gsub("miR-143$", "miR-143-3p", lbls)
lbls <- gsub("miR-34ac$", "miR-34-5p", lbls)
lbls <- gsub("miR-338$", "miR-338-3p", lbls)
lbls <- gsub("miR-377$", "miR-377-3p", lbls)
lbls <- gsub("miR-495$", "miR-495-3p", lbls)
lbls <- gsub("miR-544$", "miR-544a-3p", lbls)
lbls <- gsub("miR-18ab$", "miR-18-5p", lbls)
lbls <- gsub("miR-299$", "miR-299-3p", lbls)
lbls <- gsub("miR-328a$", "miR-328-3p", lbls)
lbls <- gsub("miR-93$", "miR-302-3p", lbls) 
lbls <- gsub("miR.181abcd$", "miR-181-5p", lbls)
lbls <- gsub("miR-539$", "miR-539-5p", lbls)
lbls <- gsub("miR-450a$", "miR-450-5p", lbls)
lbls <- gsub("miR-135ab$", "miR-135-5p", lbls)
lbls <- gsub("miR-138$", "miR-138-5p", lbls)
lbls <- gsub("miR-141$", "miR-141-3p", lbls)
lbls <- gsub("miR-146ac$", "miR-146-5p", lbls)
lbls <- gsub("miR-153$", "miR-153-3p", lbls)
lbls <- gsub("miR-155$", "miR-155-5p", lbls)
lbls <- gsub("miR-15abc$", "miR-15-5p", lbls)
lbls <- gsub("miR-184$", "miR-184", lbls)
lbls <- gsub("miR-187$", "miR-187-3p", lbls)
lbls <- gsub("miR-191$", "miR-191-5p", lbls)
lbls <- gsub("miR-196abc$", "miR-196-5p", lbls)
lbls <- gsub("miR-1ab$", "miR-1-3p", lbls)
lbls <- gsub("miR-204$", "miR-204-5p", lbls)
lbls <- gsub("miR-184$", "miR-184", lbls)
lbls <- gsub("miR-210$", "miR-210-3p", lbls)
lbls <- gsub("miR-214$", "miR-214-3p", lbls)
lbls <- gsub("miR-221$", "miR-221-3p", lbls)
lbls <- gsub("miR-31$", "miR-31-5p", lbls)
lbls <- gsub("miR-375$", "miR-375", lbls)
lbls <- gsub("miR-383$", "miR-383-5p", lbls)
lbls <- gsub("miR-551a$", "miR-551-3p", lbls)
lbls <- gsub("miR-7$", "miR-7-5p", lbls)
lbls <- gsub("miR-9$", "miR-9-5p", lbls)
lbls <- gsub("miR-99ab$", "miR-99-5p", lbls)
lbls <- gsub("miR-136$", "miR-136-5p", lbls)
lbls <- gsub("miR-185$", "miR-185-5p", lbls)
lbls <- gsub("miR-300$", "miR-381-3p", lbls) ## This is not a typo 
lbls <- gsub("miR-326$", "miR-326", lbls)
lbls <- gsub("miR-329$", "miR-329-3p", lbls)
lbls <- gsub("miR-433$", "miR-433-3p", lbls)
lbls <- gsub("miR-487b$", "miR-487-3p", lbls)
lbls <- gsub("miR-504$", "miR-4725-5p", lbls)
lbls <- gsub("miR-505$", "miR-505-3p", lbls)
lbls <- gsub("miR-543$", "miR-543", lbls)
lbls <- gsub("miR-145$", "miR-145-5p", lbls)
lbls <- gsub("miR-101$", "miR-101-3p", lbls)
lbls <- gsub("miR-194$", "miR-194-5p", lbls)
lbls <- gsub("miR-203$", "miR-203a-3p", lbls) 
lbls <- gsub("miR-193$", "miR-193-3p", lbls)
lbls <- gsub("miR-216b$", "miR-216b-5p", lbls)
lbls <- gsub("miR-24$", "miR-24-3p", lbls)
lbls <- gsub("miR-496$", "miR-496", lbls)
lbls <- gsub("miR-448$", "miR-448", lbls)
lbls <- gsub("miR-216a$", "miR-216a-5p", lbls)
lbls <- gsub("miR-200bc$", "miR-200bc-3p", lbls)
lbls <- gsub("miR-874$", "miR-874-3p", lbls)
lbls <- gsub("miR-127$", "miR-127-3p", lbls)
lbls <- gsub("miR-217$", "miR-217", lbls)
lbls <- gsub("let-7$", "let-7-5p", lbls) 
lbls <- gsub("miR-30abcdef$", "miR-30-5p", lbls)
lbls <- gsub("miR-122$", "miR-122-5p", lbls)
row.names(res_up) <- lbls
row.names(res_down) <- lbls
```

Get the data after clustering

```{r, results = FALSE}
htmp <- heatmap.2(res_down, hclust=function(x) hclust(x,method="complete"), distfun = function(x) dist(x,method = 'euclidean'))
res_clust_down <- res_down[rev(htmp$rowInd), htmp$colInd]
res_clust_up <- res_up[rev(htmp$rowInd), htmp$colInd]
```

Prepare the data tables for making the heatmaps

```{r}
res_srt_up <- data.table(cbind(row.names(res_clust_up), res_clust_up))
colnames(res_srt_up)[1] <- "miRNAs"
res_melt_up <- melt(res_srt_up, id.vars = "miRNAs")
names(res_melt_up)[2:3] <- c("Cancers", "Percent samples with significant upregulation")
dat_htmp_up <- res_melt_up
dat_htmp_up$`Percent samples with significant upregulation` <- as.numeric(as.vector(dat_htmp_up$`Percent samples with significant upregulation`))

res_srt_down <- data.table(cbind(row.names(res_clust_down), res_clust_down))
colnames(res_srt_down)[1] <- "miRNAs"
res_melt_down <- melt(res_srt_down, id.vars = "miRNAs")
names(res_melt_down)[2:3] <- c("Cancers", "Percent samples with significant downregulation")
dat_htmp_down <- res_melt_down
dat_htmp_down$`Percent samples with significant downregulation` <- as.numeric(as.vector(dat_htmp_down$`Percent samples with significant downregulation`))
```

Set the data limits and squish the data accordingly

```{r}
res_lim_1 <- 1
res_lim_2 <- 10

dat_htmp_down$`Percent samples with significant downregulation`[dat_htmp_down$`Percent samples with significant downregulation` > 0 & dat_htmp_down$`Percent samples with significant downregulation` < res_lim_1] <- res_lim_1

dat_htmp_up$`Percent samples with significant upregulation`[dat_htmp_up$`Percent samples with significant upregulation` > 0 & dat_htmp_up$`Percent samples with significant upregulation` < res_lim_1] <- res_lim_1
```

Draw the heatmaps

```{r, eval=F}
pdf(paste("TCGA_pancancer_percentSamplesWithSigUpMiRsAct_circlesHeatmap.pdf", sep=""), width=9.5, height=15)
ggplot(dat_htmp_up) +
geom_point(aes(
  x = Cancers, 
  size = `Percent samples with significant upregulation`, 
  y = factor(miRNAs, levels = rev(unique(dat_htmp_up$miRNAs))),
  color=ifelse(`Percent samples with significant upregulation`>0, 'red', 'white'))) +
scale_color_identity() +
scale_size("Percent samples with significant\nmiRNA activity upregulation", range = c(0.1, 8), breaks = seq(0, res_lim_2, 2), limits=c(0, res_lim_2))+
labs(x="",y="") +
theme(legend.key.size = unit(1.5, 'lines'), text = element_text(size=12), axis.text = element_text(size=12), axis.text.x = element_text(angle=90, size=12, vjust=0.5, hjust=1), axis.text.y = element_text(size=8), panel.background = element_rect(fill = 'white', colour = 'white')) +
theme(plot.margin = unit(c(1,1,1,1), "cm")) 
dev.off()

pdf(paste("TCGA_pancancer_percentSamplesWithSigDownMiRsAct_circlesHeatmap.pdf", sep=""), width=9.5, height=15)
ggplot(dat_htmp_down) +
geom_point(aes(
  x = Cancers, 
  size = `Percent samples with significant downregulation`, 
  y = factor(miRNAs, levels = rev(unique(dat_htmp_down$miRNAs))),
  color=ifelse(`Percent samples with significant downregulation`>0, 'blue', 'white'))) +
scale_color_identity() +
scale_size("Percent samples with significant\nmiRNA activity downregulation", range = c(0.1, 8), breaks = seq(0, res_lim_2, 2), limits=c(0, res_lim_2))+
labs(x="",y="") +
theme(legend.key.size = unit(1.5, 'lines'), text = element_text(size=12), axis.text = element_text(size=12), axis.text.x = element_text(angle=90, size=12, vjust=0.5, hjust=1), axis.text.y = element_text(size=8), panel.background = element_rect(fill = 'white', colour = 'white')) +
theme(plot.margin = unit(c(1,1,1,1), "cm")) 
dev.off()
```

![Heatmap showing the percent of tumors with significant decrease in the activity of the miRNAs](TCGA_pancancer_percentSamplesWithSigDownMiRsAct_circlesHeatmap.pdf)

![Heatmap showing the percent of tumors with significant increase in the activity of the miRNAs](TCGA_pancancer_percentSamplesWithSigUpMiRsAct_circlesHeatmap.pdf)

# Pan-cancer miRNA activity illustrated using t-SNE plots

Define the function to make the t-SNE plots. This function is derived from code shared by Ariel Madrigal Aguirre.

```{r}
custom_theme_plot <- function(size_text=15,
                              leg_pos="right",
                              size_text_titles=NULL,
                              classic=TRUE,
                              line_size=0.85,
                              line_size_color="black",
                              axis_ticks_length=0.25,
                              axis_ticks_color="black"
                              ){
    require(ggplot2)
    if( is.null(size_text_titles) ){
        size_text_titles<- size_text + 2
    }
    if( classic ){
        gg.plots<- theme_classic()
    }else{
        gg.plots<- theme_bw()
    }
    gg.plots <- gg.plots +
        theme(axis.text.x=element_text(size=size_text, angle=45, hjust=1, color=line_size_color),
              axis.line=element_line(size=line_size, colour=line_size_color),
              axis.ticks=element_line(colour=axis_ticks_color, size=line_size),
              axis.ticks.length=unit(axis_ticks_length, "cm"),
              axis.title.x=element_text(size=size_text_titles),
              legend.position=leg_pos,
              plot.title=element_text(hjust=0.5, size=size_text),
              axis.text.y=element_text(size=size_text, color="black"),
              axis.title.y=element_text(size=size_text_titles) )
}
```

Keep miRNAs (columns) and tumors (rows) that have at least one significant hit at FDR < 0.05

```{r}
rowsKeep <- which(rowSums(FDR[, -ncol(FDR)] < 0.05) > 0)
colsKeep <- c(which(colSums(FDR[, -ncol(FDR)] < 0.05) > 0), ncol(FDR))
signedLogPvals_fltr <- signedLogPvals[rowsKeep, colsKeep]
act_fltr <- act[rowsKeep, colsKeep]
FDR_fltr <- FDR[rowsKeep, colsKeep]
```

Make the data object for plotting the t-SNE plots. Code derived from one shared by Ariel Madrigal Aguirre.

```{r}
dt <- act_fltr[ , -ncol(act_fltr)]
dt <- data.matrix(t(dt))
coldata <- as.data.frame(cbind(colnames(dt), signedLogPvals_fltr[, ncol(signedLogPvals_fltr)], signedLogPvals_fltr$miR.29abcd, signedLogPvals_fltr$miR.30abcdef, signedLogPvals_fltr$miR.19ab, signedLogPvals_fltr$miR.17, signedLogPvals_fltr$miR.200bc, signedLogPvals_fltr$miR.122))
colnames(coldata) <- c("sample", "cancer", "miR-29abcd", "miR-30abcdef", "miR-19ab", "miR-17", "miR-200bc", "miR-122")
coldata$`miR-29abcd` <- as.numeric(as.vector(coldata$`miR-29abcd`))
coldata$`miR-30abcdef` <- as.numeric(as.vector(coldata$`miR-30abcdef`))
coldata$`miR-19ab` <- as.numeric(as.vector(coldata$`miR-19ab`))
coldata$`miR-17` <- as.numeric(as.vector(coldata$`miR-17`))
coldata$`miR-200bc` <- as.numeric(as.vector(coldata$`miR-200bc`))
coldata$`miR-122` <- as.numeric(as.vector(coldata$`miR-122`))

sce <- SingleCellExperiment(list(logcounts=dt), colData=coldata)
sce
```

Adding PCA

```{r}
set.seed(5431)
sce_pca<- rpca(t(assay(sce, "logcounts")), k=100)
reducedDim(sce, "pca")<- sce_pca$x
```

t-SNE

```{r}
sce_tSNE <- runTSNE(sce, dimred="pca", perplexity=50) 
```

Set limits

```{r}
lims=2
```

Plot t-SNE based on miRNA activity for all miRNAs retained, and save the file

```{r}
plotReducedDim(sce_tSNE,
               ncomponents = 2,
               dimred="TSNE",
               point_size=2,               
               colour_by ="cancer") +
      custom_theme_plot() +  
      geom_point(shape = 1, size=2 ,colour = "grey", alpha=0.5)

pdf("TCGA_pancancer_miRsignedLogPvals_tSNE_cancer.pdf", height=10, width=10)
plotReducedDim(sce_tSNE,
               ncomponents = 2,
               dimred="TSNE",
               point_size=2,               
               colour_by ="cancer") +
      custom_theme_plot() +  
      geom_point(shape = 1, size=2 ,colour = "grey", alpha=0.5)
dev.off()
```

Plot the same t-SNE but showing miR-17 activity, and save the file

```{r}
plotReducedDim(sce_tSNE,
               ncomponents = 2,
               dimred="TSNE",
               point_size=2,               
               colour_by ="miR-17") +
    custom_theme_plot() +
    scale_colour_gradient2(low="blue", high="red", mid="white", midpoint = 0, limits=c(-lims, lims), oob=scales::squish, name="miRNA activity\nsigned log10(p-value)") +
    geom_point(shape = 1, size=2 ,colour = "grey", alpha=0.5)

pdf("TCGA_pancancer_miRsignedLogPvals_tSNE_miR17.pdf", height=10, width=10)
plotReducedDim(sce_tSNE,
               ncomponents = 2,
               dimred="TSNE",
               point_size=2,               
               colour_by ="miR-17") +
    custom_theme_plot() +
    scale_colour_gradient2(low="blue", high="red", mid="white", midpoint = 0, limits=c(-lims, lims), oob=scales::squish, name="miRNA activity\nsigned log10(p-value)") +
    geom_point(shape = 1, size=2 ,colour = "grey", alpha=0.5)
dev.off()
```

Repeat for miR-29

```{r}
plotReducedDim(sce_tSNE,
               ncomponents = 2,
               dimred="TSNE",
               point_size=2,               
               colour_by ="miR-29abcd") +
    custom_theme_plot() +
    scale_colour_gradient2(low="blue", high="red", mid="white", midpoint = 0, limits=c(-lims, lims), oob=scales::squish, name="miRNA activity\nsigned log10(p-value)") +
    geom_point(shape = 1, size=2 ,colour = "grey", alpha=0.5)

pdf("TCGA_pancancer_miRsignedLogPvals_tSNE_miR29.pdf", height=10, width=10)
plotReducedDim(sce_tSNE,
               ncomponents = 2,
               dimred="TSNE",
               point_size=2,               
               colour_by ="miR-29abcd") +
    custom_theme_plot() +
    scale_colour_gradient2(low="blue", high="red", mid="white", midpoint = 0, limits=c(-lims, lims), oob=scales::squish, name="miRNA activity\nsigned log10(p-value)") +
    geom_point(shape = 1, size=2 ,colour = "grey", alpha=0.5)
dev.off()
```

Repeat for miR-19

```{r}
plotReducedDim(sce_tSNE,
               ncomponents = 2,
               dimred="TSNE",
               point_size=2,               
               colour_by ="miR-19ab") +
    custom_theme_plot() +
    scale_colour_gradient2(low="blue", high="red", mid="white", midpoint = 0, limits=c(-lims, lims), oob=scales::squish, name="miRNA activity\nsigned log10(p-value)") +
    geom_point(shape = 1, size=2 ,colour = "grey", alpha=0.5)

pdf("TCGA_pancancer_miRsignedLogPvals_tSNE_miR19.pdf", height=10, width=10)
plotReducedDim(sce_tSNE,
               ncomponents = 2,
               dimred="TSNE",
               point_size=2,               
               colour_by ="miR-19ab") +
    custom_theme_plot() +
    scale_colour_gradient2(low="blue", high="red", mid="white", midpoint = 0, limits=c(-lims, lims), oob=scales::squish, name="miRNA activity\nsigned log10(p-value)") +
    geom_point(shape = 1, size=2 ,colour = "grey", alpha=0.5)
dev.off()
```

Same for miR-30

```{r}
plotReducedDim(sce_tSNE,
               ncomponents = 2,
               dimred="TSNE",
               point_size=2,               
               colour_by ="miR-30abcdef") +
    custom_theme_plot() +
    scale_colour_gradient2(low="blue", high="red", mid="white", midpoint = 0, limits=c(-lims, lims), oob=scales::squish, name="miRNA activity\nsigned log10(p-value)") +
    geom_point(shape = 1, size=2 ,colour = "grey", alpha=0.5)

pdf("TCGA_pancancer_miRsignedLogPvals_tSNE_miR30.pdf", height=10, width=10)
plotReducedDim(sce_tSNE,
               ncomponents = 2,
               dimred="TSNE",
               point_size=2,               
               colour_by ="miR-30abcdef") +
    custom_theme_plot() +
    scale_colour_gradient2(low="blue", high="red", mid="white", midpoint = 0, limits=c(-lims, lims), oob=scales::squish, name="miRNA activity\nsigned log10(p-value)") +
    geom_point(shape = 1, size=2 ,colour = "grey", alpha=0.5)
dev.off()
```

Same for miR-200

```{r}
plotReducedDim(sce_tSNE,
               ncomponents = 2,
               dimred="TSNE",
               point_size=2,               
               colour_by ="miR-200bc") +
    custom_theme_plot() +
    scale_colour_gradient2(low="blue", high="red", mid="white", midpoint = 0, limits=c(-lims, lims), oob=scales::squish, name="miRNA activity\nsigned log10(p-value)") +
    geom_point(shape = 1, size=2 ,colour = "grey", alpha=0.5)

pdf("TCGA_pancancer_miRsignedLogPvals_tSNE_miR200.pdf", height=10, width=10)
plotReducedDim(sce_tSNE,
               ncomponents = 2,
               dimred="TSNE",
               point_size=2,               
               colour_by ="miR-200bc") +
    custom_theme_plot() +
    scale_colour_gradient2(low="blue", high="red", mid="white", midpoint = 0, limits=c(-lims, lims), oob=scales::squish, name="miRNA activity\nsigned log10(p-value)") +
    geom_point(shape = 1, size=2 ,colour = "grey", alpha=0.5)
dev.off()
```

Same for miR-122

```{r}
plotReducedDim(sce_tSNE,
               ncomponents = 2,
               dimred="TSNE",
               point_size=2,               
               colour_by ="miR-122") +
    custom_theme_plot() +
    scale_colour_gradient2(low="blue", high="red", mid="white", midpoint = 0, limits=c(-lims, lims), oob=scales::squish, name="miRNA activity\nsigned log10(p-value)") +
    geom_point(shape = 1, size=2 ,colour = "grey", alpha=0.5)

pdf("TCGA_pancancer_miRsignedLogPvals_tSNE_miR122.pdf", height=10, width=10)
plotReducedDim(sce_tSNE,
               ncomponents = 2,
               dimred="TSNE",
               point_size=2,               
               colour_by ="miR-122") +
    custom_theme_plot() +
    scale_colour_gradient2(low="blue", high="red", mid="white", midpoint = 0, limits=c(-lims, lims), oob=scales::squish, name="miRNA activity\nsigned log10(p-value)") +
    geom_point(shape = 1, size=2 ,colour = "grey", alpha=0.5)
dev.off()
```

```{r}
sessionInfo()
```
