---
title: 'Association between miRNA activity and miRNA abundance across TCGA tumors'
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
set.seed(1)
```

# Load libraries

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(data.table)
```

# List of cancer types 

```{r}
glm_results <- read.csv(file = here("TCGA", "output", "supplementary_tables", "TCGA_miRNA_activity.csv"))
cancers <- unique(glm_results$cancer_type)
cancers
```

# Load the miRNA families annotation from TargetScan (https://www.targetscan.org/cgi-bin/targetscan/data_download.vert80.cgi), and filter for human miRNAs

```{r}
fams <- fread(here("TCGA", "data", "miR_Family_Info.txt"))
fams <- fams[grep("hsa-", fams$`MiRBase ID`), ] 
```

# Load the list of miRNA families with significant changes in activity in tumor vs normal samples in at least 10 tumor samples of each cancer type, excluding GBM

```{r}
sig_mirs <- read.csv(here("TCGA", "output", "supplementary_tables", "TCGA_sig_mirs_atleast10Tsamples_allCancers.csv"))
sig_mirs <- sig_mirs[sig_mirs$cancer_type != "GBM", ] # Remove GBM
miRNAs_og <- unique(sig_mirs$miRNA)
miRNAs_og
length(miRNAs_og)
```

# Retrieve corresponding miRNA family names from the TS annotation 

```{r}
miRNAs_TS_fam <- unique(fams$`miR family`[grep(paste(miRNAs_og, collapse="|"), fams$`miR family`)])
length(miRNAs_TS_fam)

miRNAs_TS_fam <- miRNAs_TS_fam[!(miRNAs_TS_fam %in% "miR-142-3p.2")]
length(miRNAs_TS_fam)
```

# Helper functions for correlation analysis

```{r}
# Load and preprocess miRNA abundance data for a specific cancer type, select tumor samples
load_mir_abundance <- function(cancer) {
  miR_expr <- read.table(
    here("TCGA", "data", paste0(cancer, ".miRseq_mature_RPM_log2.txt")),
    header = TRUE
  )
  
  # Reformat miR names and use as row names
  miR_expr$miRseqMature <- gsub("\\|.*", "", miR_expr$miRseqMature)
  row.names(miR_expr) <- miR_expr$miRseqMature
  miR_expr <- miR_expr[, -1]
  
  # Reformat sample names, select for primary tumors (01A)
  colnames(miR_expr) <- substr(colnames(miR_expr), 1, 16)
  miR_expr <- miR_expr[, grep("\\.01A$", colnames(miR_expr))]
  
  # Format sample names by removing trailing A
  colnames(miR_expr) <- gsub("A$", "", colnames(miR_expr))
  
  miR_expr
}

# Extract miRNA activity from glm_results for a specific cancer type (tumors only)
extract_mir_activity <- function(cancer, glm_results, miRNAs_og) {
  glm_results |>
    filter(cancer_type == cancer, grepl("\\.01$", sample_name)) |>
    filter(miRNA %in% miRNAs_og) |>
    select(sample_name, miRNA, coefficient_log2)
}

# Compute correlation between single miRNA abundance and activity pair
compute_single_correlation <- function(abundance_vals, activity_vals, abundance_samples = NULL, activity_samples = NULL) {
  # Verify that sample names match (if provided)
  if (!is.null(abundance_samples) && !is.null(activity_samples)) {
    stopifnot(
      "Sample names must be provided for both datasets" = 
        length(abundance_samples) == length(activity_samples),
      "Sample names must match in the same order" = 
        all(abundance_samples == activity_samples)
    )
  }
  
  # Skip if less than 5 valid samples in either dataset
  if (sum(!is.na(abundance_vals)) < 5 || sum(!is.na(activity_vals)) < 5) {
    return(list(estimate = NA, p.value = NA))
  }
  
  cor.test(abundance_vals, activity_vals)
}

# Compute correlation matrix between all mature miRNA - miRNA family pairs
compute_correlation_matrix <- function(miR_expr_T, miR_act_long, miROIs, miRNAs_og, common_samples) {
  coeffs <- matrix(NA, nrow = length(miRNAs_og), ncol = length(miROIs))
  rownames(coeffs) <- miRNAs_og
  colnames(coeffs) <- miROIs
  
  for (i in seq_along(miRNAs_og)) {
    mir_family <- miRNAs_og[i]
    for (j in seq_along(miROIs)) {
      mir_mature <- miROIs[j]
      
      # Extract abundance values for this mature miRNA in the same order as common_samples
      spec_miR_expr <- as.numeric(miR_expr_T[mir_mature, common_samples])
      abundance_samples <- common_samples
      
      # Extract activity values for this family from long format
      spec_miR_act_df <- miR_act_long |>
        filter(miRNA == mir_family, sample_name %in% common_samples) |>
        arrange(factor(sample_name, levels = common_samples))
      
      if (nrow(spec_miR_act_df) > 0) {
        activity_samples <- spec_miR_act_df$sample_name
        spec_miR_act <- spec_miR_act_df$coefficient_log2
        
        # Verify sample names match in the same order
        stopifnot(
          "Abundance samples must match activity samples" = 
            all(abundance_samples == activity_samples)
        )
        
        test_result <- compute_single_correlation(spec_miR_expr, spec_miR_act, 
                                                  abundance_samples, activity_samples)
        coeffs[i, j] <- test_result$estimate
      }
    }
  }
  
  coeffs
}

# Classify correlations as matching or non-matching families
classify_correlations <- function(coeffs, miROIs, fams) {
  corrs_matching <- numeric()
  corrs_nonMatching <- numeric()
  
  for (i in seq_len(nrow(coeffs))) {
    for (j in seq_len(ncol(coeffs))) {
      # Get the mature miRNA name
      mir_name <- colnames(coeffs)[j]
      
      # Get family name from TargetScan
      mir_fam <- fams[grep(mir_name, fams$`MiRBase ID`), ]$`miR family`[1]
      mir_fam <- gsub("/.*", "", mir_fam)
      mir_fam <- gsub("\\.1", "", mir_fam)
      
      
      if (mir_fam == rownames(coeffs)[i]) {
        corrs_matching[length(corrs_matching) + 1] <- coeffs[i, j]
      } else {
        corrs_nonMatching[length(corrs_nonMatching) + 1] <- coeffs[i, j]
      }
    }
  }
  
  list(matching = corrs_matching, nonMatching = corrs_nonMatching)
}

# Plot density distributions and wilcox test
plot_correlation_distributions <- function(corrs_matching, corrs_nonMatching, cancer) {
  # Remove NAs
  cm_clean <- corrs_matching[!is.na(corrs_matching)]
  cnm_clean <- corrs_nonMatching[!is.na(corrs_nonMatching)]
  
  plot(density(cnm_clean), xlim = c(-0.6, 1), xlab = "Pearson correlation",
       col = "dodgerblue", main = cancer, ylim = c(0, 4), cex.main = 2, 
       cex.lab = 1.8, cex.axis = 1.6)
  lines(density(cm_clean), col = "tomato")
  
  wilcox_p <- wilcox.test(corrs_matching, corrs_nonMatching)$p.value
  text(0.6, 3.7, paste("p =", formatC(wilcox_p, format = "e", digits = 2)), cex = 1.8)
}
```

# miRNA activity vs expression correlations

For each cancer and each pair of mature miRNA and miRNA family, correlate mature miRNA abundance and miRNA (family) activity. Output plots showing the probability density functions of Pearson correlations between inferred activity and abundance pairs of mature miRNAs and miRNA family that are matching, or not, in all TCGA cancers for which activity was inferred. miRNAs with significant changes in activity in at least 10 tumor samples in at least one cancer type are included.

```{r}
# Get all the miRNAs (MiRBase IDs) included in the miRNA families of interest
miROIs <- unique(gsub("\\.1$", "", 
  fams[fams$`miR family` %in% miRNAs_TS_fam, ]$`MiRBase ID`))

# Process each cancer type
results <- lapply(cancers, function(cancer) {
  # Load abundance data, filter for tumors
  miR_expr_T <- load_mir_abundance(cancer)
  
  # Extract activity from glm_results, filter for tumors
  miR_act_long <- extract_mir_activity(cancer, glm_results, miRNAs_og)
  
  # Find common samples
  common_samples <- intersect(unique(miR_act_long$sample_name), colnames(miR_expr_T))
  if (length(common_samples) == 0) return(NULL)
  
  miR_expr_T <- miR_expr_T[, common_samples]
  
  # Compute correlations
  coeffs <- compute_correlation_matrix(miR_expr_T, miR_act_long, miROIs, miRNAs_og, common_samples)
  
  # Classify as matching/non-matching
  corr_classification <- classify_correlations(coeffs, miROIs, fams)
  
  # Calculate percentages
  pct_matching <- sum(corr_classification$matching > 0, na.rm = TRUE) /
    sum(!is.na(corr_classification$matching)) * 100
  pct_nonMatching <- sum(corr_classification$nonMatching > 0, na.rm = TRUE) /
    sum(!is.na(corr_classification$nonMatching)) * 100
  
  list(
    cancer = cancer,
    corrs_matching = corr_classification$matching,
    corrs_nonMatching = corr_classification$nonMatching,
    pct_matching = pct_matching,
    pct_nonMatching = pct_nonMatching
  )
}) |> setNames(cancers)
```

# Density plots for correlation coefficients of matching and non-matching pairs  

```{r}
pdf(here("TCGA", "output", "figures", "TCGA_miRNA_act_vs_expr_density_plots.pdf"), 
    width = 14, height = 17)
par(mfrow = c(5, 4), mar = c(4, 4, 3, 2), mgp = c(2.5, 0.8, 0))

# Plot each cancer's density distributions in alphabetical order
for (cancer in sort(cancers)) {
  if (!is.null(results[[cancer]])) {
    plot_correlation_distributions(
      results[[cancer]]$corrs_matching,
      results[[cancer]]$corrs_nonMatching,
      cancer
    )
  }
}

# Add legend on a blank panel at the end
plot.new()
legend("center", 
       legend = c("miRNAs from same family", "miRNAs from different families"),
       col = c("tomato", "dodgerblue"), 
       lty = 1, 
       cex = 1.2)

dev.off()
```

# Average % of matching mature miRNA - miRNA family pairs for which there is a significant positive correlation between the activity and abundance, across all cancer types.

```{r}
# Extract results into separate lists
pancancer_corrs_matching <- setNames(
  lapply(results, \(x) x$corrs_matching),
  names(results)
)
pancancer_corrs_nonMatching <- setNames(
  lapply(results, \(x) x$corrs_nonMatching),
  names(results)
)
percents_pos_corrs_matching <- setNames(
  sapply(results, \(x) x$pct_matching),
  names(results)
)
percents_pos_corrs_nonMatching <- setNames(
  sapply(results, \(x) x$pct_nonMatching),
  names(results)
)
```

```{r}
mean(unlist(percents_pos_corrs_matching), na.rm = T)
```

# Average % of NON-matching mature miRNA - miRNA family pairs for which there is a significant positive correlation between the activity and abundance, across all cancer types.

```{r}
mean(unlist(percents_pos_corrs_nonMatching), na.rm = T)
```

# Violin plots of correlation coefficients for matching pairs in each cancer

```{r}
# Make a table containing these mature miRNA - miRNA family pairs for all cancers, with the correlation coefficients and cancer type name. Also remove NAs
pancancer_corrs_matching_df <- data.table(cbind(names(pancancer_corrs_matching[[1]]), do.call(cbind, pancancer_corrs_matching)))
miRs_table <- melt(pancancer_corrs_matching_df,id.vars = "V1")
colnames(miRs_table) <- c("miRNAs", "cancer", "correlation")
miRs_table$correlation <- as.numeric(miRs_table$correlation)
miRs_table <- miRs_table[!(is.na(miRs_table$correlation)), ]

# Sort by the median correlation coefficient in each cancer. Save for later use

medians <- aggregate(as.vector(miRs_table$correlation), list(miRs_table$cancer), median)
order_cancers <- as.vector(medians[,1][order(medians[,2], decreasing = T)])
miRs_table$cancer <- reorder.factor(miRs_table$cancer, new.order=order_cancers)
miRs_table <- miRs_table %>% arrange(cancer)

# saveRDS(miRs_table, here("TCGA", "output", "rds",  "TCGA_miRfamsSig10T_corr_act_vs_expr_res.rds"))

# Draw the violin plot
ggplot(miRs_table, aes(x = cancer, y = correlation)) + 
  geom_violin(fill = "grey80", colour = "#3366FF") +
  geom_boxplot(width = 0.07) +
  theme_classic() +
  ylab("Pearson correlation") + 
  xlab("") + 
  theme(axis.text=element_text(size=16),
        axis.title=element_text(size=20),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "none") +
  geom_hline(yintercept=0, linetype="dashed", 
            color = "black", size=0.5)
```

# Heatmap

```{r}
# Arrange the cancer types in the same order as in the violin plot. Remove miRNAs that only have NAs and replace remaining NAs with 0s.
dat_htmp <- as.data.frame(do.call(cbind, pancancer_corrs_matching))
dat_htmp <- dat_htmp[, as.vector(unique(miRs_table$cancer))]
dat_htmp <- data.matrix(dat_htmp)
rows_keep <- which(rowSums(is.na(dat_htmp)) != ncol(dat_htmp))
dat_htmp <- dat_htmp[rows_keep, ] 
dat_htmp[is.na(dat_htmp)] <- 0

# Draw the heatmap
palette <- colorRampPalette(c("blue", "white", "red"))
pdf(here("TCGA", "output", "figures", "TCGA_miRfamsSig10T_corr_act_vs_expr_htmp.pdf"), width=18, height=7)
heatmap.2(dat_htmp, trace="none", col=palette, cexRow=0.01, cexCol=2.5,  breaks=seq(-1,1,length.out=50), mar=c(10,20), keysize=1, hclust=function(x) hclust(x,method="complete"), distfun = function(x) dist(x,method = 'euclidean'), density.info = "none", key.par = list(cex=2), lhei=c(1,2), lwid = c(1,5), Colv = F, key.title = "", key.xlab = "Pearson Correlation")
dev.off()
```

# Helper function to create scatter plots

```{r}
plot_mir_abundance_vs_activity <- function(cancer, miRNA_fam, miRNA_mature, glm_results) {
  # Load abundance
  miR_expr_T <- load_mir_abundance(cancer)
  
  # Extract activity (long format)
  miR_act_long <- extract_mir_activity(cancer, glm_results, c(miRNA_fam))
  
  # Match samples
  common_samples <- intersect(unique(miR_act_long$sample_name), colnames(miR_expr_T))
  miR_expr_T <- miR_expr_T[, common_samples]
  miR_act_long <- miR_act_long |> filter(sample_name %in% common_samples)
  
  # Check if miRNA family exists
  if (nrow(miR_act_long) == 0) {
    cat("Warning: miRNA family", miRNA_fam, "not found in activity data\n")
    return(invisible())
  }
  
  # Sort by sample name for correct alignment
  miR_act_long <- miR_act_long |> arrange(factor(sample_name, levels = common_samples))
  expr_samples <- common_samples
  expr_vals <- as.numeric(miR_expr_T[miRNA_mature, common_samples])
  act_samples <- miR_act_long$sample_name
  act_vals <- miR_act_long$coefficient_log2
  
  # Verify sample names match in the same order
  stopifnot(
    "Expression and activity samples must match in the same order" = 
      all(expr_samples == act_samples)
  )
  
  # Create plot
  plot(expr_vals, act_vals, pch = 16, cex = 0.5,
       xlab = "miRNA expression log2(RPM)", ylab = "miRNA activity")
  abline(lm(act_vals ~ expr_vals), col = 4, lwd = 2)
  
  # Add correlation
  r <- cor.test(expr_vals, act_vals)$estimate
  text(5.4, 0.09, paste("r =", round(r, 3)))
}
```

# Example scatter plot for hsa-miR-122-5p in LIHC

```{r}
plot_mir_abundance_vs_activity("LIHC", "miR-122", "hsa-miR-122-5p", glm_results)
```

# Example scatter plot for hsa-miR-204-5p in PRAD

```{r}
plot_mir_abundance_vs_activity("PRAD", "miR-204", "hsa-miR-204-5p", glm_results)
```

Show, as examples, the individual coefficients for the correlations between inferred activity of each miRNA family included above with the abundance of each of their family members, in TCGA-KIRC and TCGA-LUAD 

```{r}
cancer="KIRC"

# Make a color vector for the number of miRNA families
clrs <- c("dodgerblue","tomato1","chartreuse","maroon4","gray43","lightpink","darkorchid4", "plum","darkseagreen","darksalmon","gray77", "firebrick","orange","lightseagreen","lightslateblue","blue","magenta")

# Retrieve the miRNA families corresponding to each mature miRNA in the results
mirs_fams <- names(pancancer_corrs_matching[[cancer]]) 
for(i in mirs_fams){
  if (i %in% fams$`MiRBase ID`){
    fam_ts <- fams[fams$`MiRBase ID` == i, ]$`miR family`
    mirs_fams[mirs_fams == i] <- fam_ts
  } else if (paste(i, ".1", sep="") %in% fams$`MiRBase ID`) {
    fam_ts <- fams[fams$`MiRBase ID` == paste(i, ".1", sep=""), ]$`miR family`
    mirs_fams[mirs_fams == i] <- fam_ts
  } else {
    stop("ERROR: miR not found")
  }
}
# Define the number of mature miRNAs per family
fams_tbl <- table(mirs_fams)

# Reorder and assign the colors
fams_tbl <- fams_tbl[miRNAs_TS_fam]
colors <- rep(clrs, as.vector(fams_tbl))

# Draw the barplot
barplot(pancancer_corrs_matching[[cancer]][!(is.na(pancancer_corrs_matching[[cancer]]))], col=colors[!(is.na(pancancer_corrs_matching[[cancer]]))], ylab="Pearson correlation", las=2, cex.names = 0.6)
plot.new()
legend("center", legend = miRNAs_TS_fam, fill=clrs, title = "miRNA families", text.font=1, xpd = F, cex = 0.8)
```

Repeat the example above for LUAD

```{r}
cancer="LUAD"

# Make a color vector for the number of miRNA families
clrs <- c("dodgerblue","tomato1","chartreuse","maroon4","gray43","lightpink","darkorchid4", "plum","darkseagreen","darksalmon","gray77", "firebrick","orange","lightseagreen","lightslateblue","blue","magenta")

# Retrieve the miRNA families corresponding to each mature miRNA in the results
mirs_fams <- names(pancancer_corrs_matching[[cancer]]) 
for(i in mirs_fams){
  if (i %in% fams$`MiRBase ID`){
    fam_ts <- fams[fams$`MiRBase ID` == i, ]$`miR family`
    mirs_fams[mirs_fams == i] <- fam_ts
  } else if (paste(i, ".1", sep="") %in% fams$`MiRBase ID`) {
    fam_ts <- fams[fams$`MiRBase ID` == paste(i, ".1", sep=""), ]$`miR family`
    mirs_fams[mirs_fams == i] <- fam_ts
  } else {
    stop("ERROR: miR not found")
  }
}
# Define the number of mature miRNAs per family
fams_tbl <- table(mirs_fams)

# Reorder and assign the colors
fams_tbl <- fams_tbl[miRNAs_TS_fam]
colors <- rep(clrs, as.vector(fams_tbl))

# Draw the barplot
barplot(pancancer_corrs_matching[[cancer]][!(is.na(pancancer_corrs_matching[[cancer]]))], col=colors[!(is.na(pancancer_corrs_matching[[cancer]]))], ylab="Pearson correlation", las=2, cex.names = 0.6)
```

```{r}
sessionInfo()
```