---
title: 'Association between miRNA transcription and miRNA activity across TCGA tumors'
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
```

# Load libraries

```{r}
library(rtracklayer)
library(GenomicRanges)
library(data.table)
library(purrr)
library(tibble)
library(stringr)
library(dplyr)
library(tidymodels)
library(gplots)
library(ggplot2)
```

# Select miRNA host genes that contain miRNA loci within introns

We use the miRNA annotation from miRBase ftp://mirbase.smith.man.ac.uk/pub/mirbase/CURRENT/genomes/

```{r}
# Load the miRBase annotation
mir_annot <- import(here("TCGA", "data", "hsa.gff3"))
mir_annot <- as.data.frame(mir_annot)

# Select for pri-miRNAs
pri_mir_annot <- mir_annot %>% filter(type == "miRNA_primary_transcript")

head(pri_mir_annot)
```

```{r}
# Load the intron annotation
intron_annot <- import(here("TCGA", "data", "Homo_sapiens.GRCh38.85.chr.Introns.gtf"))
intron_annot <- as.data.frame(intron_annot)

head(intron_annot)
```

```{r}
# Use GenomicRanges to detect coordinate overlap 
pri_mir_gr <- GRanges(
  seqnames = pri_mir_annot$seqnames,
  ranges = IRanges(start = pri_mir_annot$start, end = pri_mir_annot$end),
  strand = pri_mir_annot$strand,
  pri_mir_id = pri_mir_annot$Name
)

intron_gr <- GRanges(
  seqnames = intron_annot$seqnames,
  ranges = IRanges(start = intron_annot$start, end = intron_annot$end),
  strand = intron_annot$strand,
  gene_id = intron_annot$gene_id
)

# Find all pri-miRNAs contained within introns
overlaps <- findOverlaps(pri_mir_gr, intron_gr, type = "within")
mir_gene_pairs <- as.data.frame(overlaps) |>
  mutate(
    pri_mir_id = pri_mir_gr$pri_mir_id[queryHits],
    gene_id = intron_gr$gene_id[subjectHits]
  ) |>
  select(gene_id, pri_mir_id)

# Save 
write.csv(mir_gene_pairs, file=here("TCGA", "output", "supplementary_tables", "GRCh38_miRs_in_introns.csv"))
```

# Load the estimates of changes in mRNA abundance (dfe) and transcription (dfi) across TCGA tumors, from BETAS

```{r}
# Helper function to load and format BETAS data
load_betas_data <- function(files) {
  map_df(files, function(f) {
    dt <- fread(f) |>
      as.data.frame()
    rownames(dt) <- dt[[1]]
    dt[[1]] <- NULL
    dt
  }, .id = "source") |>
    rownames_to_column("gene_id") |>
    pivot_longer(
      cols = -c(source, gene_id),
      names_to = "sample",
      values_to = "value"
    )
}

dfi_files <- Sys.glob(here("TCGA", "data", "TCGA-*.BETAS_dfi.csv"))
dfe_files <- Sys.glob(here("TCGA", "data", "TCGA-*.BETAS_dfe.csv"))

# Extract cancer type from filename
cancers <- dfi_files |>
  basename() |>
  str_extract("TCGA-[A-Z]+") |>
  str_remove("TCGA-")

# Load dfi and dfe data, indicate cancer type 
dfi_data_list <- map2(dfi_files, cancers, function(f, cancer) {
  fread(f) |>
    as.data.frame() |>
    (\(dt) {rownames(dt) <- dt[[1]]; dt[, -1]})() |>
    as.matrix() |>
    as.data.frame() |>
    rownames_to_column("gene_id") |>
    mutate(cancer = cancer)
})

dfe_data_list <- map2(dfe_files, cancers, function(f, cancer) {
  fread(f) |>
    as.data.frame() |>
    (\(dt) {rownames(dt) <- dt[[1]]; dt[, -1]})() |>
    as.matrix() |>
    as.data.frame() |>
    rownames_to_column("gene_id") |>
    mutate(cancer = cancer)
})

# Combine into long tables
# Pivot each data frame before binding 
dfi_long <- map_df(dfi_data_list, ~{
  .x |>
    pivot_longer(-c(gene_id, cancer), names_to = "sample", values_to = "dfi")
})

dfe_long <- map_df(dfe_data_list, ~{
  .x |>
    pivot_longer(-c(gene_id, cancer), names_to = "sample", values_to = "dfe")
})
```

# For each cancer type, for each miRNA host gene, test the correlation between changes in mRNA abundance (dfe) and transcription (dfi)

```{r}
# Compute correlations for each gene-cancer combination
dfe_dfi_results <- bind_cols(
  dfi_long |> select(gene_id, cancer, sample, dfi),
  dfe_long |> select(dfe)
) |>
  
  # Filter to tumor samples (those ending in -01)
  filter(str_detect(sample, "\\-01$")) |>
  
  # Remove the -1 suffix
  mutate(sample = str_remove(sample, "\\-1$")) |>
  
  # Filter to genes in mir_gene_pairs
  semi_join(
    mir_gene_pairs |> select(gene_id) |> distinct() |> rename(gene_id = gene_id),
    by = "gene_id"
  ) |>
  
  # Remove rows where all values in a gene-cancer group are NA
  group_by(gene_id, cancer) |>
  filter(!all(is.na(dfi) | is.na(dfe))) |>
  
  # Compute correlation statistics
  nest(data = -c(gene_id, cancer)) |>
  mutate(
    corr_test = map(data, ~{
      dat <- .x |> filter(!is.na(dfi), !is.na(dfe))
      if (nrow(dat) < 3) return(NA)  # Need at least 3 points for correlation
      
      corr <- cor.test(dat$dfi, dat$dfe)
      fit <- lm(dfe ~ dfi, data = dat)
      
      tibble(
        correlation = corr$estimate,
        p_value = corr$p.value,
        r_squared = summary(fit)$r.squared,
        n = nrow(dat)
      )
    }),
    .keep = "unused"
  ) |>
  unnest(corr_test)

# FDR correction
dfe_dfi_results <- dfe_dfi_results |>
  mutate(
    fdr = p.adjust(p_value, method = "fdr"),
    signed_r2 = sign(correlation) * sqrt(abs(r_squared))
  )

# Save results
# saveRDS(dfe_dfi_results, here("TCGA", "output", "rds", "TCGA_miRhostGenes_dfeVSdfi_results.rds"))
```

# Heatmap of correlation coefficients for dfe vs dfi for all genes that contain a miRNA in their introns

```{r}
# Create matrix for heatmap
signed_r2_matrix <- dfe_dfi_results |>
  select(gene_id, cancer, signed_r2) |>
  pivot_wider(names_from = cancer, values_from = signed_r2, values_fill = NA) |>
  column_to_rownames("gene_id") |>
  as.matrix()

# Remove rows and columns with only NAs
signed_r2_matrix <- signed_r2_matrix[
  rowSums(is.na(signed_r2_matrix)) != ncol(signed_r2_matrix),
  colSums(is.na(signed_r2_matrix)) != nrow(signed_r2_matrix)
]

palette <- colorRampPalette(c("blue", "white", "red"))

pdf(here("TCGA", "output", "figures", "TCGA_miRhostGenes_dfeVSdfi_heatmap.pdf"), width = 5.5, height = 6)
heatmap.2(signed_r2_matrix, trace = "none", col = palette, cexRow = 0.01, cexCol = 1.1,
  breaks = seq(-1, 1, length.out = 100), mar = c(5, 1), keysize = 0.2,
  hclust = function(x) hclust(x, method = "average"),
  distfun = function(x) dist(x, method = "euclidean"),
  density.info = "none", key.par = list(cex = 1), lhei = c(1, 3), lwid = c(1, 1.5))
dev.off()
```

# Filter miRNA-gene pairs for those where changes in abundance relative to the average (dfe) are due at least in part to changes in transcription (dfi). Specifically, retain pairs where the host gene's dfi vs dfe correlation was significant (after FDR correction) and positive in at least 80% of cancers.

```{r}
dim(dfe_dfi_results)
genes_sig_pos <- dfe_dfi_results |>
  filter(fdr < 0.05, correlation > 0) |>
  group_by(gene_id) |>
  summarise(sig_pos_freq = n(), .groups = "drop") |>
  # Keep genes with significant positive correlation in ≥80% of cancers
  filter(sig_pos_freq / length(cancers) >= 0.8) |>
  pull(gene_id)
length(genes_sig_pos)
```

# Correlation between host gene transcription changes and miRNA activity, for significant genes identified above

First make a table of host gene id, pri-miRNA id and miRNA family

```{r}
# Selected host genes and associated miRNA
mir_gene_pairs_sig <- mir_gene_pairs[mir_gene_pairs$gene_id %in% genes_sig_pos,]

# Load the miRNA families annotation from TargetScan
fams <- fread(here("TCGA", "data", "miR_Family_Info.txt"))
fams <- fams[grep("hsa-", fams$`MiRBase ID`), ] 

# Format pri-miRNA IDs to better match
mir_gene_pairs_sig$pri_mir_id <- gsub("mir-", "miR-", mir_gene_pairs_sig$pri_mir_id)

# Convert pri-miRNA names to miRNA family names compatible with the activity data
fams$miR_family_2 <- gsub("/.*", "", fams$`miR family`)
fams$pri_mir_id <- gsub("-3p", "", fams$`MiRBase ID`)
fams$pri_mir_id <- gsub("-5p", "", fams$pri_mir_id)
fams <- fams[-grep("\\.1$", fams$pri_mir_id), ]
fams <- fams[-grep("\\.2$", fams$pri_mir_id), ]
fams <- fams[, c("pri_mir_id","miR_family_2")]
fams <- unique(fams)

# Merge TS families in the gene-miRNA pairs table
mir_gene_pairs_sig <- mir_gene_pairs_sig %>%
  left_join(
    fams %>%
      select(pri_mir_id, miR_family = miR_family_2),
    by = "pri_mir_id"
  )
dim(mir_gene_pairs_sig)

# For the pairs that got no match (NA miR_family), try to remove the -1/2 at the end of the pri-mir to get a match
mir_gene_pairs_sig$pri_mir_id_2 <- mir_gene_pairs_sig$pri_mir_id
mir_gene_pairs_sig$pri_mir_id_2[is.na(mir_gene_pairs_sig$miR_family)] <- gsub("-1$", "", mir_gene_pairs_sig$pri_mir_id_2[is.na(mir_gene_pairs_sig$miR_family)])
mir_gene_pairs_sig$pri_mir_id_2[is.na(mir_gene_pairs_sig$miR_family)] <- gsub("-2$", "", mir_gene_pairs_sig$pri_mir_id_2[is.na(mir_gene_pairs_sig$miR_family)])
mir_gene_pairs_sig$pri_mir_id_2[is.na(mir_gene_pairs_sig$miR_family)] <- gsub("-3$", "", mir_gene_pairs_sig$pri_mir_id_2[is.na(mir_gene_pairs_sig$miR_family)])
mir_gene_pairs_sig$pri_mir_id_2[is.na(mir_gene_pairs_sig$miR_family)] <- gsub("-4$", "", mir_gene_pairs_sig$pri_mir_id_2[is.na(mir_gene_pairs_sig$miR_family)])
mir_gene_pairs_sig$pri_mir_id_2[is.na(mir_gene_pairs_sig$miR_family)] <- gsub("-5$", "", mir_gene_pairs_sig$pri_mir_id_2[is.na(mir_gene_pairs_sig$miR_family)])
mir_gene_pairs_sig$miR_family[is.na(mir_gene_pairs_sig$miR_family)] <- fams$miR_family_2[match(mir_gene_pairs_sig$pri_mir_id_2[is.na(mir_gene_pairs_sig$miR_family)], fams$pri_mir_id)]
mir_gene_pairs_sig[is.na(mir_gene_pairs_sig$miR_family), ]
dim(mir_gene_pairs_sig)

# Remove gene_pri-mir pairs without any found miRNA family in the TS annotation
mir_gene_pairs_sig <- mir_gene_pairs_sig[!(is.na(mir_gene_pairs_sig$miR_family)), ]
dim(mir_gene_pairs_sig)

# Load miRNA activity 
miR_act <- read.csv(here("TCGA", "output", "supplementary_tables", "TCGA_miRNA_activity.csv"))
miR_act$sample_name <- gsub("\\.", "-", miR_act$sample_name)

# Retain gene_pri-mir pairs that are present in the miRNA activity data
mir_gene_pairs_sig <- mir_gene_pairs_sig[mir_gene_pairs_sig$miR_family %in% miR_act$miRNA, ]
dim(mir_gene_pairs_sig)
```

```{r}
# Get the dfi results for the selected host genes in mir_gene_pairs_sig, merge the host miRNA info ONTO IT
dfi_sig_hosts <- dfi_long[dfi_long$gene_id %in% mir_gene_pairs_sig$gene_id, ]

dfi_sig_hosts <- dfi_sig_hosts %>%
  left_join(
    mir_gene_pairs_sig %>%
      select(gene_id, pri_mir_id, miR_family),
    by = "gene_id", 
    relationship =
  "many-to-many"
  )
dim(dfi_sig_hosts)

# Merge miRNA activity results onto that based on miRNA family, cancer, and sample name
dfi_sig_hosts <- dfi_sig_hosts %>%
  left_join(
    miR_act %>%
      select(
        miRNA,
        cancer_type,
        sample_name,
        coefficient_log2
      ),
    by = c(
      "miR_family" = "miRNA",
      "cancer" = "cancer_type",
      "sample" = "sample_name"
    )
  )
dim(dfi_sig_hosts)
```

# For each pri-miRNA-host-gene pair retained, for each cancer type, correlate miRNA activity and host gene transcriptional changes across samples

```{r}
dim(dfi_sig_hosts)
cor_results <- dfi_sig_hosts %>%
  # Keep only primary tumor samples ending in -01
  filter(grepl("-01$", sample)) %>%
  
  # Correlate within each cancer and gene/pri-miR/family pair
  group_by(gene_id, pri_mir_id, miR_family, cancer) %>%
  summarise(
    test = list(cor.test(coefficient_log2, dfi, method = "pearson")),
    .groups = "drop"
  ) %>%
  mutate(
    r = sapply(test, \(x) unname(x$estimate)),
    p = sapply(test, \(x) x$p.value),
    r_squared = r^2
  ) %>%
  select(
    gene_id, pri_mir_id, miR_family, cancer,
    r, p, r_squared
  )
dim(cor_results)
head(cor_results)

# Sort by median r in each cancer, keeping the cancers together
medians <- aggregate(cor_results$r, list(cor_results$cancer), median)
order_cancers <- as.vector(medians[,1][order(medians[,2], decreasing = T)])
cor_results$cancer <- factor(cor_results$cancer, levels = order_cancers)
cor_results <- cor_results %>% arrange(cancer)
```

# Violin plots of correlation coefficients 

```{r}
p <- ggplot(cor_results, aes(x = cancer, y = r)) + 
  geom_violin(fill = "grey80", colour = "#3366FF") +
  geom_boxplot(width = 0.07) +
  theme_classic() +
  ylab("Pearson r") + 
  xlab("") + 
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=14),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "none") +
  geom_hline(yintercept=0, linetype="dashed", 
            color = "black", size=0.5)
p

pdf(here("TCGA", "output", "figures", "TCGA_miRNA_act_vs_transcription_violin.pdf"), 7, 3.5)
p
dev.off()
```

# Corresponding heatmap 

```{r}
# Convert long to wide format with host gene/pri-mir/family pair as rows and cancers as columns
dat_htmp <- cor_results[, -(1:3)]
dat_htmp$pair <- paste(cor_results$gene_id, cor_results$pri_mir_id, cor_results$miR_family, sep = "_")

dat_htmp <- dat_htmp %>%
  select(pair, cancer, r) %>%
  pivot_wider(
    names_from = cancer,
    values_from = r
  )

dat_htmp <- column_to_rownames(dat_htmp, "pair")
dat_htmp <- data.matrix(dat_htmp)

# Draw the heatmap
palette <- colorRampPalette(c("blue", "white", "red"))
pdf(here("TCGA", "output", "figures", "TCGA_miRNA_act_vs_transcription_htmp.pdf"), width=18, height=7)
heatmap.2(dat_htmp, trace="none", col=palette, cexRow=0.01, cexCol=2,  breaks=seq(-1,1,length.out=50), mar=c(10,10), keysize=1, hclust=function(x) hclust(x,method="complete"), distfun = function(x) dist(x,method = 'euclidean'), density.info = "none", key.par = list(cex=2), lhei=c(1,1.2), lwid = c(1,2), Colv = F, key.title = "", key.xlab = "Pearson r")
dev.off()
```

# Average correlation between miRNA activity and rate of transcription of their primary transcripts

```{r}
mean(cor_results$r)
```

Example scatter plot for miR-193 in STAD

```{r}
<==
miR <- "miR-139"
gene <- "ENSG00000186642"
cancer="STAD"

# Filter dfi table for the gene of interest
dfi_gene <- lapply(dfi_data, function(x) unlist(x[gene, ]))
dfi_gene <- unlist(dfi_gene)

# Make a vector indicating the cancer type of each sample
nb_samples_cancers <- unlist(lapply(dfi_data, function(x) ncol(x)))
cancers_vector <- rep(cancers, nb_samples_cancers)

# Merge into table containing all samples of all cancer types and indicating the cancer type
dfi_complete <- as.data.frame(cbind(dfi_gene, cancers_vector))
colnames(dfi_complete) <- c("GOI_dfi", "cancer")

# Filter miRNA activity for the miRNA of interest
miRNA <- as.vector(miR)
miR_cancers <- as.data.frame(cbind(row.names(miR_act), miR_act[,miRNA], as.vector(miR_act$cancers_vector)))
colnames(miR_cancers) <- c("samples", "miR_act", "cancer")

# Make table with the samples, dfi for the host gene and miRNA activity in that cancer type, removing normal samples
cancer_data <- dfi_complete[dfi_complete$cancer== cancer, ]
row.names(cancer_data) <- gsub("-1$", "", row.names(cancer_data))
cancer_data <- cancer_data[grep("\\-01$", row.names(cancer_data)), ]
cancer_data$miRNA_act <- NA
cancer_miR <- miR_cancers[miR_cancers$cancer == cancer, ]
cancer_data$miRNA_act <- cancer_miR[match(row.names(cancer_data), cancer_miR$samples),]$miR_act

# Correlation
corr <- cor.test(as.numeric(as.vector(cancer_data$GOI_dfi)), as.numeric(as.vector(cancer_data$miRNA_act)))

# Scatter plot
plot(as.numeric(as.vector(cancer_data$GOI_dfi)), as.numeric(as.vector(cancer_data$miRNA_act)), pch = 16, cex=0.5, xlab="Host gene transcription log(FC)", ylab="miRNA activity")
abline(lm(as.numeric(as.vector(cancer_data$miRNA_act)) ~ as.numeric(as.vector(cancer_data$GOI_dfi))), col = 4, lwd = 2)
text(-4, 0.025, paste("r = ", round(corr$estimate, 3), sep=""))
```


Example scatter plot for miR-590 in LIHC

```{r}
miR <- "miR-590"
gene <- "ENSG00000106682"
cancer="LIHC"

# Filter dfi table for the gene of interest
dfi_gene <- lapply(dfi_data, function(x) unlist(x[gene, ]))
dfi_gene <- unlist(dfi_gene)

# Make a vector indicating the cancer type of each sample
nb_samples_cancers <- unlist(lapply(dfi_data, function(x) ncol(x)))
cancers_vector <- rep(cancers, nb_samples_cancers)

# Merge into table containing all samples of all cancer types and indicating the cancer type
dfi_complete <- as.data.frame(cbind(dfi_gene, cancers_vector))
colnames(dfi_complete) <- c("GOI_dfi", "cancer")

# Filter miRNA activity for the miRNA of interest
miRNA <- as.vector(miR)
miR_cancers <- as.data.frame(cbind(row.names(miR_act), miR_act[,miRNA], as.vector(miR_act$cancers_vector)))
colnames(miR_cancers) <- c("samples", "miR_act", "cancer")

# Make table with the samples, dfi for the host gene and miRNA activity in that cancer type, removing normal samples
cancer_data <- dfi_complete[dfi_complete$cancer== cancer, ]
row.names(cancer_data) <- gsub("-1$", "", row.names(cancer_data))
cancer_data <- cancer_data[grep("\\-01$", row.names(cancer_data)), ]
cancer_data$miRNA_act <- NA
cancer_miR <- miR_cancers[miR_cancers$cancer == cancer, ]
cancer_data$miRNA_act <- cancer_miR[match(row.names(cancer_data), cancer_miR$samples),]$miR_act

# Correlation
corr <- cor.test(as.numeric(as.vector(cancer_data$GOI_dfi)), as.numeric(as.vector(cancer_data$miRNA_act)))

# Scatter plot
plot(as.numeric(as.vector(cancer_data$GOI_dfi)), as.numeric(as.vector(cancer_data$miRNA_act)), pch = 16, cex=0.5, xlab="Host gene transcription log(FC)", ylab="miRNA activity")
abline(lm(as.numeric(as.vector(cancer_data$miRNA_act)) ~ as.numeric(as.vector(cancer_data$GOI_dfi))), col = 4, lwd = 2)
text(1, 0.08, paste("r = ", round(corr$estimate, 3), sep=""))

```

```{r}
sessionInfo()
```
