---
title: 'Association between miRNA transcription and miRNA activity across TCGA tumors'
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
```

# Load libraries

```{r}
library(rtracklayer)
library(GenomicRanges)
library(data.table)
library(purrr)
library(stringr)
library(dplyr)
library(tidymodels)
library(gplots)
library(ggplot2)
```

# Select miRNA host genes that contain miRNA loci within introns

We use the miRNA annotation from miRBase ftp://mirbase.smith.man.ac.uk/pub/mirbase/CURRENT/genomes/

```{r}
# Load the miRBase annotation
mir_annot <- import(here("TCGA", "data", "hsa.gff3"))
mir_annot <- as.data.frame(mir_annot)

# Select for pri-miRNAs
pri_mir_annot <- mir_annot %>% filter(type == "miRNA_primary_transcript")

# Add column with a new unique miRNA name to the pri-miRNA annotaion
pri_mir_annot$pri_mir_id <- make.unique(pri_mir_annot$Name)
head(pri_mir_annot)
```

```{r}
# Load the intron annotation
intron_annot <- import(here("TCGA", "data", "Homo_sapiens.GRCh38.85.chr.Introns.gtf"))
intron_annot <- as.data.frame(intron_annot)

# Add column containing a new unique intron id
intron_annot$intron_id <- make.unique(intron_annot$gene_id)
head(intron_annot)
```

```{r}
# Use GenomicRanges to detect coordinate overlap 
pri_mir_gr <- GRanges(
  seqnames = pri_mir_annot$seqnames,
  ranges = IRanges(start = pri_mir_annot$start, end = pri_mir_annot$end),
  strand = pri_mir_annot$strand,
  pri_mir_id = pri_mir_annot$pri_mir_id
)

intron_gr <- GRanges(
  seqnames = intron_annot$seqnames,
  ranges = IRanges(start = intron_annot$start, end = intron_annot$end),
  strand = intron_annot$strand,
  intron_id = intron_annot$intron_id
)

# Find all pri-miRNAs contained within introns
overlaps <- findOverlaps(pri_mir_gr, intron_gr, type = "within")
mir_gene_pairs <- as.data.frame(overlaps) |>
  mutate(
    pri_mir_id = pri_mir_gr$pri_mir_id[queryHits],
    intron_id = intron_gr$intron_id[subjectHits]
  ) |>
  select(intron_id, pri_mir_id)

# Save 
write.csv(mir_gene_pairs, file=here("TCGA", "output", "supplementary_tables", "GRCh38_miRs_in_introns.csv"))
```

# Load the estimates of changes in mRNA abundance (dfe) and transcription (dfi) across TCGA tumors, from BETAS

```{r}
# Helper function to load and format BETAS data
load_betas_data <- function(files) {
  map_df(files, function(f) {
    dt <- fread(f) |>
      as.data.frame()
    rownames(dt) <- dt[[1]]
    dt[[1]] <- NULL
    dt
  }, .id = "source") |>
    rownames_to_column("gene_id") |>
    pivot_longer(
      cols = -c(source, gene_id),
      names_to = "sample",
      values_to = "value"
    )
}

dfi_files <- Sys.glob(here("TCGA", "data", "TCGA-*.BETAS_dfi.csv"))
dfe_files <- Sys.glob(here("TCGA", "data", "TCGA-*.BETAS_dfe.csv"))

# Extract cancer type from filename
cancers <- dfi_files |>
  basename() |>
  str_extract("TCGA-[A-Z]+") |>
  str_remove("TCGA-")

# Load dfi and dfe data, indicate cancer type 
dfi_data_list <- map2(dfi_files, cancers, function(f, cancer) {
  fread(f) |>
    as.data.frame() |>
    (\(dt) {rownames(dt) <- dt[[1]]; dt[, -1]})() |>
    as.matrix() |>
    as.data.frame() |>
    rownames_to_column("gene_id") |>
    mutate(cancer = cancer)
})

dfe_data_list <- map2(dfe_files, cancers, function(f, cancer) {
  fread(f) |>
    as.data.frame() |>
    (\(dt) {rownames(dt) <- dt[[1]]; dt[, -1]})() |>
    as.matrix() |>
    as.data.frame() |>
    rownames_to_column("gene_id") |>
    mutate(cancer = cancer)
})

# Combine into long tables
# Pivot each data frame before binding 
dfi_long <- map_df(dfi_data_list, ~{
  .x |>
    pivot_longer(-c(gene_id, cancer), names_to = "sample", values_to = "dfi")
})

dfe_long <- map_df(dfe_data_list, ~{
  .x |>
    pivot_longer(-c(gene_id, cancer), names_to = "sample", values_to = "dfe")
})
```

# For each cancer type, for each miRNA contained in introns, test the correlation between the corresponding gene's changes in mRNA abundance (dfe) and transcription (dfi) across tumors.

```{r}
# Compute correlations for each gene-cancer combination
dfe_dfi_results <- bind_cols(
  dfi_long |> select(gene_id, cancer, sample, dfi),
  dfe_long |> select(dfe)
) |>
  
  # Filter to tumor samples (those ending in -01)
  filter(str_detect(sample, "\\-01$")) |>
  
  # Remove the -1 suffix
  mutate(sample = str_remove(sample, "\\-1$")) |>
  
  # Filter to genes in mir_gene_pairs
  semi_join(
    mir_gene_pairs |> select(intron_id) |> distinct() |> rename(gene_id = intron_id),
    by = "gene_id"
  ) |>
  
  # Remove rows where all values in a gene-cancer group are NA
  group_by(gene_id, cancer) |>
  filter(!all(is.na(dfi) | is.na(dfe))) |>
  
  # Compute correlation statistics
  nest(data = -c(gene_id, cancer)) |>
  mutate(
    corr_test = map(data, ~{
      dat <- .x |> filter(!is.na(dfi), !is.na(dfe))
      if (nrow(dat) < 3) return(NA)  # Need at least 3 points for correlation
      
      corr <- cor.test(dat$dfi, dat$dfe)
      fit <- lm(dfe ~ dfi, data = dat)
      
      tibble(
        correlation = corr$estimate,
        p_value = corr$p.value,
        r_squared = summary(fit)$r.squared,
        n = nrow(dat)
      )
    }),
    .keep = "unused"
  ) |>
  unnest(corr_test)

# FDR correction
dfe_dfi_results <- dfe_dfi_results |>
  mutate(
    fdr = p.adjust(p_value, method = "fdr"),
    signed_r2 = sign(correlation) * sqrt(abs(r_squared))
  )

# Save results
saveRDS(dfe_dfi_results, "TCGA_miRhostGenes_dfeVSdfi_results.rds")
```

# Filter miRNA-gene pairs for those where changes in abundance relative to the average (dfe) are due at least in part to changes in transcription (dfi). Specifically, retain pairs where the host gene's dfi vs dfe correlation was significant (after FDR correction) and positive in at least 80% of cancers.

```{r}
genes_sig_pos <- dfe_dfi_results |>
  filter(fdr < 0.05, correlation > 0) |>
  group_by(gene_id) |>
  summarise(sig_pos_freq = n(), .groups = "drop") |>
  # Keep genes with significant positive correlation in ≥80% of cancers
  filter(sig_pos_freq / length(cancers) >= 0.8) |>
  pull(gene_id)
```

# Heatmap of correlation coefficients for dfe vs dfi for all genes that contain a miRNA in their introns

```{r}
# Create matrix for heatmap
signed_r2_matrix <- dfe_dfi_results |>
  select(gene_id, cancer, signed_r2) |>
  pivot_wider(names_from = cancer, values_from = signed_r2, values_fill = NA) |>
  column_to_rownames("gene_id") |>
  as.matrix()

# Remove rows and columns with only NAs
signed_r2_matrix <- signed_r2_matrix[
  rowSums(is.na(signed_r2_matrix)) != ncol(signed_r2_matrix),
  colSums(is.na(signed_r2_matrix)) != nrow(signed_r2_matrix)
]

palette <- colorRampPalette(c("blue", "white", "red"))

pdf("TCGA_miRhostGenes_dfeVSdfi_heatmap.pdf", width = 5.5, height = 6)
heatmap.2(signed_r2_matrix, trace = "none", col = palette, cexRow = 0.01, cexCol = 1.1,
  breaks = seq(-1, 1, length.out = 100), mar = c(5, 1), keysize = 0.2,
  hclust = function(x) hclust(x, method = "average"),
  distfun = function(x) dist(x, method = "euclidean"),
  density.info = "none", key.par = list(cex = 1), lhei = c(1, 3), lwid = c(1, 1.5))
dev.off()
```

# Correlation between host gene transcription changes and miRNA activity

For the genes identified in the previous section

```{r}
mir_gene_pairs <- read.csv("GRCh38_miRs_in_introns_pairs.csv", row.names = 1)
mir_gene_pairs <- mir_gene_pairs[mir_gene_pairs$intron_id %in% genes_sig_pos,]
```

Convert pri-miRNA names to miRNA family names compatible with the activity data

```{r}
mir_gene_pairs[,2] <- gsub("hsa-", "", mir_gene_pairs[,2])
mir_gene_pairs[,2] <- gsub("mir-", "miR-", mir_gene_pairs[,2])
mir_gene_pairs <- as.data.frame(mir_gene_pairs)
```

Load the miRNA families annotation from TargetScan (https://www.targetscan.org/cgi-bin/targetscan/data_download.vert80.cgi)

```{r}
fams <- fread("miR_Family_Info.txt")
fams <- fams[grep("hsa-", fams$`MiRBase ID`), c(1,4,6)] # Filter for human and useful columns
```

Load miRNA activity (output from TCGA_miRNA_activity.Rmd)

```{r}
miR_act <- as.data.frame(read.csv("TCGA_pancancer_miR_act_onlyT.csv", row.names=1))
row.names(miR_act) <- gsub("\\.", "-", row.names(miR_act))
colnames(miR_act) <- gsub("\\.", "-", colnames(miR_act))
```

Format the miRNA family names in the activity data to match the annotation

```{r}
x <- gsub("-5p", "", colnames(miR_act))
x <- gsub("-3p", "", x)
rm_vals <- x[which(duplicated(x))]
rm_cols <- which(x %in% rm_vals)
miR_act <- miR_act[, -rm_cols]
colnames(miR_act) <- x[-rm_cols]
```

Match the miRNA names in the activity data and the pri-miRNA-host-gene pairs identified previously

```{r}
act_mirs <- colnames(miR_act)[-ncol(miR_act)]
act_mirs_prefix <- paste(gsub("-.*", "", act_mirs), "-", sep="")
act_mirs_nbs <- gsub(".*-", "", act_mirs)
act_mirs_nbs <- gsub("[a-z]", "", act_mirs_nbs)
act_mirs <- paste(act_mirs_prefix, act_mirs_nbs, sep="")
act_mirs_ltrs <- gsub(".*-", "", colnames(miR_act)[-ncol(miR_act)])
act_mirs_ltrs <- gsub("[0-9]", "", act_mirs_ltrs)

act_mirs_lst <- list()
i=1
for(m in 1:(length(act_mirs))){
  if(act_mirs_ltrs[m] == ""){
    act_mirs_lst[[i]] <- c(colnames(miR_act)[m], colnames(miR_act)[m])
    i=i+1
  } else {
    for(l in unlist(strsplit(act_mirs_ltrs[m], ""))){
      act_mirs_lst[[i]] <- c(paste(act_mirs[m], l, sep=""), colnames(miR_act)[m])
      i=i+1
    }
  }
}
act_mirs_matches <- do.call("rbind", act_mirs_lst)
mir_gene_pairs$miR_act_family <- act_mirs_matches[match(mir_gene_pairs$pri_mir_id, act_mirs_matches[,1]), 2]
```

For miRNAs for which a family was not found, find it in the TargetScan annotation

```{r}
for (i in 1:nrow(mir_gene_pairs)){
  if(is.na(mir_gene_pairs$miR_act_family[i])){
    mir <- mir_gene_pairs$pri_mir_id[i]
    mir_nb <- gsub(".*-", "", mir)
    ts_fams <- fams[grep(paste("-", mir_nb, "-", sep=""), fams$`MiRBase ID`), ]
    ts_fam_alt <- ts_fams[grep("/", ts_fams$`miR family`), ]
    if (nrow(ts_fams) == 0 | nrow(ts_fam_alt) == 0){
      next
    }
    ts_fam_first_mir <- gsub("/.*", "", ts_fam_alt$`miR family`)
    ts_fam_first_mir <- gsub("-3p", "", ts_fam_first_mir)
    ts_fam_first_mir <- gsub("-5p", "", ts_fam_first_mir)
    if(length(ts_fam_first_mir %in% act_mirs_matches[,1]) > 1){
      if(sum(ts_fam_alt$`Family Conservation?` >= 1) == 0){
        next
      } else if (sum(ts_fam_alt$`Family Conservation?` >= 1) == 1){
        ts_fam_alt <- ts_fam_alt[ts_fam_alt$`Family Conservation?` >= 1, ]
        ts_fam_first_mir <- gsub("/.*", "", ts_fam_alt$`miR family`)
        ts_fam_first_mir <- gsub("-3p", "", ts_fam_first_mir)
        ts_fam_first_mir <- gsub("-5p", "", ts_fam_first_mir)
      } else { # More than one possible miR fam, skipping
        next
      }
    }
    if(ts_fam_first_mir %in% act_mirs_matches[,1]){
      if(ts_fam_alt$`Family Conservation?` >= 1){
        mir_gene_pairs$miR_act_family[i] <- act_mirs_matches[act_mirs_matches[,1] == ts_fam_first_mir, 2]
      }
    }
  }
}                                                               
```

Filter for miRNAs for which we have activity data and remove duplicate gene-miR-family pairs

```{r}
mir_gene_pairs <- mir_gene_pairs[!(is.na(mir_gene_pairs$miR_act_family)), ]
mir_gene_pairs <- as.data.frame(cbind(mir_gene_pairs$intron_id, mir_gene_pairs$miR_act_family))
mir_gene_pairs <- mir_gene_pairs[!(duplicated(mir_gene_pairs)), ]
colnames(mir_gene_pairs) <- c("gene", "miR_act_family")
```

Initialize tables that will be filled with the correlation coefficients, p-values and r-squared 

```{r}
coeffs <- matrix(NA, nrow=nrow(mir_gene_pairs), ncol = length(cancers))
row.names(coeffs) <- paste(mir_gene_pairs$intron_id, mir_gene_pairs$miR_act_family, sep="_")
colnames(coeffs) <- cancers

pvals <- coeffs
r2 <- coeffs
```

For for each pri-miRNA-host-gene pair selected above, for each cancer type, correlate miRNA activity and host gene transcriptional changes across samples

```{r}
for (i in 1:nrow(mir_gene_pairs))
{
  # Filter dfi table for the gene of interest
  dfi_gene <- lapply(dfi_data, function(x) unlist(x[mir_gene_pairs$intron_id[i], ]))
  dfi_gene <- unlist(dfi_gene)

  if(sum(is.na(dfi_gene)) == length(dfi_gene))
  {
    next
  }
  
  # Make a vector indicating the cancer type of each sample
  nb_samples_cancers <- unlist(lapply(dfi_data, function(x) ncol(x)))
  cancers_vector <- rep(cancers, nb_samples_cancers)
  
  # Merge into table containing all samples of all cancer types and indicating the cancer type
  dfi_complete <- as.data.frame(cbind(dfi_gene, cancers_vector))
  colnames(dfi_complete) <- c("GOI_dfi", "cancer")
  
  # Filter miRNA activity for the miRNA of interest
  miRNA <- as.vector(mir_gene_pairs$miR_act_family[i])
  miR_cancers <- as.data.frame(cbind(row.names(miR_act), miR_act[,miRNA], as.vector(miR_act$cancers_vector)))
  colnames(miR_cancers) <- c("samples", "miR_act", "cancer")
  
  # In each cancer type, compute the pearson correlation coefficient and p-value
  for (cancer in cancers)
  {
    # Make table with the samples, dfi for the host gene and miRNA activity in that cancer type, removing normal samples
    cancer_data <- dfi_complete[dfi_complete$cancer== cancer, ]
    row.names(cancer_data) <- gsub("-1$", "", row.names(cancer_data))
    cancer_data <- cancer_data[grep("\\-01$", row.names(cancer_data)), ]
    cancer_data$miRNA_act <- NA
    cancer_miR <- miR_cancers[miR_cancers$cancer == cancer, ]
    if (nrow(cancer_miR) == 0)
    {
      next
    }
    cancer_data$miRNA_act <- cancer_miR[match(row.names(cancer_data), cancer_miR$samples),]$miR_act
    
    # Correlation
    corr <- cor.test(as.numeric(as.vector(cancer_data$GOI_dfi)), as.numeric(as.vector(cancer_data$miRNA_act)))
    coeffs[i, cancer] <- corr$estimate
    pvals[i, cancer] <- corr$p.value
    fit <- lm(as.numeric(as.vector(cancer_data$miRNA_act)) ~ as.numeric(as.vector(cancer_data$GOI_dfi)))
    r2[i, cancer] <- summary(fit)$r.squared
    
    
  }
}
```

Remove rows and columns with only NAs
```{r}
rows_keep <- which(rowSums(is.na(coeffs)) != ncol(coeffs))
cols_keep <- which(colSums(is.na(coeffs)) != nrow(coeffs))
coeffs <- coeffs[rows_keep, cols_keep]
pvals <- pvals[rows_keep, cols_keep]
r2 <- r2[rows_keep, cols_keep]
```

Compute signed r-squared

```{r}
signs <- sign(coeffs)
signed_r2 <- abs(r2) * signs
```

In preparation to make the violin plots, make a table with all host-gene-miRNA pairs and all cancer types, indicating the correlation coefficients 

```{r}
signed_r2 <- data.table(signed_r2, keep.rownames = T)
miRs_table <- melt(signed_r2, id.vars=1)
colnames(miRs_table) <- c("hostGene_miR_pairs", "cancer", "signed_r2")
```

Sort by average signed_r2 in each cancer, keeping the cancers together

```{r}
means <- aggregate(as.vector(miRs_table$signed_r2), list(miRs_table$cancer), mean)
order_cancers <- means[,1][order(means[,2], decreasing = T)]
miRs_table$cancer <- reorder.factor(miRs_table$cancer, new.order=order_cancers)
miRs_table <- miRs_table %>% arrange(cancer)
miRs_table <- as.data.frame(miRs_table %>% 
        group_by(cancer) %>%
        arrange(desc(signed_r2), .by_group = TRUE))
```

Add the correlation coefficients to the table 

```{r}
miRs_table_coeff <- miRs_table
miRs_table_coeff$corr <- sign(miRs_table_coeff$signed_r2) * sqrt(abs(miRs_table_coeff$signed_r2))
miRs_table_coeff <- miRs_table_coeff[,-3]
saveRDS(miRs_table, "TCGA_pancancer_miRgeneDfiVSmiRact_fltrGenesSigPosCorrDfeDfi_res.rds")
```

Draw the violin plots of correlation coefficients (can be modified to draw signed r2 as well)

```{r}
ggplot(miRs_table_coeff, aes(x = cancer, y = corr)) + 
  geom_violin(fill = "grey80", colour = "#3366FF") +
  geom_boxplot(width = 0.07) +
  theme_classic() +
  ylab("Pearson correlation") + 
  xlab("Cancer Type") + 
  theme(axis.text=element_text(size=16),
        axis.title=element_text(size=20),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  theme(legend.position = "none") +
  geom_hline(yintercept=0, linetype="dashed", 
              color = "black", size=0.5)
```

Draw the corresponding heatmap (can also be modified to show signed r2)

```{r}
palette <- colorRampPalette(c("blue", "white", "red"))
dat_htmp <- data.matrix(signed_r2)[, as.vector(unique(miRs_table$cancer))]
dat_htmp_coeffs <- sign(dat_htmp) * sqrt(abs(dat_htmp))
pdf("TCGA_pancancer_miRgeneDfiVSmiRact_fltrGenesSigPosCorrDfeDfi_pearson_htmp.pdf", width=18, height=7)
heatmap.2(dat_htmp_coeffs, trace="none", col=palette, cexRow=0.01, cexCol=2.5,  breaks=seq(-1,1,length.out=50), mar=c(10,20), keysize=1, hclust=function(x) hclust(x,method="complete"), distfun = function(x) dist(x,method = 'euclidean'), density.info = "none", key.par = list(cex=2), lhei=c(1,2), lwid = c(1,5), Colv = F, key.title = "", key.xlab = "Pearson correlation")
dev.off()
```

![Heatmap of associations between miRNA activity and transcription](TCGA_pancancer_miRgeneDfiVSmiRact_fltrGenesSigPosCorrDfeDfi_pearson_htmp.pdf)

Average correlation between miRNA activity and rate of transcription of their primary transcripts

```{r}
mean(miRs_table_coeff$corr)
```

Example scatter plot for miR-193 in STAD

```{r}
miR <- "miR-139"
gene <- "ENSG00000186642"
cancer="STAD"

# Filter dfi table for the gene of interest
dfi_gene <- lapply(dfi_data, function(x) unlist(x[gene, ]))
dfi_gene <- unlist(dfi_gene)

# Make a vector indicating the cancer type of each sample
nb_samples_cancers <- unlist(lapply(dfi_data, function(x) ncol(x)))
cancers_vector <- rep(cancers, nb_samples_cancers)

# Merge into table containing all samples of all cancer types and indicating the cancer type
dfi_complete <- as.data.frame(cbind(dfi_gene, cancers_vector))
colnames(dfi_complete) <- c("GOI_dfi", "cancer")

# Filter miRNA activity for the miRNA of interest
miRNA <- as.vector(miR)
miR_cancers <- as.data.frame(cbind(row.names(miR_act), miR_act[,miRNA], as.vector(miR_act$cancers_vector)))
colnames(miR_cancers) <- c("samples", "miR_act", "cancer")

# Make table with the samples, dfi for the host gene and miRNA activity in that cancer type, removing normal samples
cancer_data <- dfi_complete[dfi_complete$cancer== cancer, ]
row.names(cancer_data) <- gsub("-1$", "", row.names(cancer_data))
cancer_data <- cancer_data[grep("\\-01$", row.names(cancer_data)), ]
cancer_data$miRNA_act <- NA
cancer_miR <- miR_cancers[miR_cancers$cancer == cancer, ]
cancer_data$miRNA_act <- cancer_miR[match(row.names(cancer_data), cancer_miR$samples),]$miR_act

# Correlation
corr <- cor.test(as.numeric(as.vector(cancer_data$GOI_dfi)), as.numeric(as.vector(cancer_data$miRNA_act)))

# Scatter plot
plot(as.numeric(as.vector(cancer_data$GOI_dfi)), as.numeric(as.vector(cancer_data$miRNA_act)), pch = 16, cex=0.5, xlab="Host gene transcription log(FC)", ylab="miRNA activity")
abline(lm(as.numeric(as.vector(cancer_data$miRNA_act)) ~ as.numeric(as.vector(cancer_data$GOI_dfi))), col = 4, lwd = 2)
text(-4, 0.025, paste("r = ", round(corr$estimate, 3), sep=""))
```


Example scatter plot for miR-590 in LIHC

```{r}
miR <- "miR-590"
gene <- "ENSG00000106682"
cancer="LIHC"

# Filter dfi table for the gene of interest
dfi_gene <- lapply(dfi_data, function(x) unlist(x[gene, ]))
dfi_gene <- unlist(dfi_gene)

# Make a vector indicating the cancer type of each sample
nb_samples_cancers <- unlist(lapply(dfi_data, function(x) ncol(x)))
cancers_vector <- rep(cancers, nb_samples_cancers)

# Merge into table containing all samples of all cancer types and indicating the cancer type
dfi_complete <- as.data.frame(cbind(dfi_gene, cancers_vector))
colnames(dfi_complete) <- c("GOI_dfi", "cancer")

# Filter miRNA activity for the miRNA of interest
miRNA <- as.vector(miR)
miR_cancers <- as.data.frame(cbind(row.names(miR_act), miR_act[,miRNA], as.vector(miR_act$cancers_vector)))
colnames(miR_cancers) <- c("samples", "miR_act", "cancer")

# Make table with the samples, dfi for the host gene and miRNA activity in that cancer type, removing normal samples
cancer_data <- dfi_complete[dfi_complete$cancer== cancer, ]
row.names(cancer_data) <- gsub("-1$", "", row.names(cancer_data))
cancer_data <- cancer_data[grep("\\-01$", row.names(cancer_data)), ]
cancer_data$miRNA_act <- NA
cancer_miR <- miR_cancers[miR_cancers$cancer == cancer, ]
cancer_data$miRNA_act <- cancer_miR[match(row.names(cancer_data), cancer_miR$samples),]$miR_act

# Correlation
corr <- cor.test(as.numeric(as.vector(cancer_data$GOI_dfi)), as.numeric(as.vector(cancer_data$miRNA_act)))

# Scatter plot
plot(as.numeric(as.vector(cancer_data$GOI_dfi)), as.numeric(as.vector(cancer_data$miRNA_act)), pch = 16, cex=0.5, xlab="Host gene transcription log(FC)", ylab="miRNA activity")
abline(lm(as.numeric(as.vector(cancer_data$miRNA_act)) ~ as.numeric(as.vector(cancer_data$GOI_dfi))), col = 4, lwd = 2)
text(1, 0.08, paste("r = ", round(corr$estimate, 3), sep=""))

```

