---
title: 'Association between miRNA transcription and miRNA abundance across TCGA tumors'
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
```

# Load libraries

```{r}
library(dplyr)
library(tidyr)
library(gplots)
library(ggplot2)
library(beeswarm)
library(tibble)
```

# Load the list of miRNA host genes and corresponding pri-miRNA and miRNA family 

```{r}
mir_gene_pairs <- readRDS(here("TCGA", "output", "rds", "TCGA_selected_miRNA_host_genes.rds"))

# Remove all -1/-2/... from the end of the pri-miR names
mir_gene_pairs$pri_mir_id_3 <- gsub(
  "(-[0-9]+[A-Za-z]*)-[0-9]+$",
  "\\1",
  mir_gene_pairs$pri_mir_id
)

# Filter for unique host gene - pri-miRNA pairs (we don't deal with family here)
mir_gene_pairs_fltr <- unique(mir_gene_pairs[, c("gene_id", "pri_mir_id_3")]) 
```

# Since each pri-miRNA will be processed into two mature miRNAs (arms), duplicate each entry and add -5p or -3p to the names

```{r}
mir_gene_pairs_fltr$mir_mature <- paste(mir_gene_pairs_fltr$pri_mir_id_3, "-3p", sep="")
mir_gene_pairs_fltr_5p <- mir_gene_pairs_fltr
mir_gene_pairs_fltr_5p$mir_mature <- gsub("-3p", "-5p", mir_gene_pairs_fltr_5p$mir_mature)
mir_gene_pairs_fltr <- rbind(mir_gene_pairs_fltr, mir_gene_pairs_fltr_5p)
mir_gene_pairs_fltr <- mir_gene_pairs_fltr[order(mir_gene_pairs_fltr$gene_id), ]
dim(mir_gene_pairs_fltr)
```

# Load the changes in transcription

```{r}
dfi_data <- readRDS(here("TCGA", "output", "rds", "TCGA_dfi_long.rds"))
```

# For each mature miRNA, correlate abundance and transcription of host gene across tumor samples

```{r}
# Load miRNA expression files
miR_expr_data <- sapply(
  unique(dfi_data$cancer),
  function(cancer) {
    miR_expr <- read.table(
      here("data", paste0(cancer, ".miRseq_mature_RPM_log2.txt")),
      header = TRUE
    )
    
    miR_expr$miRseqMature <- gsub("\\|.*|,.*", "", miR_expr$miRseqMature)
    
    miR_expr
  },
  simplify = FALSE
)
```

```{r}
# Helper function to process a single gene-miRNA pair
corr_single_pair <- function(i, mir_gene_pairs_fltr, dfi_data, miR_expr_data) {
  gene_id <- mir_gene_pairs_fltr$gene_id[i]
  mir_mature <- mir_gene_pairs_fltr$mir_mature[i]
  
  # Filter dfi table for the gene of interest
  dfi_gene <- dfi_data[dfi_data$gene_id == gene_id, ]
  
  if (nrow(dfi_gene) == 0) {
    return(NULL)
  }
  
  # Apply correlations across cancers 
  cancer_results <- lapply(unique(dfi_gene$cancer), function(cancer) {
    miR_expr <- miR_expr_data[[cancer]]
    miR_expr <- column_to_rownames(miR_expr, "miRseqMature")
    
    # Check if miRNA exists in the abundance data
    if (!(mir_mature %in% row.names(miR_expr))) {
      return(NULL)
    }
    
    # Extract and format miRNA abundance, select tumor samples
    miR_cancer <- data.frame(
      sample = colnames(miR_expr),
      miR_expr = as.numeric(miR_expr[mir_mature, ]),
      stringsAsFactors = FALSE
    )
    miR_cancer$sample <- gsub("\\.", "-", miR_cancer$sample)
    miR_cancer <- miR_cancer[grep("-01A", miR_cancer$sample), ]
    miR_cancer$sample <- gsub("-01A.*", "-01", miR_cancer$sample)
    
    # Skip if too many missing values (>20%)
    if (sum(is.na(miR_cancer$miR_expr)) / nrow(miR_cancer) > 0.2) {
      return(NULL)
    }
    
    # Combine with host gene transcription data
    cancer_data <- dfi_gene[dfi_gene$cancer == cancer, ]
    cancer_data <- cancer_data[grep("\\-01$", cancer_data$sample), ]
    cancer_data$miRNA_expr <- as.numeric(
      miR_cancer$miR_expr[match(cancer_data$sample, miR_cancer$sample)]
    )
    
    # Compute correlation
    complete <- complete.cases(cancer_data$dfi, cancer_data$miRNA_expr)

    corr <- cor.test(
      as.numeric(cancer_data$dfi[complete]),
      cancer_data$miRNA_expr[complete]
    )
    
    # Return result as a data frame row
    data.frame(
      miRNA = mir_mature,
      gene_id = gene_id, 
      cancer = cancer,
      coefficient = corr$estimate,
      p_value = corr$p.value,
      r_squared = (corr$estimate)^2,
      stringsAsFactors = FALSE
    )
  })
  
  # Combine results for this pair
  do.call(rbind, Filter(Negate(is.null), cancer_results))
}

# Apply across all gene-miRNA pairs and combine into a single long df
results_list <- lapply(
  1:nrow(mir_gene_pairs_fltr),
  corr_single_pair,
  mir_gene_pairs_fltr = mir_gene_pairs_fltr,
  dfi_data = dfi_data,
  miR_expr_data = miR_expr_data
)

corr_results_long <- do.call(rbind, Filter(Negate(is.null), results_list))
row.names(corr_results_long) <- NULL
```

```{r}
# Add additional metrics 
corr_results_long <- corr_results_long |>
  mutate(
    signed_r2 = sign(coefficient) * r_squared,
    host_miR = paste(gene_id, miRNA, sep="_"),
    .after = r_squared
  )

# FDR correction across all tests
pvalue_fdr <- p.adjust(corr_results_long$p_value, method = "fdr")

corr_results_long <- corr_results_long |>
  mutate(FDR = pvalue_fdr, .after = p_value)
```

# Filter for miRNA-cancer pairs where the miRNA is deferentially expressed 
Compute differential expression (tumor vs paired normal) for each miRNA in each cancer type using paired t-tests.

```{r}
# Helper function to test differential expression for a single cancer
get_DE_cancer <- function(cancer, miR_expr_data, miR_all) {
  miR_expr <- column_to_rownames(miR_expr_data[[cancer]], "miRseqMature")
  
  # Identify paired tumor and normal samples
  patients_normal <- substr(
    colnames(miR_expr)[grep("\\.11A$", substr(colnames(miR_expr), 1, 16))],
    1, 12
  )
  
  if (length(patients_normal) == 0) return(NULL)
  
  patients_tumor <- substr(
    colnames(miR_expr)[grep("\\.01A$", substr(colnames(miR_expr), 1, 16))],
    1, 12
  )
  
  patients_matched <- intersect(patients_normal, patients_tumor)
  if (length(patients_matched) == 0) return(NULL)
    
  # Filter to matched patients
  miR_expr_matched <- miR_expr[
    , substr(colnames(miR_expr), 1, 12) %in% patients_matched
  ]

  sample_info <- data.frame(
    sample = colnames(miR_expr_matched),
    patient = substr(colnames(miR_expr_matched), 1, 12),
    sample_type = substr(colnames(miR_expr_matched), 14, 16),
    stringsAsFactors = FALSE
  )
  
  # Identify tumor and normal samples for the same patients
  tumor_info <- sample_info |>
    filter(sample_type == "01A") |>
    arrange(patient)
  
  normal_info <- sample_info |>
    filter(sample_type == "11A") |>
    arrange(patient)
  
  # Keep only patients with exactly one tumor and one normal sample
  paired_info <- inner_join(
    tumor_info |> rename(tumor_sample = sample),
    normal_info |> rename(normal_sample = sample),
    by = "patient"
  )
  
  # Extract expression matrices in matching patient order
  expr_tumor <- miR_expr_matched[
    , paired_info$tumor_sample,
    drop = FALSE
  ]
  
  expr_normal <- miR_expr_matched[
    , paired_info$normal_sample,
    drop = FALSE
  ]
  
  # Apply paired t-tests using sapply for each miRNA
  de_results <- sapply(miR_all, function(mir) {
    if (!(mir %in% rownames(expr_tumor))) {
      return(c(log2FC = NA, p_value = NA))
    }
    
    tumor_vals <- as.numeric(expr_tumor[mir, ])
    normal_vals <- as.numeric(expr_normal[mir, ])
    
    # Skip if too many NAs (>20%)
    if (mean(is.na(c(tumor_vals, normal_vals))) > 0.2) {
      return(c(log2FC = NA, p_value = NA))
    }
    
    t_test <- t.test(tumor_vals, normal_vals, paired = TRUE, na.action = na.omit)
    c(log2FC = unname(t_test$estimate), p_value = t_test$p.value)
  }, simplify = TRUE)
  
  # Convert to long format
  DE_long <- data.frame(
    miRNA = colnames(de_results),
    log2FC = as.numeric(de_results["log2FC", ]),
    p_value = as.numeric(de_results["p_value", ]),
    cancer = cancer,
    stringsAsFactors = FALSE
  )
  
  DE_long
}

# Get unique miRNAs from correlation results
miR_all <- unique(corr_results_long$miRNA)

# Compute DE for these miRNAs for all cancers and combine into long format
DE_res_long <- lapply(
  unique(corr_results_long$cancer),
  get_DE_cancer,
  miR_expr_data = miR_expr_data,
  miR_all = miR_all
) |>
  Filter(Negate(is.null), x = _) |>
  do.call(rbind, args = _)

row.names(DE_res_long) <- NULL

# Add FDR correction within each cancer type
DE_res_long <- DE_res_long |>
  group_by(cancer) |>
  mutate(FDR = p.adjust(p_value, method = "fdr")) |>
  ungroup() |>
  as.data.frame()

# Filter for miRNA-cancer pairs with significant differential miRNA expression 
DE_sig <- DE_res_long |>
  filter(abs(log2FC) > 1, FDR < 0.05) |>
  select(miRNA, cancer)
```

Create table with correlation results added on to these significantly DE miRNAs

```{r}
miRs_table <- corr_results_long |>
  inner_join(DE_sig, by = c("miRNA", "cancer")) |>
  select(miRNA, host_miR, cancer, coefficient, signed_r2) 

medians <- aggregate(miRs_table$signed_r2, list(miRs_table$cancer), median)
medians <- medians[order(medians[,2], decreasing = T), ]
miRs_table$cancer <- factor(miRs_table$cancer, levels = medians[,1])
miRs_table <- miRs_table %>% arrange(cancer)

saveRDS(miRs_table, here("TCGA", "output", "rds", "TCGA_miRhostGenes_transc_vs_abundance_fltrSigDEmiRs.rds"))
```

# Average correlation between miRNA abundance and rate of transcription of their primary transcripts, for miRNAs that are differentially expressed in at least one cancer type

```{r}
mean(miRs_table$coefficient)
```

# Average signed r2 

```{r}
mean(miRs_table$signed_r2)
```

# Violin plot of signed r-squared by cancer type for these miRNAs

```{r}
p <- ggplot(miRs_table, aes(x = cancer, y = signed_r2)) + 
  geom_violin(fill = "grey80", colour = "#3366FF", width = 1.2) +
  geom_boxplot(width = 0.07) +
  theme_classic() +
  ylab(expression("Signed " * r^2)) + 
  xlab("") + 
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=14),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "none") +
  geom_hline(yintercept=0, linetype="dashed", 
            color = "black", size=0.5)
p

pdf(here("TCGA", "output", "figures", "TCGA_miRNA_abundance_vs_transcription_violin.pdf"), 7, 3.5)
p
dev.off()
```

# Corresponding heatmap 

```{r}
# Filter full data for all significant miRNAs and cancer types retained above
corr_results_fltr <- corr_results_long[corr_results_long$miRNA %in% miRs_table$miRNA & corr_results_long$cancer %in% miRs_table$cancer, ]
corr_results_fltr$cancer <- factor(corr_results_fltr$cancer, levels = medians[,1])
corr_results_fltr <- corr_results_fltr %>% arrange(cancer)

# Convert to wide format for clustering
dat_htmp <- corr_results_fltr |>
  select(host_miR, cancer, signed_r2) |>
  pivot_wider(
    names_from = cancer,
    values_from = signed_r2
  )

dat_htmp <- column_to_rownames(dat_htmp, "host_miR")
dat_htmp <- data.matrix(dat_htmp)

# Replace NAs with 0s
dat_htmp[is.na(dat_htmp)] <- 0

# Draw the heatmap
palette <- colorRampPalette(c("blue", "white", "red"))
pdf(here("TCGA", "output", "figures", "TCGA_miRNA_abundance_vs_transcription_htmp.pdf"),
    width = 18, height = 7)
heatmap.2(
  dat_htmp,
  trace = "none",
  col = palette,
  cexRow = 0.01,
  cexCol = 2,
  breaks = seq(-1, 1, length.out = 51),
  mar = c(10, 10),
  keysize = 1,
  hclust = function(x) hclust(x, method = "complete"), 
  distfun = function(x) dist(x, method = "euclidean"),
  density.info = "none",
  Colv = FALSE,
  key.par = list(cex = 2),
  lhei = c(1, 1.2),
  lwid = c(1, 2),
  key.title = "",
  key.xlab = expression("Signed " * r^2)
)

dev.off()
```

# Comparison of three associations: A) miRNA abundance vs transcription, B) activity vs abundance and C) activity vs transcription

```{r}
expr_transc <- miRs_table
act_expr <- readRDS(here("TCGA", "output", "rds", "TCGA_miRhostGenes_abundance_vs_activity.rds"))
act_transc <- readRDS(here("TCGA", "output", "rds", "TCGA_miRhostGenes_transc_vs_activity.rds"))
```

Data for the beeswarm plot

```{r}
dt <- list(expr_transc$coefficient, act_expr$correlation, act_transc$r)
names(dt) <- c("miRNA abundance\nvs transcription", "miRNA activity\nvs abundance", "miRNA activity\nvs transcription")
```

Draw the beeswarm plot

```{r}
pdf(here("TCGA", "output", "figures", "TCGA_miRNA_abundance_vs_transcription_vs_activity_beeswarm.pdf"),
    width = 4.5, height = 3.5)
beeswarm(dt,
                col = c("olivedrab3", "slateblue3", "dodgerblue"), pch = c(16,16,16),
                method = "swarm",
                cex=0.2,
                ylab="Pearson r",
                xlab = "",
                cex.axis=0.7
)
boxplot(dt, add = T, outline=F, col=rgb(0,0,0, alpha=0.001), boxlwd = 1, boxwex=0.5, whisklty = 1, staplelwd = 1, medlwd=1, whisklwd=1, names=c("","",""), yaxt="n") # See https://stackoverflow.com/questions/28889977/r-boxplot-how-to-customize-the-appearance-of-the-box-and-whisker-plots-e-g-r
dev.off()
```

```{r}
sessionInfo()
```