---
title: 'Association between miRNA transcription and miRNA abundance across TCGA tumors'
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
```

# Load libraries

```{r}
library(dplyr)
library(gplots)
library(ggplot2)
library(vioplot)
library(data.table)
library(beeswarm)
```

# Load the list of miRNA host genes and corresponding pri-miRNA and miRNA family 

```{r}
mir_gene_pairs <- readRDS(here("TCGA", "output", "rds", "TCGA_selected_miRNA_host_genes.rds"))
mir_gene_pairs_fltr <- unique(mir_gene_pairs[, c("gene_id", "pri_mir_id", "pri_mir_id_2")]) 
```

# Since each pri-miRNA will be processed into two mature miRNAs (arms), duplicate each entry and add -5p or -3p to the names

```{r}
mir_gene_pairs_fltr$miR <- paste(mir_gene_pairs_fltr$pri_mir_id, "-3p", sep="")
mir_gene_pairs_fltr_5p <- mir_gene_pairs_fltr
mir_gene_pairs_fltr_5p$miR <- gsub("-3p", "-5p", mir_gene_pairs_fltr_5p$miR)
mir_gene_pairs_fltr <- rbind(mir_gene_pairs_fltr, mir_gene_pairs_fltr_5p)
mir_gene_pairs_fltr <- mir_gene_pairs_fltr[order(mir_gene_pairs_fltr$gene), ]
dim(mir_gene_pairs_fltr)
```

# Load the changes in transcription

```{r}
dfi_data <- readRDS(here("TCGA", "output", "rds", "TCGA_dfi_long.rds"))
```

# For each mature miRNA, correlate abundance and transcription of host gene across tumor samples

```{r}
# Load miRNA expression files
miR_expr_data <- sapply(
  unique(dfi_data$cancer),
  function(cancer) {
    read.table(
      here("TCGA", "data", paste0(cancer, ".miRseq_mature_RPM_log2.txt")),
      header = TRUE
    )
  },
  simplify = FALSE
)
```

```{r}
# Helper function to process a single gene-miRNA pair
corr_single_pair <- function(i, mir_gene_pairs_fltr, dfi_data, miR_expr_data) {
  gene_id <- mir_gene_pairs_fltr$gene_id[i]
  miRNA <- mir_gene_pairs_fltr$miR[i]
  
  # Filter dfi table for the gene of interest
  dfi_gene <- dfi_data[dfi_data$gene_id == gene_id, ]
  
  if (nrow(dfi_gene) == 0) {
    return(NULL)
  }
  
  # Apply correlation computation across cancers 
  cancer_results <- lapply(unique(dfi_gene$cancer), function(cancer) {
    miR_expr <- miR_expr_data[[cancer]]
    
    # Check if miRNA exists in the abundance data
    if (!(miRNA %in% row.names(miR_expr))) {
      return(NULL)
    }
    
    # Extract and format miRNA abundance
    miR_cancer <- as.data.frame(cbind(colnames(miR_expr), unlist(miR_expr[miRNA, ])))
    colnames(miR_cancer) <- c("sample", "miR_expr")
    miR_cancer$sample <- gsub("\\.", "-", miR_cancer$sample)
    miR_cancer <- miR_cancer[grep("-01$", miR_cancer$sample), ]
    
    # Skip if too many missing values (>20%)
    if (sum(is.na(miR_cancer$miR_expr)) / nrow(miR_cancer) > 0.2) {
      return(NULL)
    }
    
    # Combine with host gene transcription data
    cancer_data <- dfi_gene[dfi_gene$cancer == cancer, ]
    cancer_data <- cancer_data[grep("\\-01$", cancer_data$sample), ]
    cancer_data$miRNA_expr <- as.numeric(
      miR_cancer[match(row.names(cancer_data), miR_cancer$sample), ]$miR_expr
    )
    cancer_data <- cancer_data[rowSums(is.na(cancer_data)) == 0, ]
    
    if (nrow(cancer_data) == 0) {
      return(NULL)
    }
    
    # Compute correlation
    corr <- cor.test(
      as.numeric(cancer_data$dfi),
      cancer_data$miRNA_expr
    )
    
    # Return result as a data frame row
    data.frame(
      miRNA = miRNA,
      cancer = cancer,
      coefficient = corr$estimate,
      p_value = corr$p.value,
      r_squared = (corr$estimate)^2,
      stringsAsFactors = FALSE
    )
  })
  
  # Combine results for this pair
  do.call(rbind, Filter(Negate(is.null), cancer_results))
}

# Apply across all gene-miRNA pairs and combine into a single long df
results_list <- lapply(
  1:nrow(mir_gene_pairs_fltr),
  corr_single_pair,
  mir_gene_pairs_fltr = mir_gene_pairs_fltr,
  dfi_data = dfi_data,
  miR_expr_data = miR_expr_data
)

corr_results_long <- do.call(rbind, Filter(Negate(is.null), results_list))
row.names(corr_results_long) <- NULL
```

```{r}
# Remove rows with all NAs and miRNAs with >20% missing data
corr_results_long <- corr_results_long |>
  filter(!is.na(coefficient)) |>
  group_by(miRNA) |>
  filter(sum(!is.na(coefficient)) / n() > 0.8) |>
  ungroup() |>
  group_by(cancer) |>
  filter(sum(!is.na(coefficient)) > 0) |>
  ungroup() |>
  distinct(miRNA, cancer, .keep_all = TRUE)

# Add additional metrics 
corr_results_long <- corr_results_long |>
  mutate(
    signed_r2 = sign(coefficient) * r_squared,
    neg_log10_pvalue = abs(log10(p_value)),
    .after = r_squared
  )

# FDR correction across all tests
pvalue_fdr <- p.adjust(corr_results_long$p_value, method = "fdr")

corr_results_long <- corr_results_long |>
  mutate(FDR = pvalue_fdr, .after = p_value)
```

# Filter for miRNAs that are differentially expressed in each cancer type

First initialize a list to store the differentially expressed miRNAs.

```{r}
<==
miRs <- row.names(coeffs)
miRs_cancers <- as.list(rep(NA, ncol(coeffs)))
names(miRs_cancers) <- colnames(coeffs)
```

Fill in the list. See comments within code for specific steps.

```{r}
for (cancer in colnames(coeffs))
{
  # Load miRNA expression data for this cancer, retrived from Firebrowse 
  miR_expr <- read.table(paste(cancer, ".miRseq_mature_RPM_log2.txt", sep=""), header=T, row.names = 1)
  
  # Select for paired samples
  patients_normal <- colnames(miR_expr)[grep("\\.11A$", substr(colnames(miR_expr),1,16))]
  patients_normal <- substr(patients_normal, 1, 12)
  
  if (length(patients_normal) == 0)
  {
    next
  }
  
  patients_tumor <- colnames(miR_expr)[grep("\\.01A$", substr(colnames(miR_expr),1,16))]
  patients_tumor <- substr(patients_tumor, 1, 12)
  
  patients_matched <- patients_normal[patients_normal %in% patients_tumor]
       
  miR_expr_matched <- miR_expr[, substr(colnames(miR_expr), 1, 12) %in% patients_matched]
  row.names(miR_expr_matched) <- gsub("\\|.*", "", row.names(miR_expr_matched))
  colnames(miR_expr_matched) <- substr(colnames(miR_expr_matched), 1, 16)
  
  # Separate tumor from normal samples, put the columns
  normal <- colnames(miR_expr_matched)[grep("\\.11A$", colnames(miR_expr_matched))]
  tumor <- colnames(miR_expr_matched)[grep("\\.01A$", colnames(miR_expr_matched))]
  
  miRNA_expression_T <- miR_expr_matched[, sort(tumor)]
  miRNA_expression_N <- miR_expr_matched[, sort(normal)]
  
  # Create a table to store the differential expression data for each miRNA (tumor vs paired normal samples)
  result_table <- data.frame(matrix(ncol = 2, nrow = length(miRs)))
  row.names(result_table) <- miRs
  colnames(result_table) <- c("log2FC", "t-test_pval")
  
  # Compute the differential expression of the miRNAs using a paired t-test for matched samples
  for (miR in miRs)
  {
    specific_miR_expr_T <- unlist(miRNA_expression_T[miR,])
    specific_miR_expr_N <- unlist(miRNA_expression_N[miR,])
    t_test <- t.test(as.numeric(as.vector(specific_miR_expr_T)), as.numeric(as.vector(specific_miR_expr_N)), paired=TRUE)
    
    result_table[miR,1] <- t_test$estimate # log2FC
    result_table[miR,2] <- t_test$p.value 
  }  
  
  # FDR correction
  result_table$FDR <- p.adjust(result_table$`t-test_pval`, method="fdr")
  
  # Filter for miRNAs that are significantly differentially expressed
  sig_mirs <- result_table[abs(result_table)$log2FC > 1 & result_table$FDR < 0.05, ]
  
  # Record in list
  miRs_cancers[[cancer]] <- row.names(sig_mirs)
}
```

Remove cancers types with NAs, meaning they don't have normal samples in the miRNA expression data. 

```{r}
miRs_cancers <- miRs_cancers[!(is.na(miRs_cancers))]
```

In preparation to make the violin plot, make a table containing all miRNAs and all cancer types, with columns to contain the signed r-squared (miRNA transcription vs abundance), p-valye, and cancer type.

```{r}
miRs_table <- as.data.frame(matrix(NA, nrow=sum(unlist(lapply(miRs_cancers, length))), ncol=5))
colnames(miRs_table) <- c("miRs", "signed_r2", "-log10(p-value)", "FDR", "cancer")
miRs_table$miRs <- unlist(miRs_cancers)
miRs_table$cancer <- rep(names(miRs_cancers), unlist(lapply(miRs_cancers, length)))
```

Fill the table

```{r}
for (i in 1:nrow(miRs_table))
{
  miRs_table[i, ]$signed_r2 <- signed_r2[miRs_table$miRs[i], miRs_table$cancer[i]]
  miRs_table[i, ]$`-log10(p-value)` <- minusLogPvals[miRs_table$miRs[i], miRs_table$cancer[i]]
  miRs_table[i, ]$FDR <- FDR[miRs_table$miRs[i], miRs_table$cancer[i]]
}
```

Remove rows with NAs, corresponding to miRNAs that did not have (sufficient) expression data

```{r}
miRs_table <- miRs_table[rowSums(is.na(miRs_table)) == 0, ]
```

Sort by average signed_r2 in each cancer type

```{r}
medians <- aggregate(miRs_table$signed_r2, list(miRs_table$cancer), median)
order_cancers <- medians[,1][order(medians[,2], decreasing = T)]
miRs_table$cancer <- reorder.factor(miRs_table$cancer, new.order=order_cancers)
miRs_table <- miRs_table %>% arrange(cancer)
miRs_table <- as.data.frame(miRs_table %>% 
        group_by(cancer) %>%
        arrange(desc(signed_r2), .by_group = TRUE))
```

Add correlation coefficients for later use and save

```{r}
miRs_table$corr <- sign(miRs_table$signed_r2) * sqrt(abs(miRs_table$signed_r2))
saveRDS(miRs_table, "TCGA_pancancer_miRgeneDfiVSmiRexpr_fltrGenesSigPosCorrDfeDfi_fltrSigDEmiRs_res.rds")
```

Draw the violin plot

```{r}
ggplot(miRs_table, aes(x = cancer, y = signed_r2)) + 
  geom_violin(fill = "grey80", colour = "#3366FF", width=1.2) +
  geom_boxplot(width = 0.07) +
  theme_classic() +
  ylab("Signed r-squared") + 
  xlab("Cancer Type") + 
  theme(axis.text=element_text(size=16),
        axis.title=element_text(size=20)) +
  theme(legend.position = "none") +
  geom_hline(yintercept=0, linetype="dashed", 
              color = "black", size=0.5)
```

Get signed r2 data after clustering and filtering for miRNAs that have significant differential expression in at least one cancer type

```{r}
signed_r2_fltr <- signed_r2[row.names(signed_r2) %in% unlist(miRs_cancers), ]
htmp <- heatmap.2(data.matrix(signed_r2_fltr), hclust=function(x) hclust(x,method="complete"), distfun = function(x) dist(x,method = 'euclidean'))
signed_r2_clust <- signed_r2_fltr[rev(htmp$rowInd), htmp$colInd]
minusLogPvals_clust <- minusLogPvals[row.names(signed_r2_clust), colnames(signed_r2_clust)]
FDR_clust <- FDR[row.names(signed_r2_clust), colnames(signed_r2_clust)]
```

Order the columns in the same way as for the violin plot

```{r}
dat_htmp <- data.matrix(signed_r2_fltr)[, as.vector(unique(miRs_table$cancer))]
```

Draw the heatmap

```{r}
palette <- colorRampPalette(c("blue", "white", "red"))
pdf("TCGA_pancancer_miRgeneDfiVSmiRexpr_fltrGenesSigPosCorrDfeDfi_signed_r2_fltrSigDEmiRs_htmp.pdf", width=25, height=10)
heatmap.2(dat_htmp, trace="none", col=palette, cexRow=0.01, cexCol=2.5,  breaks=seq(-1,1,length.out=50), mar=c(10,20), keysize=1, hclust=function(x) hclust(x,method="complete"), distfun = function(x) dist(x,method = 'euclidean'), density.info = "none", key.par = list(cex=2), lhei=c(1,2), lwid = c(1,5), Colv = F, key.title = "", key.xlab = "Signed r-squared")
dev.off()
```

# Comparison of three associations: A) miRNA abundance vs transcription, B) activity vs abundance and C) activity vs transcription

Load the correlation coefficients from the 3 analyses: A) miRNA abundance vs transcription (output from this current notebook), B) activity vs abundance (output from TCGA_miRNA_activity_vs_abundance.Rmd) and C) activity vs transcription (output from TCGA_miRNA_transcription_vs_activity.Rmd)

```{r}
expr_transc <- readRDS("TCGA_pancancer_miRgeneDfiVSmiRexpr_fltrGenesSigPosCorrDfeDfi_fltrSigDEmiRs_res.rds")
act_expr <- readRDS("TCGA_pancancer_miRfamiliesSigActAtLeast10T_corrFamActVSfamMembersExpr_res.rds")
act_transc <- readRDS("TCGA_pancancer_miRgeneDfiVSmiRact_fltrGenesSigPosCorrDfeDfi_res.rds")
```

Add correlation coefficients where not already indicated

```{r}
act_transc$correlation <- sqrt(abs(act_transc$signed_r2)) * sign(act_transc$signed_r2)
```

Make a data table for the beeswarm plot

```{r}
dt <- list(expr_transc$corr, act_expr$correlation, act_transc$correlation)
names(dt) <- c("miRNA abundance\nvs transcription", "miRNA activity\nvs abundance", "miRNA activity\nvs transcription")
```

Draw the beeswarm plot

```{r}
plt <- beeswarm(dt,
                col = c("olivedrab3", "slateblue3", "dodgerblue"), pch = c(16,16,16),
                method = "swarm",
                cex=0.3, corral = "random",
                ylab="Pearson correlation",
                xlab = "",
                cex.axis=0.8
)
boxplot(dt, add = T, outline=F, col=rgb(0,0,0, alpha=0.001), boxlwd = 2, boxwex=0.5, whisklty = 1, staplelwd = 2, medlwd=2, whisklwd=2, names=c("","",""), yaxt="n") # See https://stackoverflow.com/questions/28889977/r-boxplot-how-to-customize-the-appearance-of-the-box-and-whisker-plots-e-g-r
```

```{r}
sessionInfo()
```