---
title: 'Associations between RBP expression and miRNA activity'
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
```

# Load libraries

```{r}
library(here)
library(tidyverse)
library(scater)
library(SingleCellExperiment)
library(scales)
library(data.table)
library(ppcor)
```

# Load the TCGA gene expression data (log2RSEM) obtained from Firebrowse

```{r}
logRSEM_data <- list.files(here("data"), pattern = "log2RSEM.csv", full.names = TRUE) |>
  lapply(fread) |>
  lapply(as.data.frame) |>
  lapply(function(x) {
    rownames(x) <- x[[1]]
    x[-1]
  })

head(logRSEM_data[[1]][, 1:5])
```

# Define the genes of interest for the t-SNE plots

```{r}
gene_names_tSNE <- c("DDX55", "PTBP1", "SUPV3L1")
```

# Normalize gene expression tables by subtracting the average of normal samples and filter for the genes of interest and for tumor samples  

```{r}
log2FC_gois <- lapply(logRSEM_data, function(x) {
  # Skip datasets without normal samples (ending in -11)
  if (!any(grepl("-11$", colnames(x)))) {
    return(NA)
  }
  
  # Filter to genes of interest
  x <- x[rownames(x) %in% gene_names_tSNE, ]
  
  # Subtract mean of normal samples from all samples
  normal_cols <- grepl("-11$", colnames(x))
  row_means_normal <- rowMeans(x[, normal_cols])
  x <- sweep(x, 1, row_means_normal, "-")
  
  # Transpose and keep only tumor samples (ending in -01)
  x <- t(x)
  as.data.frame(x[grepl("-01$", rownames(x)), ])
})

head(log2FC_gois[[2]])
```

# Extract cancer types and combine data

```{r}
# Extract cancer type from file paths
logRSEM_files <- list.files(here("data"), pattern = "log2RSEM.csv", full.names = TRUE)
cancer_types <- str_extract(logRSEM_files, "[A-Z]+(?=_log2RSEM)")

# Combine data from all cancers into a long-format data frame
log2FC_gois_all <- log2FC_gois |>
  imap_dfr(
    ~ if (is.data.frame(.x)) {
      tibble::rownames_to_column(.x, "sample") |>
        mutate(cancer = cancer_types[.y], .before = 1)
    }
  ) |>
  mutate(sample = str_replace_all(sample, "-", "."))
``` 

# Load the miRNA activity data and filter for tumors

```{r}
miR_act <- read.csv(here("TCGA", "output", "supplementary_tables", "TCGA_miRNA_activity.csv"))
miR_act <- miR_act[miR_act$sample_type == "tumor", ]
```

# Visualization of pan-cancer RBP expression using t-SNE plots

The t-SNE embedding of pan-cancer miRNA activity was already computed and
plotted for miR-17-5p, miR-29-3p, and miR-19-3p in
`09_TCGA_miRNA_activity.Rmd`, which caches the resulting
`SingleCellExperiment` to `TCGA/output/rds/sce_tSNE.rds`. Here we reuse that same embedding (`reducedDim(sce_tSNE, "TSNE")`) and colour it by RBP
expression instead of recomputing the t-SNE.

```{r}
sce_tSNE <- readRDS(here("TCGA", "output", "rds", "sce_tSNE.rds"))
sce_tSNE
```

Merge selected RBP expression log2FC values into the t-SNE object

```{r}
rbp_expr <- log2FC_gois_all |>
  dplyr::select(sample, all_of(gene_names_tSNE))

coldata_tsne <- as.data.frame(colData(sce_tSNE)) |>
  left_join(rbp_expr, by = "sample")

colData(sce_tSNE)[gene_names_tSNE] <- coldata_tsne[gene_names_tSNE]
```

Theme function for t-SNE plots

```{r}
custom_theme_plot <- function(size_text = 15,
                               leg_pos = "right",
                               size_text_titles = NULL,
                               classic = TRUE,
                               line_size = 0.85,
                               line_size_color = "black",
                               axis_ticks_length = 0.25,
                               axis_ticks_color = "black",
                               size_legend_text = 12,
                               size_legend_title = 14) {
  if (is.null(size_text_titles)) {
    size_text_titles <- size_text + 2
  }
  gg.plots <- if (classic) theme_classic() else theme_bw()
  gg.plots +
    theme(
      axis.text.x = element_text(size = size_text, angle = 45, hjust = 1, color = line_size_color),
      axis.line = element_line(size = line_size, colour = line_size_color),
      axis.ticks = element_line(colour = axis_ticks_color, size = line_size),
      axis.ticks.length = unit(axis_ticks_length, "cm"),
      axis.title.x = element_text(size = size_text_titles),
      legend.position = leg_pos,
      plot.title = element_text(hjust = 0.5, size = size_text),
      axis.text.y = element_text(size = size_text, color = "black"),
      axis.title.y = element_text(size = size_text_titles),
      legend.text = element_text(size = size_legend_text),
      legend.title = element_text(size = size_legend_title),
      legend.key.size = unit(0.8, "cm")
    )
}
```

Helper function to create a t-SNE plot colored by RBP expression

```{r}
rbp_tsne_plot <- function(gene, sce_obj, limits = 1) {
  plotReducedDim(sce_obj,
                 ncomponents = 2,
                 dimred = "TSNE",
                 point_size = 2,
                 colour_by = gene) +
    custom_theme_plot(size_text = 27, size_text_titles = 30, size_legend_text = 21, size_legend_title = 24) +
    scale_colour_gradient2(
      low = "blue", high = "red", mid = "white",
      midpoint = 0, limits = c(-limits, limits),
      oob = scales::squish,
      name = paste0(gene, " expression\nlog2FC(RSEM T vs avg N)")
    ) +
    theme(legend.key.size = unit(1.2, "cm"))
}
```

Plot t-SNE for each RBP, coloured by expression log2FC(tumor vs average normal)

```{r, fig.dim = c(12,10), fig.align = 'center'}
rbp_lims <- 1

rbp_pdf_names <- setNames(
  paste0("TCGA_", gene_names_tSNE, "_expr_tSNE.pdf"),
  gene_names_tSNE
)

for (gene in gene_names_tSNE) {
  print(rbp_tsne_plot(gene, sce_tSNE, limits = rbp_lims))

  pdf_file <- here("TCGA", "output", "figures", rbp_pdf_names[[gene]])
  pdf(pdf_file, height = 10, width = 12.5)
  print(rbp_tsne_plot(gene, sce_tSNE, limits = rbp_lims))
  dev.off()

  cat("Saved:", rbp_pdf_names[[gene]], "\n")
}
```

# Correlations between miRNA activity and RBP expression for all candidate RBP-miRNA interactions

Using the RBP-miRNA interaction list from the intersection of candidates identified with the RBP KD experiments from ENCODE and validated using the tissue and cell line samples from ENCODE. 

```{r}
# Load interactions
interactions <- readRDS(here("ENCODE_tissues_cell_lines", "output", "rds", "ENCODE_tissues_cell_lines_high_conf_ints.rds"))

# Add cancer names to expression data list
names(logRSEM_data) <- cancer_types

# Reformat sample names in expression data
logRSEM_data <- lapply(logRSEM_data, function(x) {
  colnames(x) <- str_replace_all(colnames(x), "-", ".")
  x
})
```

```{r}
# Helper function to compute the correlation between the expression of a given RBP and the activity of a given miRNA, within a given cancer type
compute_correlation_cancer <- function(rbp, mirna, cancer) {
  expr_mat <- logRSEM_data[[cancer]]

  # Skip if the RBP is not present in this cancer's expression data
  if (!(rbp %in% rownames(expr_mat))) {
    return(NULL)
  }

  rbp_expr <- as.numeric(expr_mat[rbp, ])
  names(rbp_expr) <- colnames(expr_mat)

  mirna_data <- miR_act |>
    filter(cancer_type == cancer, miRNA == mirna)

  if (nrow(mirna_data) == 0) {
    return(NULL)
  }

  # Match samples between expression and miRNA activity data
  common_samples <- intersect(names(rbp_expr), mirna_data$sample_name)
  if (length(common_samples) == 0) {
    return(NULL)
  }

  rbp_matched <- rbp_expr[common_samples]
  mirna_matched <- mirna_data$coefficient_log2[match(common_samples, mirna_data$sample_name)]

  corr <- cor.test(rbp_matched, mirna_matched)
  tibble(r = corr$estimate, pval = corr$p.value)
}
```

```{r}
# Compute the correlations for every RBP-miRNA interaction, in every cancer type
corr_table <- interactions |>
  dplyr::select(RBP, miRNA, sign) |>
  # Match samples across expression, miRNA activity, and purity data
  tidyr::crossing(cancer = cancer_types) |>
  mutate(
    interaction = paste(RBP, miRNA, sep = " - "),
    corr_data = pmap(list(RBP, miRNA, cancer), compute_correlation_cancer)
  ) |>
  unnest(corr_data, keep_empty = TRUE) |>
  filter(!is.na(r)) |>
  mutate(
    padj = p.adjust(pval, method = "fdr"),
    signedlog10pvals = abs(log10(pval))
  ) |>
  dplyr::select(interaction, RBP, miRNA, sign, cancer, r, pval, padj, signedlog10pvals) |>
  as.data.frame()

head(corr_table)
nrow(corr_table)
```

Cluster RBP-miRNA interactions and cancer types by their correlation coefficients

```{r}
r_mat <- corr_table |>
  dplyr::select(interaction, cancer, r) |>
  pivot_wider(names_from = cancer, values_from = r) |>
  column_to_rownames("interaction") |>
  as.matrix()

row_clust <- hclust(dist(r_mat, method = "euclidean"), method = "complete")
col_clust <- hclust(dist(t(r_mat), method = "euclidean"), method = "complete")

interaction_levels <- rownames(r_mat)[row_clust$order]
cancer_levels <- colnames(r_mat)[col_clust$order]
```

Make the data table for the heatmap, clustered and converting saturated values

```{r}
pval_lim <- 5
coeff_lim <- 0.3

dat_htmp <- corr_table |>
  transmute(
    Interactions = factor(interaction, levels = interaction_levels),
    Cancers = factor(cancer, levels = cancer_levels),
    `-log10(p-value)` = pmin(signedlog10pvals, pval_lim),
    `Pearson r` = pmax(pmin(r, coeff_lim), -coeff_lim),
    FDR = as.integer(padj < 0.05)
  )
```

Assign interaction labels colors according to each interaction's expected sign

```{r}
sign_by_interaction <- corr_table |>
  distinct(interaction, sign) |>
  deframe()

colors_sign <- ifelse(sign_by_interaction[interaction_levels] == "-", "blue", "red")
```

Draw the dotplot

```{r, fig.dim = c(10.2, 7), fig.align = 'center'}
dotplot <- ggplot(dat_htmp, aes(x = Cancers, y = Interactions)) +
  geom_point(aes(size = `-log10(p-value)`, fill = `Pearson r`, color = FDR),
             shape = 21, stroke = 1) +
  scale_fill_gradient2(limits = c(-coeff_lim, coeff_lim), low = "blue", mid = "white", high = "red") +
  scale_colour_gradient(limits = c(0, 1), low = "white", high = "black") +
  scale_size("-log10(p-value)", range = c(0.1, 10), breaks = seq(0, pval_lim, 1), limits = c(0, pval_lim)) +
  guides(size = guide_legend(override.aes = list(fill = "black"))) +
  labs(x = "", y = "") +
  theme(
    axis.text.y = element_text(color = colors_sign, size = 17),
    legend.key.size = unit(1.5, "lines"),
    text = element_text(size = 16),
    axis.text = element_text(size = 17),
    axis.text.x = element_text(angle = 90, size = 17, vjust = 0.5, hjust = 1),
    axis.title = element_text(size = 18),
    legend.title = element_text(size = 17),
    legend.text = element_text(size = 15),
    panel.background = element_rect(fill = "white", colour = "white"),
    plot.margin = unit(c(1, 1, 1, 1), "cm")
  )

dotplot

pdf(here("TCGA", "output", "figures", "TCGA_high_conf_ints_corrRBPexprMirAct_dotplot.pdf"), width = 11, height = 8)
print(dotplot)
dev.off()
```

# Identify interactions for which the direction of correlation is in agreement with the predicted type of interaction (activating / inhibitory) in at least 70% of TCGA cancers

```{r}
length(unique(corr_table$interaction))
sign_consistent <- corr_table %>%
  group_by(interaction, RBP, miRNA, sign) %>%
  summarise(
    prop_agree = mean(ifelse(sign == "+", r > 0, r < 0), na.rm = TRUE),
    n_cancers = n(),
    .groups = "drop"
  ) %>%
  filter(prop_agree >= 0.70) %>%
  dplyr::select(interaction, RBP, miRNA, sign)
sign_consistent
nrow(sign_consistent)

write.csv(sign_consistent, here("TCGA", "output", "supplementary_tables", "interactions_RBPKD_ENCODEcellsTissues_TCGA.csv"))
```

# Example scatter plot for SUPV3L1-miR-19 in BRCA

Helper function to make a base R scatter plot of RBP expression vs. miRNA activity, for a given RBP, miRNA, and cancer type

```{r}
rbp_mirna_scatter <- function(rbp, mirna, cancer, save_pdf = FALSE, pdf_dir = here("TCGA", "output", "figures")) {
  expr_mat <- logRSEM_data[[cancer]]

  rbp_expr_vals <- as.numeric(expr_mat[rbp, ])
  names(rbp_expr_vals) <- colnames(expr_mat)

  mirna_data <- miR_act |>
    filter(cancer_type == cancer, miRNA == mirna)

  # Match samples between expression and miRNA activity data
  common_samples <- intersect(names(rbp_expr_vals), mirna_data$sample_name)

  rbp_matched <- rbp_expr_vals[common_samples]
  mirna_matched <- mirna_data$coefficient_log2[match(common_samples, mirna_data$sample_name)]

  corr <- cor.test(rbp_matched, mirna_matched)
  n_tumors <- sum(complete.cases(rbp_matched, mirna_matched))

  draw_plot <- function() {
    plot(rbp_matched, mirna_matched,
         pch = 19,
         cex = 0.35,
         main = cancer,
         xlab = paste0(rbp, " expression (log2RSEM)"),
         ylab = paste0(mirna, " activity"))
    abline(lm(mirna_matched ~ rbp_matched), col = "dodgerblue", lwd = 2)
    if (sign(corr$estimate) == 1){
      legend("bottomright",
             legend = paste0("r = ", round(corr$estimate, 3), "\np = ", signif(corr$p.value, 3), "\nn = ", n_tumors, " tumors"),
             bty = "n",
             inset = c(0, 0.1),
             cex = 0.9)      
    } else {
      legend("topleft",
             legend = paste0("r = ", round(corr$estimate, 3), "\np = ", signif(corr$p.value, 3), "\nn = ", n_tumors, " tumors"),
             bty = "n",
             inset = c(-0.05, -0.05),
             cex = 0.9)           
    }

  }

  draw_plot()

  if (save_pdf) {
    pdf_file <- file.path(pdf_dir, paste0("TCGA_", cancer, "_", rbp, "_", mirna, "_scatter.pdf"))
    pdf(pdf_file, width = 4, height = 4)
    draw_plot()
    dev.off()
  }

  invisible(corr)
}
```

Scatter plot for SUPV3L1 expression vs. miR-19-3p activity in BRCA

```{r, fig.dim = c(6, 6), fig.align = 'center'}
rbp_mirna_scatter("SUPV3L1", "miR-19-3p", "BRCA", save_pdf = TRUE)
```

# Example scatter plot for DDX55-miR-29 in STAD

```{r, fig.dim = c(6, 6), fig.align = 'center'}
rbp_mirna_scatter("DDX55", "miR-29-3p", "STAD", save_pdf = TRUE)
```

# Example bar plot for DDX55 - miR-29-3p across cancers

Helper function to make bar plot of RBP vs. miRNA activity correlation across cancer types, for a given RBP-miRNA interaction

```{r}
rbp_mirna_barplot <- function(rbp, mirna, corr_data = corr_table,
                               sig_col = "pval", sig_threshold = 0.05,
                               bar_color = "#5B9BD5", ns_color = "grey70",
                               save_pdf = FALSE,
                               pdf_dir = here("TCGA", "output", "figures"),
                               width = 5.2, height = 4) {
  dat <- corr_data |>
    filter(RBP == rbp, miRNA == mirna) |>
    mutate(
      Significant = factor(
        ifelse(.data[[sig_col]] < sig_threshold, "p < 0.05", "NS"),
        levels = c("p < 0.05", "NS")
      )
    )

  # Order cancers by strength of correlation, matching the expected sign
  expected_sign <- unique(dat$sign)
  dat <- dat |>
    arrange(if (expected_sign == "-") r else dplyr::desc(r)) |>
    mutate(cancer = factor(cancer, levels = cancer))

  y_lab <- paste0(
    rbp, " expression vs ", mirna, " activity\n",
    "across tumors (Pearson r)"
  )

  p <- ggplot(dat, aes(x = cancer, y = r, fill = Significant)) +
    geom_col(colour = "black", linewidth = 0.3, width = 0.8) +
    scale_x_discrete(expand = expansion(add = 0.6)) +
    scale_fill_manual(values = c("p < 0.05" = bar_color, "NS" = ns_color), drop = FALSE) +
    labs(x = NULL, y = y_lab, fill = NULL) +
    theme_classic(base_size = 14) +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, colour = "black"),
      axis.text.y = element_text(colour = "black"),
      axis.title.y = element_text(size = 13),
      axis.line = element_line(linewidth = 0.4),
      axis.ticks = element_line(linewidth = 0.4),
      legend.position = if (expected_sign == "+") c(0.98, 0.98) else c(0.98, 0.02),
      legend.justification = if (expected_sign == "+") c(1, 1) else c(1, 0),
      legend.background = element_rect(colour = "black", linewidth = 0.3),
      legend.margin = margin(4, 6, 4, 6),
      plot.margin = margin(10, 10, 10, 10)
    )

  print(p)

  if (save_pdf) {
    pdf_file <- file.path(pdf_dir, paste0("TCGA_", rbp, "_", mirna, "_barplot.pdf"))
    ggsave(pdf_file, plot = p, width = width, height = height)
  }

  invisible(p)
}
```

# Bar plot for DDX55 expression vs. miR-29-3p activity correlation across cancers

```{r, fig.dim = c(6, 5), fig.align = 'center'}
rbp_mirna_barplot("DDX55", "miR-29-3p", save_pdf = TRUE)
```

# Bar plot for PTBP1 expression vs. miR-29-3p activity

```{r, fig.dim = c(6, 5), fig.align = 'center'}
rbp_mirna_barplot("PTBP1", "miR-29-3p", save_pdf = TRUE)
```

# Bar plot for RPLP0 expression vs. miR-29-3p activity

```{r, fig.dim = c(6, 5), fig.align = 'center'}
rbp_mirna_barplot("RPLP0", "miR-29-3p", save_pdf = TRUE)
```

# Bar plot for FAM120A expression vs. miR-29-3p activity

```{r, fig.dim = c(6, 5), fig.align = 'center'}
rbp_mirna_barplot("FAM120A", "miR-29-3p", save_pdf = TRUE)
```

# Correction of associations for tumor purity

Essentially the same analysis as above, but with an added correction for tumor purity, by computing partial correlations between RBP expression and miRNA activity while controlling for tumor purity as a covariate.

## Load tumor purity estimates

```{r}
purity_files <- list.files(here("data"), pattern = "^purity_.*\\.txt$", full.names = TRUE)

purity_data <- purity_files |>
  setNames(str_extract(basename(purity_files), "(?<=purity_)[A-Z]+(?=\\.txt)")) |>
  lapply(function(f) {
    dat <- read.delim(f, stringsAsFactors = FALSE)
    # Keep only primary tumor, vial A samples (-01A-)
    dat <- dat[grepl("-01A-", dat$SampleName), ]
    dat$sample <- str_extract(dat$SampleName, "^TCGA-[^-]+-[^-]+-01") |>
      str_replace_all("-", ".")
    dat[, c("sample", "Purity")]
  })

head(purity_data[["UCEC"]])
```

## Compute partial correlations controlling for tumor purity

```{r}
# Helper function to compute the partial correlation between the expression of a given RBP and the activity of a given miRNA, within a given cancer type, controlling for tumor purity
compute_partial_correlation_cancer <- function(rbp, mirna, cancer) {
  # Skip cancer types without a purity file
  if (!(cancer %in% names(purity_data))) {
    return(NULL)
  }

  expr_mat <- logRSEM_data[[cancer]]

  # Skip if the RBP is not present in this cancer's expression data
  if (!(rbp %in% rownames(expr_mat))) {
    return(NULL)
  }

  rbp_expr <- as.numeric(expr_mat[rbp, ])
  names(rbp_expr) <- colnames(expr_mat)

  mirna_data <- miR_act |>
    filter(cancer_type == cancer, miRNA == mirna)

  if (nrow(mirna_data) == 0) {
    return(NULL)
  }

  purity_cancer <- purity_data[[cancer]]

  # Match samples across expression, miRNA activity, and purity data
  common_samples <- Reduce(intersect, list(names(rbp_expr), mirna_data$sample_name, purity_cancer$sample))
  if (length(common_samples) < 4) {
    return(NULL)
  }

  rbp_matched <- rbp_expr[common_samples]
  mirna_matched <- mirna_data$coefficient_log2[match(common_samples, mirna_data$sample_name)]
  purity_matched <- purity_cancer$Purity[match(common_samples, purity_cancer$sample)]

  pcorr <- pcor.test(rbp_matched, mirna_matched, purity_matched)
  tibble(r = pcorr$estimate, pval = pcorr$p.value, n = pcorr$n)
}
```

```{r}
# Compute the purity-corrected partial correlations for every RBP-miRNA interaction, in every cancer type
corr_table_purity <- interactions |>
  dplyr::select(RBP, miRNA, sign) |>
  # Cross every interaction with every cancer type
  tidyr::crossing(cancer = cancer_types) |>
  mutate(
    interaction = paste(RBP, miRNA, sep = " - "),
    corr_data = pmap(list(RBP, miRNA, cancer), compute_partial_correlation_cancer)
  ) |>
  unnest(corr_data, keep_empty = TRUE) |>
  filter(!is.na(r)) |>
  mutate(
    padj = p.adjust(pval, method = "fdr"),
    signedlog10pvals = abs(log10(pval))
  ) |>
  dplyr::select(interaction, RBP, miRNA, sign, cancer, r, pval, padj, signedlog10pvals, n) |>
  as.data.frame()

head(corr_table_purity)
nrow(corr_table_purity)
```

## Dotplot of purity-corrected correlations

```{r}
r_mat_purity <- corr_table_purity |>
  dplyr::select(interaction, cancer, r) |>
  pivot_wider(names_from = cancer, values_from = r) |>
  column_to_rownames("interaction") |>
  as.matrix()

row_clust_purity <- hclust(dist(r_mat_purity, method = "euclidean"), method = "complete")
col_clust_purity <- hclust(dist(t(r_mat_purity), method = "euclidean"), method = "complete")

interaction_levels_purity <- rownames(r_mat_purity)[row_clust_purity$order]
cancer_levels_purity <- colnames(r_mat_purity)[col_clust_purity$order]
```

Make the data table for the dotplot, clustered and converting saturated values

```{r}
dat_htmp_purity <- corr_table_purity |>
  transmute(
    Interactions = factor(interaction, levels = interaction_levels_purity),
    Cancers = factor(cancer, levels = cancer_levels_purity),
    `-log10(p-value)` = pmin(signedlog10pvals, pval_lim),
    `Pearson r` = pmax(pmin(r, coeff_lim), -coeff_lim),
    FDR = as.integer(padj < 0.05)
  )
```

Assign interaction labels colors according to each interaction's expected sign

```{r}
sign_by_interaction_purity <- corr_table_purity |>
  distinct(interaction, sign) |>
  deframe()

colors_sign_purity <- ifelse(sign_by_interaction_purity[interaction_levels_purity] == "-", "blue", "red")
```

Draw the dotplot

```{r, fig.dim = c(10.2, 7), fig.align = 'center'}
dotplot_purity <- ggplot(dat_htmp_purity, aes(x = Cancers, y = Interactions)) +
  geom_point(aes(size = `-log10(p-value)`, fill = `Pearson r`, color = FDR),
             shape = 21, stroke = 1) +
  scale_fill_gradient2(name = "Partial correlation\ncoefficient", limits = c(-coeff_lim, coeff_lim), low = "blue", mid = "white", high = "red") +
  scale_colour_gradient(limits = c(0, 1), low = "white", high = "black") +
  scale_size("-log10(p-value)", range = c(0.1, 10), breaks = seq(0, pval_lim, 1), limits = c(0, pval_lim)) +
  guides(size = guide_legend(override.aes = list(fill = "black"))) +
  labs(x = "", y = "") +
  theme(
    axis.text.y = element_text(color = colors_sign_purity, size = 17),
    legend.key.size = unit(1.5, "lines"),
    text = element_text(size = 16),
    axis.text = element_text(size = 17),
    axis.text.x = element_text(angle = 90, size = 17, vjust = 0.5, hjust = 1),
    axis.title = element_text(size = 18),
    legend.title = element_text(size = 17),
    legend.text = element_text(size = 15),
    panel.background = element_rect(fill = "white", colour = "white"),
    plot.margin = unit(c(1, 1, 1, 1), "cm")
  )

dotplot_purity

pdf(here("TCGA", "output", "figures", "TCGA_high_conf_ints_corrRBPexprMirAct_purityCorrected_dotplot.pdf"), width = 10, height = 8)
print(dotplot_purity)
dev.off()
```

## Identify interactions consistent with the predicted direction in at least 70% of TCGA cancers, after purity correction

```{r}
length(unique(corr_table_purity$interaction))
sign_consistent_purity <- corr_table_purity %>%
  group_by(interaction, RBP, miRNA, sign) %>%
  summarise(
    prop_agree = mean(ifelse(sign == "+", r > 0, r < 0), na.rm = TRUE),
    n_cancers = n(),
    .groups = "drop"
  ) %>%
  filter(prop_agree >= 0.70) %>%
  dplyr::select(interaction, RBP, miRNA, sign)
sign_consistent_purity
nrow(sign_consistent_purity)

write.csv(sign_consistent_purity, here("TCGA", "output", "supplementary_tables", "interactions_RBPKD_ENCODEcellsTissues_TCGA_purityCorrected.csv"))
```

```{r}
sessionInfo()
```
