---
title: 'Associations between RBP expression and miRNA activity'
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
```

# Load libraries

```{r}
library(tidyverse)
library(stringr)
library(scater)
library(SingleCellExperiment)
library(rsvd)
library(scales)
library(data.table)
# library(ppcor)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(gplots)
```

# Load the TCGA gene expression data (log2RSEM) obtained from Firebrowse

```{r}
logRSEM_data <- list.files(here("TCGA", "data"), pattern = "log2RSEM.csv", full.names = TRUE) |>
  lapply(fread) |>
  lapply(as.data.frame) |>
  lapply(function(x) {
    rownames(x) <- x[[1]]
    x[-1]
  })

head(logRSEM_data[[1]][, 1:5])
```

# Define the gene IDs and gene names of interest

```{r}
gene_IDs <- c("ENSG00000111364", "ENSG00000011304", "ENSG00000156502")
gene_names <- c("DDX55", "PTBP1", "SUPV3L1")
```

# Filter gene expression tables for the genes of interest and for tumor samples, and normalize by subtracting the average of normal samples

```{r}
log2FC_gois <- lapply(logRSEM_data, function(x) {
  # Skip datasets without normal samples (ending in -11)
  if (!any(grepl("-11$", colnames(x)))) {
    return(NA)
  }
  
  # Filter to genes of interest
  x <- x[rownames(x) %in% gene_names, ]
  
  # Subtract mean of normal samples from all samples
  normal_cols <- grepl("-11$", colnames(x))
  row_means_normal <- rowMeans(x[, normal_cols])
  x <- sweep(x, 1, row_means_normal, "-")
  
  # Transpose and keep only tumor samples (ending in -01)
  x <- t(x)
  as.data.frame(x[grepl("-01$", rownames(x)), ])
})

head(log2FC_gois[[2]])
```

# Extract cancer types and combine data

```{r}
# Extract cancer type from file paths
logRSEM_files <- list.files(here("TCGA", "data"), pattern = "log2RSEM.csv", full.names = TRUE)
cancer_types <- str_extract(logRSEM_files, "[A-Z]+(?=_log2RSEM)")

# Combine all data into a long-format data frame
log2FC_all <- log2FC_gois |>
  imap_dfr(
    ~ if (is.data.frame(.x)) {
      tibble::rownames_to_column(.x, "sample") |>
        mutate(cancer = cancer_types[.y], .before = 1) |>
        rename_with(tolower, all_of(gene_names))
    }
  ) |>
  mutate(sample = str_replace_all(sample, "-", "."))
``` 

# Load the miRNA activity data and filter for tumors

```{r}
miR_act <- read.csv(here("TCGA", "output", "supplementary_tables", "TCGA_miRNA_activity.csv"))
miR_act <- miR_act[miR_act$sample_type == "tumor", ]
```

# Combine miRNA activity and RBP expression data 

```{r}
<==
common_samples <- intersect(signedLogPvals_fltr$X, row.names(log2FC_all))
log2FC_all <- log2FC_all[common_samples, ]
signedLogPvals_fltr <- signedLogPvals_fltr[signedLogPvals_fltr$X %in% common_samples, ]
act_fltr <- act_fltr[act_fltr$X %in% common_samples, ]
all(row.names(log2FC_all) == act_fltr$X)
```

# Visualization of pan-cancer RBP expression using t-SNE plots 

Function used to make the t-SNE plot. The scripts are derived from code shared by Ariel Madrigal Aguirre

```{r}
custom_theme_plot <- function(size_text=15,
                              leg_pos="right",
                              size_text_titles=NULL,
                              classic=TRUE,
                              line_size=0.85,
                              line_size_color="black",
                              axis_ticks_length=0.25,
                              axis_ticks_color="black"
                              ){
    require(ggplot2)
    if( is.null(size_text_titles) ){
        size_text_titles<- size_text + 2
    }
    if( classic ){
        gg.plots<- theme_classic()
    }else{
        gg.plots<- theme_bw()
    }
    gg.plots <- gg.plots +
        theme(axis.text.x=element_text(size=size_text, angle=45, hjust=1, color=line_size_color),
              axis.line=element_line(size=line_size, colour=line_size_color),
              axis.ticks=element_line(colour=axis_ticks_color, size=line_size),
              axis.ticks.length=unit(axis_ticks_length, "cm"),
              axis.title.x=element_text(size=size_text_titles),
              legend.position=leg_pos,
              plot.title=element_text(hjust=0.5, size=size_text),
              axis.text.y=element_text(size=size_text, color="black"),
              axis.title.y=element_text(size=size_text_titles) )
}
```

Create the data object to make the t-SNE plots

```{r}
dt <- act_fltr[ , c(-1, -ncol(act_fltr))]
dt <- data.matrix(t(dt))
coldata <- as.data.frame(cbind(as.vector(act_fltr$X), act_fltr$miR.29abcd, act_fltr$miR.19ab, act_fltr$miR.17, log2FC_all))
colnames(coldata) <- c("sample", "miR-29abcd", "miR-19ab", "miR-17", gene_names, "cancer")
for(i in 2:(ncol(coldata)-1)){
  coldata[,i] <- as.numeric(as.vector(coldata[,i]))
}
sce <- SingleCellExperiment(list(logcounts=dt), colData=coldata, checkDimnames=FALSE)
sce
```

Add PCA

```{r}
set.seed(18374)
sce_pca<- rpca(t(assay(sce, "logcounts")), k=100)
reducedDim(sce, "pca")<- sce_pca$x
```

t-SNE
```{r}
sce_tSNE <- runTSNE(sce, dimred="pca", perplexity=50) # perplexity=5, n_dimred=10
```

Set limits

```{r}
mir_lims=0.05
rbp_lims=1
```

Plot t-SNE showing the cancer type

```{r}
plotReducedDim(sce_tSNE,
               ncomponents = 2,
               dimred="TSNE",
               point_size=1.5,               
               colour_by ="cancer") +
      custom_theme_plot() +  
      geom_point(shape = 1, size=1.5 ,colour = "grey", alpha=0.5)
```

Show DDX55 expression

```{r}
plotReducedDim(sce_tSNE,
               ncomponents = 2,
               dimred="TSNE",
               point_size=1.5,               
               colour_by ="DDX55") +
    custom_theme_plot() +
    scale_colour_gradient2(low="blue", high="red", mid="white", midpoint = 0, limits=c(-rbp_lims, rbp_lims), oob=squish, name="RBP expression\nlog2FC(RSEM T vs avg N)") +
    geom_point(shape = 1, size=1.5 ,colour = "grey", alpha=0.5)
```

Show PTBP1 expression

```{r}
plotReducedDim(sce_tSNE,
               ncomponents = 2,
               dimred="TSNE",
               point_size=1.5,               
               colour_by ="PTBP1") +
    custom_theme_plot() +
    scale_colour_gradient2(low="blue", high="red", mid="white", midpoint = 0, limits=c(-rbp_lims, rbp_lims), oob=squish, name="RBP expression\nlog2FC(RSEM T vs avg N)") +
    geom_point(shape = 1, size=1.5 ,colour = "grey", alpha=0.5)
```

Show SUPV3L1 expression

```{r}
plotReducedDim(sce_tSNE,
               ncomponents = 2,
               dimred="TSNE",
               point_size=1.5,               
               colour_by ="SUPV3L1") +
    custom_theme_plot() +
    scale_colour_gradient2(low="blue", high="red", mid="white", midpoint = 0, limits=c(-rbp_lims, rbp_lims), oob=squish, name="RBP expression\nlog2FC(RSEM T vs avg N)") +
    geom_point(shape = 1, size=1.5 ,colour = "grey", alpha=0.5)
```

Show miR-17 activity

```{r}
plotReducedDim(sce_tSNE,
               ncomponents = 2,
               dimred="TSNE",
               point_size=1.5,               
               colour_by ="miR-17") +
    custom_theme_plot() +
    scale_colour_gradient2(low="blue", high="red", mid="white", midpoint = 0, limits=c(-mir_lims, mir_lims), oob=squish, name="miRNA activity") +
    geom_point(shape = 1, size=1.5 ,colour = "grey", alpha=0.5)
```

Show miR-29 activity

```{r}
plotReducedDim(sce_tSNE,
               ncomponents = 2,
               dimred="TSNE",
               point_size=1.5,               
               colour_by ="miR-29abcd") +
    custom_theme_plot() +
    scale_colour_gradient2(low="blue", high="red", mid="white", midpoint = 0, limits=c(-mir_lims, mir_lims), oob=squish, name="miRNA activity") +
    geom_point(shape = 1, size=1.5 ,colour = "grey", alpha=0.5)
```

Show miR-19 activity

```{r}
plotReducedDim(sce_tSNE,
               ncomponents = 2,
               dimred="TSNE",
               point_size=1.5,               
               colour_by ="miR-19ab") +
    custom_theme_plot() +
    scale_colour_gradient2(low="blue", high="red", mid="white", midpoint = 0, limits=c(-mir_lims, mir_lims), oob=squish, name="miRNA activity") +
    geom_point(shape = 1, size=1.5 ,colour = "grey", alpha=0.5)

```

# Systematic analysis of correlations between miRNA activity and RBP expression for all candidate RBP-miRNA interactions

Using the RBP-miRNA interaction list from the intersection of candidates identified with the RBP KD experiments from ENCODE and validated usisng the tissue and cell line samples from ENCODE. Define RBPs and miRNAs of interest

```{r}
interactions <- read.csv("RBPKD_miRNAs_newSigInts_corrShRNA6nts7ntsSigPosFishers_intersectRBPKDtotalRNAseq.csv")[,-1]
interactions$miRNA <- as.vector(interactions$miRNA)
interactions$RBP <- as.vector(interactions$RBP)
miRNAs <- as.vector(interactions$miRNA)
```

Reformat the miRNA activity table

```{r}
row.names(act) <- act$X
act <- act[,-1]
row.names(act) <- gsub("\\.", "-", row.names(act))
colnames(act) <- gsub("\\.", "-", colnames(act))
```

Initiallize tables where the correlation coefficients and p-values will be stored

```{r}
coeffs <- matrix(NA, nrow=nrow(interactions), ncol = length(cancers))
row.names(coeffs) <- paste(interactions$RBP, miRNAs, sep = "_")
colnames(coeffs) <- cancers

pvals <- coeffs
```

For each interaction pair, compute the correlations and record the results. See comments within the code for descriptions of the steps.

```{r}
for (i in 1:nrow(interactions))
{
  # Filter all gene expression tables for the RBP of interest
  RBP <- as.vector(interactions$RBP[i])
  
  logRSEM_gene <- lapply(logRSEM_data, function(x) unlist(x[RBP, ]))
  logRSEM_gene <- unlist(logRSEM_gene)
  
  if(sum(is.na(logRSEM_gene)) == length(logRSEM_gene) | length(logRSEM_gene) == 0)
  {
    next
  }

  # Make a vector indicating the cancer type of each sample
  nb_samples_cancers <- unlist(lapply(logRSEM_data, function(x) ncol(x)))
  cancers_vector <- rep(cancers, nb_samples_cancers)
  
  # Merge into table containing all samples of all cancer types and indicating the cancer type
  logRSEM_complete <- as.data.frame(cbind(logRSEM_gene, cancers_vector))
  colnames(logRSEM_complete) <- c("GOI_expr", "cancer")

  # Filter miRNA activity table for the miRNA of interest
  miRNA <- interactions$miRNA[i]
  miR_cancers <- as.data.frame(cbind(row.names(act), act[,miRNA], as.vector(act$cancers_vector)))
  colnames(miR_cancers) <- c("samples", "miR_act", "cancer")
  
  # In each cancer, compute the pearson correlation coefficient and p-value
  for (cancer in cancers)
  {
    # Make a table with samples, GOI expression, and miRNA activity in that cancer type, removing normal samples 
    cancer_data <- logRSEM_complete[logRSEM_complete$cancer== cancer, ]
    row.names(cancer_data) <- gsub("-1$", "", row.names(cancer_data))
    cancer_data <- cancer_data[grep("\\-01$", row.names(cancer_data)), ]
    cancer_data$miRNA_act <- NA
    cancer_miR <- miR_cancers[miR_cancers$cancer == cancer, ]
    if (nrow(cancer_miR) == 0)
    {
      next
    }
    cancer_data$miRNA_act <- cancer_miR[match(row.names(cancer_data), cancer_miR$samples),]$miR_act
    
    cancer_data <- cancer_data[rowSums(is.na(cancer_data)) == 0, ]
    
    # Correlation
    corr <- cor.test(as.numeric(as.vector(cancer_data$GOI_expr)), as.numeric(as.vector(cancer_data$miRNA_act)))
    coeffs[i, cancer] <- corr$estimate
    pvals[i, cancer] <- corr$p.value
  }
  
  # Add + or - sign to the interaction name to indicate the type of interaction
  if (interactions$sign[i] == "-")
  {
    row.names(coeffs)[i] <- paste("-", row.names(coeffs)[i], sep="  ")
  } else {
    row.names(coeffs)[i] <- paste("+", row.names(coeffs)[i], sep="  ")
  }
  row.names(pvals) <- row.names(coeffs)
}
```

Remove NA rows and columns

```{r}
rows_keep <- which(rowSums(is.na(coeffs)) != ncol(coeffs))
cols_keep <- which(colSums(is.na(coeffs)) != nrow(coeffs))
coeffs <- coeffs[rows_keep, cols_keep]
pvals <- pvals[rows_keep, cols_keep]
```

FDR correction

```{r}
pvalues_vector <- c(pvals)
pvalues_vector_FDR <- p.adjust(pvalues_vector, method="fdr")
FDR <- matrix(pvalues_vector_FDR, nrow = nrow(pvals), ncol = ncol(pvals))
row.names(FDR) <- row.names(pvals)
colnames(FDR) <- colnames(pvals)
```

Compute -log10 p-values 

```{r}
signedlog10pvals <- abs(log10(pvals)) 
```

Get data after clustering

```{r, results=F}
htmp <- heatmap.2(coeffs, hclust=function(x) hclust(x,method="complete"), distfun = function(x) dist(x,method = 'euclidean'))
coeffs_clust <- coeffs[rev(htmp$rowInd), htmp$colInd]
signedlog10pvals_clust <- signedlog10pvals[row.names(coeffs_clust), colnames(coeffs_clust)]
FDR_clust <- FDR[row.names(coeffs_clust), colnames(coeffs_clust)]
```

Make data tables to make the heatmap

```{r}
signedLogPval_srt <- data.table(cbind(row.names(signedlog10pvals_clust), signedlog10pvals_clust))
colnames(signedLogPval_srt)[1] <- "Interactions"
signedLogPval_melt <- melt(signedLogPval_srt, id.vars = "Interactions")
names(signedLogPval_melt)[2:3] <- c("Cancers", "signedLog10pval")

coeffs_srt <- data.table(cbind(row.names(coeffs_clust), coeffs_clust))
colnames(coeffs_srt)[1] <- "Interactions"
coeffs_melt <- melt(coeffs_srt, id.vars = "Interactions")
names(coeffs_melt)[2:3] <- c("Cancers", "coeffs")

FDR_srt <- data.table(cbind(row.names(FDR_clust), FDR_clust))
colnames(FDR_srt)[1] <- "Interactions"
FDR_melt <- melt(FDR_srt, id.vars = "Interactions")
names(FDR_melt)[2:3] <- c("Cancers", "FDR")
```

Combine into a single table

```{r}
dat_htmp <- cbind(signedLogPval_melt, coeffs_melt$coeffs, FDR_melt$FDR)
colnames(dat_htmp)[3:5] <- c("-log10(p-value)", "coeffs", "padj")
dat_htmp[, "-log10(p-value)"] <- as.numeric(as.vector(unlist(dat_htmp[, "-log10(p-value)"])))
dat_htmp$coeffs <- as.numeric(as.vector(dat_htmp$coeffs))
dat_htmp$padj <- as.numeric(as.vector(dat_htmp$padj))

dat_htmp$FDR <- ifelse(dat_htmp$padj<0.05, 1, 0)
```

Set -log10(p-value) and coefficients limits and squish saturated values

```{r}
pval_lim <- 5
dat_htmp$`-log10(p-value)`[abs(dat_htmp$`-log10(p-value)`) > pval_lim] <- pval_lim
coeff_lim <- 0.3
dat_htmp$coeffs[abs(dat_htmp$coeffs) > coeff_lim] <- coeff_lim*sign(dat_htmp$coeffs[abs(dat_htmp$coeffs) > coeff_lim])
```

Assign colors corresponding to interaction types

```{r}
colors_sign <- rev(row.names(coeffs_clust))
colors_sign <- gsub(" .*", "", colors_sign)
colors_sign[grep("-", colors_sign)] <- "blue"
colors_sign[grep("\\+", colors_sign)] <- "red"
```

Improve label and miRNA names format

```{r}
colnames(dat_htmp)[4] <- "Pearson correlation"
dat_htmp$Interactions <- gsub(".* ", "", dat_htmp$Interactions)
signedLogPval_srt$Interactions <- gsub(".* ", "", signedLogPval_srt$Interactions)
row.names(coeffs_clust) <- gsub(".* ", "", row.names(coeffs_clust) )

lbls <- row.names(coeffs_clust)
lbls <- gsub("miR-33ab$"," - miR-33-5p", lbls)
lbls <- gsub("miR-26ab$"," - miR-26-5p", lbls)
lbls <- gsub("miR-29abcd$"," - miR-29-3p", lbls)
lbls <- gsub("miR-27abc$"," - miR-27-3p", lbls)
lbls <- gsub("miR-133abc$"," - miR-133-3p", lbls)
lbls <- gsub("miR-199ab-5p$"," - miR-199-5p", lbls)
lbls <- gsub("miR-376abd$"," - miR-376-3p", lbls)
lbls <- gsub("miR-23abc$"," - miR-23-3p", lbls)
lbls <- gsub("miR-19ab$"," - miR-19-3p", lbls)
lbls <- gsub("miR-182$"," - miR-182-5p", lbls)
lbls <- gsub("miR-758$"," - miR-758-3p", lbls)
lbls <- gsub("miR-22$"," - miR-22-3p", lbls)
lbls <- gsub("miR-150$"," - miR-150-5p", lbls)
lbls <- gsub("miR-124$"," - miR-124-3p", lbls)
lbls <- gsub("miR-96$"," - miR-96-5p", lbls)
lbls <- gsub("miR-370$"," - miR-370-3p", lbls)
lbls <- gsub("miR-17$"," - miR-17-5p", lbls)
lbls <- gsub("miR-218$"," - miR-218-5p", lbls)
lbls <- gsub("miR-21$"," - miR-21-5p", lbls)
lbls <- gsub("miR-25$"," - miR-25-3p", lbls)
lbls <- gsub("_ -"," -", lbls)
lbls <- gsub("_"," - ", lbls)
```

Draw the heatmap

```{r, eval = F}
pdf(paste("heatmap_TCGA_intsConfENCODEtotalRNAseq_corrRBPexprMirAct_signedLog10PvaluesANDcoeffs.pdf", sep=""), width=10.2, height=7)
ggplot(dat_htmp) +
geom_point(aes(
  x = Cancers, 
  size = `-log10(p-value)`, 
  y = factor(Interactions, levels = rev(as.character(signedLogPval_srt$Interactions))),
  shape = 21, 
  stroke = 1,
  fill=`Pearson correlation`, 
  color=FDR)) +
scale_shape_identity() +
scale_y_discrete(label = rev(lbls)) +
theme(axis.text.y = element_text(color=colors_sign)) +
scale_fill_gradient2(limits=c(-coeff_lim,coeff_lim), low="blue", mid="white", high="red") + 
scale_colour_gradient( limits=c(0,1), low="white", high="black") + 
scale_size("-log10(p-value)", range = c(0.1, 10), breaks = seq(0, pval_lim, 1), limits=c(0, pval_lim))+
guides(size = guide_legend(override.aes = list(fill = "red")))+
labs(x="",y="") +
theme(legend.key.size = unit(1.5, 'lines'), text = element_text(size=12), axis.text = element_text(size=12), axis.text.x = element_text(angle=90, size=12, vjust=0.5, hjust=1), panel.background = element_rect(fill = 'white', colour = 'white')) +
theme(plot.margin = unit(c(1,1,1,1), "cm")) 
dev.off()
```

![Heatmap of associations between RBP expression and miRNA activity per cancer type](heatmap_TCGA_intsConfENCODEtotalRNAseq_corrRBPexprMirAct_signedLog10PvaluesANDcoeffs.pdf){width=70%}

Make a list of interactions for which the direction of correlation is in agreement with the predicted type of interaction (activating / inhibitory) in at least 70% of TCGA cancers

```{r,}
toKeep <- rep(NA, nrow(coeffs))
for (i in 1:nrow(coeffs))
{
  int_sign <- gsub(" .*", "", row.names(coeffs)[i])
  if(int_sign == "-")
  {
    int_sign <- "-1"
  } else {
    int_sign <- "1"
  }
  toKeep[i] <- sum(sign(coeffs[i,]) == int_sign)/ncol(coeffs) >= 0.7
}

row.names(coeffs)[toKeep]
```

Filter the previous table of high-confidence interactions to retain these, and save

```{r}
new_ints <- interactions[paste(interactions$RBP, miRNAs, sep = "_") %in% gsub(".* ", "", row.names(signedlog10pvals)[toKeep]), ]
write.csv(new_ints, "RBPKD_miRNAs_newSigInts_corrShRNA6nts7ntsSigPosFishers_intersectRBPKDtotalRNAseqANDtcgaCorrs70percent.csv")

```


Example scatter plot for SUPV3L1-miR-19 in BRCA

```{r, echo=F}
RBP <- "SUPV3L1"
miRNA <- "miR-19ab"
cancer="BRCA"

# Filter all gene expression tables for the RBP of interest
logRSEM_gene <- lapply(logRSEM_data, function(x) unlist(x[RBP, ]))
logRSEM_gene <- unlist(logRSEM_gene)

# Make a vector indicating the cancer type of each sample
nb_samples_cancers <- unlist(lapply(logRSEM_data, function(x) ncol(x)))
cancers_vector <- rep(cancers, nb_samples_cancers)

# Merge into table containing all samples of all cancer types and indicating the cancer type
logRSEM_complete <- as.data.frame(cbind(logRSEM_gene, cancers_vector))
colnames(logRSEM_complete) <- c("GOI_expr", "cancer")

# Filter miRNA activity table for the miRNA of interest
miR_cancers <- as.data.frame(cbind(row.names(act), act[,miRNA], as.vector(act$cancers_vector)))
colnames(miR_cancers) <- c("samples", "miR_act", "cancer")

# Make a table with samples, GOI expression, and miRNA activity in that cancer type, removing normal samples 
cancer_data <- logRSEM_complete[logRSEM_complete$cancer== cancer, ]
row.names(cancer_data) <- gsub("-1$", "", row.names(cancer_data))
cancer_data <- cancer_data[grep("\\-01$", row.names(cancer_data)), ]
cancer_data$miRNA_act <- NA
cancer_miR <- miR_cancers[miR_cancers$cancer == cancer, ]

cancer_data$miRNA_act <- cancer_miR[match(row.names(cancer_data), cancer_miR$samples),]$miR_act

cancer_data <- cancer_data[rowSums(is.na(cancer_data)) == 0, ]

# Correlation
corr <- cor.test(as.numeric(as.vector(cancer_data$GOI_expr)), as.numeric(as.vector(cancer_data$miRNA_act)))

plot(as.numeric(as.vector(cancer_data$GOI_expr)), as.numeric(as.vector(cancer_data$miRNA_act)), pch = 16, cex=0.5, xlab=paste(RBP, " expression log2(RSEM)", sep=""), ylab="miR-19-3p activity")
abline(lm(as.numeric(as.vector(cancer_data$miRNA_act)) ~ as.numeric(as.vector(cancer_data$GOI_expr))), col = 4, lwd = 2)
text(13, -0.03, paste("r = ", round(corr$estimate, 3), sep=""))
```

Example scatter plot for DDX55-miR-29 in STAD

```{r, echo=F}
RBP <- "DDX55"
miRNA <- "miR-29abcd"
cancer="STAD"

# Filter all gene expression tables for the RBP of interest
logRSEM_gene <- lapply(logRSEM_data, function(x) unlist(x[RBP, ]))
logRSEM_gene <- unlist(logRSEM_gene)

# Make a vector indicating the cancer type of each sample
nb_samples_cancers <- unlist(lapply(logRSEM_data, function(x) ncol(x)))
cancers_vector <- rep(cancers, nb_samples_cancers)

# Merge into table containing all samples of all cancer types and indicating the cancer type
logRSEM_complete <- as.data.frame(cbind(logRSEM_gene, cancers_vector))
colnames(logRSEM_complete) <- c("GOI_expr", "cancer")

# Filter miRNA activity table for the miRNA of interest
miR_cancers <- as.data.frame(cbind(row.names(act), act[,miRNA], as.vector(act$cancers_vector)))
colnames(miR_cancers) <- c("samples", "miR_act", "cancer")

# Make a table with samples, GOI expression, and miRNA activity in that cancer type, removing normal samples 
cancer_data <- logRSEM_complete[logRSEM_complete$cancer== cancer, ]
row.names(cancer_data) <- gsub("-1$", "", row.names(cancer_data))
cancer_data <- cancer_data[grep("\\-01$", row.names(cancer_data)), ]
cancer_data$miRNA_act <- NA
cancer_miR <- miR_cancers[miR_cancers$cancer == cancer, ]

cancer_data$miRNA_act <- cancer_miR[match(row.names(cancer_data), cancer_miR$samples),]$miR_act

cancer_data <- cancer_data[rowSums(is.na(cancer_data)) == 0, ]

# Correlation
corr <- cor.test(as.numeric(as.vector(cancer_data$GOI_expr)), as.numeric(as.vector(cancer_data$miRNA_act)))

plot(as.numeric(as.vector(cancer_data$GOI_expr)), as.numeric(as.vector(cancer_data$miRNA_act)), pch = 16, cex=0.5, xlab=paste(RBP, " expression log2(RSEM)", sep=""), ylab="miR-19-3p activity")
abline(lm(as.numeric(as.vector(cancer_data$miRNA_act)) ~ as.numeric(as.vector(cancer_data$GOI_expr))), col = 4, lwd = 2)
text(10, 0.05, paste("r = ", round(corr$estimate, 3), sep=""))
```

Example bar plot for DDX55-miR-29

```{r, echo=F, out.width = "50%"}
RBP="DDX55"
miRNA="miR-29abcd"
miRNA_name="miR-29-3p"

# Filter all gene expression tables for the RBP of interest
logRSEM_gene <- lapply(logRSEM_data, function(x) unlist(x[RBP, ]))
logRSEM_gene <- unlist(logRSEM_gene)

# Make a vector indicating the cancer type of each sample
nb_samples_cancers <- unlist(lapply(logRSEM_data, function(x) ncol(x)))
cancers_vector <- rep(cancers, nb_samples_cancers)

# Merge into table containing all samples of all cancer types and indicating the cancer type
logRSEM_complete <- as.data.frame(cbind(logRSEM_gene, cancers_vector))
colnames(logRSEM_complete) <- c("GOI_expr", "cancer")

# Filter miRNA activity table for the miRNA of interest
miR_cancers <- as.data.frame(cbind(row.names(act), act[,miRNA], as.vector(act$cancers_vector)))
colnames(miR_cancers) <- c("samples", "miR_act", "cancer")

# Initialize a table to contain the correlation results
correlations <- as.data.frame(matrix(NA, nrow=length(cancers), ncol=2))
row.names(correlations) <- cancers
colnames(correlations) <- c("coeff", "pval")

# In each cancer, compute the pearson correlation coefficient and p-value
for (cancer in cancers)
{
  # Make a table with samples, GOI expression, and miRNA activity in that cancer type, removing normal samples 
  cancer_data <- logRSEM_complete[logRSEM_complete$cancer== cancer, ]
  row.names(cancer_data) <- gsub("-1$", "", row.names(cancer_data))
  cancer_data <- cancer_data[grep("\\-01$", row.names(cancer_data)), ]
  cancer_data$miRNA_act <- NA
  cancer_miR <- miR_cancers[miR_cancers$cancer == cancer, ]
  if (nrow(cancer_miR) == 0)
  {
    next
  }
  cancer_data$miRNA_act <- cancer_miR[match(row.names(cancer_data), cancer_miR$samples),]$miR_act
  
  cancer_data <- cancer_data[rowSums(is.na(cancer_data)) == 0, ]
  
  # Correlation
  corr <- cor.test(as.numeric(as.vector(cancer_data$GOI_expr)), as.numeric(as.vector(cancer_data$miRNA_act)))
  correlations[cancer, "coeff"] <- corr$estimate
  correlations[cancer, "pval"] <- corr$p.value
}

# Remove rows with  NAs
correlations <- correlations[!(is.na(correlations$coeff)),]

# Sort according to coefficients
correlations <- correlations[order(correlations$coeff),]

# Barplot
par(mar=c(5,5,5,5))
x <- barplot(correlations$coeff, 
        ylab = paste("Correlation of ", miRNA_name, " activity and ", RBP, "\n expression across tumors (Pearson)", sep=""),
        ylim = c(-0.45, 0.15),
        las=2,
        cex.names = 0.9,
        cex.axis = 0.7,
        cex.lab=0.8,
        col=ifelse(correlations$pval<0.05, "cornflowerblue","gray"))
legend(14,-0.3, legend = c("p < 0.05", "NS"), fill = c("cornflowerblue","gray"), cex=0.8)
text(x=x+0.3, y=-0.43, labels = row.names(correlations), srt=90, xpd=TRUE, cex=0.8, adj=c(1,0))
```

Example bar plots for the interactions between miR29 and PTBP1, RPLP0 or FAM120A

```{r, echo=F, out.width = "50%"}
RBP="PTBP1"
miRNA="miR-29abcd"
miRNA_name="miR-29-3p"

# Filter all gene expression tables for the RBP of interest
logRSEM_gene <- lapply(logRSEM_data, function(x) unlist(x[RBP, ]))
logRSEM_gene <- unlist(logRSEM_gene)

# Make a vector indicating the cancer type of each sample
nb_samples_cancers <- unlist(lapply(logRSEM_data, function(x) ncol(x)))
cancers_vector <- rep(cancers, nb_samples_cancers)

# Merge into table containing all samples of all cancer types and indicating the cancer type
logRSEM_complete <- as.data.frame(cbind(logRSEM_gene, cancers_vector))
colnames(logRSEM_complete) <- c("GOI_expr", "cancer")

# Filter miRNA activity table for the miRNA of interest
miR_cancers <- as.data.frame(cbind(row.names(act), act[,miRNA], as.vector(act$cancers_vector)))
colnames(miR_cancers) <- c("samples", "miR_act", "cancer")

# Initialize a table to contain the correlation results
correlations <- as.data.frame(matrix(NA, nrow=length(cancers), ncol=2))
row.names(correlations) <- cancers
colnames(correlations) <- c("coeff", "pval")

# In each cancer, compute the pearson correlation coefficient and p-value
for (cancer in cancers)
{
  # Make a table with samples, GOI expression, and miRNA activity in that cancer type, removing normal samples 
  cancer_data <- logRSEM_complete[logRSEM_complete$cancer== cancer, ]
  row.names(cancer_data) <- gsub("-1$", "", row.names(cancer_data))
  cancer_data <- cancer_data[grep("\\-01$", row.names(cancer_data)), ]
  cancer_data$miRNA_act <- NA
  cancer_miR <- miR_cancers[miR_cancers$cancer == cancer, ]
  if (nrow(cancer_miR) == 0)
  {
    next
  }
  cancer_data$miRNA_act <- cancer_miR[match(row.names(cancer_data), cancer_miR$samples),]$miR_act
  
  cancer_data <- cancer_data[rowSums(is.na(cancer_data)) == 0, ]
  
  # Correlation
  corr <- cor.test(as.numeric(as.vector(cancer_data$GOI_expr)), as.numeric(as.vector(cancer_data$miRNA_act)))
  correlations[cancer, "coeff"] <- corr$estimate
  correlations[cancer, "pval"] <- corr$p.value
}

# Remove rows with  NAs
correlations <- correlations[!(is.na(correlations$coeff)),]

# Sort according to coefficients
correlations <- correlations[order(correlations$coeff),]

# Barplot
par(mar=c(5,5,5,5))
x <- barplot(correlations$coeff, 
        ylab = paste("Correlation of ", miRNA_name, " activity and ", RBP, "\n expression across tumors (Pearson)", sep=""),
        ylim = c(-0.45, 0.15),
        las=2,
        cex.names = 0.9,
        cex.axis = 0.7,
        cex.lab=0.8,
        col=ifelse(correlations$pval<0.05, "cornflowerblue","gray"))
legend(14,-0.3, legend = c("p < 0.05", "NS"), fill = c("cornflowerblue","gray"), cex=0.8)
text(x=x+0.3, y=-0.43, labels = row.names(correlations), srt=90, xpd=TRUE, cex=0.8, adj=c(1,0))
```

```{r, echo=F, out.width = "50%"}
RBP="RPLP0"
miRNA="miR-29abcd"
miRNA_name="miR-29-3p"

# Filter all gene expression tables for the RBP of interest
logRSEM_gene <- lapply(logRSEM_data, function(x) unlist(x[RBP, ]))
logRSEM_gene <- unlist(logRSEM_gene)

# Make a vector indicating the cancer type of each sample
nb_samples_cancers <- unlist(lapply(logRSEM_data, function(x) ncol(x)))
cancers_vector <- rep(cancers, nb_samples_cancers)

# Merge into table containing all samples of all cancer types and indicating the cancer type
logRSEM_complete <- as.data.frame(cbind(logRSEM_gene, cancers_vector))
colnames(logRSEM_complete) <- c("GOI_expr", "cancer")

# Filter miRNA activity table for the miRNA of interest
miR_cancers <- as.data.frame(cbind(row.names(act), act[,miRNA], as.vector(act$cancers_vector)))
colnames(miR_cancers) <- c("samples", "miR_act", "cancer")

# Initialize a table to contain the correlation results
correlations <- as.data.frame(matrix(NA, nrow=length(cancers), ncol=2))
row.names(correlations) <- cancers
colnames(correlations) <- c("coeff", "pval")

# In each cancer, compute the pearson correlation coefficient and p-value
for (cancer in cancers)
{
  # Make a table with samples, GOI expression, and miRNA activity in that cancer type, removing normal samples 
  cancer_data <- logRSEM_complete[logRSEM_complete$cancer== cancer, ]
  row.names(cancer_data) <- gsub("-1$", "", row.names(cancer_data))
  cancer_data <- cancer_data[grep("\\-01$", row.names(cancer_data)), ]
  cancer_data$miRNA_act <- NA
  cancer_miR <- miR_cancers[miR_cancers$cancer == cancer, ]
  if (nrow(cancer_miR) == 0)
  {
    next
  }
  cancer_data$miRNA_act <- cancer_miR[match(row.names(cancer_data), cancer_miR$samples),]$miR_act
  
  cancer_data <- cancer_data[rowSums(is.na(cancer_data)) == 0, ]
  
  # Correlation
  corr <- cor.test(as.numeric(as.vector(cancer_data$GOI_expr)), as.numeric(as.vector(cancer_data$miRNA_act)))
  correlations[cancer, "coeff"] <- corr$estimate
  correlations[cancer, "pval"] <- corr$p.value
}

# Remove rows with  NAs
correlations <- correlations[!(is.na(correlations$coeff)),]

# Sort according to coefficients
correlations <- correlations[order(correlations$coeff),]

# Barplot
par(mar=c(5,5,5,5))
x <- barplot(correlations$coeff, 
        ylab = paste("Correlation of ", miRNA_name, " activity and ", RBP, "\n expression across tumors (Pearson)", sep=""),
        ylim = c(-0.45, 0.15),
        las=2,
        cex.names = 0.9,
        cex.axis = 0.7,
        cex.lab=0.8,
        col=ifelse(correlations$pval<0.05, "cornflowerblue","gray"))
legend(14,-0.3, legend = c("p < 0.05", "NS"), fill = c("cornflowerblue","gray"), cex=0.8)
text(x=x+0.3, y=-0.43, labels = row.names(correlations), srt=90, xpd=TRUE, cex=0.8, adj=c(1,0))
```

```{r, echo=F, out.width = "60%"}
RBP="FAM120A"
miRNA="miR-29abcd"
miRNA_name="miR-29-3p"

# Filter all gene expression tables for the RBP of interest
logRSEM_gene <- lapply(logRSEM_data, function(x) unlist(x[RBP, ]))
logRSEM_gene <- unlist(logRSEM_gene)

# Make a vector indicating the cancer type of each sample
nb_samples_cancers <- unlist(lapply(logRSEM_data, function(x) ncol(x)))
cancers_vector <- rep(cancers, nb_samples_cancers)

# Merge into table containing all samples of all cancer types and indicating the cancer type
logRSEM_complete <- as.data.frame(cbind(logRSEM_gene, cancers_vector))
colnames(logRSEM_complete) <- c("GOI_expr", "cancer")

# Filter miRNA activity table for the miRNA of interest
miR_cancers <- as.data.frame(cbind(row.names(act), act[,miRNA], as.vector(act$cancers_vector)))
colnames(miR_cancers) <- c("samples", "miR_act", "cancer")

# Initialize a table to contain the correlation results
correlations <- as.data.frame(matrix(NA, nrow=length(cancers), ncol=2))
row.names(correlations) <- cancers
colnames(correlations) <- c("coeff", "pval")

# In each cancer, compute the pearson correlation coefficient and p-value
for (cancer in cancers)
{
  # Make a table with samples, GOI expression, and miRNA activity in that cancer type, removing normal samples 
  cancer_data <- logRSEM_complete[logRSEM_complete$cancer== cancer, ]
  row.names(cancer_data) <- gsub("-1$", "", row.names(cancer_data))
  cancer_data <- cancer_data[grep("\\-01$", row.names(cancer_data)), ]
  cancer_data$miRNA_act <- NA
  cancer_miR <- miR_cancers[miR_cancers$cancer == cancer, ]
  if (nrow(cancer_miR) == 0)
  {
    next
  }
  cancer_data$miRNA_act <- cancer_miR[match(row.names(cancer_data), cancer_miR$samples),]$miR_act
  
  cancer_data <- cancer_data[rowSums(is.na(cancer_data)) == 0, ]
  
  # Correlation
  corr <- cor.test(as.numeric(as.vector(cancer_data$GOI_expr)), as.numeric(as.vector(cancer_data$miRNA_act)))
  correlations[cancer, "coeff"] <- corr$estimate
  correlations[cancer, "pval"] <- corr$p.value
}

# Remove rows with  NAs
correlations <- correlations[!(is.na(correlations$coeff)),]

# Sort according to coefficients
correlations <- correlations[order(correlations$coeff),]

# Barplot
par(mar=c(5,5,10,5))
x <- barplot(correlations$coeff, 
        ylab = paste("Correlation of ", miRNA_name, " activity and ", RBP, "\n expression across tumors (Pearson)", sep=""),
        ylim = c(-0.45, 0.15),
        las=2,
        cex.names = 0.9,
        cex.axis = 0.7,
        cex.lab=0.8,
        col=ifelse(correlations$pval<0.05, "cornflowerblue","gray"))
legend(14,-0.3, legend = c("p < 0.05", "NS"), fill = c("cornflowerblue","gray"), cex=0.8)
text(x=x+0.3, y=-0.43, labels = row.names(correlations), srt=90, xpd=TRUE, cex=0.8, adj=c(1,0))
```

# Correction of associations for tumor purity

Initialize tables that will be filled with the correlation coefficients and p-values

```{r}
coeffs <- matrix(NA, nrow=nrow(interactions), ncol = length(cancers))
row.names(coeffs) <- paste(interactions$RBP, miRNAs, sep = "_")
colnames(coeffs) <- cancers

pvals <- coeffs
```

Essentially the same analysis as above, but with an added correction for tumor purity. See the comments within the code for descriptions of the steps.

```{r}
# For each candidate interaction
for (i in 1:nrow(interactions))
{
  RBP <- as.vector(interactions$RBP[i])
  
  logRSEM_gene <- lapply(logRSEM_data, function(x) unlist(x[RBP, ]))
  logRSEM_gene <- unlist(logRSEM_gene)
  
  if(sum(is.na(logRSEM_gene)) == length(logRSEM_gene) | length(logRSEM_gene) == 0)
  {
    next
  }

  # Make a vector indicating the cancer type of each sample
  nb_samples_cancers <- unlist(lapply(logRSEM_data, function(x) ncol(x)))
  cancers_vector <- rep(cancers, nb_samples_cancers)
  
  # Merge into table containing all samples of all cancer types and indicating the cancer type
  logRSEM_complete <- as.data.frame(cbind(logRSEM_gene, cancers_vector))
  colnames(logRSEM_complete) <- c("GOI_expr", "cancer")

  # Filter miRNA activity table for the miRNA of interest
  miRNA <- interactions$miRNA[i]
  miR_cancers <- as.data.frame(cbind(row.names(act), act[,miRNA], as.vector(act$cancers_vector)))
  colnames(miR_cancers) <- c("samples", "miR_act", "cancer")
  
  # For each cancer type, compute the partial pearson correlation coefficient and p-value
  for (cancer in cancers)
  {
    # Make table with sample names, RBP expression, tumor purity and miRNA activity for that cancer type, removing normal samples 
    cancer_data <- logRSEM_complete[logRSEM_complete$cancer== cancer, ]
    row.names(cancer_data) <- gsub("-1$", "", row.names(cancer_data))
    cancer_data <- cancer_data[grep("\\-01$", row.names(cancer_data)), ]
    cancer_data$miRNA_act <- NA
    cancer_miR <- miR_cancers[miR_cancers$cancer == cancer, ]
    if (nrow(cancer_miR) == 0)
    {
      next
    }
    cancer_data$miRNA_act <- cancer_miR[match(row.names(cancer_data), cancer_miR$samples),]$miR_act
    
    if (!(file.exists(paste("counts_samples_", cancer, ".txt", sep = ""))))
    {
      next
    }
    purity <- read.table(paste("counts_samples_", cancer, ".txt", sep = ""), header = T)
    purity <- purity[grep("-01A-", purity$SampleName), ]
    purity$SampleName <- substr(purity$SampleName, 1, 15)
    purity$SampleName <- gsub("\\.", "-", purity$SampleName)
    
    cancer_data$purity <- purity[match(row.names(cancer_data), purity$SampleName),]$Purity
    
    cancer_data <- cancer_data[rowSums(is.na(cancer_data)) == 0, ]
    
    # Partial correlation
    corr <- pcor.test(as.numeric(as.vector(cancer_data$GOI_expr)), as.numeric(as.vector(cancer_data$miRNA_act)), as.numeric(as.vector(cancer_data$purity)))
    coeffs[i, cancer] <- corr$estimate
    pvals[i, cancer] <- corr$p.value
  }
  # Add + or - sign to the interaction name to indicate the type of interaction
  if (interactions$sign[i] == "-")
  {
    row.names(coeffs)[i] <- paste("-", row.names(coeffs)[i], sep="  ")
  } else {
    row.names(coeffs)[i] <- paste("+", row.names(coeffs)[i], sep="  ")
  }
  row.names(pvals) <- row.names(coeffs)
}
```

Remove rows and columns with NAs

```{r}
rows_keep <- which(rowSums(is.na(coeffs)) != ncol(coeffs))
cols_keep <- which(colSums(is.na(coeffs)) != nrow(coeffs))
coeffs <- coeffs[rows_keep, cols_keep]
pvals <- pvals[rows_keep, cols_keep]
```

FDR correction

```{r}
pvalues_vector <- c(pvals)
pvalues_vector_FDR <- p.adjust(pvalues_vector, method="fdr")
FDR <- matrix(pvalues_vector_FDR, nrow = nrow(pvals), ncol = ncol(pvals))
row.names(FDR) <- row.names(pvals)
colnames(FDR) <- colnames(pvals)
```

Compute -log10(p-values)

```{r}
signedlog10pvals <- abs(log10(pvals)) 
```

Get data after clustering

```{r}
htmp <- heatmap.2(coeffs, hclust=function(x) hclust(x,method="complete"), distfun = function(x) dist(x,method = 'euclidean'))
coeffs_clust <- coeffs[rev(htmp$rowInd), htmp$colInd]
signedlog10pvals_clust <- signedlog10pvals[row.names(coeffs_clust), colnames(coeffs_clust)]
FDR_clust <- FDR[row.names(coeffs_clust), colnames(coeffs_clust)]
```

Prepare the data for making the heatmap

```{r}
signedLogPval_srt <- data.table(cbind(row.names(signedlog10pvals_clust), signedlog10pvals_clust))
colnames(signedLogPval_srt)[1] <- "Interactions"
signedLogPval_melt <- melt(signedLogPval_srt, id.vars = "Interactions")
names(signedLogPval_melt)[2:3] <- c("Cancers", "signedLog10pval")

coeffs_srt <- data.table(cbind(row.names(coeffs_clust), coeffs_clust))
colnames(coeffs_srt)[1] <- "Interactions"
coeffs_melt <- melt(coeffs_srt, id.vars = "Interactions")
names(coeffs_melt)[2:3] <- c("Cancers", "coeffs")

FDR_srt <- data.table(cbind(row.names(FDR_clust), FDR_clust))
colnames(FDR_srt)[1] <- "Interactions"
FDR_melt <- melt(FDR_srt, id.vars = "Interactions")
names(FDR_melt)[2:3] <- c("Cancers", "FDR")

dat_htmp <- cbind(signedLogPval_melt, coeffs_melt$coeffs, FDR_melt$FDR)
colnames(dat_htmp)[3:5] <- c("-log10(p-value)", "coeffs", "padj")
dat_htmp[, "-log10(p-value)"] <- as.numeric(as.vector(unlist(dat_htmp[, "-log10(p-value)"])))
dat_htmp$coeffs <- as.numeric(as.vector(dat_htmp$coeffs))
dat_htmp$padj <- as.numeric(as.vector(dat_htmp$padj))

dat_htmp$FDR <- ifelse(dat_htmp$padj<0.05, 1, 0)
```

Set the limits for effect size and -log10(p-valuse) and squish the saturated values

```{r}
pval_lim <- 5
dat_htmp$`-log10(p-value)`[abs(dat_htmp$`-log10(p-value)`) > pval_lim] <- pval_lim
coeff_lim <- 0.3
dat_htmp$coeffs[abs(dat_htmp$coeffs) > coeff_lim] <- coeff_lim*sign(dat_htmp$coeffs[abs(dat_htmp$coeffs) > coeff_lim])
```

Make a color vector to indicate the type of interaction

```{r}
colors_sign <- rev(as.character(signedLogPval_srt$Interactions))
colors_sign <- gsub(" .*", "", colors_sign)
colors_sign[grep("-", colors_sign)] <- "blue"
colors_sign[grep("\\+", colors_sign)] <- "red"
```

Improve the labels

```{r}
dat_htmp$Interactions <- gsub(".* ", "", dat_htmp$Interactions)
signedLogPval_srt$Interactions <- gsub(".* ", "", signedLogPval_srt$Interactions)
row.names(coeffs_clust) <- gsub(".* ", "", row.names(coeffs_clust) )

lbls <- row.names(coeffs_clust)
lbls <- gsub("miR-33ab$"," - miR-33-5p", lbls)
lbls <- gsub("miR-26ab$"," - miR-26-5p", lbls)
lbls <- gsub("miR-29abcd$"," - miR-29-3p", lbls)
lbls <- gsub("miR-27abc$"," - miR-27-3p", lbls)
lbls <- gsub("miR-133abc$"," - miR-133-3p", lbls)
lbls <- gsub("miR-199ab-5p$"," - miR-199-5p", lbls)
lbls <- gsub("miR-376abd$"," - miR-376-3p", lbls)
lbls <- gsub("miR-23abc$"," - miR-23-3p", lbls)
lbls <- gsub("miR-19ab$"," - miR-19-3p", lbls)
lbls <- gsub("miR-182$"," - miR-182-5p", lbls)
lbls <- gsub("miR-758$"," - miR-758-3p", lbls)
lbls <- gsub("miR-22$"," - miR-22-3p", lbls)
lbls <- gsub("miR-150$"," - miR-150-5p", lbls)
lbls <- gsub("miR-124$"," - miR-124-3p", lbls)
lbls <- gsub("miR-96$"," - miR-96-5p", lbls)
lbls <- gsub("miR-370$"," - miR-370-3p", lbls)
lbls <- gsub("miR-17$"," - miR-17-5p", lbls)
lbls <- gsub("miR-218$"," - miR-218-5p", lbls)
lbls <- gsub("miR-21$"," - miR-21-5p", lbls)
lbls <- gsub("miR-25$"," - miR-25-3p", lbls)
lbls <- gsub("_ -"," -", lbls)
lbls <- gsub("_"," - ", lbls)
```

Plot the heatmap

```{r}
pdf(paste("heatmap_TCGA_intsConfENCODEtotalRNAseq_corrRBPexprMirAct_signedLog10PvaluesANDcoeffs_purityCorrected2.pdf", sep=""), width=9.5, height=7)
ggplot(dat_htmp) +
geom_point(aes(
  x = Cancers, 
  size = `-log10(p-value)`, 
  y = factor(Interactions, levels = rev(as.character(signedLogPval_srt$Interactions))),
  shape = 21, 
  stroke = 1,
  fill=coeffs, 
  color=FDR)) +
scale_shape_identity() +
scale_y_discrete(label = rev(lbls)) +
theme(axis.text.y = element_text(color=colors_sign)) +
scale_fill_gradient2(limits=c(-coeff_lim,coeff_lim), low="blue", mid="white", high="red") +
scale_colour_gradient( limits=c(0,1), low="white", high="black") +
scale_size("-log10(p-value)", range = c(0.1, 10), breaks = seq(0, pval_lim, 1), limits=c(0, pval_lim))+
guides(size = guide_legend(override.aes = list(fill = "red")))+
labs(x="",y="") +
theme(legend.key.size = unit(1.5, 'lines'), text = element_text(size=12), axis.text = element_text(size=12), axis.text.x = element_text(angle=90, size=12, vjust=0.5, hjust=1), panel.background = element_rect(fill = 'white', colour = 'white')) +
theme(plot.margin = unit(c(1,1,1,1), "cm")) 
dev.off()
```

```{r}
sessionInfo()
```