---
title: 'Associations between RBP expression and miRNA activity'
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
```

# Load libraries

```{r}
library(tidyverse)
library(stringr)
library(scater)
library(SingleCellExperiment)
library(scales)
library(data.table)
# library(ppcor)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
```

# Load the TCGA gene expression data (log2RSEM) obtained from Firebrowse

```{r}
logRSEM_data <- list.files(here("TCGA", "data"), pattern = "log2RSEM.csv", full.names = TRUE) |>
  lapply(fread) |>
  lapply(as.data.frame) |>
  lapply(function(x) {
    rownames(x) <- x[[1]]
    x[-1]
  })

head(logRSEM_data[[1]][, 1:5])
```

# Define the genes of interest for the t-SNE plots

```{r}
gene_names_tSNE <- c("DDX55", "PTBP1", "SUPV3L1")
```

# Filter gene expression tables for the genes of interest and for tumor samples, and normalize by subtracting the average of normal samples

```{r}
log2FC_gois <- lapply(logRSEM_data, function(x) {
  # Skip datasets without normal samples (ending in -11)
  if (!any(grepl("-11$", colnames(x)))) {
    return(NA)
  }
  
  # Filter to genes of interest
  x <- x[rownames(x) %in% gene_names_tSNE, ]
  
  # Subtract mean of normal samples from all samples
  normal_cols <- grepl("-11$", colnames(x))
  row_means_normal <- rowMeans(x[, normal_cols])
  x <- sweep(x, 1, row_means_normal, "-")
  
  # Transpose and keep only tumor samples (ending in -01)
  x <- t(x)
  as.data.frame(x[grepl("-01$", rownames(x)), ])
})

head(log2FC_gois[[2]])
```

# Extract cancer types and combine data

```{r}
# Extract cancer type from file paths
logRSEM_files <- list.files(here("TCGA", "data"), pattern = "log2RSEM.csv", full.names = TRUE)
cancer_types <- str_extract(logRSEM_files, "[A-Z]+(?=_log2RSEM)")

# Combine data from all cancers into a long-format data frame
log2FC_gois_all <- log2FC_gois |>
  imap_dfr(
    ~ if (is.data.frame(.x)) {
      tibble::rownames_to_column(.x, "sample") |>
        mutate(cancer = cancer_types[.y], .before = 1)
    }
  ) |>
  mutate(sample = str_replace_all(sample, "-", "."))
``` 

# Load the miRNA activity data and filter for tumors

```{r}
miR_act <- read.csv(here("TCGA", "output", "supplementary_tables", "TCGA_miRNA_activity.csv"))
miR_act <- miR_act[miR_act$sample_type == "tumor", ]
```

# Visualization of pan-cancer RBP expression using t-SNE plots

The t-SNE embedding of pan-cancer miRNA activity was already computed and
plotted for miR-17-5p, miR-29-3p, and miR-19-3p in
`09_TCGA_miRNA_activity.Rmd`, which caches the resulting
`SingleCellExperiment` to `TCGA/output/rds/sce_tSNE.rds`. Here we reuse that same embedding (`reducedDim(sce_tSNE, "TSNE")`) and colour it by RBP
expression instead of recomputing the t-SNE.

```{r}
sce_tSNE <- readRDS(here("TCGA", "output", "rds", "sce_tSNE.rds"))
sce_tSNE
```

Merge selected RBP expression log2FC values into the t-SNE object

```{r}
rbp_expr <- log2FC_gois_all |>
  select(sample, all_of(gene_names_tSNE))

coldata_tsne <- as.data.frame(colData(sce_tSNE)) |>
  left_join(rbp_expr, by = "sample")

colData(sce_tSNE)[gene_names_tSNE] <- coldata_tsne[gene_names_tSNE]
```

Theme function for t-SNE plots

```{r}
custom_theme_plot <- function(size_text = 15,
                               leg_pos = "right",
                               size_text_titles = NULL,
                               classic = TRUE,
                               line_size = 0.85,
                               line_size_color = "black",
                               axis_ticks_length = 0.25,
                               axis_ticks_color = "black",
                               size_legend_text = 12,
                               size_legend_title = 14) {
  require(ggplot2)
  if (is.null(size_text_titles)) {
    size_text_titles <- size_text + 2
  }
  gg.plots <- if (classic) theme_classic() else theme_bw()
  gg.plots +
    theme(
      axis.text.x = element_text(size = size_text, angle = 45, hjust = 1, color = line_size_color),
      axis.line = element_line(size = line_size, colour = line_size_color),
      axis.ticks = element_line(colour = axis_ticks_color, size = line_size),
      axis.ticks.length = unit(axis_ticks_length, "cm"),
      axis.title.x = element_text(size = size_text_titles),
      legend.position = leg_pos,
      plot.title = element_text(hjust = 0.5, size = size_text),
      axis.text.y = element_text(size = size_text, color = "black"),
      axis.title.y = element_text(size = size_text_titles),
      legend.text = element_text(size = size_legend_text),
      legend.title = element_text(size = size_legend_title),
      legend.key.size = unit(0.8, "cm")
    )
}
```

Helper function to create a t-SNE plot coloured by RBP expression

```{r}
rbp_tsne_plot <- function(gene, sce_obj, limits = 1) {
  plotReducedDim(sce_obj,
                 ncomponents = 2,
                 dimred = "TSNE",
                 point_size = 2,
                 colour_by = gene) +
    custom_theme_plot(size_text = 27, size_text_titles = 30, size_legend_text = 21, size_legend_title = 24) +
    scale_colour_gradient2(
      low = "blue", high = "red", mid = "white",
      midpoint = 0, limits = c(-limits, limits),
      oob = scales::squish,
      name = paste0(gene, " expression\nlog2FC(RSEM T vs avg N)")
    ) +
    theme(legend.key.size = unit(1.2, "cm"))
}
```

Plot t-SNE for each RBP, coloured by expression log2FC(tumor vs average normal)

```{r, fig.dim = c(12,10), fig.align = 'center'}
rbp_lims <- 1

rbp_pdf_names <- setNames(
  paste0("TCGA_", gene_names_tSNE, "_expr_tSNE.pdf"),
  gene_names_tSNE
)

for (gene in gene_names_tSNE) {
  print(rbp_tsne_plot(gene, sce_tSNE, limits = rbp_lims))

  pdf_file <- here("TCGA", "output", "figures", rbp_pdf_names[[gene]])
  pdf(pdf_file, height = 10, width = 12.5)
  print(rbp_tsne_plot(gene, sce_tSNE, limits = rbp_lims))
  dev.off()

  cat("Saved:", rbp_pdf_names[[gene]], "\n")
}
```

# Correlations between miRNA activity and RBP expression for all candidate RBP-miRNA interactions

Using the RBP-miRNA interaction list from the intersection of candidates identified with the RBP KD experiments from ENCODE and validated using the tissue and cell line samples from ENCODE. 

```{r}
# Load interactions
interactions <- readRDS(here("ENCODE_tissues_cell_lines", "output", "rds", "ENCODE_tissues_cell_lines_high_conf_ints.rds"))

# Add cancer names to expression data list
names(logRSEM_data) <- cancer_types

# Reformat sample names in expression data
logRSEM_data <- lapply(logRSEM_data, function(x) {
  colnames(x) <- str_replace_all(colnames(x), "-", ".")
  x
})
```

```{r}
# Helper function to compute the correlation between the expression of a given RBP and the activity of a given miRNA, within a given cancer type
compute_correlation_cancer <- function(rbp, mirna, cancer) {
  expr_mat <- logRSEM_data[[cancer]]

  # Skip if the RBP is not present in this cancer's expression data
  if (!(rbp %in% rownames(expr_mat))) {
    return(NULL)
  }

  rbp_expr <- as.numeric(expr_mat[rbp, ])
  names(rbp_expr) <- colnames(expr_mat)

  mirna_data <- miR_act |>
    filter(cancer_type == cancer, miRNA == mirna)

  if (nrow(mirna_data) == 0) {
    return(NULL)
  }

  # Match samples between expression and miRNA activity data
  common_samples <- intersect(names(rbp_expr), mirna_data$sample_name)
  if (length(common_samples) == 0) {
    return(NULL)
  }

  rbp_matched <- rbp_expr[common_samples]
  mirna_matched <- mirna_data$coefficient_log2[match(common_samples, mirna_data$sample_name)]

  corr <- cor.test(rbp_matched, mirna_matched)
  tibble(r = corr$estimate, pval = corr$p.value)
}
```

```{r}
# Compute the correlations for every RBP-miRNA interaction, in every cancer type
corr_table <- interactions |>
  select(RBP, miRNA, sign) |>
  # Cross every interaction with every cancer type
  tidyr::crossing(cancer = cancer_types) |>
  mutate(
    interaction = paste(RBP, miRNA, sep = " - "),
    corr_data = pmap(list(RBP, miRNA, cancer), compute_correlation_cancer)
  ) |>
  unnest(corr_data, keep_empty = TRUE) |>
  filter(!is.na(r)) |>
  mutate(
    padj = p.adjust(pval, method = "fdr"),
    signedlog10pvals = abs(log10(pval))
  ) |>
  select(interaction, RBP, miRNA, sign, cancer, r, pval, padj, signedlog10pvals) |>
  as.data.frame()

head(corr_table)
nrow(corr_table)
```

Cluster RBP-miRNA interactions and cancer types by their correlation coefficients

```{r}
r_mat <- corr_table |>
  select(interaction, cancer, r) |>
  pivot_wider(names_from = cancer, values_from = r) |>
  column_to_rownames("interaction") |>
  as.matrix()

row_clust <- hclust(dist(r_mat, method = "euclidean"), method = "complete")
col_clust <- hclust(dist(t(r_mat), method = "euclidean"), method = "complete")

interaction_levels <- rownames(r_mat)[row_clust$order]
cancer_levels <- colnames(r_mat)[col_clust$order]
```

Make the data table for the heatmap, clustered and converting saturated values

```{r}
pval_lim <- 5
coeff_lim <- 0.3

dat_htmp <- corr_table |>
  transmute(
    Interactions = factor(interaction, levels = interaction_levels),
    Cancers = factor(cancer, levels = cancer_levels),
    `-log10(p-value)` = pmin(signedlog10pvals, pval_lim),
    `Pearson r` = pmax(pmin(r, coeff_lim), -coeff_lim),
    FDR = as.integer(padj < 0.05)
  )
```

Assign interaction labels colors according to each interaction's expected sign

```{r}
sign_by_interaction <- corr_table |>
  distinct(interaction, sign) |>
  deframe()

colors_sign <- ifelse(sign_by_interaction[interaction_levels] == "-", "blue", "red")
```

Draw the dotplot

```{r, fig.dim = c(10.2, 7), fig.align = 'center'}
dotplot <- ggplot(dat_htmp, aes(x = Cancers, y = Interactions)) +
  geom_point(aes(size = `-log10(p-value)`, fill = `Pearson r`, color = FDR),
             shape = 21, stroke = 1) +
  scale_fill_gradient2(limits = c(-coeff_lim, coeff_lim), low = "blue", mid = "white", high = "red") +
  scale_colour_gradient(limits = c(0, 1), low = "white", high = "black") +
  scale_size("-log10(p-value)", range = c(0.1, 10), breaks = seq(0, pval_lim, 1), limits = c(0, pval_lim)) +
  guides(size = guide_legend(override.aes = list(fill = "black"))) +
  labs(x = "", y = "") +
  theme(
    axis.text.y = element_text(color = colors_sign, size = 17),
    legend.key.size = unit(1.5, "lines"),
    text = element_text(size = 16),
    axis.text = element_text(size = 17),
    axis.text.x = element_text(angle = 90, size = 17, vjust = 0.5, hjust = 1),
    axis.title = element_text(size = 18),
    legend.title = element_text(size = 17),
    legend.text = element_text(size = 15),
    panel.background = element_rect(fill = "white", colour = "white"),
    plot.margin = unit(c(1, 1, 1, 1), "cm")
  )

dotplot

pdf(here("TCGA", "output", "figures", "TCGA_high_conf_ints_corrRBPexprMirAct_dotplot.pdf"), width = 11, height = 8)
print(dotplot)
dev.off()
```

# Make a list of interactions for which the direction of correlation is in agreement with the predicted type of interaction (activating / inhibitory) in at least 70% of TCGA cancers

```{r,}
<== based on corr_table

toKeep <- rep(NA, nrow(coeffs))
for (i in 1:nrow(coeffs))
{
  int_sign <- gsub(" .*", "", row.names(coeffs)[i])
  if(int_sign == "-")
  {
    int_sign <- "-1"
  } else {
    int_sign <- "1"
  }
  toKeep[i] <- sum(sign(coeffs[i,]) == int_sign)/ncol(coeffs) >= 0.7
}

row.names(coeffs)[toKeep]
```

Filter the previous table of high-confidence interactions to retain these, and save

```{r}
new_ints <- interactions[paste(interactions$RBP, miRNAs, sep = "_") %in% gsub(".* ", "", row.names(signedlog10pvals)[toKeep]), ]
write.csv(new_ints, "RBPKD_miRNAs_newSigInts_corrShRNA6nts7ntsSigPosFishers_intersectRBPKDtotalRNAseqANDtcgaCorrs70percent.csv")

```


Example scatter plot for SUPV3L1-miR-19 in BRCA

```{r, echo=F}
RBP <- "SUPV3L1"
miRNA <- "miR-19ab"
cancer="BRCA"

# Filter all gene expression tables for the RBP of interest
logRSEM_gene <- lapply(logRSEM_data, function(x) unlist(x[RBP, ]))
logRSEM_gene <- unlist(logRSEM_gene)

# Make a vector indicating the cancer type of each sample
nb_samples_cancers <- unlist(lapply(logRSEM_data, function(x) ncol(x)))
cancers_vector <- rep(cancers, nb_samples_cancers)

# Merge into table containing all samples of all cancer types and indicating the cancer type
logRSEM_complete <- as.data.frame(cbind(logRSEM_gene, cancers_vector))
colnames(logRSEM_complete) <- c("GOI_expr", "cancer")

# Filter miRNA activity table for the miRNA of interest
miR_cancers <- as.data.frame(cbind(row.names(act), act[,miRNA], as.vector(act$cancers_vector)))
colnames(miR_cancers) <- c("samples", "miR_act", "cancer")

# Make a table with samples, GOI expression, and miRNA activity in that cancer type, removing normal samples 
cancer_data <- logRSEM_complete[logRSEM_complete$cancer== cancer, ]
row.names(cancer_data) <- gsub("-1$", "", row.names(cancer_data))
cancer_data <- cancer_data[grep("\\-01$", row.names(cancer_data)), ]
cancer_data$miRNA_act <- NA
cancer_miR <- miR_cancers[miR_cancers$cancer == cancer, ]

cancer_data$miRNA_act <- cancer_miR[match(row.names(cancer_data), cancer_miR$samples),]$miR_act

cancer_data <- cancer_data[rowSums(is.na(cancer_data)) == 0, ]

# Correlation
corr <- cor.test(as.numeric(as.vector(cancer_data$GOI_expr)), as.numeric(as.vector(cancer_data$miRNA_act)))

plot(as.numeric(as.vector(cancer_data$GOI_expr)), as.numeric(as.vector(cancer_data$miRNA_act)), pch = 16, cex=0.5, xlab=paste(RBP, " expression log2(RSEM)", sep=""), ylab="miR-19-3p activity")
abline(lm(as.numeric(as.vector(cancer_data$miRNA_act)) ~ as.numeric(as.vector(cancer_data$GOI_expr))), col = 4, lwd = 2)
text(13, -0.03, paste("r = ", round(corr$estimate, 3), sep=""))
```

Example scatter plot for DDX55-miR-29 in STAD

```{r, echo=F}
RBP <- "DDX55"
miRNA <- "miR-29abcd"
cancer="STAD"

# Filter all gene expression tables for the RBP of interest
logRSEM_gene <- lapply(logRSEM_data, function(x) unlist(x[RBP, ]))
logRSEM_gene <- unlist(logRSEM_gene)

# Make a vector indicating the cancer type of each sample
nb_samples_cancers <- unlist(lapply(logRSEM_data, function(x) ncol(x)))
cancers_vector <- rep(cancers, nb_samples_cancers)

# Merge into table containing all samples of all cancer types and indicating the cancer type
logRSEM_complete <- as.data.frame(cbind(logRSEM_gene, cancers_vector))
colnames(logRSEM_complete) <- c("GOI_expr", "cancer")

# Filter miRNA activity table for the miRNA of interest
miR_cancers <- as.data.frame(cbind(row.names(act), act[,miRNA], as.vector(act$cancers_vector)))
colnames(miR_cancers) <- c("samples", "miR_act", "cancer")

# Make a table with samples, GOI expression, and miRNA activity in that cancer type, removing normal samples 
cancer_data <- logRSEM_complete[logRSEM_complete$cancer== cancer, ]
row.names(cancer_data) <- gsub("-1$", "", row.names(cancer_data))
cancer_data <- cancer_data[grep("\\-01$", row.names(cancer_data)), ]
cancer_data$miRNA_act <- NA
cancer_miR <- miR_cancers[miR_cancers$cancer == cancer, ]

cancer_data$miRNA_act <- cancer_miR[match(row.names(cancer_data), cancer_miR$samples),]$miR_act

cancer_data <- cancer_data[rowSums(is.na(cancer_data)) == 0, ]

# Correlation
corr <- cor.test(as.numeric(as.vector(cancer_data$GOI_expr)), as.numeric(as.vector(cancer_data$miRNA_act)))

plot(as.numeric(as.vector(cancer_data$GOI_expr)), as.numeric(as.vector(cancer_data$miRNA_act)), pch = 16, cex=0.5, xlab=paste(RBP, " expression log2(RSEM)", sep=""), ylab="miR-19-3p activity")
abline(lm(as.numeric(as.vector(cancer_data$miRNA_act)) ~ as.numeric(as.vector(cancer_data$GOI_expr))), col = 4, lwd = 2)
text(10, 0.05, paste("r = ", round(corr$estimate, 3), sep=""))
```

Example bar plot for DDX55-miR-29

```{r, echo=F, out.width = "50%"}
RBP="DDX55"
miRNA="miR-29abcd"
miRNA_name="miR-29-3p"

# Filter all gene expression tables for the RBP of interest
logRSEM_gene <- lapply(logRSEM_data, function(x) unlist(x[RBP, ]))
logRSEM_gene <- unlist(logRSEM_gene)

# Make a vector indicating the cancer type of each sample
nb_samples_cancers <- unlist(lapply(logRSEM_data, function(x) ncol(x)))
cancers_vector <- rep(cancers, nb_samples_cancers)

# Merge into table containing all samples of all cancer types and indicating the cancer type
logRSEM_complete <- as.data.frame(cbind(logRSEM_gene, cancers_vector))
colnames(logRSEM_complete) <- c("GOI_expr", "cancer")

# Filter miRNA activity table for the miRNA of interest
miR_cancers <- as.data.frame(cbind(row.names(act), act[,miRNA], as.vector(act$cancers_vector)))
colnames(miR_cancers) <- c("samples", "miR_act", "cancer")

# Initialize a table to contain the correlation results
correlations <- as.data.frame(matrix(NA, nrow=length(cancers), ncol=2))
row.names(correlations) <- cancers
colnames(correlations) <- c("coeff", "pval")

# In each cancer, compute the pearson correlation coefficient and p-value
for (cancer in cancers)
{
  # Make a table with samples, GOI expression, and miRNA activity in that cancer type, removing normal samples 
  cancer_data <- logRSEM_complete[logRSEM_complete$cancer== cancer, ]
  row.names(cancer_data) <- gsub("-1$", "", row.names(cancer_data))
  cancer_data <- cancer_data[grep("\\-01$", row.names(cancer_data)), ]
  cancer_data$miRNA_act <- NA
  cancer_miR <- miR_cancers[miR_cancers$cancer == cancer, ]
  if (nrow(cancer_miR) == 0)
  {
    next
  }
  cancer_data$miRNA_act <- cancer_miR[match(row.names(cancer_data), cancer_miR$samples),]$miR_act
  
  cancer_data <- cancer_data[rowSums(is.na(cancer_data)) == 0, ]
  
  # Correlation
  corr <- cor.test(as.numeric(as.vector(cancer_data$GOI_expr)), as.numeric(as.vector(cancer_data$miRNA_act)))
  correlations[cancer, "coeff"] <- corr$estimate
  correlations[cancer, "pval"] <- corr$p.value
}

# Remove rows with  NAs
correlations <- correlations[!(is.na(correlations$coeff)),]

# Sort according to coefficients
correlations <- correlations[order(correlations$coeff),]

# Barplot
par(mar=c(5,5,5,5))
x <- barplot(correlations$coeff, 
        ylab = paste("Correlation of ", miRNA_name, " activity and ", RBP, "\n expression across tumors (Pearson)", sep=""),
        ylim = c(-0.45, 0.15),
        las=2,
        cex.names = 0.9,
        cex.axis = 0.7,
        cex.lab=0.8,
        col=ifelse(correlations$pval<0.05, "cornflowerblue","gray"))
legend(14,-0.3, legend = c("p < 0.05", "NS"), fill = c("cornflowerblue","gray"), cex=0.8)
text(x=x+0.3, y=-0.43, labels = row.names(correlations), srt=90, xpd=TRUE, cex=0.8, adj=c(1,0))
```

Example bar plots for the interactions between miR29 and PTBP1, RPLP0 or FAM120A

```{r, echo=F, out.width = "50%"}
RBP="PTBP1"
miRNA="miR-29abcd"
miRNA_name="miR-29-3p"

# Filter all gene expression tables for the RBP of interest
logRSEM_gene <- lapply(logRSEM_data, function(x) unlist(x[RBP, ]))
logRSEM_gene <- unlist(logRSEM_gene)

# Make a vector indicating the cancer type of each sample
nb_samples_cancers <- unlist(lapply(logRSEM_data, function(x) ncol(x)))
cancers_vector <- rep(cancers, nb_samples_cancers)

# Merge into table containing all samples of all cancer types and indicating the cancer type
logRSEM_complete <- as.data.frame(cbind(logRSEM_gene, cancers_vector))
colnames(logRSEM_complete) <- c("GOI_expr", "cancer")

# Filter miRNA activity table for the miRNA of interest
miR_cancers <- as.data.frame(cbind(row.names(act), act[,miRNA], as.vector(act$cancers_vector)))
colnames(miR_cancers) <- c("samples", "miR_act", "cancer")

# Initialize a table to contain the correlation results
correlations <- as.data.frame(matrix(NA, nrow=length(cancers), ncol=2))
row.names(correlations) <- cancers
colnames(correlations) <- c("coeff", "pval")

# In each cancer, compute the pearson correlation coefficient and p-value
for (cancer in cancers)
{
  # Make a table with samples, GOI expression, and miRNA activity in that cancer type, removing normal samples 
  cancer_data <- logRSEM_complete[logRSEM_complete$cancer== cancer, ]
  row.names(cancer_data) <- gsub("-1$", "", row.names(cancer_data))
  cancer_data <- cancer_data[grep("\\-01$", row.names(cancer_data)), ]
  cancer_data$miRNA_act <- NA
  cancer_miR <- miR_cancers[miR_cancers$cancer == cancer, ]
  if (nrow(cancer_miR) == 0)
  {
    next
  }
  cancer_data$miRNA_act <- cancer_miR[match(row.names(cancer_data), cancer_miR$samples),]$miR_act
  
  cancer_data <- cancer_data[rowSums(is.na(cancer_data)) == 0, ]
  
  # Correlation
  corr <- cor.test(as.numeric(as.vector(cancer_data$GOI_expr)), as.numeric(as.vector(cancer_data$miRNA_act)))
  correlations[cancer, "coeff"] <- corr$estimate
  correlations[cancer, "pval"] <- corr$p.value
}

# Remove rows with  NAs
correlations <- correlations[!(is.na(correlations$coeff)),]

# Sort according to coefficients
correlations <- correlations[order(correlations$coeff),]

# Barplot
par(mar=c(5,5,5,5))
x <- barplot(correlations$coeff, 
        ylab = paste("Correlation of ", miRNA_name, " activity and ", RBP, "\n expression across tumors (Pearson)", sep=""),
        ylim = c(-0.45, 0.15),
        las=2,
        cex.names = 0.9,
        cex.axis = 0.7,
        cex.lab=0.8,
        col=ifelse(correlations$pval<0.05, "cornflowerblue","gray"))
legend(14,-0.3, legend = c("p < 0.05", "NS"), fill = c("cornflowerblue","gray"), cex=0.8)
text(x=x+0.3, y=-0.43, labels = row.names(correlations), srt=90, xpd=TRUE, cex=0.8, adj=c(1,0))
```

```{r, echo=F, out.width = "50%"}
RBP="RPLP0"
miRNA="miR-29abcd"
miRNA_name="miR-29-3p"

# Filter all gene expression tables for the RBP of interest
logRSEM_gene <- lapply(logRSEM_data, function(x) unlist(x[RBP, ]))
logRSEM_gene <- unlist(logRSEM_gene)

# Make a vector indicating the cancer type of each sample
nb_samples_cancers <- unlist(lapply(logRSEM_data, function(x) ncol(x)))
cancers_vector <- rep(cancers, nb_samples_cancers)

# Merge into table containing all samples of all cancer types and indicating the cancer type
logRSEM_complete <- as.data.frame(cbind(logRSEM_gene, cancers_vector))
colnames(logRSEM_complete) <- c("GOI_expr", "cancer")

# Filter miRNA activity table for the miRNA of interest
miR_cancers <- as.data.frame(cbind(row.names(act), act[,miRNA], as.vector(act$cancers_vector)))
colnames(miR_cancers) <- c("samples", "miR_act", "cancer")

# Initialize a table to contain the correlation results
correlations <- as.data.frame(matrix(NA, nrow=length(cancers), ncol=2))
row.names(correlations) <- cancers
colnames(correlations) <- c("coeff", "pval")

# In each cancer, compute the pearson correlation coefficient and p-value
for (cancer in cancers)
{
  # Make a table with samples, GOI expression, and miRNA activity in that cancer type, removing normal samples 
  cancer_data <- logRSEM_complete[logRSEM_complete$cancer== cancer, ]
  row.names(cancer_data) <- gsub("-1$", "", row.names(cancer_data))
  cancer_data <- cancer_data[grep("\\-01$", row.names(cancer_data)), ]
  cancer_data$miRNA_act <- NA
  cancer_miR <- miR_cancers[miR_cancers$cancer == cancer, ]
  if (nrow(cancer_miR) == 0)
  {
    next
  }
  cancer_data$miRNA_act <- cancer_miR[match(row.names(cancer_data), cancer_miR$samples),]$miR_act
  
  cancer_data <- cancer_data[rowSums(is.na(cancer_data)) == 0, ]
  
  # Correlation
  corr <- cor.test(as.numeric(as.vector(cancer_data$GOI_expr)), as.numeric(as.vector(cancer_data$miRNA_act)))
  correlations[cancer, "coeff"] <- corr$estimate
  correlations[cancer, "pval"] <- corr$p.value
}

# Remove rows with  NAs
correlations <- correlations[!(is.na(correlations$coeff)),]

# Sort according to coefficients
correlations <- correlations[order(correlations$coeff),]

# Barplot
par(mar=c(5,5,5,5))
x <- barplot(correlations$coeff, 
        ylab = paste("Correlation of ", miRNA_name, " activity and ", RBP, "\n expression across tumors (Pearson)", sep=""),
        ylim = c(-0.45, 0.15),
        las=2,
        cex.names = 0.9,
        cex.axis = 0.7,
        cex.lab=0.8,
        col=ifelse(correlations$pval<0.05, "cornflowerblue","gray"))
legend(14,-0.3, legend = c("p < 0.05", "NS"), fill = c("cornflowerblue","gray"), cex=0.8)
text(x=x+0.3, y=-0.43, labels = row.names(correlations), srt=90, xpd=TRUE, cex=0.8, adj=c(1,0))
```

```{r, echo=F, out.width = "60%"}
RBP="FAM120A"
miRNA="miR-29abcd"
miRNA_name="miR-29-3p"

# Filter all gene expression tables for the RBP of interest
logRSEM_gene <- lapply(logRSEM_data, function(x) unlist(x[RBP, ]))
logRSEM_gene <- unlist(logRSEM_gene)

# Make a vector indicating the cancer type of each sample
nb_samples_cancers <- unlist(lapply(logRSEM_data, function(x) ncol(x)))
cancers_vector <- rep(cancers, nb_samples_cancers)

# Merge into table containing all samples of all cancer types and indicating the cancer type
logRSEM_complete <- as.data.frame(cbind(logRSEM_gene, cancers_vector))
colnames(logRSEM_complete) <- c("GOI_expr", "cancer")

# Filter miRNA activity table for the miRNA of interest
miR_cancers <- as.data.frame(cbind(row.names(act), act[,miRNA], as.vector(act$cancers_vector)))
colnames(miR_cancers) <- c("samples", "miR_act", "cancer")

# Initialize a table to contain the correlation results
correlations <- as.data.frame(matrix(NA, nrow=length(cancers), ncol=2))
row.names(correlations) <- cancers
colnames(correlations) <- c("coeff", "pval")

# In each cancer, compute the pearson correlation coefficient and p-value
for (cancer in cancers)
{
  # Make a table with samples, GOI expression, and miRNA activity in that cancer type, removing normal samples 
  cancer_data <- logRSEM_complete[logRSEM_complete$cancer== cancer, ]
  row.names(cancer_data) <- gsub("-1$", "", row.names(cancer_data))
  cancer_data <- cancer_data[grep("\\-01$", row.names(cancer_data)), ]
  cancer_data$miRNA_act <- NA
  cancer_miR <- miR_cancers[miR_cancers$cancer == cancer, ]
  if (nrow(cancer_miR) == 0)
  {
    next
  }
  cancer_data$miRNA_act <- cancer_miR[match(row.names(cancer_data), cancer_miR$samples),]$miR_act
  
  cancer_data <- cancer_data[rowSums(is.na(cancer_data)) == 0, ]
  
  # Correlation
  corr <- cor.test(as.numeric(as.vector(cancer_data$GOI_expr)), as.numeric(as.vector(cancer_data$miRNA_act)))
  correlations[cancer, "coeff"] <- corr$estimate
  correlations[cancer, "pval"] <- corr$p.value
}

# Remove rows with  NAs
correlations <- correlations[!(is.na(correlations$coeff)),]

# Sort according to coefficients
correlations <- correlations[order(correlations$coeff),]

# Barplot
par(mar=c(5,5,10,5))
x <- barplot(correlations$coeff, 
        ylab = paste("Correlation of ", miRNA_name, " activity and ", RBP, "\n expression across tumors (Pearson)", sep=""),
        ylim = c(-0.45, 0.15),
        las=2,
        cex.names = 0.9,
        cex.axis = 0.7,
        cex.lab=0.8,
        col=ifelse(correlations$pval<0.05, "cornflowerblue","gray"))
legend(14,-0.3, legend = c("p < 0.05", "NS"), fill = c("cornflowerblue","gray"), cex=0.8)
text(x=x+0.3, y=-0.43, labels = row.names(correlations), srt=90, xpd=TRUE, cex=0.8, adj=c(1,0))
```

# Correction of associations for tumor purity

Initialize tables that will be filled with the correlation coefficients and p-values

```{r}
coeffs <- matrix(NA, nrow=nrow(interactions), ncol = length(cancers))
row.names(coeffs) <- paste(interactions$RBP, miRNAs, sep = "_")
colnames(coeffs) <- cancers

pvals <- coeffs
```

Essentially the same analysis as above, but with an added correction for tumor purity. See the comments within the code for descriptions of the steps.

```{r}
# For each candidate interaction
for (i in 1:nrow(interactions))
{
  RBP <- as.vector(interactions$RBP[i])
  
  logRSEM_gene <- lapply(logRSEM_data, function(x) unlist(x[RBP, ]))
  logRSEM_gene <- unlist(logRSEM_gene)
  
  if(sum(is.na(logRSEM_gene)) == length(logRSEM_gene) | length(logRSEM_gene) == 0)
  {
    next
  }

  # Make a vector indicating the cancer type of each sample
  nb_samples_cancers <- unlist(lapply(logRSEM_data, function(x) ncol(x)))
  cancers_vector <- rep(cancers, nb_samples_cancers)
  
  # Merge into table containing all samples of all cancer types and indicating the cancer type
  logRSEM_complete <- as.data.frame(cbind(logRSEM_gene, cancers_vector))
  colnames(logRSEM_complete) <- c("GOI_expr", "cancer")

  # Filter miRNA activity table for the miRNA of interest
  miRNA <- interactions$miRNA[i]
  miR_cancers <- as.data.frame(cbind(row.names(act), act[,miRNA], as.vector(act$cancers_vector)))
  colnames(miR_cancers) <- c("samples", "miR_act", "cancer")
  
  # For each cancer type, compute the partial pearson correlation coefficient and p-value
  for (cancer in cancers)
  {
    # Make table with sample names, RBP expression, tumor purity and miRNA activity for that cancer type, removing normal samples 
    cancer_data <- logRSEM_complete[logRSEM_complete$cancer== cancer, ]
    row.names(cancer_data) <- gsub("-1$", "", row.names(cancer_data))
    cancer_data <- cancer_data[grep("\\-01$", row.names(cancer_data)), ]
    cancer_data$miRNA_act <- NA
    cancer_miR <- miR_cancers[miR_cancers$cancer == cancer, ]
    if (nrow(cancer_miR) == 0)
    {
      next
    }
    cancer_data$miRNA_act <- cancer_miR[match(row.names(cancer_data), cancer_miR$samples),]$miR_act
    
    if (!(file.exists(paste("counts_samples_", cancer, ".txt", sep = ""))))
    {
      next
    }
    purity <- read.table(paste("counts_samples_", cancer, ".txt", sep = ""), header = T)
    purity <- purity[grep("-01A-", purity$SampleName), ]
    purity$SampleName <- substr(purity$SampleName, 1, 15)
    purity$SampleName <- gsub("\\.", "-", purity$SampleName)
    
    cancer_data$purity <- purity[match(row.names(cancer_data), purity$SampleName),]$Purity
    
    cancer_data <- cancer_data[rowSums(is.na(cancer_data)) == 0, ]
    
    # Partial correlation
    corr <- pcor.test(as.numeric(as.vector(cancer_data$GOI_expr)), as.numeric(as.vector(cancer_data$miRNA_act)), as.numeric(as.vector(cancer_data$purity)))
    coeffs[i, cancer] <- corr$estimate
    pvals[i, cancer] <- corr$p.value
  }
  # Add + or - sign to the interaction name to indicate the type of interaction
  if (interactions$sign[i] == "-")
  {
    row.names(coeffs)[i] <- paste("-", row.names(coeffs)[i], sep="  ")
  } else {
    row.names(coeffs)[i] <- paste("+", row.names(coeffs)[i], sep="  ")
  }
  row.names(pvals) <- row.names(coeffs)
}
```

Remove rows and columns with NAs

```{r}
rows_keep <- which(rowSums(is.na(coeffs)) != ncol(coeffs))
cols_keep <- which(colSums(is.na(coeffs)) != nrow(coeffs))
coeffs <- coeffs[rows_keep, cols_keep]
pvals <- pvals[rows_keep, cols_keep]
```

FDR correction

```{r}
pvalues_vector <- c(pvals)
pvalues_vector_FDR <- p.adjust(pvalues_vector, method="fdr")
FDR <- matrix(pvalues_vector_FDR, nrow = nrow(pvals), ncol = ncol(pvals))
row.names(FDR) <- row.names(pvals)
colnames(FDR) <- colnames(pvals)
```

Compute -log10(p-values)

```{r}
signedlog10pvals <- abs(log10(pvals)) 
```

Get data after clustering

```{r}
htmp <- heatmap.2(coeffs, hclust=function(x) hclust(x,method="complete"), distfun = function(x) dist(x,method = 'euclidean'))
coeffs_clust <- coeffs[rev(htmp$rowInd), htmp$colInd]
signedlog10pvals_clust <- signedlog10pvals[row.names(coeffs_clust), colnames(coeffs_clust)]
FDR_clust <- FDR[row.names(coeffs_clust), colnames(coeffs_clust)]
```

Prepare the data for making the heatmap

```{r}
signedLogPval_srt <- data.table(cbind(row.names(signedlog10pvals_clust), signedlog10pvals_clust))
colnames(signedLogPval_srt)[1] <- "Interactions"
signedLogPval_melt <- melt(signedLogPval_srt, id.vars = "Interactions")
names(signedLogPval_melt)[2:3] <- c("Cancers", "signedLog10pval")

coeffs_srt <- data.table(cbind(row.names(coeffs_clust), coeffs_clust))
colnames(coeffs_srt)[1] <- "Interactions"
coeffs_melt <- melt(coeffs_srt, id.vars = "Interactions")
names(coeffs_melt)[2:3] <- c("Cancers", "coeffs")

FDR_srt <- data.table(cbind(row.names(FDR_clust), FDR_clust))
colnames(FDR_srt)[1] <- "Interactions"
FDR_melt <- melt(FDR_srt, id.vars = "Interactions")
names(FDR_melt)[2:3] <- c("Cancers", "FDR")

dat_htmp <- cbind(signedLogPval_melt, coeffs_melt$coeffs, FDR_melt$FDR)
colnames(dat_htmp)[3:5] <- c("-log10(p-value)", "coeffs", "padj")
dat_htmp[, "-log10(p-value)"] <- as.numeric(as.vector(unlist(dat_htmp[, "-log10(p-value)"])))
dat_htmp$coeffs <- as.numeric(as.vector(dat_htmp$coeffs))
dat_htmp$padj <- as.numeric(as.vector(dat_htmp$padj))

dat_htmp$FDR <- ifelse(dat_htmp$padj<0.05, 1, 0)
```

Set the limits for effect size and -log10(p-valuse) and squish the saturated values

```{r}
pval_lim <- 5
dat_htmp$`-log10(p-value)`[abs(dat_htmp$`-log10(p-value)`) > pval_lim] <- pval_lim
coeff_lim <- 0.3
dat_htmp$coeffs[abs(dat_htmp$coeffs) > coeff_lim] <- coeff_lim*sign(dat_htmp$coeffs[abs(dat_htmp$coeffs) > coeff_lim])
```

Make a color vector to indicate the type of interaction

```{r}
colors_sign <- rev(as.character(signedLogPval_srt$Interactions))
colors_sign <- gsub(" .*", "", colors_sign)
colors_sign[grep("-", colors_sign)] <- "blue"
colors_sign[grep("\\+", colors_sign)] <- "red"
```

Improve the labels

```{r}
dat_htmp$Interactions <- gsub(".* ", "", dat_htmp$Interactions)
signedLogPval_srt$Interactions <- gsub(".* ", "", signedLogPval_srt$Interactions)
row.names(coeffs_clust) <- gsub(".* ", "", row.names(coeffs_clust) )

lbls <- row.names(coeffs_clust)
lbls <- gsub("miR-33ab$"," - miR-33-5p", lbls)
lbls <- gsub("miR-26ab$"," - miR-26-5p", lbls)
lbls <- gsub("miR-29abcd$"," - miR-29-3p", lbls)
lbls <- gsub("miR-27abc$"," - miR-27-3p", lbls)
lbls <- gsub("miR-133abc$"," - miR-133-3p", lbls)
lbls <- gsub("miR-199ab-5p$"," - miR-199-5p", lbls)
lbls <- gsub("miR-376abd$"," - miR-376-3p", lbls)
lbls <- gsub("miR-23abc$"," - miR-23-3p", lbls)
lbls <- gsub("miR-19ab$"," - miR-19-3p", lbls)
lbls <- gsub("miR-182$"," - miR-182-5p", lbls)
lbls <- gsub("miR-758$"," - miR-758-3p", lbls)
lbls <- gsub("miR-22$"," - miR-22-3p", lbls)
lbls <- gsub("miR-150$"," - miR-150-5p", lbls)
lbls <- gsub("miR-124$"," - miR-124-3p", lbls)
lbls <- gsub("miR-96$"," - miR-96-5p", lbls)
lbls <- gsub("miR-370$"," - miR-370-3p", lbls)
lbls <- gsub("miR-17$"," - miR-17-5p", lbls)
lbls <- gsub("miR-218$"," - miR-218-5p", lbls)
lbls <- gsub("miR-21$"," - miR-21-5p", lbls)
lbls <- gsub("miR-25$"," - miR-25-3p", lbls)
lbls <- gsub("_ -"," -", lbls)
lbls <- gsub("_"," - ", lbls)
```

Plot the heatmap

```{r}
pdf(paste("heatmap_TCGA_intsConfENCODEtotalRNAseq_corrRBPexprMirAct_signedLog10PvaluesANDcoeffs_purityCorrected2.pdf", sep=""), width=9.5, height=7)
ggplot(dat_htmp) +
geom_point(aes(
  x = Cancers, 
  size = `-log10(p-value)`, 
  y = factor(Interactions, levels = rev(as.character(signedLogPval_srt$Interactions))),
  shape = 21, 
  stroke = 1,
  fill=coeffs, 
  color=FDR)) +
scale_shape_identity() +
scale_y_discrete(label = rev(lbls)) +
theme(axis.text.y = element_text(color=colors_sign)) +
scale_fill_gradient2(limits=c(-coeff_lim,coeff_lim), low="blue", mid="white", high="red") +
scale_colour_gradient( limits=c(0,1), low="white", high="black") +
scale_size("-log10(p-value)", range = c(0.1, 10), breaks = seq(0, pval_lim, 1), limits=c(0, pval_lim))+
guides(size = guide_legend(override.aes = list(fill = "red")))+
labs(x="",y="") +
theme(legend.key.size = unit(1.5, 'lines'), text = element_text(size=12), axis.text = element_text(size=12), axis.text.x = element_text(angle=90, size=12, vjust=0.5, hjust=1), panel.background = element_rect(fill = 'white', colour = 'white')) +
theme(plot.margin = unit(c(1,1,1,1), "cm")) 
dev.off()
```

```{r}
sessionInfo()
```
