---
title: 'Association between miRNA activity and tumor purity in TCGA tumors'
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
```

# Load libraries

```{r}
library(tidyverse)
```

# Load the miRNA activity data and filter for tumors

```{r}
miR_act <- read.csv(here("TCGA", "output", "supplementary_tables", "TCGA_miRNA_activity.csv"))
miR_act <- miR_act[miR_act$sample_type == "tumor", ]
```

# Load the list of miRNAs significantly dysregulated in at least 10 tumor samples in at least one cancer type

```{r}
sig_mirs <- read.csv(here("TCGA", "output", "supplementary_tables", "TCGA_sig_mirs_atleast10Tsamples_allCancers.csv")) |>
  pull(miRNA) |>
  unique()

sig_mirs
```

# Load tumor purity estimates

```{r}
purity_files <- list.files(here("data"), pattern = "^purity_.*\\.txt$", full.names = TRUE)

purity_data <- purity_files |>
  setNames(str_extract(basename(purity_files), "(?<=purity_)[A-Z]+(?=\\.txt)")) |>
  lapply(function(f) {
    dat <- read.delim(f, stringsAsFactors = FALSE)
    # Keep only primary tumor, vial A samples (-01A-)
    dat <- dat[grepl("-01A-", dat$SampleName), ]
    dat$sample <- str_extract(dat$SampleName, "^TCGA-[^-]+-[^-]+-01") |>
      str_replace_all("-", ".")
    dat[, c("sample", "Purity")]
  })

head(purity_data[["UCEC"]])
```

# Restrict the analysis to cancer types for which both miRNA activity and tumor purity estimates are available.

```{r}
cancer_types <- intersect(names(purity_data), unique(miR_act$cancer_type))
cancer_types
```

# Correlations between miRNA activity and tumor purity, across cancer types

```{r}
# Helper function to compute the correlation between the activity of a given miRNA and tumor purity, within a given cancer type
compute_purity_correlation <- function(mirna, cancer) {
  mirna_data <- miR_act |>
    filter(cancer_type == cancer, miRNA == mirna)

  purity_cancer <- purity_data[[cancer]]

  # Match samples between miRNA activity and purity data
  common_samples <- intersect(mirna_data$sample_name, purity_cancer$sample)
  if (length(common_samples) == 0) {
    return(NULL)
  }

  mirna_matched <- mirna_data$coefficient_log2[match(common_samples, mirna_data$sample_name)]
  purity_matched <- purity_cancer$Purity[match(common_samples, purity_cancer$sample)]

  corr <- cor.test(mirna_matched, purity_matched)
  tibble(r = corr$estimate, pval = corr$p.value, n = length(common_samples))
}
```

```{r}
# Compute the correlation for every significant miRNA, in every cancer type with purity data available
corr_table_purity <- crossing(miRNA = sig_mirs, cancer = cancer_types) |>
  mutate(corr_data = map2(miRNA, cancer, compute_purity_correlation)) |>
  unnest(corr_data, keep_empty = TRUE) |>
  filter(!is.na(r)) |>
  mutate(
    padj = p.adjust(pval, method = "fdr"),
    minuslog10pvals = -log10(pval)
  ) |>
  dplyr::select(miRNA, cancer, r, pval, padj, minuslog10pvals, n) |>
  as.data.frame()

head(corr_table_purity)
nrow(corr_table_purity)
```

# Cluster miRNAs and cancer types by their correlation coefficients

```{r}
r_mat <- corr_table_purity |>
  dplyr::select(miRNA, cancer, r) |>
  pivot_wider(names_from = cancer, values_from = r) |>
  column_to_rownames("miRNA") |>
  as.matrix()

row_clust <- hclust(dist(r_mat, method = "euclidean"), method = "complete")
col_clust <- hclust(dist(t(r_mat), method = "euclidean"), method = "complete")

mirna_levels <- rownames(r_mat)[row_clust$order]
cancer_levels <- colnames(r_mat)[col_clust$order]
```

# Make the data table for the dotplot, clustered and converting saturated values

```{r}
pval_lim <- 10
coeff_lim <- 0.4

dat_htmp_purity <- corr_table_purity |>
  transmute(
    miRNAs = factor(miRNA, levels = mirna_levels),
    Cancers = factor(cancer, levels = cancer_levels),
    `-log10(p-value)` = pmin(minuslog10pvals, pval_lim),
    `Pearson r` = pmax(pmin(r, coeff_lim), -coeff_lim),
    FDR = as.integer(padj < 0.05)
  )
```

# Dotplot

```{r, fig.dim = c(9.5, 7), fig.align = 'center'}
dotplot_purity <- ggplot(dat_htmp_purity, aes(x = Cancers, y = miRNAs)) +
  geom_point(aes(size = `-log10(p-value)`, fill = `Pearson r`, color = FDR),
             shape = 21, stroke = 1) +
  scale_fill_gradient2(limits = c(-coeff_lim, coeff_lim), low = "blue", mid = "white", high = "red") +
  scale_colour_gradient(limits = c(0, 1), low = "white", high = "black") +
  scale_size("-log10(p-value)", range = c(0.1, 10), breaks = seq(0, pval_lim, 2), limits = c(0, pval_lim)) +
  guides(size = guide_legend(override.aes = list(fill = "black"))) +
  labs(x = "", y = "") +
  theme(
    legend.key.size = unit(1.5, "lines"),
    text = element_text(size = 12),
    axis.text = element_text(size = 12),
    axis.text.x = element_text(angle = 90, size = 12, vjust = 0.5, hjust = 1),
    panel.background = element_rect(fill = "white", colour = "white"),
    plot.margin = unit(c(1, 1, 1, 1), "cm")
  )

dotplot_purity

pdf(here("TCGA", "output", "figures", "TCGA_corrMirActVsPurity_dotplot.pdf"), width = 8, height = 7)
print(dotplot_purity)
dev.off()
```

```{r}
sessionInfo()
```
