---
title: 'Association between DDX55 expression and survival in TCGA patients'
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
```

# Load libraries

```{r}
library(data.table)
library(dplyr)
library(purrr)
library(stringr)
library(tibble)
library(ggplot2)
library(survival)
library(survminer)
```

# Locate the data files

```{r}
logRSEM_files <- list.files(here("TCGA", "data"), pattern = "log2RSEM.csv", full.names = TRUE)
clin_files <- Sys.glob(here("TCGA", "data", "*.merged_only_clinical_clin_format.txt"))
purity_files <- list.files(here("TCGA", "data"), pattern = "^purity_.*\\.txt$", full.names = TRUE)

cancers <- str_remove(basename(logRSEM_files), "_log2RSEM\\.csv$")
```

# Helper functions

```{r}
# Long table of expression for one gene, for one tumor sample per patient, across all cancer types found in `logRSEM_files`.
load_gene_expression <- function(gene_name, logRSEM_files) {
  map_dfr(logRSEM_files, function(file) {
    df <- fread(file, header = TRUE)
    tibble(
      sample = colnames(df)[-1],
      cancer = str_remove(basename(file), "_log2RSEM\\.csv$"),
      expr = as.numeric(df[df[[1]] == gene_name, -1, with = FALSE])
    )
  }) |>
    filter(str_detect(sample, "-01$")) |>
    transmute(patient = str_remove(sample, "-01$"), cancer, expr)
}
```

```{r}
# TCGA pathologic stage strings (e.g. "Stage IIIA") to numeric stage (1-4).
stage_to_numeric <- function(stage) {
  roman <- str_extract(toupper(stage), "IV|III|II|I")
  suppressWarnings(as.numeric(as.roman(roman)))
}

# Clinical data files store one column per patient; transpose to one row per patient with the original column headers as variable names.
read_clinical_raw <- function(cancer, clin_files) {
  file <- clin_files[str_detect(clin_files, cancer)]
  if (length(file) == 0) return(NULL)
  clinical <- t(fread(file, header = TRUE, sep = "\t"))
  colnames(clinical) <- clinical[1, ]
  as.data.frame(clinical[-1, , drop = FALSE], stringsAsFactors = FALSE)
}

# Minimal survival variables (time, status), used for the Kaplan-Meier plot.
load_survival_data <- function(cancer, clin_files) {
  clinical <- read_clinical_raw(cancer, clin_files)
  if (is.null(clinical)) return(NULL)
  clinical |>
    transmute(
      patient = toupper(patient.bcr_patient_barcode),
      time = coalesce(as.numeric(patient.days_to_death), as.numeric(patient.days_to_last_followup)),
      status = case_when(
        toupper(patient.vital_status) %in% c("ALIVE", "LIVING") ~ 0,
        toupper(patient.vital_status) == "DEAD" ~ 1
      )
    ) |>
    filter(!is.na(time), !is.na(status))
}

# Survival variables plus age and stage, for the COX models
load_clinical_data <- function(cancer, clin_files) {
  clinical <- read_clinical_raw(cancer, clin_files)
  if (is.null(clinical) || !"patient.stage_event.pathologic_stage" %in% colnames(clinical)) {
    return(NULL)
  }

  age_vars <- intersect(
    c("patient.age_at_initial_pathologic_diagnosis",
      "patient.primary_pathology.age_at_initial_pathologic_diagnosis"),
    colnames(clinical)
  )
  if (length(age_vars) == 0) return(NULL)

  clinical$age <- if (length(age_vars) == 2) {
    rowSums(sapply(clinical[age_vars], as.numeric), na.rm = TRUE)
  } else {
    as.numeric(clinical[[age_vars]])
  }

  clinical |>
    transmute(
      patient = toupper(patient.bcr_patient_barcode),
      time = coalesce(as.numeric(patient.days_to_death), as.numeric(patient.days_to_last_followup)),
      status = case_when(
        toupper(patient.vital_status) %in% c("ALIVE", "LIVING") ~ 0,
        toupper(patient.vital_status) == "DEAD" ~ 1
      ),
      age = age,
      stage = stage_to_numeric(patient.stage_event.pathologic_stage)
    ) |>
    filter(!is.na(time), !is.na(status), !is.na(age), !is.na(stage))
}
```

```{r}
# Tumor purity for one cancer type
load_purity_data <- function(cancer, purity_files) {
  file <- purity_files[str_detect(purity_files, cancer)]
  if (length(file) == 0) return(NULL)
  purity <- read.table(file, header = TRUE, row.names = 1)
  if (nrow(purity) == 0) return(NULL)
  purity |>
    rownames_to_column("sample") |>
    filter(str_detect(sample, "-01A-")) |>
    transmute(patient = substr(sample, 1, 12), purity = Purity)
}
```

```{r}
extract_cox_stats <- function(fit) {
  s <- coef(summary(fit))
  tibble(coeff = unname(fit$coefficients[1]), pval = s[1, 5], se = s[1, 3])
}

# For one cancer type, fit a COX model for gene expression (one gene), correcting for stage only, and (when tumor purity data is available) a second model also correcting for age and tumor purity. Returns NULL if there isn't enough data to fit even the stage-only model
run_cox_regression <- function(cancer, expr, clin_files, purity_files, min_patients = 10) {
  clinical <- load_clinical_data(cancer, clin_files)
  if (is.null(clinical)) return(NULL)

  data <- inner_join(clinical, filter(expr, cancer == !!cancer), by = "patient")
  if (nrow(data) < min_patients) return(NULL)

  fit_wo_conf <- coxph(Surv(time, status) ~ expr + stage, data = data)
  stats_wo_conf <- extract_cox_stats(fit_wo_conf) |> rename_with(~ paste0(., "_wo_conf"))

  stats_conf <- tibble(coeff_conf = NA_real_, pval_conf = NA_real_, se_conf = NA_real_)
  purity <- load_purity_data(cancer, purity_files)
  if (!is.null(purity)) {
    data_conf <- inner_join(data, purity, by = "patient")
    if (nrow(data_conf) >= min_patients) {
      fit_conf <- coxph(Surv(time, status) ~ expr + age + stage + purity, data = data_conf, iter.max = 30)
      stats_conf <- extract_cox_stats(fit_conf) |> rename_with(~ paste0(., "_conf"))
    }
  }

  bind_cols(tibble(cancer = cancer), stats_wo_conf, stats_conf)
}
```

```{r}
# Bar plot of COX coefficients per cancer type, colored by the signed -log10(p-value) (with saturation limits), with the standard error as the error bars
plot_cox_results <- function(data, coeff_col, se_col, signed_log10p_col, ylim, lim = 5) {
  data <- data |>
    arrange(desc(.data[[coeff_col]])) |>
    mutate(
      cancer = factor(cancer, levels = cancer),
      color_val = pmin(pmax(.data[[signed_log10p_col]], -lim), lim)
    )

  ggplot(data, aes(x = cancer, y = .data[[coeff_col]], fill = color_val)) +
    geom_col(color = "black", linewidth = 0.2) +
    geom_errorbar(
      aes(ymin = .data[[coeff_col]] - .data[[se_col]], ymax = .data[[coeff_col]] + .data[[se_col]]),
      width = 0.2, linewidth = 0.3
    ) +
    scale_fill_gradientn(colors = c("darkblue", "blue", "white", "red", "darkred"), limits = c(-lim, lim)) +
    coord_cartesian(ylim = ylim) +
    labs(x = NULL, y = "COX regression coefficient", fill = "signed\nlog10(p)") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
      panel.grid = element_blank(),
      axis.line.y = element_line(color = "black", linewidth = 0.5),
      axis.ticks.y = element_line(color = "black", linewidth = 0.5)
    )
}
```

```{r}
# Kaplan-Meier plot for one cancer type, splitting patients into "high"/"low" expression groups at the median (patients equal to the median are dropped)
plot_survival_by_expression <- function(cancer, gene_name, expr, clin_files) {
  data <- load_survival_data(cancer, clin_files) |>
    inner_join(filter(expr, cancer == !!cancer), by = "patient") |>
    mutate(
      expr_group = case_when(
        expr > median(expr) ~ "High expression",
        expr < median(expr) ~ "Low expression",
        TRUE ~ NA_character_
      )
    ) |>
    filter(!is.na(expr_group))

  fit <- survfit(Surv(time, status) ~ expr_group, data = data)
  p <- ggsurvplot(
    fit, data = data, pval = TRUE,
    title = cancer,
    font.main = c(10, "bold", "black"),
    xlab = "Time (days)",
    legend = "none",
    palette = c("brown1", "dodgerblue")
  )
  p$plot <- p$plot + theme(plot.title = element_text(hjust = 0.5))
  p
}
```

# Run the analysis for the gene of interest

```{r}
gene_name <- "DDX55"

gene_expr <- load_gene_expression(gene_name, logRSEM_files)
head(gene_expr)
```

# COX regression across all cancer types

For each cancer type, fit a COX model correcting for stage only, and a second model also correcting for age and tumor purity where available

```{r}
cox_results <- map(
  cancers, run_cox_regression,
  expr = gene_expr, clin_files = clin_files, purity_files = purity_files
) |>
  list_rbind() |>
  filter(!is.na(coeff_wo_conf)) |>
  mutate(
    signedLog10pval_wo_conf = sign(coeff_wo_conf) * -log10(pval_wo_conf),
    signedLog10pval_w_conf = sign(coeff_conf) * -log10(pval_conf)
  )
```

## Correcting for stage only

```{r}
p <- plot_cox_results(cox_results, "coeff_wo_conf", "se_wo_conf", "signedLog10pval_wo_conf", ylim = c(-1, 5))
ggsave(here("TCGA", "output", "figures", "TCGA_DDX55_cox_stage_only.pdf"), plot = p, width = 4, height = 3)

p
```

## Additionally correcting for age and tumor purity

```{r}
p <- cox_results |>
  filter(!is.na(coeff_conf)) |>
  plot_cox_results("coeff_conf", "se_conf", "signedLog10pval_w_conf", ylim = c(-1, 5))
ggsave(here("TCGA", "output", "figures", "TCGA_DDX55_cox_with_confounders.pdf"), plot = p, width = 4, height = 3)

p
```

# Kaplan-Meier survival curve in KIRC

```{r}
p <- plot_survival_by_expression("KIRC", gene_name, gene_expr, clin_files)
ggsave(here("TCGA", "output", "figures", "TCGA_DDX55_km_KIRC.pdf"), plot = p$plot, width = 4, height = 4)

p
```

```{r}
sessionInfo()
```
