---
title: "Effect of DDX55 knockdown (ENCODE) on functional miR-29 targets defined using miR-29 inhibition and over-expression experiments"
author: "Gabrielle Perron"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
```

# Load libraries

```{r}
library(venn)
library(fgsea)
```

# Load the miRNA-binding matrix

```{r}
binding <- read.csv(here("ENCODE_RBP_KD", "data", "miRNA_seed_matches_3UTRs2000bp.csv"), row.names = 1, header = TRUE)
```

# Define miR-29 predicted targets

```{r}
pred_tars <- row.names(binding)[binding[, "miR.29abcd"] > 0]
```

# Load the differential gene expression data for the DDX55 KD experiment from ENCODE. Retain only non-NA entries

```{r}
DDX55_KD <- readRDS("ENCODE_DDX55_KD_DESeq2_res.rds")
DDX55_KD <- DDX55_KD[ !is.na(DDX55_KD$padj), ] 
```

Load the differential expression analysis results for the miR-29 mimic experiment in 786-O cells (from the DiffRAC paper https://doi.org/10.1038/s42003-022-03796-w). Retain only non-NA entries

```{r}
mimic_786O <- readRDS("siDDX55_res_condition_Mimic.miR_vs_Mimic.neg.rds")
mimic_786O <- mimic_786O[ !is.na(mimic_786O$padj), ]
```

Load the differential expression analysis results for the miR-29 mimic experiment in A-498 cells (from the DiffRAC paper https://doi.org/10.1038/s42003-022-03796-w). Retain only non-NA entries

```{r}
mimic_A498 <- readRDS("DiffRAC_miR29_validation_mimic_res_condition_miR29_mimic_vs_miR29_mimic_ctrl.rds")
mimic_A498 <- mimic_A498[ !is.na(mimic_A498$padj), ]
```

Load the differential expression analysis results for the miR-29 inhibition experiment in ACHN cells (from the DiffRAC paper https://doi.org/10.1038/s42003-022-03796-w). Retain only non-NA entries

```{r}
inhib_ACHN <- readRDS("DiffRAC_miR29_validation_inhibition_res_condition_miR29_inhibition_vs_miR29_inhibition_ctrl.rds")
inhib_ACHN <- inhib_ACHN[ !is.na(inhib_ACHN$padj), ]
```

Define functional miR-29 targets: Genes that are predicted targets of miR-29 and at least one of these categories:
- significantly downregulated in a miR-29
- significantly upregulated in the inhibition experiment

```{r}
down_786O <- row.names(mimic_786O)[mimic_786O$padj < 0.05 & mimic_786O$log2FoldChange < 0]
down_A498 <- row.names(mimic_A498)[mimic_A498$padj < 0.05 & mimic_A498$log2FoldChange < 0]
up_ACHN <- row.names(inhib_ACHN)[mimic_A498$padj < 0.05 & mimic_A498$log2FoldChange > 0]
tars <- pred_tars[pred_tars %in% down_786O | pred_tars %in% down_A498 | pred_tars %in% up_ACHN]
```

GSEA to define the enrichment of these miR-29 targets among up/down-regulated genes in the DDX55 KD from ENCODE. We

```{r}
genelist <- list(tars)
stat <- DDX55_KD$stat
names(stat) <- row.names(DDX55_KD)
names(genelist) <- "miR-29 functional targets"
fgseaRes <- fgsea(pathways = genelist,
  stats = stat,
  minSize=10)
print(plotEnrichment(genelist[[1]], stat, gseaParam = 1, ticksSize = 0.05))
fgseaRes$ES
fgseaRes$pval
```

Also save the figure

```{r}
pdf("GSEA_miR-29_targets_mimic_inhibition_DDX55_KD.pdf", height=3, width=4)
plotEnrichment(genelist[[1]], stat, gseaParam = 1, ticksSize = 0.05)
dev.off()
```

```{r}
sessionInfo()
```
